### 2013-03-01 Paper.Rnw
initial commit

\documentclass[letterpaper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{figure/}}
\usepackage{epstopdf}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}

\newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike
\newcommand{\svp}[1]{{\color{red}#1}} %green: Susan (It's more fun that way - comment roulette)

%opening
\title{Binning Strategies and Related Loss for Graphics of Large Data}
\author{Susan VanderPlas$^1$, Karsten Maurer$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}

\begin{document}
\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
\maketitle

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. Large data demands require us to re-think strategies of visualizing data. Binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much. This comes with the price of loss of information inherent to any binning scheme. In this paper we discuss properties of two commonly used binning algorithms. We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the frame work of a simulation and two case studies.
\end{abstract}

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)

connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
                   password="R R0cks", port=3306, dbname="baseball", 
                   host="headnode.stat.iastate.edu")
pitch <- new("dataDB", co=connect, table="Pitching")

d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs

xtd.data <- function(x) {
# replicate each line as often as given by the last entry
  k <- dim(x)[2]

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, dependson='setup', echo=F, fig.keep='none', results='hide'>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

# Sunflower plot
# d1.hex$id <- seq(1:nrow(d1.hex))
# d1.hex2 <-  ddply(d1.hex, .(G, SO, id), function(i) return(data.frame(r = cos((floor(i$Freq/10))*seq(0, 2*pi, .05)), theta=seq(0, 2*pi, 0.05), Freq=i$Freq)))
# ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))

@

Technological advances have  facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online. According to estimates published by the IDC in 2009 the worldwide digital content  doubles about every 18 months -- in 2011 the digital amount was estimated to be at 1.8 zeta bytes (1 ZB = $2^{70}$ bytes $\approx 10^{21}$ bytes).
This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: 
\begin{quote}``It’s a revolution ... The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government. There is no area that is going to be untouched"\end{quote} (Gary King, Harvard Institute).

% which challenges?
Data sets with millions of records and thousands of variables are not uncommon. 
Such data sets are often too large to fit into a single computer's working memory, and  single-user machines cannot easily deal with them. We can employ database tools to extract relevant variables, thereby vertically subsetting the data. 
However, with millions of records, it can be very difficult to work with data even at that level.

\citet{Friedman97} proposed in his paper on datamining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it". The same holds for visualizations.  
With a 100-1000 fold increase in the amount of data, the utility of some of  graphs changes dramatically, with some of our most-used tools such as the scatterplot deteriorating quickly \citep{gold}.

% binning as a way out
Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots, which utilize  raw data. 
This fact is being exploited to make scatterplots again useful instruments for large data settings  by employing binning strategies \citep{gold}.

In this paper we describe first the problem scatterplots are exposed to in large-data situations. We discuss two different binning methods and introduce the  {\it loss of information} inherent to binning. We conclude with a detailed example.

% It is important to be able to describe the contents of the database in a quick and easily comprehensible manner, similar to the use of the mean and standard deviation to describe univariate data, or the mean, and variance matrix to describe multivariate data. Graphical summaries of the data contained within the database are likely to more fully describe large, complex data sets which may not originate from a single multivariate distribution than a set of numerical summaries (i.e. the mean, median, mode, and range used in introductory statistics classes).

\section{Scatterplots for Large Data Sets}
In the case of medium sized data scatterplots are great tools for showing relationships in two dimensions. For large data, scatterplots suffer from over-plotting -- i.e. more and more points are drawn in close-by places, thereby masking relevant structure.
Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball players as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games (G) played in a season is plotted versus number of strikeouts (SO), i.e. a batter receives three strikes during his time at bat. Pitchers want to throw as many strikeouts as possible. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
Figure~\ref{fig:scatter-alpha} shows a traditional scatterplot on the left. Each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts. Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points visible that are close together. This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points. 
A modern alternative to open circles is alpha-blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/Overplotting.pdf}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/OverplottingCircles.pdf}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/OverplottingAlpha.pdf}}

  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{figure/HexBinning.pdf}}
  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{figure/BinScatter.pdf}}

	\caption{Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha-blending ($\alpha$=0.05). Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
  \label{fig:scatter-alpha}
\end{figure}

All of these methods fall short in the example.
As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help some, but are as prone to over-plotting as solid circles when the data set is large. Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. The data set is large enough that neither alpha-blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

One approach to reduce the graphical complexity is to bin the displayed data. This has the additional advantage of reducing the size of the stored data, as only the bin centers and the frequency of points which correspond to that bin must be stored.

Histograms are an example of plots which use binned variables. To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to create a two-dimensional graph which uses some other attribute (color, 3D rendering) to provide joint density information. A {\it tile plot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.

Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, modified scatterplots, and kernel density smoothing of tonal variation \citep{martin-gold}. Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin. Sunflower plots are particularly useful when the number of points in each bin remains reasonably small.

When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.  When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to tile plots or $\alpha$-blended scatterplots in a more smooth, continuous fashion. These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.

As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonal tile plot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e). The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c). The tile plot is less prone to the Hermann-grid illusion than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.

% \begin{figure}[hbtp]
% <<conclusions, out.width='\\textwidth', fig.width=5, fig.height=3>>=
% library(dbData)
% connect <- dbConnect(dbDriver("MySQL"), user="2009Expo",
% password="R R0cks", port=3306, dbname="baseball",
% host="headnode.stat.iastate.edu")
% pitch <- new("dataDB", co=connect, table="Pitching")
% pitch.stats <- dbData(vars=list("G", "SO", "yearID"), pitch)
% qplot(yearID, SO/G, alpha=I(0.2), weight=Freq, data=pitch.stats, colour=G) + xlab("season")
% @
% \caption{\caption{strikeout-rate}Scatterplot of rate of strikeouts per game over time. Each dot corresponds to a pitcher's seasonal average. Coloring indicates number of games played, alpha-blending is used additionally to emphasize high-density regions. There's clear break points in the strikeout rates over time. In 1902 XXXX}
% \end{figure}
Only in the tile and the bubble plot the inner structure of the data becomes apparent: 
the joint density consists two distinct ridges following two lines with very different slopes. The lower slope corresponds to the --modern-- average strike out rate of pitchers of just under one strike-out per game. The other line has a slope of about four times that rate. This high rate is also associated with fewer games played. Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and  bats.

% two centers - almost figure~\ref{strikeout-rate} is numerically loss-free (binning is to integers, reduces the data by 16\%)

%\hh{... This is the place to put some background literature: include the discussion of use of binning in the literature: Martin Theus's chapter in the GOLD book, Dan Carr's hexagonal binning, sunflower plots, ... }

For extremely large data sets, tile plots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be  rendered separately.

As with histograms, the width of bins (or the number of bins) is an important factor in the detail of the binned data and the resulting plot: if the bin width is too small in comparison to the amount of data available, there is little advantage to binning, but if the bin width is too large, interesting features of the joint distribution may be obscured by over-smoothing.

%\hh{Main motivation: we want to investigate how much information we lose when we bin.}

%\hh{I also would like to pick up the idea of sufficient statistics for scatterplots again - as a mathematical basis for measuring the amount of information that goes into a plot (since we can't expect to ever get more out of a plot). .... I'm not sure, yet, where in the paper we should place this discussion.}

\section{Binning Data}
We will only consider binning in two dimensions, $X$ and $Y$. The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.
Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to  a fraction of the original.

Binning is a two-step procedure: we first assign each data tuple $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.

 We will proceed with rectangular bins for simplicity, but other binning schemes, such as those which utilize hexagonal bins are also common \citep{scatterplots}. Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two dimensions.

\subsection{Standard Rectangular Binning Algorithm}
Let $x_{(i)}$ correspond to the ordered values of $X$, where $i=1, ..., n$ and $n$ is the number of total observations. Then $x_{(1)}$ and $x_{(n)}$ then denote the observed minimum and the maximum of $X$, respectively.

We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le n_X$, where $n_X$ is the overall number of bins in dimension $X$. Binning can either be determined by the number of bins, $n_X$, or, equivalently, by the bin width, $b_X$. For standard binning the relationship between these two parameters is  given as 
\[
b_X = (x_{(n)} - x_{(1)})/n_X
\]

Mathematically, binning corresponds to a function $b(.)$ that assigns to a value $x$ the closest bin center $x^\ast \in \{ x_j^\ast \}, j \in \{1, ..., n_X\}$, i.e. the binning function   $b(.) : x \rightarrow x^\ast$ is defined as
\begin{eqnarray*}\label{stdbin}
x^\ast = b(x) =  b_X \cdot \left\lfloor (x-O_X)/ b_X + b_X/2\right\rfloor + O_X
\end{eqnarray*}
where $O_X$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. 
%to allow for bins of width $b_X$ centered at any sequence $O_X, O_X + b_X, ..., O_X + n b_X$.

For illustration, let us assume that we have observed values 1.23, 1.55, and 2.35, and bin centers $\{1, 2, 3\}$, corresponding to a bin width of $b_X = 1$ and a first bin center $O_X = 1$.
Intuitively it is clear, that value 1.23 is assigned to bin center 1, while values 1.55 and 2.35 are assigned to bin center 2.
Applying equation (\ref{stdbin}) to value 1.23 we get: $1 \cdot \lfloor (1.23-1)/ 1 + 0.5\rfloor + 1$ = $\lfloor 0.73 \rfloor + 1$ =  0 + 1 = 1. Similarly, 1.55 is assigned to $\lfloor 0.55 + 0.5\rfloor + 1$ = 1+1 = 2.

Alternate binning parameterizations include dividing up the range of data by a specified number of bins of equal width, but even these strategies fit into the framework of assigning values to the closest bin center.
% , with smallest bin
% $b_X   \cdot \lfloor (x_{(1)}-O_X)/b_X + 0.5\rfloor + O_X$ and largest bin 
% $b_X   \cdot \lfloor (x_{(n)}-O_X)/b_X + 0.5\rfloor + O_X$.

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo,echo=F, fig.keep='none', cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

      d$xbin <- NA
      d$ybin <- NA
      idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
      d$xbin[idx] <- dnew$x[idx]
      d$ybin[idx] <- 25

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
@

The method described above is the standard binning algorithm. As a different approach, we define the random binning algorithm below.

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinningDataPoints2.pdf}}
  \caption{Points are assigned to bin (0, 20) and bin (10,20). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE,fig.keep='none', cache=TRUE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@
An interesting variation which results in statistically nice properties is a random, non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x$ to bin center $x^\ast$ with a probability depending on the distance to that bin center;  the closer a value is to a bin center, the higher the probability that the value is assigned to this bin center. 
More formally,
\begin{eqnarray}\label{randbin}
b^r(x) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x)}/b_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x - x^\ast_{j})}/b_X
  \end{array}\right.
\end{eqnarray}

% In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin. 
This method is easily extensible to non-uniform bin sizes and can also map $x$ to more than two bins, unlike the standard algorithm.

To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.

Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.

The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin. ``Linear" binning  \citep{martin-gold} is a --computationally intensive-- alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.  With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning. Specifically, the expectation of the random binning algorithm is the same as for linear binning.
%However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
In the next section, we examine the loss of visual and numerical information due to binning.
% linear binning in Gold book - random algorithm as a less computational method of linear binning

\section{Loss due to Binning}
While binning data has some notable advantages, it does result in loss of some information. Using a small number of bins removes much of the relevant information in the data, while an extremely large number of bins may not reduce the data size or complexity to a sufficient degree. 
Figure~\ref{binning} gives an overview of a data set and binned representations using different numbers of bins, demonstrating the loss of information with increasing bin size.

\begin{figure}[hbtp]
<<binning_setup, echo=FALSE,fig.width=2, fig.height=2, out.width='\\linewidth'>>=
set.seed(46)
x <- round(rnorm(200), digits=2)
y <- round(rnorm(200), digits=2)
library(ggplot2)
X <- data.frame(x,y)
library(plyr)
@
\begin{minipage}{.35\linewidth}
<<first, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
qplot(x,y, asp=1, main='Unbinned Data')+theme_bw()
@
\end{minipage}
\hfil\begin{minipage}{.645\linewidth}
\begin{minipage}{.495\linewidth}
<<bin001, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
binwidth=0.1
X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey90") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey90") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
@
\end{minipage}\hfil
\begin{minipage}{.495\linewidth}
<<bin025, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
binwidth=0.25
X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey80") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey80")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
@
\end{minipage}\\
\hfil\begin{minipage}{.495\linewidth}
<<bin05, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
binwidth=0.5
X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey70") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey70")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
@
\end{minipage}\hfil
\begin{minipage}{.495\linewidth}
<<bin1, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
binwidth=1
X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey60") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey60")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
@
\end{minipage}
\end{minipage}
\caption{\label{binning}Series of scatterplots showing the original data (scatterplot, left), and versions of the binned data for different bin widths. The visual loss from binning at 0.1 is minimal, while a bin width of 1 gives a rough approximation.}
\end{figure}
In the first set of binned data, the bins and the points on the scatterplot are nearly identical, but the scatterplot contains information about overlapping points. The second and third sets of binned data, with bin width=0.25 and 0.50 respectively, show higher-level summaries of the data that contain some numerical and visual loss but which may also provide more visually accessible information about the shape of the two-dimensional density between $x$ and $y$. The fourth set of binned data with bin width = 1.0 is nearly unrecognizable, because the bins are large enough that they provide very little additional information.

Loss of information occurs on multiple levels during the binning and rendering process. We distinguish three sources:
\begin{itemize}
\item {\it Numerical Loss} occurs when the data set is reduced  to bin centers, i.e. this is the difference between a data point $(x,y)$ and the corresponding bin center $(x^\ast, y^\ast)$.

\item {\it Frequency Loss}  results from our inability to render the frequency information accurately. Different rendering methods for the frequency information will lead to different losses. For the remainder of the paper we will assume that we are using colors in tile plots to represent this information. The number of colors that can be differentiated and mapped back to frequency information accurately is fairly limited. Note that even though this loss will turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where there is no frequency information, so that points which are over-plotted are not at all distinguishable.

% 
% , so we must balance perceptual limits of frequency information with the limits of overplotting single points.

\item {\it Visual Loss}: While the visual loss is most salient, it is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile. This will generally different from the center of the data corresponding to that bin as given by the mean. This causes some bias in the perception.

% which may not be the center of the data within the bin, causing bias and numerical loss.
% 
% Visual loss occurs in the graphical display of the data, as individual points are represented as tiles, and tiles are perceived according to their visual center, rather than according to the center of the data.
% 
%  
\end{itemize}

\noindent
Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.

<<tablelossexample, echo=FALSE, results='asis',cache=FALSE>>=
loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)

loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))

table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
                               c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
                               c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
                               c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
                               c(binwidth=1, loss1[5:7]*100, loss1[8])))
names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")

@

We explore the different types of loss in the remainder of this section in more detail.

\subsection{Loss in Spatial Dimensions}
Reducing data points $x$ to bin centers $\hat x$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid. Let us denote the difference between point $x_i$ from its bin center as $S_i = \left|x_i-b(x_i)\right|$ for each point in the data set $i=1, ..., N$. The total loss from binning an entire data set is defined as

\[L_s(x) = 1/S^2_{\emptyset} \cdot \sum_{i=1}^{N} S^2_i\] 
where $S_\emptyset^2$ is the loss which results when the entire data set is reduced to a single bin,
$S^2_{\emptyset} = \sum_i (x_i-\tilde{x})^2$ for bin center $\tilde{x}$.
The loss depends on the choice of bin center: candidates for bin centers are the mean of the points within the bin or the visual center of the bin. 
If we consider the bin center to be the numerical mean, then the loss is the sum of the variance of the data in each bin, compared to the variance of the data set as a whole, and $L_s(x) \in [0,1]$.

The disadvantage to this choice is that this introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of he symbol representing the data.
Therefore we distinguish two contributions to this loss:
Numerical loss ($L^\text{N}$) results from the binning process as points are reduced to the numerical mean center, and visual loss ($L^\text{V}$) is due to the difference between the mean of points within the bin and the visual center of the bin. This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.

Fortunately we can show that the total loss perceived is partitioned by these two components of loss, and it holds (see appendix \ref{proof:partition} for a proof):
\begin{eqnarray}\label{eqn:partition}
L_s(X) =  1/S^2_{\emptyset} \cdot \left(\sum_i L^N_i + \sum_i L^V_i\right)
\end{eqnarray}
<<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
set.seed(53022)
data <- data.frame(x=rnorm(200, 0, 2), y=rnorm(200, 0, 2), Freq=1)
data$xbin <- round(data$x)
data$ybin <- round(data$y)
data.means <- ddply(data, .(xbin,ybin), summarise, xmean=mean(x), ymean=mean(y))
data <- merge(data, data.means)

ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
@

\noindent
Note that the discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms. The decomposition also holds for the random binning algorithm as well as the standard binning algorithm. 
%The expected total loss for either algorithm depends on the distribution the underlying data.

%The total loss due to binning is $\sum_{i=1}^n (x_i - b(x_i))^2$, where in the standard algorithm, $b(x)$ maps $x$ to the closest bin center, and in the random algorithm, $b(x)$ is assigned probabilistically.

\paragraph{Loss in the Random Assignment Algorithm:}
An interesting feature of the random binning algorithm is that the loss that we experience by doubling the bin width is deterministic, i.e. independent even though the assignment of bins is not. This holds for repeated doubling of the bin width. A proof of this can be found in appendix \ref{proof:constantloss}. Doubling bin widths is, particularly in the framework of visualizations, a very natural step, making this an important property of the random binning.

The loss as a function of bin width for both algorithms is shown in Figure~\ref{fig:LossComparison1}.

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./Binning-DataVisualNumericalCenter.pdf}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

\subsection{Loss of Frequency Information} 
Loss of frequency information happens because of our inability to render elements of a third dimension efficiently. By employing a color scheme for filling tiles according to the observed frequencies corresponds to a binning of the frequencies into a few different levels, with losses stemming from the same source as the previously discussed numerical loss.
We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_j\}$ with $1 \le j \le n_C$, and that an observed  bin frequency $c$ is assigned to the closest frequency level $c^\ast = b(c)$.  
The --unscaled-- frequency loss is then defined as
\begin{eqnarray}\label{eqn:frequency}
L_\text{Freq} = \sum_{i=1}^{n_X n_Y} \left(c_i - b(c_i)\right)^2
\end{eqnarray}

% When we consider loss of frequency information, we must consider both dimensions simultaneously, as frequency information is based on the combination of binning in $x$ and  $y$ direction. We define loss of frequency information as: \hh{what's that in words?}
% $$ L_\text{Freq} = \sum_j \left(c_j - \frac{1}{n_X n_Y}\sum_j c_j\right)^2,$$ 
% where $n_X, n_Y$ are the number of bins in $x$ and $y$, respectively.

The loss of frequency information due to binning results from tiling over larger areas, thus results from {\it smoothing} the frequency variation (see Figure~\ref{fig:stdBinTutorial2}).

Frequency data consists of counts, which most commonly exhibit skew densities, i.e. there are usually a lot of cells with small cell counts and a few cells with extremely large counts.

A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. This suggests that frequency loss be calculated as 
\begin{eqnarray}\label{eqn:logfrequency}
L_{\log \text{Freq}} = \sum_{i=1}^{n_X n_Y} \left(\log(c_i+1) - \log(b(c_i)+1)\right)^2
\end{eqnarray}

Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

This numerical assessment of frequency loss does not account for limitations in human perceptual ability. Research suggests that under optimal conditions, we can effectively compare about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.

To examine this further, we simulated points from three different distributions: Uniform, Normal, and Exponential to determine under what circumstances log frequency scaling is performing better than linear frequency scaling. Figure~\ref{fig:logcolor} gives an overview of this simulation. In the top row tile plots with a color scheme based on linear frequency scaling is used, the bottom row shows the same data based on log frequency scaled colors.

For uniformly distributed data (right column) the choice of color scheme does not have a huge impact, but under both the normal and exponential distribution, log frequency scaling results in categories which account for more uniform proportions of the data (as can be seen by the color strips in the second and third row of Figure~\ref{fig:logcolor}). This is desirable because it provides more discernible frequency information. 
\begin{figure}[hbtp]
\centering
<<FrequencyCategories,echo=FALSE, out.width='\\textwidth', fig.width=8.5, fig.height=11>>=
options(stringsAsFactors=FALSE)
library(plyr)
library(ggplot2)
library(RColorBrewer)
set.seed(5032)
data <- rbind(data.frame(x=runif(1000, 0, 20), y=runif(1000, 0, 20), 
                         Freq=1, Distribution="Uniform"), 
              data.frame(x=rnorm(1000, 10, 3), y=rnorm(1000, 10, 3), 
                         Freq=1, Distribution="Normal"),
              data.frame(x=rexp(1000, .3), y=rexp(1000, .3), 
                         Freq=1, Distribution="Exponential"))
data <- data[which(abs(data$x-10)<10 & abs(data$y-10)<10),]
# qplot(data=data, x=x, y=y, geom="point")+facet_wrap(~Distribution)

data.bin <- data
data.bin[,1:3] <- round(data[,1:3])
data.bin <- ddply(data.bin, .(Distribution, x, y), summarise, Freq=sum(Freq))
data.bin <- data.bin[order(data.bin$Distribution, data.bin$Freq),]
# 
# 
# qplot(data=data.bin, x=Freq,geom="blank") + geom_density(aes(y=..scaled..),adjust=2) + facet_wrap(~Distribution, scales="free") + ylab("Density")
# 
# qplot(data=data.bin, x=Freq, geom="blank") + geom_density(aes(y=..scaled..), adjust=2) + facet_wrap(~Distribution,scales="free") + scale_x_continuous(trans="log", breaks=c(1, 5, 10, 20, 40, 60)) + ylab("Density")
#

get.upper <- function(x){
  num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][2])
  num <- sapply(num, function(i) substr(i, 1, nchar(i)-1))
  return(as.numeric(num))
}
get.lower <- function(x){
  num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][1])
  num <- sapply(num, function(i) substr(i, 2, nchar(i)))
  return(as.numeric(num))
}

labelsToValues <- function(x){
  return(data.frame(lower=get.lower(x), upper=get.upper(x)))
}

linearbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(x$Freq, 7)), label=as.character(cut(x$Freq,7)), id=1:nrow(x))))[,2:4])
linearbins$fill <- factor(linearbins$fill)

x <- subset(linearbins, Distribution=="Uniform" & fill==1)
linearbins2 <- ddply(linearbins, .(Distribution, fill),
                     function(x){
                       temp <- labelsToValues(x$label[1])
                       return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
                     })
linearbins2 <- ddply(linearbins2, .(Distribution), transform, height=height/sum(height))

logbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(log(x$Freq), 7)), label=as.character(cut(log(x$Freq),7)), id=1:nrow(x))))[,2:4])
logbins$fill <- factor(logbins$fill)

x <- subset(logbins, Distribution=="Uniform" & fill==1)
logbins2 <- ddply(logbins, .(Distribution, fill),
                  function(x){
                    temp <- exp(labelsToValues(x$label[1]))
                    newlabel <- paste(c("(", round(temp[1], 3), ",", round(temp[2], 3), "]"), sep="", collapse="")
                    return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
                  })

logbins2 <- ddply(logbins2, .(Distribution), transform, height=height/sum(height))

colors <- brewer.pal(9, "Blues")[-c(1:2)]

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, heights=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    nrow = ceiling(numPlots/cols)

    if (is.null(heights)) heights = unit(rep(1, numPlots), "null")
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = nrow)
  }

  if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout), heights=heights)))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

p1 <- qplot(data=linearbins, x=x, y=y, geom="tile", fill=fill)+facet_wrap(~Distribution) +
  scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
  ggtitle("(a) Joint Density: Linear Binning") +  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))

X <- data.frame(fill=with(linearbins, rep(fill, Freq)), Distribution = with(linearbins, rep(Distribution, Freq)))
X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
p2 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(1)) + 
  scale_fill_manual(values=colors, name="Frequency", guide="none") + 
  facet_grid(.~Distribution, scales="free") +
  xlab("Count") + theme_bw() + ylab("") +
  ggtitle("(b) Color Frequency: Linear Binning") + 
  theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
# linear binning
pdf(file="figure/p2.pdf", width=8, height=3)
p2
dev.off()

# p2 <- qplot(data=linearbins2, xmin=lower, xmax=upper, ymin=0, ymax=height*1000, fill=fill, geom="blank", asp=1.5) + 
#   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
#   scale_fill_manual("Tile Color", values=colors)+
#   facet_wrap(~Distribution, scales="free") + 
#   theme_bw() + scale_x_continuous("Bin Frequency") + 
#   scale_y_continuous("") + ggtitle("(b) Distribution of Frequency: Linear Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))

# p4 <- qplot(data=logbins2, xmin=lower, xmax=upper, ymin=0, ymax=1000*height, fill=fill,  geom="blank", asp=1.5) + 
#   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
#   scale_fill_manual("Tile Color", values=colors)+
#   facet_wrap(~Distribution, scales="free") + 
#   theme_bw() + scale_x_continuous("Bin Frequency") + 
#   scale_y_continuous("") + ggtitle("(c) Distribution of Frequency: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))

X <- data.frame(fill=with(logbins, rep(fill, Freq)), Distribution = with(logbins, rep(Distribution, Freq)))
X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
p4 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(NULL)) + 
  scale_fill_manual(values=colors, name="Frequency", guide="none") + 
  facet_grid(.~Distribution, scales="free") +
  xlab("Count") + theme_bw() + ylab("") +
  ggtitle("(c) Color Frequency: Log Binning") + 
  theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))

p3 <- qplot(data=logbins, x=x, y=y, geom="tile", fill=fill, asp=1.0)+facet_wrap(~Distribution) +
  scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
  ggtitle("(d) Joint Density: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))

multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
@
%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
\caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
\end{figure}

Frequency loss is much more difficult to normalize, as we must balance the partial loss of frequency information due to over-plotting with the total loss of frequency information that results if we use only a single bin to categorize the data. 
For the purposes of numerical assessment of frequency loss, we calculate the baseline frequency loss as the loss corresponding to a minimal bin width given by the smallest non-zero absolute difference of successive values in the original data set, as shown in Figure~\ref{binning}.

\paragraph{Traditional Scatterplots and Frequency Information:}
using a minimal bin width, a tile display is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin. Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency. The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information. 
By explicitly shading bins (or scaling size) according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points. This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha} d and e under the same framework as plots a-c.

By additionally increasing the bin width, we  provide increasingly higher-level summaries of the data by  smoothing over local structures. For small bin widths these structures are likely to be  {\it noise} inherent in any real data set, while removing structures associated with  larger bin widths likely masks real signal in the data.

\subsection{Recovery of Numerical Information}
% The total loss from reducing a value $x$ to bin center $x^\ast$ is at least as large as the loss from reducing points $x_i$ to $\overline x_i$, the numerical center of the binned data.
% % that's only true for the expected value - not for all individuals, but you don't need - we just use the partition.

As shown before, the overall loss partitions naturally into visual and numeric loss (see equation \ref{eqn:partition}). We can make use of this and
 recover some of the numerical information lost due to binning by storing the numerical center $\overline x$ instead of the bin center $x^\ast$ for use in subsequent calculations. Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.

<<recover,echo=FALSE,results='asis', eval=FALSE>>=
setwd("./figure")
set.seed(90)
x <- rnorm(100, 5, 2)
y <- rnorm(100, 5, 2)
d <- data.frame(x=x, y=y, c=1)
d <- d[order(d$x, d$y),]
print(xtable(head(d), digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)
dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
dbin <- dbin[order(dbin$x, dbin$y),]
print(xtable(head(dbin), digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')
names(d) <- c("x", "y", "Freq")
drecover <- unique(lossCalc(d, c(1,1), "standard", newData=TRUE)$NewData)
names(drecover) <- c("x", "y", "c")
drecover <- drecover[order(drecover$x, drecover$y),]
print(xtable(head(drecover), digits=c(0, 4, 4, 0), caption=paste('Recovered Data\\newline', nrow(drecover), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
setwd("../")
@
\begin{table}[hbtp]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 100 rows]{
\begin{tabular}{rrr}\hline
       x &      y & c \\ \hline
 -0.3843 & 5.6508 & 1 \\ 
 -0.2582 & 1.5497 & 1 \\ 
  1.0597 & 2.4308 & 1 \\ 
  1.6054 & 6.4125 & 1 \\ 
  1.6788 & 5.0260 & 1 \\ 
  2.2680 & 1.5877 & 1 \\ \hline
\end{tabular}}\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data, 49 rows]{
\begin{tabularx}{.7\textwidth}{XXc} \hline
  x & y & c \\ \hline
  0 & 6 & 1 \\ 
  0 & 2 & 1 \\ 
  1 & 2 & 1 \\ 
  2 & 5 & 2 \\ 
  2 & 6 & 3 \\ 
  2 & 2 & 1 \\ 
   \hline
\end{tabularx}
}\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Recovered Data, 49 rows]{
\begin{tabular}{rrc}\hline
       x &      y & c \\ \hline
 -0.3843 & 5.6508 & 1 \\ 
 -0.2582 & 1.5497 & 1 \\ 
  1.0597 & 2.4308 & 1 \\ 
  2.0890 & 4.8260 & 2 \\ 
  2.1526 & 6.4341 & 3 \\ 
  2.2680 & 1.5877 & 1 \\ \hline
\end{tabular}
}\end{minipage}
\caption{\label{RecoveryComparison}Six rows from Original, Binned, and Recovered Data Tables, with data storage sizes}
\end{table}

Information recovery comes with a two-fold cost: the additional storage of numeric bin centers triples the amount of information stored, and the computational cost to compute numeric bin centers is O($nm^2$), where $n$ is the number of data points and $m^2$, the total number of bins.
O($nm^2$) $\approx$ O($n$), as $m$ is generally negligible compared to $n$.

In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable. Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning. %but creates an additional complexity in that the geometric center of the tile (the visual bin center) does not correspond to the stored, binned data. 
% The random binning algorithm serves as a less computationally intensive, unbiased alternative for 
% data sets which are too large to partition the loss in this manner.
% \hh{is that true, or does it require the expected value to be the visual center?}

\section{Discussion and Examples}
\subsection{Comparison of Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
LossTable <- NULL
# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
bins <- rbind(c(2,2), c(5,25))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
    qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
#     qplot(G, SO, fill = log10(Freq), geom="tile", data = lMean$NewData, 
#       main=paste("Standard Binning (", binning[1], ", ", binning[2],")", sep="")) + 
#       scale_fill_gradient(low="#56B1F7", high="#132B43", guide="none", trans="log") + theme_bw()
#     ggsave(paste("./figure/Standard", binning[1], binning[2], ".pdf", sep=""), width=4.5, height=4.5, units="in")
  return(rbind(
    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$LossAll), 
    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lRandom$LossAll)))
}

LossTable <- do.call("rbind", lapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
LossTable <- as.data.frame(LossTable)
names(LossTable) <- c("binX", "binY", "Function", "G", "SO", "Freq")
tab <- xtable(LossTable[order(LossTable[,3], LossTable[,1]),], digits=c(0, 0, 0, 0, 8, 8, 8))
@

At the example of the pitching data, we will %consider the variables Games and Strikeouts, which are strongly correlated. This example 
highlight an important advantage of random binning. The two tile plots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident: while the standard binning algorithm has rounding artifacts that are clearly visible for even bin widths, the random algorithm produces a graph that better represents the true shape of the data.
\begin{figure}[hbtp]
\centering
\subfloat[(2, 2) tiling of G, SO data]{
\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
}\\
\subfloat[(5, 25) tiling of G, SO data]{
\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
}
\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chose bin width is not a multiple of the data resolution.}
\end{figure}

In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable tile plots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.

The spatial loss is a result of our desire to reduce and summarize the data until it is more manageable in size, and hence is a factor in our decision to bin the data in the first place. More importantly, frequency loss has a larger effect on the utility of the graphical data summary. Figure~\ref{fig:binning-comparison} shows clearly that the standard algorithm can produce misleading graphical summaries. In order to quantify the advantages of each algorithm, we compare the spatial and frequency loss at different bin widths in $x$ (Games). The random algorithm has higher spatial loss than the standard algorithm, but in many cases the tile graphs are visually indistinguishable. The numerical loss is a function of the distribution and bin width, and once the bin width is chosen, it is entirely fixed. As the visual loss is the difference between the numerical center of points in the bin and the geometric center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm). We would not expect to notice the difference in the spatial loss in either algorithm, so long as the bin width is constant. The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins.

<<LossComparison1, echo=FALSE, eval=T, fig.cap='Spatial Loss increases with bin width for each variable. The random algorithm has higher spatial loss than the standard algorithm.', fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth', fig.align='center', fig.pos='H'>>=

# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                      password="R R0cks", port=3306, dbname="baseball", 
#                      host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# d1 <- dbData(pitch, vars=c( "G", "SO"))
# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
# 
# library(multicore)
# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
# 
# lossesRdm <- cbind(bins, lossesRdm)
# lossesStd <- cbind(bins, lossesStd)
# 
# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)

losses <- read.csv("../dbData/data/losses.csv", stringsAsFactors=FALSE)

qplot(x=x, y=TotalLoss.G*100, group=interaction(y, Algorithm), data=subset(losses, y<=10), geom="line", colour=Algorithm, ylab="Percent Spatial Loss", xlab="Bin  Width  (Games)")+ theme_bw() + theme(legend.position="bottom") + ggtitle("Spatial Loss:  Baseball Games") + guides(colour = guide_legend(override.aes = list(size = 2))) + facet_wrap(~Algorithm)
# qplot(x=y, y=VisLoss.SO*100, group=interaction(x, Algorithm), data=subset(losses, x==1), geom="line", colour=Algorithm, ylab="Percent Visual Loss", xlab="Bin Width (Strike Outs)") + theme_bw() + theme(legend.position="bottom") + ggtitle("Visual  Loss  with  increasing  Bin  Width: Baseball Strike Outs\n Games not binned") + guides(colour = guide_legend(override.aes = list(size = 2)))
@
The standard algorithm is preferable because it has lower spatial loss, but the two algorithms are very similar when we compare frequency loss. The random algorithm does have more variability, but the frequency loss values are much more similar, and when bins are of size two in either or both dimensions, the random algorithm has lower frequency loss than the standard algorithm. This holds for bins which are powers of two, as long as the random binning algorithm is applied iteratively to take advantage of the constant loss property noted above.  
<<LossComparison, echo=FALSE, fig.cap="Frequency Loss increases as bin width increases in Games and Strike Outs. The random binning algorithm has slightly higher frequency loss as bin width grows large, but the loss is comparable when bin widths are small.", fig.width=8, fig.height=4, out.width='\\linewidth', out.height='.5\\linewidth', fig.pos='H'>>=
# 
# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                      password="R R0cks", port=3306, dbname="baseball", 
#                      host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# d1 <- dbData(pitch, vars=c( "G", "SO"))
# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 15)))
# 
# library(multicore)
# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
# 
# lossesRdm <- cbind(bins, lossesRdm)
# lossesStd <- cbind(bins, lossesStd)
# 
# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
losses <- read.csv("../dbData/data/losses.csv", stringsAsFactors=FALSE)

qplot(x=y, y=TotalLoss.LogFreq, group=interaction(x, Algorithm), data=losses, geom="line", colour=Algorithm, ylab="Frequency Loss", xlab="Bin Width (Strike Outs)")+ facet_grid(.~Algorithm) + scale_colour_discrete(guide="none") + theme_bw() + ggtitle("Frequency Loss: Baseball Games and Strike-Outs")
@

\subsection{Big Data: Airline Departure Times}
The Federal Aviation Association (FAA) requires all airlines based in the United States to report   details for every single flight. These are published online by the Bureau of Transportation Services at \url{http://www.transtats.bts.gov/DataIndex.asp}. 
Every day there are about 16,000 flights across the United States adding up to almost 6 Million flights a year. Scheduled and actual departure times for all flights in 2011 make up --in uncompressed form-- a file of about 450 MB. A comparison of scheduled and actual departure times allows us an investigation of on-time performance of air carriers.

%We obtained airline scheduled departure times and actual departure times from the Department of Transportation for the year 2011. These times do not include actual departure date, however, so a flight scheduled to leave at 23:00 which is delayed by two hours is recorded as leaving at 01:00. The uncompressed data file is about 450 MB, and consists of two columns of almost six million records.

Figure~\ref{airline-scatter} shows examples of two scatterplots of scheduled versus actual departure times. The plot on the left shows a sample of one million of those records. Even while using alpha blending this results in a severely over-plotted graph. On the right is a tile plot of the complete data binned at 1-minute intervals. This does not have any spatial loss, as the times are recorded by minute. The 1-minute binning also reduces the data file to 176,384 individual records, less than 3\% of the original data. Both plots show the same big picture patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the line of identity.  
Scheduled departure times past 6 am in the morning are much more common than earlier flights. It is much more likely for a  flight to be delayed than to leave early,
leading to the wash-out effect above the line, that is getting thinner with increasing delays. The range of delays on usual days starts at about one hour at 6 am and increases during the day to about 2 hours. The few number of flights before 6 am are also visible in both plots.
The triangle of observations on the bottom right visible in both plots is nothing but an artifact of the data collection consisting of flights that are scheduled before midnight, but are delayed to departures past midnight. The cloud of outliers halfway between the two main structures is potentially interesting, since no immediate explanation comes to mind, and would be worthy of a follow-up investigation.
 What is not apparent in the plot on the left, is some fine-level structure that the plot based on all of the data shows. A close inspection of the plot on the right hand side reveals  darker colored vertical lines at 30 minute intervals. It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.

\begin{figure}[hbtp]\centering
\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{./images/MillionSample.pdf}}\hfil
\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{./images/AirlineStdBinning1min.pdf}}
\caption{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011. The plot on the left is based on a sample of the data, the plot on the right shows all flights. The big patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}
\end{figure}

\begin{figure}[hbtp]\centering
\includegraphics[keepaspectratio=true, height=.43\textheight]{./images/AirlineStdBinning5mins.pdf}
\includegraphics[keepaspectratio=true, height=.43\textheight]{./images/AirlineStdBinning15mins.pdf}
\caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b. 15-minute bins produce an even more coarse summary of the data.}\label{airline5mins-binning}
\end{figure}
Binning data by five-minute intervals produces a more high-level summary of the relationship between actual and scheduled departure time, though it necessarily obscures some of the finer details. In addition, binning data by five minute intervals reduces the size of the data set to a much more manageable 19,787 observations  which can be easily manipulated on probably any modern computer. Binning by 15-minute intervals reduces the data set to a nearly-trivial 3,575 observations, but the graphical summary becomes granular and less appealing at that resolution.

<<airlinesetup,echo=FALSE, eval=FALSE>>=
setwd("./data/")
# airline <- do.call("rbind", lapply(list.files(), function(i) read.csv(i)))
# write.csv(airline, "airline.csv", row.names=FALSE)
airline <- read.csv("airline.csv")
na.departure <- (is.na(airline[,8]) + is.na(airline[,9]))> 0
airline <- airline[!na.departure,8:9]

airline$CRS_DEP_TIME <- floor(airline$CRS_DEP_TIME/100)*60 + airline$CRS_DEP_TIME%%100
airline$DEP_TIME <- floor(airline$DEP_TIME/100)*60 + airline$DEP_TIME%%100

library(lubridate)
convert.time <- function(x){
  floor(x/60) + (x%%60)/60
}

library(multicore)

airline.freq <- ddply(airline, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=nrow(i)))
write.csv(airline.freq, "ReducedAirlineData.csv", row.names=FALSE)
airline.freq <- read.csv("ReducedAirlineData.csv")
airline.freq.plot <- airline.freq
airline.freq.plot[,1] <- convert.time(airline.freq.plot[,1])
airline.freq.plot[,2] <- convert.time(airline.freq.plot[,2])
airline.freq.plot <- airline.freq.plot[order(airline.freq.plot$Freq),]
airline.dummy <- airline.freq.plot[1,]
airline.dummy[1,1] <- NA
qplot(data=airline.freq.plot, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="blank", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (1 minute bins)") + geom_rect(aes(xmin=CRS_DEP_TIME-3/60, xmax=CRS_DEP_TIME+3/60, ymin=DEP_TIME-3/60, ymax=DEP_TIME+3/60, fill=Freq), alpha=I(.5)) + geom_rect(data=airline.dummy, aes(xmin=CRS_DEP_TIME-3/60, xmax=CRS_DEP_TIME+3/60, ymin=DEP_TIME-3/60, ymax=DEP_TIME+3/60, fill=Freq), aes.inherit=FALSE) + theme_bw()+ theme(legend.position="right") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
ggsave("AirlineStdBinning1min.pdf", width=7.5, height=6, dpi=2304)

library(ggplot2)
library(dbData)
air.sample <- airline.freq.plot[sample(1:nrow(airline.freq.plot), 1000000, replace=TRUE, prob=airline.freq.plot$Freq),1:2]

qplot(data=air.sample, x=CRS_DEP_TIME, y=DEP_TIME, geom="point", alpha=I(.05), xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Sampled Airline Departure Times") + theme_bw()+ theme(legend.position="right") + scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
ggsave("MillionSample.pdf", width=6, height=6, dpi=576)

air.binned5 <- binStd(airline.freq, c(5, 5))
air.binned5 <- ddply(air.binned5, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(convert.time(unique(i[,1:2])), Freq=sum(i$Freq)))
write.csv(air.binned5, "airlineBin5Std.csv", row.names=FALSE)

# air.binned5.rdm <- binRdm(cbind(airline, Freq=1), c(5, 5))
# air.binned5.rdm2 <- ddply(air.binned5.rdm, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(convert.time(unique(i[,1:2]), Freq=sum(i$Freq)))
# write.csv(air.binned5.rdm2, "airlineBin5Rdm.csv", row.names=FALSE)

air.binned5 <- read.csv("airlineBin5Std.csv")
qplot(data=air.binned5, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (5 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
ggsave("AirlineStdBinning5mins.pdf", width=6, height=6, dpi=576)
# 
# qplot(data=air.binned5.rdm2, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (5 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B4FB", high="#183347", guide="legend", trans="log")
# ggsave("AirlineRdmBinning5mins.png", binned, width=6, height=6)

# air.binned <- data.frame(apply(airline, 2, function(i) 15*round(i/15, 0)))
# air.binned.rdm <- binRdm(airline.freq, c(15, 15))
# air.binned2 <- ddply(air.binned, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=sum(i$Freq)))
# write.csv(air.binned2, "airlineBin15.csv", row.names=FALSE)
air.binned <- read.csv("airlineBin15.csv")
air.binned2 <- air.binned
air.binned2[,1] <- convert.time(air.binned[,1])
air.binned2[,2] <- convert.time(air.binned[,2])
air.binned.rdm <- read.csv("airlineBin15Rdm.csv")
air.binned2.rdm <- air.binned.rdm
air.binned2.rdm[,1] <- convert.time(air.binned.rdm[,1])
air.binned2.rdm[,2] <- convert.time(air.binned.rdm[,2])

# air.binned2.rdm <- ddply(air.binned.rdm, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=sum(i$Freq)))
# write.csv(air.binned2.rdm, "airlineBin15Rdm.csv", row.names=FALSE)

binned <- qplot(data=air.binned2, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
ggsave("AirlineStdBinning15mins.pdf", binned, width=6, height=6, dpi=576)

# binnedRdm <- qplot(data=air.binned2.rdm, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
# ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=600)
@
\section{Conclusion}
Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into tile plots which are more appropriate for the human visual system. The loss due to binning reduces the data set to a more manageable size, but some of that loss can be recovered with additional computational investment.

We have presented two algorithms for binning; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient. The choice of algorithm is a trade-off between speed and accuracy, and may be highly data set dependent. It is possible that there are computational modifications which could improve the accuracy of the random algorithm, but these modifications are likely to require additional computational operations.

% \begin{subsection}{Numerical Comparison of Binning Algorithms}
<<PrintTable,echo=F,include=FALSE>>=
  print(tab, include.rownames=FALSE)
@
% \end{subsection}

\begin{appendix}
\section{Mathematical Considerations}
\subsection{Partition of Spatial Loss}\label{proof:partition}
We want to show that spatial loss can be written as a partition of the form 
\[L_s(x) = 1/S^2_{\emptyset}\sum_i S^2_i = 1/S^2_{\emptyset}\left[\sum_i L^N_i + \sum_i L^V_i\right]\]

For that let us consider the values $\{x_{i}: b(x_i) = x^\ast\}$ in a single bin with center $x^\ast$.
The numerical center of these points is given as  $1/n \sum x_i = \overline x$.
Then the total loss for each element $x_i$ in this bin is:

\begin{align*}
S_i^2 = \sum (x_i - x_i^\ast)^2 & = \sum (x_i - \overline x + \overline x - x_i^\ast)^2\\
& = \sum(x_i-\overline x)^2 + 2\sum (x_i-\overline x)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
& = \sum(x_i - \overline x)^2 + 2\sum(0)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
& = \sum(x_i - \overline x)^2 + \sum(\overline x-x_i^\ast)^2\\
& = L^N_i + L^V_i
\end{align*}
Adding over all elements in the bin and all bins gives the formula above.

\subsection{Constancy of Loss under Binary Splits}\label{proof:constantloss}
We want to show that under random binning the loss due to doubling of bin width is deterministic.

Let us consider the contribution $L_i$ of a single point $x_i$ to the loss when starting with the minimal bin width, i.e. each unique point is in a separate bin:
doubling the bin width leads to two  scenarios: 
a point is either at  the bin center or directly between two bins.
In the case that the point is at the bin center, its contribution to the total overall loss is zero,  $L_i = 0$, and the associated probability $p_i = 1$.

Alternately, when the point is exactly half-way between two bin centers, the point is attributed with probability $p_i = \frac{1}{2}$ to either bin, leading to a loss of $L_i = \frac{1}{2} c_i$ regardless of which bin is assigned. Hence, in this case, loss is also entirely independent of which bin is assigned.
\end{appendix}
\bibliographystyle{asa}
\bibliography{references}
\end{document}

### 2013-03-01 Paper.Rnw
some calls to ggsave removed

@@
 @
-<<scatterplots, dependson='setup', echo=F, fig.keep='none', results='hide'>>=
+<<scatterplots, dependson='setup', echo=F, results='hide'>>=
 qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
@@
 Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
-<<binningDemo,echo=F, fig.keep='none', cache=T>>=
+<<binningDemo,echo=F, results='none', cache=T>>=
       library(mvtnorm)
@@
                  xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
-      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
+      #ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
 
@@
         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
+      #ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
+
 @
@@
   \centering
-  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
+  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/binningDemo1.pdf}}
 %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
@@
   \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
-  {./figure/RandomBinningDataPoints2.pdf}}
+  {./figure/RandomBinDemo.pdf}}
   \caption{Points are assigned to bin (0, 20) and bin (10,20).
@@
 \subsection{Random Binning Algorithm}
-<<RandomBinDemo,echo=F,include=FALSE,fig.keep='none', cache=TRUE>>=
+<<RandomBinDemo,echo=F,include=FALSE,results='hide',  cache=TRUE>>=
 nl <- ncol(d.rand)
@@
   annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
-ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
+#ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
 @
@@
     \centering
-		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./Binning-DataVisualNumericalCenter.pdf}}
+		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{binningDemo2}}
     \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}

### 2013-03-06 Paper.Rnw
Trying out GIT
next commit
Fixing pictures, etc. May depend on newest dbData package from github.

@@
 
+%Karsten was here
+%Heike, too
+
 \newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
+% \SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
 \maketitle
@@
 library(dbData)
-#library(multicore)
 library(devtools)
@@
 library(RColorBrewer)
+opts_chunk$set(concordance=TRUE, tidy=TRUE, cache=TRUE, message=FALSE, warning=FALSE)
 
@@
 @
-<<scatterplots, dependson='setup', echo=F, results='hide'>>=
+<<scatterplots, dependson='setup', echo=F, include=FALSE>>=
+library(hexbin)
 qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
@@
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
-  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
+  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
   theme(legend.position="bottom")
@@
 # ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))
-
 @
@@
 Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
-<<binningDemo,echo=F, results='none', cache=T>>=
+<<binningDemo, echo=FALSE, include=FALSE>>=
       library(mvtnorm)
@@
 \subsection{Random Binning Algorithm}
-<<RandomBinDemo,echo=F,include=FALSE,results='hide',  cache=TRUE>>=
+<<RandomBinDemo,echo=F,include=FALSE,cache=TRUE>>=
 nl <- ncol(d.rand)
@@
 
-loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
+loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
 
@@
 \end{eqnarray}
-<<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
+
+<<smoothing, echo=FALSE, include=FALSE, eval=FALSE>>=
+library(plyr)
+library(ggplot2)
+library(reshape)
 set.seed(53022)
@@
 
-ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
+ggplot()+
+  geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + 
+  geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + 
+  theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
 ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
@@
 \centering
-<<FrequencyCategories,echo=FALSE, out.width='\\textwidth', fig.width=8.5, fig.height=11>>=
+<<FrequencyCategories, echo=FALSE, out.width='\\linewidth', fig.width=8.5, fig.height=11>>=
 options(stringsAsFactors=FALSE)
@@
 # linear binning
-pdf(file="figure/p2.pdf", width=8, height=3)
-p2
-dev.off()
+# pdf(file="figure/p2.pdf", width=8, height=3)
+# p2
+# dev.off()
 
@@
 multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
-#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
+# ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
 @
@@
 <<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
+# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
+#                    password="R R0cks", port=3306, dbname="baseball", 
+#                    host="headnode.stat.iastate.edu")
+# pitch <- new("dataDB", co=connect, table="Pitching")
+# 
+# d1 <- dbData(pitch, vars=c( "G", "SO"))
+
 LossTable <- NULL
-# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
 bins <- rbind(c(2,2), c(5,25))
@@
   qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
-        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
-        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
-    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
-#     qplot(G, SO, fill = log10(Freq), geom="tile", data = lMean$NewData, 
-#       main=paste("Standard Binning (", binning[1], ", ", binning[2],")", sep="")) + 
-#       scale_fill_gradient(low="#56B1F7", high="#132B43", guide="none", trans="log") + theme_bw()
-#     ggsave(paste("./figure/Standard", binning[1], binning[2], ".pdf", sep=""), width=4.5, height=4.5, units="in")
+      main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + 
+    facet_grid(.~Algorithm) +
+    scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", 
+                        breaks=c(1, 5, 50, 500, 5000)) + 
+    theme_bw() + theme(legend.position="bottom")
+  ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), 
+         width=8, height=5, units="in", dpi=1200)
+
   return(rbind(
-    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$LossAll), 
-    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lRandom$LossAll)))
+    data.frame(binX = binning[1], binY = binning[2], Function = "Random", loss=lRandom$Loss[,5:7]), 
+    data.frame(binX = binning[1], binY = binning[2], Function = "Standard", loss=lMean$Loss[,5:7])))
 }
@@
 LossTable <- as.data.frame(LossTable)
-names(LossTable) <- c("binX", "binY", "Function", "G", "SO", "Freq")
+names(LossTable2) <- c("binX", "binY", "Function", "G", "SO", "Freq")
 tab <- xtable(LossTable[order(LossTable[,3], LossTable[,1]),], digits=c(0, 0, 0, 0, 8, 8, 8))
@@
 <<PrintTable,echo=F,include=FALSE>>=
-  print(tab, include.rownames=FALSE)
+#   print(tab, include.rownames=FALSE)
 @

### 2013-03-06 Paper.Rnw
Fix extra 2 in the LossFunction() call.

@@
 computeLoss <- function(binning, data){
-  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
+  lRandom <- lossCalc(data, binning=binning, type="random", newData=TRUE)
   lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
@@
 LossTable <- as.data.frame(LossTable)
-names(LossTable2) <- c("binX", "binY", "Function", "G", "SO", "Freq")
+names(LossTable) <- c("binX", "binY", "Function", "G", "SO", "Freq")
 tab <- xtable(LossTable[order(LossTable[,3], LossTable[,1]),], digits=c(0, 0, 0, 0, 8, 8, 8))

### 2013-03-06 Paper.Rnw
Fixed std/random "plaid" graph.

@@
 LossTable <- NULL
-bins <- rbind(c(2,2), c(5,25))
+bins <- rbind(c(2,10), c(5,25))
 computeLoss <- function(binning, data){
@@
   datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
+  datplot <- ddply(datplot, .(G, SO, Algorithm), summarise, Freq=sum(Freq))
   qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
@@
 \subfloat[(2, 2) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd210.pdf}
 }\\

### 2013-03-07 Paper.Rnw
Fixed figure problems I caused during the meeting yesterday.

@@
                  xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
-      #ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
+      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
 
@@
         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-      #ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
+      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
 
@@
   \centering
-  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/binningDemo1.pdf}}
+  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
 %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
@@
     \centering
-		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{binningDemo2}}
+		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter}}
     \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
@@
 \subsection{Comparison of Binning Algorithms}
-<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
+<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none'>>=
 # connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
@@
 \subfloat[(2, 2) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd210.pdf}
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{figure/RandomStd210.pdf}
 }\\
 \subfloat[(5, 25) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{figure/RandomStd525.pdf}
 }

### 2013-03-13 Paper.Rnw
Karsten's changes

--- 2013-03-06 Paper.Rnw
@@
 \usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
-\graphicspath{{figure/}}
+\graphicspath{{images/}}
 \usepackage{epstopdf}
@@
 
-%Karsten was here
-%Heike, too
-
 \newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike
@@
 %opening
-\title{Binning Strategies and Related Loss for Graphics of Large Data}
-\author{Susan VanderPlas$^1$, Karsten Maurer$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
+\title{Binning Strategies and Related Loss for Large Data}
+\author{Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
 
 \begin{document}
-% \SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
+\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
 \maketitle
@@
 Dealing with the data deluge of the Big Data Age is both exciting and challenging.
-Large data demands require us to re-think strategies of visualizing data.
-Binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much.
+The demands of large data require us to re-think strategies of visualizing data.
+Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much.
 This comes with the price of loss of information inherent to any binning scheme.
 In this paper we discuss properties of two commonly used binning algorithms.
-We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the frame work of a simulation and two case studies.
+We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the framework of a simulation and two case studies.
 \end{abstract}
@@
 library(dbData)
+#library(multicore)
 library(devtools)
@@
 library(RColorBrewer)
-opts_chunk$set(concordance=TRUE, tidy=TRUE, cache=TRUE, message=FALSE, warning=FALSE)
+# setwd("/home/susan/Dropbox/GraphicsGroup/")
+setwd("/Users/heike/Dropbox/papers/")
+
+setwd("./dbaccess/NewWriteUp/")
 
@@
 @
-<<scatterplots, dependson='setup', echo=F, include=FALSE>>=
-library(hexbin)
+<<scatterplots, echo=F,include=F,eval=F>>=
 qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
-  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
+  ggsave("./images/OverplottingAlpha.pdf", width=5, height=5)
 qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
-  ggsave("./figure/Overplotting.pdf", width=5, height=5)
+  ggsave("./images/Overplotting.pdf", width=5, height=5)
 qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
-  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
+  ggsave("./images/OverplottingCircles.pdf", width=5, height=5)
 qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
-  ggsave("./figure/HexBinning.pdf", width=5, height=5)
+  ggsave("./images/HexBinning.pdf", width=5, height=5)
 d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
@@
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
-  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
+  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
   theme(legend.position="bottom")
-  ggsave("./figure/BinScatter.pdf", width=5, height=5)
+  ggsave("./images/BinScatter.pdf", width=5, height=5)
 
@@
 # ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))
+
 @
@@
 
-\citet{Friedman97} proposed in his paper on datamining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it".
+\citet{Friedman97} proposed in his paper on data mining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it".
 The same holds for visualizations.
-With a 100-1000 fold increase in the amount of data, the utility of some of  graphs changes dramatically, with some of our most-used tools such as the scatterplot deteriorating quickly \citep{gold}.
+With a 100-1000 fold increase in the amount of data, the utility of some of our most used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.
 
 % binning as a way out
-Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots, which utilize  raw data. 
-This fact is being exploited to make scatterplots again useful instruments for large data settings  by employing binning strategies \citep{gold}.
+Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots that utilize raw data. 
+By using binning strategies and the principles for displaying information in area plots, scatterplots can again become useful instruments for large data settings \citep{gold}.
 
@@
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
-The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball players as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
-The number of games (G) played in a season is plotted versus number of strikeouts (SO), i.e.
-a batter receives three strikes during his time at bat.
-Pitchers want to throw as many strikeouts as possible.
+The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
+The number of games played in a season is plotted versus number of strikeouts a pitchr threw over the course of a season.
 While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
@@
 \begin{figure}[hbtp]
-  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/Overplotting.pdf}}
-  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/OverplottingCircles.pdf}} 
-	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{figure/OverplottingAlpha.pdf}}
-
-  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{figure/HexBinning.pdf}}
-  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{figure/BinScatter.pdf}}
+  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/Overplotting.eps}}
+  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingCircles.eps}} 
+	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingAlpha.eps}}
+
+  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/HexBinning.eps}}
+  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/BinScatter.eps}}
 
@@
 Histograms are an example of plots which use binned variables.
-To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to create a two-dimensional graph which uses some other attribute (color, 3D rendering) to provide joint density information.
-A {\it tile plot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.
+To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tesselated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile.
+A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.
 
@@
 When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}.
-Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to tile plots or $\alpha$-blended scatterplots in a more smooth, continuous fashion.
+Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or $\alpha$-blended scatterplots in a more smooth, continuous fashion.
 These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.
 
-As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonal tile plot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e).
+As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e).
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The tile plot is less prone to the Hermann-grid illusion than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
+The binned scatterplot is less prone to the Hermann-grid illusion than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
 
@@
 % \end{figure}
-Only in the tile and the bubble plot the inner structure of the data becomes apparent: 
+Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: 
 the joint density consists two distinct ridges following two lines with very different slopes.
-The lower slope corresponds to the --modern-- average strike out rate of pitchers of just under one strike-out per game.
+The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game.
 The other line has a slope of about four times that rate.
@@
 
-For extremely large data sets, tile plots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be  rendered separately.
+For extremely large data sets, binned scatterplots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be  rendered separately.
 
@@
 Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
-<<binningDemo, echo=FALSE, include=FALSE>>=
+<<binningDemo,echo=F,eval=F,include=F,cache=T>>=
       library(mvtnorm)
@@
                  xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
-      #ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
+      ggsave("./images/Binning-DataAndBin.pdf", width=5, height=5, units="in")
 
@@
         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-      #ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
-
+      ggsave("./images/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
 @
@@
   \centering
-  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/binningDemo1.pdf}}
+  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./images/Binning-DataAndBin.eps}}
 %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
-%   {./figure/RandomBinningDataPoints1}}
+%   {./images/RandomBinningDataPoints1}}
   \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
-  {./figure/RandomBinDemo.pdf}}
-  \caption{Points are assigned to bin (0, 20) and bin (10,20).
+  {./images/RandomBinningDataPoints2.eps}}
+  \caption{Points are assigned to bins centered at (5, 25) an (15,25).
 Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
@@
 \subsection{Random Binning Algorithm}
-<<RandomBinDemo,echo=F,include=FALSE,cache=TRUE>>=
+<<RandomBinDemo,echo=F,include=FALSE,eval=F>>=
 nl <- ncol(d.rand)
@@
   annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
-#ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
+ggsave("./images/RandomBinningDataPoints2.pdf",width=5, height=5)
 @
@@
 % In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
-This method is easily extensible to non-uniform bin sizes and can also map $x$ to more than two bins, unlike the standard algorithm.
-
-To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
-Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
-1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
+This method is easily extensible to also map $x$ into one of more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
+
+%To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
+%Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
+%1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
 
@@
 The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
-``Linear" binning  \citep{martin-gold} is a --computationally intensive-- alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
+``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
 With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
@@
 \begin{itemize}
-\item {\it Numerical Loss} occurs when the data set is reduced  to bin centers, i.e. this is the difference between a data point $(x,y)$ and the corresponding bin center $(x^\ast, y^\ast)$.
+
+\item {\it Visual Loss} is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
+This visual center will generally different from the numerical center, or the mean of the data corresponding to that bin, which can be thought of as the center of mass.
+Thus visual loss is the difference between the visual center $(x^\ast, y^\ast)$ and the numerical center $(\bar{x}, \bar{y})$.
+This causes some bias in the perception.
+
+\item {\it Numerical Loss} occurs when points in the data set are reduced to a single point at the mean, i.e.
+this is the collective difference between all data points $(x_i,y_i)$ and their corresponding numerical centers $(\bar{x}, \bar{y})$.
 
@@
 Different rendering methods for the frequency information will lead to different losses.
-For the remainder of the paper we will assume that we are using colors in tile plots to represent this information.
+For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent this information.
 The number of colors that can be differentiated and mapped back to frequency information accurately is fairly limited.
-Note that even though this loss will turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where there is no frequency information, so that points which are over-plotted are not at all distinguishable.
+Note that even though this loss will turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where frequency information is masked in large data situations due to over-plotting of points.
 
@@
 % , so we must balance perceptual limits of frequency information with the limits of overplotting single points.
-
-\item {\it Visual Loss}: While the visual loss is most salient, it is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
-This will generally different from the center of the data corresponding to that bin as given by the mean.
-This causes some bias in the perception.
 
@@
 
-\noindent
-Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
-
-<<tablelossexample, echo=FALSE, results='asis',cache=FALSE>>=
-loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
-loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
-loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
-loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
-
-loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-
-table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
-                               c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
-                               c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
-                               c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
-                               c(binwidth=1, loss1[5:7]*100, loss1[8])))
-names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
-print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
-
-@
-
 We explore the different types of loss in the remainder of this section in more detail.
-
-\subsection{Loss in Spatial Dimensions}
-Reducing data points $x$ to bin centers $\hat x$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
-Let us denote the difference between point $x_i$ from its bin center as $S_i = \left|x_i-b(x_i)\right|$ for each point in the data set $i=1, ..., N$.
+For clarity we will again be specifying loss using notation for a one dimensional binning process.
+This notation can then be extended to higher dimensions in applications.
+
+\subsection{Numerical and Visual Loss}
+Reducing data points $\mathbf{x}$ to the visual bin centers $x^\ast$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
+Let us denote the difference between point $x_i$ from its bin center as $S_i = \left|x_i-b(x_i)\right|$ for each point in the data set $i=1, ..., n$.
 The total loss from binning an entire data set is defined as
 
-\[L_s(x) = 1/S^2_{\emptyset} \cdot \sum_{i=1}^{N} S^2_i\] 
+\[L_s(\mathbf{x}) = 1/S^2_{\emptyset} \cdot \sum_{i=1}^{n} S^2_i\] 
 where $S_\emptyset^2$ is the loss which results when the entire data set is reduced to a single bin,
-$S^2_{\emptyset} = \sum_i (x_i-\tilde{x})^2$ for bin center $\tilde{x}$.
-The loss depends on the choice of bin center: candidates for bin centers are the mean of the points within the bin or the visual center of the bin. 
-If we consider the bin center to be the numerical mean, then the loss is the sum of the variance of the data in each bin, compared to the variance of the data set as a whole, and $L_s(x) \in [0,1]$.
-
-The disadvantage to this choice is that this introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of he symbol representing the data.
-Therefore we distinguish two contributions to this loss:
-Numerical loss ($L^\text{N}$) results from the binning process as points are reduced to the numerical mean center, and visual loss ($L^\text{V}$) is due to the difference between the mean of points within the bin and the visual center of the bin.
+$S^2_{\emptyset} = \sum_i (x_i-\tilde{x})^2$ for the single bin center $\tilde{x}$.
+%The loss depends on the choice of bin center: candidates for bin centers are the mean of the points within the bin or the visual center of the bin. 
+If the visual bin centers are the same as the numerical centers for each bin, then any loss could be strictly attributed to difference between points and the bin centers. 
+The problem is that the visual center and the numerical center for bins will rarely match, which introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of the symbol representing the data.
+Therefore we distinguish two contributions to the total loss.
+Numerical loss results from the binning process as points are reduced to the numerical center point and is thus the sum of numerical losses for individual points ($\sum_{i=1}^n L_i^\text{N}$).
+Visual loss is due to the difference between the numerical centers and the visual centers and is thus the sum of visual losses for each bin ( $\sum_{j=1}^{N_x} L_j^\text{V}$).
 This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.
@@
 \begin{eqnarray}\label{eqn:partition}
-L_s(X) =  1/S^2_{\emptyset} \cdot \left(\sum_i L^N_i + \sum_i L^V_i\right)
+L_s(\mathbf{x}) =  1/S^2_{\emptyset} \cdot \left(\sum_i^n L^N_i + \sum_j^{N_x} L^V_j\right)
 \end{eqnarray}
-
-<<smoothing, echo=FALSE, include=FALSE, eval=FALSE>>=
-library(plyr)
-library(ggplot2)
-library(reshape)
+<<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
 set.seed(53022)
@@
 
-ggplot()+
-  geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + 
-  geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + 
-  theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
-ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
+ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
+ggsave("./images/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
 @
@@
 %The expected total loss for either algorithm depends on the distribution the underlying data. 
-
 %The total loss due to binning is $\sum_{i=1}^n (x_i - b(x_i))^2$, where in the standard algorithm, $b(x)$ maps $x$ to the closest bin center, and in the random algorithm, $b(x)$ is assigned probabilistically.
 
-\paragraph{Loss in the Random Assignment Algorithm:}
-An interesting feature of the random binning algorithm is that the loss that we experience by doubling the bin width is deterministic, i.e.
-independent even though the assignment of bins is not.
+It is worth noting that the loss function is a monotonically increasing function of binwidth.
+An interesting feature of the random binning algorithm is that the increase in loss that we experience by doubling the bin width is deterministic, even though the assignment of bins is not.
 This holds for repeated doubling of the bin width.
@@
 
-The loss as a function of bin width for both algorithms is shown in Figure~\ref{fig:LossComparison1}.
+The loss as a function of bin width was computed for both algorithms for the baseball data and is shown in Figure~\ref{fig:LossComparison1}.
 
@@
     \centering
-		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{binningDemo2}}
-    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
+		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./Binning-DataVisualNumericalCenter.eps}}
+    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./images/FreqLoss.eps}}
 \caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.
@@
 \subsection{Loss of Frequency Information} 
-Loss of frequency information happens because of our inability to render elements of a third dimension efficiently.
+Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently.
 By employing a color scheme for filling tiles according to the observed frequencies corresponds to a binning of the frequencies into a few different levels, with losses stemming from the same source as the previously discussed numerical loss.
-We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_j\}$ with $1 \le j \le n_C$, and that an observed  bin frequency $c$ is assigned to the closest frequency level $c^\ast = b(c)$.
-The --unscaled-- frequency loss is then defined as
+We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_k\}$ with $1 \le k \le n_C$.
+An observed frequency $c_j$ with $1 \le j \le N_x \times N_y$, where $N_x$ and $N_y$ are the number of bin divisions in the X and Y directions respectively,  is assigned to the closest frequency level $c^\ast_k = b(c_j)$.
+The \textit{unscaled} frequency loss is then defined as
 \begin{eqnarray}\label{eqn:frequency}
-L_\text{Freq} = \sum_{i=1}^{n_X n_Y} \left(c_i - b(c_i)\right)^2
+L_\text{Freq} = \sum_{j=1}^{N_x N_y} \left(c_j - b(c_j)\right)^2
 \end{eqnarray}
@@
 \begin{eqnarray}\label{eqn:logfrequency}
-L_{\log \text{Freq}} = \sum_{i=1}^{n_X n_Y} \left(\log(c_i+1) - \log(b(c_i)+1)\right)^2
+L_{\log \text{Freq}} = \sum_{j=1}^{N_x N_y} \left(\log(c_j+1) - \log(b(c_j)+1)\right)^2
 \end{eqnarray}
@@
 Figure~\ref{fig:logcolor} gives an overview of this simulation.
-In the top row tile plots with a color scheme based on linear frequency scaling is used, the bottom row shows the same data based on log frequency scaled colors.
+In the top row binned scatterplots with a color scheme based on linear frequency scaling is used, the bottom row shows the same data based on log frequency scaled colors.
 
@@
 \centering
-<<FrequencyCategories, echo=FALSE, out.width='\\linewidth', fig.width=8.5, fig.height=11>>=
+<<FrequencyCategories,echo=FALSE, results='hide'>>=
 options(stringsAsFactors=FALSE)
@@
 # linear binning
-# pdf(file="figure/p2.pdf", width=8, height=3)
-# p2
-# dev.off()
+pdf(file="p2.pdf", width=8, height=3)
+p2
+dev.off()
 
@@
 
-multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
-# ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
-@
-%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
+#multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
+#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
+@
+\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
 \caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
@@
 \paragraph{Traditional Scatterplots and Frequency Information:}
-using a minimal bin width, a tile display is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
+
+Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
 Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
 The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information.
-By explicitly shading bins (or scaling size) according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points.
-This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha} d and e under the same framework as plots a-c.
+By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points.
+This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c.
 
@@
 
-As shown before, the overall loss partitions naturally into visual and numeric loss (see equation \ref{eqn:partition}). We can make use of this and
- recover some of the numerical information lost due to binning by storing the numerical center $\overline x$ instead of the bin center $x^\ast$ for use in subsequent calculations.
+As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \ref{eqn:partition}). We can make use of this and
+ recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations.
 Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.
@@
 d <- d[order(d$x, d$y),]
-print(xtable(head(d), digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)
+print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)
 dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
@@
 \begin{tabular}{rrr}\hline
-       x &      y & c \\ \hline
+       $x$ &      $y$ & c \\ \hline
+ -0.2582 & 1.5497 & 1 \\ 
+ -0.3843 & 5.6508 & 1 \\ 
+ 1.0597 & 2.4308 & 1 \\ 
+ 2.2680 & 1.5877 & 1 \\ 
+ 2.4992 & 4.6260 & 1 \\ 
+ 1.6788 & 5.0260 & 1 \\ 
+ 1.6054 & 6.4125 & 1 \\ 
+ 2.4180 & 6.4438 & 1 \\ 
+ 2.4345 & 6.4462 & 1 \\ 
+ 2.3620 & 10.3119 & 1 \\  \hline
+\end{tabular}}\end{minipage}\hfil
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Visual Centers, 49 rows]{
+\begin{tabularx}{.7\textwidth}{XXc} \hline
+  $x^\ast$ & $y^\ast$ & c \\ \hline
+  0 & 2 & 1 \\ 
+  0 & 6 & 1 \\ 
+  1 & 2 & 1 \\
+  2 & 2 & 1 \\
+  2 & 5 & 2 \\ 
+  2 & 6 & 3 \\ 
+  2 & 10 & 1 \\ 
+   \hline
+\end{tabularx}
+}\end{minipage}\hfil
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Numerical Centers, 49 rows]{
+\begin{tabular}{rrc}\hline
+  $\bar{x}$   &    $\bar{y}$ & c \\ \hline
  -0.3843 & 5.6508 & 1 \\ 
@@
   1.0597 & 2.4308 & 1 \\ 
-  1.6054 & 6.4125 & 1 \\ 
-  1.6788 & 5.0260 & 1 \\ 
-  2.2680 & 1.5877 & 1 \\ \hline
-\end{tabular}}\end{minipage}\hfil
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data, 49 rows]{
-\begin{tabularx}{.7\textwidth}{XXc} \hline
-  x & y & c \\ \hline
-  0 & 6 & 1 \\ 
-  0 & 2 & 1 \\ 
-  1 & 2 & 1 \\ 
-  2 & 5 & 2 \\ 
-  2 & 6 & 3 \\ 
-  2 & 2 & 1 \\ 
-   \hline
-\end{tabularx}
-}\end{minipage}\hfil
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Recovered Data, 49 rows]{
-\begin{tabular}{rrc}\hline
-       x &      y & c \\ \hline
- -0.3843 & 5.6508 & 1 \\ 
- -0.2582 & 1.5497 & 1 \\ 
-  1.0597 & 2.4308 & 1 \\ 
+  2.2680 & 1.5877 & 1 \\  
   2.0890 & 4.8260 & 2 \\ 
@@
 }\end{minipage}
-\caption{\label{RecoveryComparison}Six rows from Original, Binned, and Recovered Data Tables, with data storage sizes}
+\caption{\label{RecoveryComparison}ten rows from Original and Binned Data Tables, with data storage sizes}
 \end{table}
@@
 \section{Discussion and Examples}
-\subsection{Comparison of Binning Algorithms}
-<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
-# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-#                    password="R R0cks", port=3306, dbname="baseball", 
-#                    host="headnode.stat.iastate.edu")
-# pitch <- new("dataDB", co=connect, table="Pitching")
-# 
-# d1 <- dbData(pitch, vars=c( "G", "SO"))
-
+\subsection{Comparison of Loss in Binning Algorithms}
+<<LossTableGraphs,echo=F,results="hide",cache=TRUE,eval=FALSE>>=
 LossTable <- NULL
+# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
 bins <- rbind(c(2,2), c(5,25))
 computeLoss <- function(binning, data){
-  lRandom <- lossCalc(data, binning=binning, type="random", newData=TRUE)
+  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
   lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
@@
     qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
-      main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + 
-    facet_grid(.~Algorithm) +
-    scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", 
-                        breaks=c(1, 5, 50, 500, 5000)) + 
-    theme_bw() + theme(legend.position="bottom")
-  ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), 
-         width=8, height=5, units="in", dpi=1200)
-
+        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
+        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
+    ggsave(paste("./images/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
+#     qplot(G, SO, fill = log10(Freq), geom="tile", data = lMean$NewData, 
+#       main=paste("Standard Binning (", binning[1], ", ", binning[2],")", sep="")) + 
+#       scale_fill_gradient(low="#56B1F7", high="#132B43", guide="none", trans="log") + theme_bw()
+#     ggsave(paste("./images/Standard", binning[1], binning[2], ".pdf", sep=""), width=4.5, height=4.5, units="in")
   return(rbind(
-    data.frame(binX = binning[1], binY = binning[2], Function = "Random", loss=lRandom$Loss[,5:7]), 
-    data.frame(binX = binning[1], binY = binning[2], Function = "Standard", loss=lMean$Loss[,5:7])))
-}
-
-LossTable <- do.call("rbind", lapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
+    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$LossAll), 
+    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lRandom$LossAll)))
+}
+
+LossTable <- do.call("rbind", mclapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
 LossTable <- as.data.frame(LossTable)
@@
 
-At the example of the pitching data, we will %consider the variables Games and Strikeouts, which are strongly correlated. This example 
-highlight an important advantage of random binning.
-The two tile plots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident: while the standard binning algorithm has rounding artifacts that are clearly visible for even bin widths, the random algorithm produces a graph that better represents the true shape of the data.
-\begin{figure}[hbtp]
-\centering
-\subfloat[(2, 2) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
-}\\
-\subfloat[(5, 25) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
-}
-\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chose bin width is not a multiple of the data resolution.}
-\end{figure}
-
-In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable tile plots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.
-
-The spatial loss is a result of our desire to reduce and summarize the data until it is more manageable in size, and hence is a factor in our decision to bin the data in the first place.
-More importantly, frequency loss has a larger effect on the utility of the graphical data summary.
-Figure~\ref{fig:binning-comparison} shows clearly that the standard algorithm can produce misleading graphical summaries.
-In order to quantify the advantages of each algorithm, we compare the spatial and frequency loss at different bin widths in $x$ (Games).
-The random algorithm has higher spatial loss than the standard algorithm, but in many cases the tile graphs are visually indistinguishable.
+In order to quantify the advantages of each algorithm, we first compare the three sources of loss at different bin widths.
+The numerical and visual losses inherent in creating a binned scatterplot are a result of our desire to reduce and summarize the data until it is more manageable in size, and hence is a factor in our decision to bin the data in the first place.
+%Figure~\ref{fig:binning-comparison} shows clearly that the standard algorithm can produce misleading graphical summaries. 
+ The random algorithm has higher numeric and visual loss than the standard algorithm, but in many cases the binned scatterplots are visually indistinguishable.
+
  The numerical loss is a function of the distribution and bin width, and once the bin width is chosen, it is entirely fixed.
-As the visual loss is the difference between the numerical center of points in the bin and the geometric center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm).
-We would not expect to notice the difference in the spatial loss in either algorithm, so long as the bin width is constant.
-The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins.
+The random binning algorithm will inherently have a larger numerical loss than the standard algorithm because it allows for random assignment of points to numeric bin centers further from the original data than does the standard algorithm.
+
+ As the visual loss is the difference between the numerical center and visual center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm).
+We would not expect to notice the difference in the  visual loss in either algorithm, so long as the bin width is constant, because the expected value of the random process that results in shifting numerical bin centers is zero.
+
+ These characteristics of the numerical and visual losses are shown using the baseball data in the total combined visual and numerical loss at different bin widths in $x$ (Games) in Figure ~\ref{fig:LossComparison1}.
 
@@
 
-qplot(x=x, y=TotalLoss.G*100, group=interaction(y, Algorithm), data=subset(losses, y<=10), geom="line", colour=Algorithm, ylab="Percent Spatial Loss", xlab="Bin  Width  (Games)")+ theme_bw() + theme(legend.position="bottom") + ggtitle("Spatial Loss:  Baseball Games") + guides(colour = guide_legend(override.aes = list(size = 2))) + facet_wrap(~Algorithm)
-# qplot(x=y, y=VisLoss.SO*100, group=interaction(x, Algorithm), data=subset(losses, x==1), geom="line", colour=Algorithm, ylab="Percent Visual Loss", xlab="Bin Width (Strike Outs)") + theme_bw() + theme(legend.position="bottom") + ggtitle("Visual  Loss  with  increasing  Bin  Width: Baseball Strike Outs\n Games not binned") + guides(colour = guide_legend(override.aes = list(size = 2)))
-@
-The standard algorithm is preferable because it has lower spatial loss, but the two algorithms are very similar when we compare frequency loss.
+qplot(x=x, y=TotalLoss.G*100, group=interaction(y, Algorithm), data=subset(losses, y<=10), geom="line", colour=Algorithm, ylab="Percent Spatial Loss", xlab="Bin  Width  (Games)")+ theme_bw() + theme(legend.position="bottom") + ggtitle("Total Combined Numerical and Visual Loss:  Baseball Games") + guides(colour = guide_legend(override.aes = list(size = 2))) + facet_wrap(~Algorithm)
+#% qplot(x=y, y=VisLoss.SO*100, group=interaction(x, Algorithm), data=subset(losses, x==1), geom="line", colour=Algorithm, ylab="Percent Visual Loss", xlab="Bin Width (Strike Outs)") + theme_bw() + theme(legend.position="bottom") + ggtitle("Visual  Loss  with  increasing  Bin  Width: Baseball Strike Outs\n Games not binned") + guides(colour = guide_legend(override.aes = list(size = 2)))
+@
+
+The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary.
+The standard algorithm is preferable because it has lower total combined numerical and visual loss, but the two algorithms are very similar when we compare frequency loss.
 The random algorithm does have more variability, but the frequency loss values are much more similar, and when bins are of size two in either or both dimensions, the random algorithm has lower frequency loss than the standard algorithm.
 This holds for bins which are powers of two, as long as the random binning algorithm is applied iteratively to take advantage of the constant loss property noted above.
+%~\ref{fig:LossComparison}
+
 <<LossComparison, echo=FALSE, fig.cap="Frequency Loss increases as bin width increases in Games and Strike Outs.
@@
 
+Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues.
+Random binning is especially advantagous when discrete data values are recorded at a consistant increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths.
+In this senario, we see that the non-synchronous data increments and bin widths will cause bins to contain unconsistant numbers of possible data value.
+Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency.
+Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artifical frequency discrepancy that manifests in a striped pattern in the visualization.
+The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.
+
+%A similar situation can arise when discrete data values are recorded at consistant increments and binwidths are chosen that are a odd integer multiple of the data incremental widths.
+This causes the situation where every other bin will have a data value located on the upper bound.
+Since the standard binning algorithm always bins data values on the boundary between bins to the lower bin center we will have an additional possible data value contributing to the frequency of every other bin, causing a striped pattern in our visualization.
+Again, the random binning algorithm mitigates this problem because it will allocate points on bin boundaries to adjacent bin centers with probability 0.5, leading to a more accurate display of frequency information.
+
+\subsection{Binning Loss in Baseball Data: Strikeout and Game Counts}
+
+For this data example we will revisit the baseball data used earlier in this paper.
+
+\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
+\noindent
+Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
+
+<<tablelossexample, echo=FALSE, results='asis',cache=FALSE>>=
+loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
+loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
+loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
+loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
+
+loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
+loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
+loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
+loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
+
+table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
+                               c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
+                               c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
+                               c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
+                               c(binwidth=1, loss1[5:7]*100, loss1[8])))
+names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
+print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
+
+@
+
+We may also use this example to highlight the advantages of random binning.
+The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident: while the standard binning algorithm has rounding artifacts that are clearly visible for even bin widths, the random algorithm produces a graph that better represents the true shape of the data.
+This is because the In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.
+
+\begin{figure}[hbtp]
+\centering
+\subfloat[(2, 2) tiling of G, SO data]{
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.eps}
+}\\
+\subfloat[(5, 25) tiling of G, SO data]{
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.eps}
+}
+\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chose bin width is not a multiple of the data resolution.}
+\end{figure}
+
 \subsection{Big Data: Airline Departure Times}
@@
 Even while using alpha blending this results in a severely over-plotted graph.
-On the right is a tile plot of the complete data binned at 1-minute intervals.
+On the right is a binned scatterplot of the complete data binned at 1-minute intervals.
 This does not have any spatial loss, as the times are recorded by minute.
@@
 \begin{figure}[hbtp]\centering
-\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{./images/MillionSample.pdf}}\hfil
-\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{./images/AirlineStdBinning1min.pdf}}
+\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{./data/MillionSample.eps}}\hfil
+\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{./data/AirlineStdBinning1min.eps}}
 \caption{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011.
@@
 \begin{figure}[hbtp]\centering
-\includegraphics[keepaspectratio=true, height=.43\textheight]{./images/AirlineStdBinning5mins.pdf}
-\includegraphics[keepaspectratio=true, height=.43\textheight]{./images/AirlineStdBinning15mins.pdf}
+\includegraphics[keepaspectratio=true, height=.43\textheight]{./data/AirlineStdBinning5mins.eps}
+\includegraphics[keepaspectratio=true, height=.43\textheight]{./data/AirlineStdBinning15mins.eps}
 \caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
@@
 Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
-Binning allows for the visualization and manipulation of large data sets, and easily translates into tile plots which are more appropriate for the human visual system.
-The loss due to binning reduces the data set to a more manageable size, but some of that loss can be recovered with additional computational investment.
+Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
+The loss due to binning is the cost of reducing the data set to a more manageable size, but some of that loss can be recovered with additional computational investment.
 
@@
 <<PrintTable,echo=F,include=FALSE>>=
-#   print(tab, include.rownames=FALSE)
+  print(tab, include.rownames=FALSE)
 @

### 2013-03-22 Paper.Rnw
fixed paths to charts: figure folder created on the fly, charts in images folder are permanent got rid of multicore calls

@@
 \usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
-\graphicspath{{images/}}
-\usepackage{epstopdf}
+\graphicspath{{figure/}}
 %\usepackage{csquotes}
@@
 \title{Binning Strategies and Related Loss for Large Data}
-\author{Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
+\author{Susan VanderPlas$^1$, Karsten Maurer$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
 
 \begin{document}
-\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
 \maketitle
@@
 \end{abstract}
+%\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
+
+% <<options-knitr,echo=F,include=F>>=
+% require(knitr)
+% options(replace.assign=TRUE,width=70)
+% opts_chunk$set(fig.path='figure/graphics-', cache.path='cache/cache-', fig.align='center', fig.width=5, fig.height=5, fig.show='hold', cache=TRUE, par=TRUE)
+% @
 
@@
 library(RColorBrewer)
-# setwd("/home/susan/Dropbox/GraphicsGroup/")
-setwd("/Users/heike/Dropbox/papers/")
-
-setwd("./dbaccess/NewWriteUp/")
 
@@
 @
-<<scatterplots, echo=F,include=F,eval=F>>=
+<<scatterplots, echo=F,include=F>>=
 qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
-  ggsave("./images/OverplottingAlpha.pdf", width=5, height=5)
+  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
 qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
-  ggsave("./images/Overplotting.pdf", width=5, height=5)
+  ggsave("./figure/Overplotting.pdf", width=5, height=5)
 qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
-  ggsave("./images/OverplottingCircles.pdf", width=5, height=5)
+  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
 qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
-  ggsave("./images/HexBinning.pdf", width=5, height=5)
+  ggsave("./figure/HexBinning.pdf", width=5, height=5)
 d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
@@
   theme(legend.position="bottom")
-  ggsave("./images/BinScatter.pdf", width=5, height=5)
+  ggsave("./figure/BinScatter.pdf", width=5, height=5)
 
@@
 \begin{figure}[hbtp]
-  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/Overplotting.eps}}
-  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingCircles.eps}} 
-	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{images/OverplottingAlpha.eps}}
-
-  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/HexBinning.eps}}
-  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{images/BinScatter.eps}}
-
-	\caption{Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha-blending ($\alpha$=0.05).
+  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{Overplotting.pdf}}
+  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingCircles.pdf}} 
+	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingAlpha.pdf}}
+
+  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{HexBinning.pdf}}
+  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{BinScatter.pdf}}
+
+	\caption{\label{fig:scatter-alpha} Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha-blending ($\alpha$=0.05).
 Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
-  \label{fig:scatter-alpha}
 \end{figure}
@@
 Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
-<<binningDemo,echo=F,eval=F,include=F,cache=T>>=
+<<binningDemo,echo=F,include=F,cache=T>>=
       library(mvtnorm)
@@
                  xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
-      ggsave("./images/Binning-DataAndBin.pdf", width=5, height=5, units="in")
+      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
 
@@
         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-      ggsave("./images/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
+      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
 @
@@
   \centering
-  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./images/Binning-DataAndBin.eps}}
+  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
 %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
-%   {./images/RandomBinningDataPoints1}}
+%   {./figure/RandomBinningDataPoints1}}
   \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
-  {./images/RandomBinningDataPoints2.eps}}
+  {./figure/RandomBinningDataPoints2.pdf}}
   \caption{Points are assigned to bins centered at (5, 25) an (15,25).
@@
 \subsection{Random Binning Algorithm}
-<<RandomBinDemo,echo=F,include=FALSE,eval=F>>=
+<<RandomBinDemo,echo=F,include=FALSE>>=
 nl <- ncol(d.rand)
@@
   annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
-ggsave("./images/RandomBinningDataPoints2.pdf",width=5, height=5)
+ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
 @
@@
 ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
-ggsave("./images/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
+ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
 @
@@
     \centering
-		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./Binning-DataVisualNumericalCenter.eps}}
-    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./images/FreqLoss.eps}}
+		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
+    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
 \caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.
@@
 
-#multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
+multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
 #ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
 @
-\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
+%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure.pdf}
 \caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
@@
 
-<<recover,echo=FALSE,results='asis', eval=FALSE>>=
-setwd("./figure")
+<<recover,echo=FALSE,results='asis'>>=
 set.seed(90)
@@
 d <- d[order(d$x, d$y),]
-print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)
+print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
 dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
@@
 print(xtable(head(drecover), digits=c(0, 4, 4, 0), caption=paste('Recovered Data\\newline', nrow(drecover), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
-setwd("../")
 @
@@
 \subsection{Comparison of Loss in Binning Algorithms}
-<<LossTableGraphs,echo=F,results="hide",cache=TRUE,eval=FALSE>>=
-LossTable <- NULL
+<<LossTableGraphs,echo=F,results="hide",cache=TRUE>>=
 # bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
-bins <- rbind(c(2,2), c(5,25))
 computeLoss <- function(binning, data){
@@
         scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
-    ggsave(paste("./images/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
-#     qplot(G, SO, fill = log10(Freq), geom="tile", data = lMean$NewData, 
-#       main=paste("Standard Binning (", binning[1], ", ", binning[2],")", sep="")) + 
-#       scale_fill_gradient(low="#56B1F7", high="#132B43", guide="none", trans="log") + theme_bw()
-#     ggsave(paste("./images/Standard", binning[1], binning[2], ".pdf", sep=""), width=4.5, height=4.5, units="in")
+    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
   return(rbind(
-    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$LossAll), 
-    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lRandom$LossAll)))
-}
-
-LossTable <- do.call("rbind", mclapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
+    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$Loss), 
+    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lMean$Loss)))
+}
+
+bins <- rbind(c(2,2), c(5,25))
+LossTable <- NULL      
+LossTable <- do.call("rbind", lapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
 LossTable <- as.data.frame(LossTable)
-names(LossTable) <- c("binX", "binY", "Function", "G", "SO", "Freq")
-tab <- xtable(LossTable[order(LossTable[,3], LossTable[,1]),], digits=c(0, 0, 0, 0, 8, 8, 8))
+names(LossTable) <- c("binX", "binY", "Function", "Numerical Loss G", "Numerical Loss SO", "Visual Loss G", "Visual Loss SO", "Total Loss G", "Total Loss SO", "Frequency Loss")
+tab <- xtable(LossTable, digits=c(0, 0, 0, 0, 8, 8, 8, 8, 8, 8, 8))
 @
@@
 
-<<LossComparison1, echo=FALSE, eval=T, fig.cap='Spatial Loss increases with bin width for each variable.
-The random algorithm has higher spatial loss than the standard algorithm.', fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth', fig.align='center', fig.pos='H'>>=
-
-# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-#                      password="R R0cks", port=3306, dbname="baseball", 
-#                      host="headnode.stat.iastate.edu")
-# pitch <- new("dataDB", co=connect, table="Pitching")
-# d1 <- dbData(pitch, vars=c( "G", "SO"))
-# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
-# 
-# library(multicore)
-# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
-# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
-# 
-# lossesRdm <- cbind(bins, lossesRdm)
-# lossesStd <- cbind(bins, lossesStd)
-# 
-# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
-# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
+\begin{figure}[hbtp]
+\centering
+<<LossComparison1, echo=FALSE, eval=T, fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth'>>=
 
@@
 qplot(x=x, y=TotalLoss.G*100, group=interaction(y, Algorithm), data=subset(losses, y<=10), geom="line", colour=Algorithm, ylab="Percent Spatial Loss", xlab="Bin  Width  (Games)")+ theme_bw() + theme(legend.position="bottom") + ggtitle("Total Combined Numerical and Visual Loss:  Baseball Games") + guides(colour = guide_legend(override.aes = list(size = 2))) + facet_wrap(~Algorithm)
-#% qplot(x=y, y=VisLoss.SO*100, group=interaction(x, Algorithm), data=subset(losses, x==1), geom="line", colour=Algorithm, ylab="Percent Visual Loss", xlab="Bin Width (Strike Outs)") + theme_bw() + theme(legend.position="bottom") + ggtitle("Visual  Loss  with  increasing  Bin  Width: Baseball Strike Outs\n Games not binned") + guides(colour = guide_legend(override.aes = list(size = 2)))
-@
-
+@
+\caption{\label{fig:LossComparison1} Spatial Loss increases with bin width for each variable. The random algorithm has higher spatial loss than the standard algorithm.}
+\end{figure}
 The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary.
@@
 \subfloat[(2, 2) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.eps}
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
 }\\
 \subfloat[(5, 25) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.eps}
+\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
 }
@@
 \begin{figure}[hbtp]\centering
-\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{./data/MillionSample.eps}}\hfil
-\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{./data/AirlineStdBinning1min.eps}}
+\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfil
+\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}
 \caption{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011.
@@
 \begin{figure}[hbtp]\centering
-\includegraphics[keepaspectratio=true, height=.43\textheight]{./data/AirlineStdBinning5mins.eps}
-\includegraphics[keepaspectratio=true, height=.43\textheight]{./data/AirlineStdBinning15mins.eps}
+\includegraphics[keepaspectratio=true, height=.43\textheight]{images/AirlineStdBinning5mins.pdf}
+\includegraphics[keepaspectratio=true, height=.43\textheight]{images/AirlineStdBinning15mins.pdf}
 \caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
@@
 % \begin{subsection}{Numerical Comparison of Binning Algorithms}
-<<PrintTable,echo=F,include=FALSE>>=
+<<PrintTable,echo=F,include=FALSE, dependson='LossTableGraphs'>>=
   print(tab, include.rownames=FALSE)

### 2013-03-30 Paper.Rnw
major overhaul of the baseball loss plots, color scheme still needs some work, surrounding paragraphs to be edited this evening

@@
 library(dbData)
+library(reshape2)
 #library(multicore)
@@
 
-<<recover,echo=FALSE,results='asis'>>=
+<<recover,echo=FALSE,eval=FALSE,results='asis'>>=
 set.seed(90)
@@
 @
+
 \begin{table}[hbtp]
@@
  2.3620 & 10.3119 & 1 \\  \hline
-\end{tabular}}\end{minipage}\hfil
+\end{tabular}}
+\end{minipage}\hfil
 \begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Visual Centers, 49 rows]{
@@
    \hline
-\end{tabularx}
-}\end{minipage}\hfil
+\end{tabularx}}
+\end{minipage}\hfil
 \begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Numerical Centers, 49 rows]{
@@
   2.2680 & 1.5877 & 1 \\ \hline
-\end{tabular}
-}\end{minipage}
+\end{tabular}}
+\end{minipage}
 \caption{\label{RecoveryComparison}ten rows from Original and Binned Data Tables, with data storage sizes}
@@
 
- These characteristics of the numerical and visual losses are shown using the baseball data in the total combined visual and numerical loss at different bin widths in $x$ (Games) in Figure ~\ref{fig:LossComparison1}.
-
-\begin{figure}[hbtp]
-\centering
-<<LossComparison1, echo=FALSE, eval=T, fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth'>>=
-
-losses <- read.csv("../dbData/data/losses.csv", stringsAsFactors=FALSE)
-
-qplot(x=x, y=TotalLoss.G*100, group=interaction(y, Algorithm), data=subset(losses, y<=10), geom="line", colour=Algorithm, ylab="Percent Spatial Loss", xlab="Bin  Width  (Games)")+ theme_bw() + theme(legend.position="bottom") + ggtitle("Total Combined Numerical and Visual Loss:  Baseball Games") + guides(colour = guide_legend(override.aes = list(size = 2))) + facet_wrap(~Algorithm)
-@
-\caption{\label{fig:LossComparison1} Spatial Loss increases with bin width for each variable. The random algorithm has higher spatial loss than the standard algorithm.}
-\end{figure}
 The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary.
@@
 %~\ref{fig:LossComparison}
-
-<<LossComparison, echo=FALSE, fig.cap="Frequency Loss increases as bin width increases in Games and Strike Outs.
-The random binning algorithm has slightly higher frequency loss as bin width grows large, but the loss is comparable when bin widths are small.", fig.width=8, fig.height=4, out.width='\\linewidth', out.height='.5\\linewidth', fig.pos='H'>>=
-# 
-# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-#                      password="R R0cks", port=3306, dbname="baseball", 
-#                      host="headnode.stat.iastate.edu")
-# pitch <- new("dataDB", co=connect, table="Pitching")
-# d1 <- dbData(pitch, vars=c( "G", "SO"))
-# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 15)))
-# 
-# library(multicore)
-# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
-# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
-# 
-# lossesRdm <- cbind(bins, lossesRdm)
-# lossesStd <- cbind(bins, lossesStd)
-# 
-# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
-# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
-losses <- read.csv("../dbData/data/losses.csv", stringsAsFactors=FALSE)
-
-qplot(x=y, y=TotalLoss.LogFreq, group=interaction(x, Algorithm), data=losses, geom="line", colour=Algorithm, ylab="Frequency Loss", xlab="Bin Width (Strike Outs)")+ facet_grid(.~Algorithm) + scale_colour_discrete(guide="none") + theme_bw() + ggtitle("Frequency Loss: Baseball Games and Strike-Outs")
-@
 
@@
 For this data example we will revisit the baseball data used earlier in this paper.
+The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}.
+
+\begin{figure}[hbtp]
+\centering
+<<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
+
+losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
+lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
+lossmelt <- lossmelt[which(lossmelt$y <= 10),]
+lossmelt$y <- as.factor(lossmelt$y)
+levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
+qplot(x=x, y=value,
+      group=interaction(y, Algorithm),
+      data=lossmelt, geom="line",
+      colour=y,
+      xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
+        theme_bw() + theme(legend.position="bottom") + 
+        guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
+        facet_grid(variable ~ Algorithm, scales = "free")+
+        ggtitle("Loss for Baseball Data under Standard and Random Binning")
+
+@
+\caption{\label{fig:LossComparison1} Loss funcion }
+\end{figure}
 

### 2013-03-31 Paper.Rnw
Edits made to plots and figures so address issues we discussed
Also, some code chunks are effectively commented out using "eval=FALSE"
because their code is not running properly and may not be used in
revised version, if needed they will be reworked.

@@
 <<scatterplots, echo=F,include=F>>=
-qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
+qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
-qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
+qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/Overplotting.pdf", width=5, height=5)
-qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
+qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
 qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
-  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
   ggsave("./figure/HexBinning.pdf", width=5, height=5)
@@
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
-  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
-  theme(legend.position="bottom")
+  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
   ggsave("./figure/BinScatter.pdf", width=5, height=5)
@@
                  colour="blue4") +
-        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
+#        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
 #         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
@@
         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
+      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
 @
@@
 %   {./figure/RandomBinningDataPoints1}}
-  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
-  {./figure/RandomBinningDataPoints2.pdf}}
+  \subfloat[Bin Attribute Diagram]{\includegraphics[keepaspectratio=TRUE, width=4in]
+  {./figure/Binning-DataVisualNumericalCenter.pdf}}
   \caption{Points are assigned to bins centered at (5, 25) an (15,25).
@@
 \subsection{Comparison of Loss in Binning Algorithms}
-<<LossTableGraphs,echo=F,results="hide",cache=TRUE>>=
+<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=FALSE>>=
 # bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
@@
 \noindent
-Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
-
-<<tablelossexample, echo=FALSE, results='asis',cache=FALSE>>=
+%Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
+
+<<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
 loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
@@
 print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
-
 @

### 2013-04-03 Paper.Rnw
edits to binning and discussion sections

@@
 \begin{eqnarray*}\label{stdbin}
-x^\ast = b(x) =  b_X \cdot \left\lfloor (x-O_X)/ b_X + b_X/2\right\rfloor + O_X
+x^\ast = b(x) =  b_X \cdot \left\lfloor (x-x_1^\ast)/ b_X + 1/2\right\rfloor + x_1^\ast
 \end{eqnarray*}
-where $O_X$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. 
+where $x_{(1)}$ is the center of the first bin and $\lfloor .
+\rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.
+An intuitive understanding of the formula is developed by noting that $(x-x_1^\ast)/ b_X$ is the number of bin widths that $x$ is above $x_1^\ast$ and that $\frac{b_X/2}{b_X}=\frac{1}{2}$ is the number of bin widths that $x_1^\ast$ is above $x_{(1)}$.
+The value $\lfloor (x-x_1^\ast)/ b_X + 1/2\rfloor$ then represents the number of bins centers that lie between $x_{(1)}$ and $x$.
+Thus, the binning function is simply finding the location of the bin center for $x$ by adding the distance between the bin centers for $x_{(1)}$ and $x$ to the known bin center for $x_{(1)}$.
 %to allow for bins of width $b_X$ centered at any sequence $O_X, O_X + b_X, ..., O_X + n b_X$.
 
-For illustration, let us assume that we have observed values 1.23, 1.55, and 2.35, and bin centers $\{1, 2, 3\}$, corresponding to a bin width of $b_X = 1$ and a first bin center $O_X = 1$.
-Intuitively it is clear, that value 1.23 is assigned to bin center 1, while values 1.55 and 2.35 are assigned to bin center 2.
-Applying equation (\ref{stdbin}) to value 1.23 we get: $1 \cdot \lfloor (1.23-1)/ 1 + 0.5\rfloor + 1$ = $\lfloor 0.73 \rfloor + 1$ =  0 + 1 = 1.
-Similarly, 1.55 is assigned to $\lfloor 0.55 + 0.5\rfloor + 1$ = 1+1 = 2.
-
 Alternate binning parameterizations include dividing up the range of data by a specified number of bins of equal width, but even these strategies fit into the framework of assigning values to the closest bin center.
+
 % , with smallest bin
@@
 
-Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
+Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process extended to a two dimensional situation.
 <<binningDemo,echo=F,include=F,cache=T>>=
@@
 
-The method described above is the standard binning algorithm. As a different approach, we define the random binning algorithm below.
-
 \begin{figure}[hbtp]
@@
   {./figure/Binning-DataVisualNumericalCenter.pdf}}
-  \caption{Points are assigned to bins centered at (5, 25) an (15,25).
-Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
+  \caption{In illustration (a) points are assigned to bins centered at (5, 25) an (15,25).
+Points in the standard algorithm are all contained within the bin region.
+Illustration (b) displays the points assigned to a single bin along with key bin attributes: the visual bin center, ($x_1^\ast$, $y_1^\ast$), and the numerical bin center, ($\bar{x}$,$\bar{y}$).
+}
   \label{fig:BinningTutorial}
@@
 @
-An interesting variation which results in statistically nice properties is a random, non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$.
+An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
+A random binning algorithm utilizes a non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$.
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
@@
 
-Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
+%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
 
@@
 
-The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary.
-The standard algorithm is preferable because it has lower total combined numerical and visual loss, but the two algorithms are very similar when we compare frequency loss.
-The random algorithm does have more variability, but the frequency loss values are much more similar, and when bins are of size two in either or both dimensions, the random algorithm has lower frequency loss than the standard algorithm.
-This holds for bins which are powers of two, as long as the random binning algorithm is applied iteratively to take advantage of the constant loss property noted above.
+Frequency loss is argueably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary.
+The standard and random algorithms are very similar when we compare frequency loss.
+The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticable than the increase seen in numerical loss.
 %~\ref{fig:LossComparison}
@@
 \subsection{Binning Loss in Baseball Data: Strikeout and Game Counts}
+
+THIS SECTION UNDER CONSTRUCTION
 

### 2013-04-03 Paper.Rnw
small changes in code

@@
 \end{abstract}
-%\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
-
-% <<options-knitr,echo=F,include=F>>=
-% require(knitr)
-% options(replace.assign=TRUE,width=70)
-% opts_chunk$set(fig.path='figure/graphics-', cache.path='cache/cache-', fig.align='center', fig.width=5, fig.height=5, fig.show='hold', cache=TRUE, par=TRUE)
-% @
 
 \section{Introduction}
+
 <<setup,echo=F,include=F,eval=T>>=
@@
 \centering
-<<FrequencyCategories,echo=FALSE, results='hide'>>=
+<<FrequencyCategories,echo=FALSE, results='hide', cache=TRUE>>=
 options(stringsAsFactors=FALSE)
@@
 \subsection{Comparison of Loss in Binning Algorithms}
-<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=FALSE>>=
+<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=
 # bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
@@
       data=lossmelt, geom="line",
-      colour=y,
+      colour=y, size=I(1),
       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
@@
         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-        facet_grid(variable ~ Algorithm, scales = "free")+
-        ggtitle("Loss for Baseball Data under Standard and Random Binning")
-
-@
-\caption{\label{fig:LossComparison1} Loss funcion }
+        facet_grid(variable ~ Algorithm, scales = "free")
+
+# qplot(x=x, y=value,
+#       group=interaction(y, Algorithm),
+#       data=lossmelt, geom="line",
+#       colour=y, size=I(1), linetype=Algorithm,
+#       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
+#         theme_bw() + theme(legend.position="bottom") + 
+#         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
+#         facet_grid(. ~ variable, scales = "free")
+
+@
+\caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
 \end{figure}

### 2013-04-04 Paper.Rnw
minor edits to figures
Note: I commented out the enormous airline plot that was slowing
compilation and making it too large to send as an attachment and added a
tiny pdf plot as a placeholder

@@
 %   {./figure/RandomBinningDataPoints1}}
-  \subfloat[Bin Attribute Diagram]{\includegraphics[keepaspectratio=TRUE, width=4in]
-  {./figure/Binning-DataVisualNumericalCenter.pdf}}
-  \caption{In illustration (a) points are assigned to bins centered at (5, 25) an (15,25).
-Points in the standard algorithm are all contained within the bin region.
-Illustration (b) displays the points assigned to a single bin along with key bin attributes: the visual bin center, ($x_1^\ast$, $y_1^\ast$), and the numerical bin center, ($\bar{x}$,$\bar{y}$).
-}
+  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
+  {./figure/RandomBinningDataPoints1.pdf}}
+  \caption{Points are assigned to bins centered at (5, 25) an (15,25).
+Points in the standard algorithm are contained within the defined bin region, where points from the random algorith may be assigned to bins from outside the defined bin regions.}
   \label{fig:BinningTutorial}
@@
     \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
-\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.
+\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.
 (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
@@
 
+# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
+#                      password="R R0cks", port=3306, dbname="baseball", 
+#                      host="headnode.stat.iastate.edu")
+# pitch <- new("dataDB", co=connect, table="Pitching")
+# d1 <- dbData(pitch, vars=c( "G", "SO"))
+# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
+# 
+# library(multicore)
+# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
+# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
+# 
+# lossesRdm <- cbind(bins, lossesRdm)
+# lossesStd <- cbind(bins, lossesStd)
+# 
+# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
+# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
+
 losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
 lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
-lossmelt <- lossmelt[which(lossmelt$y <= 10),]
+lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
 lossmelt$y <- as.factor(lossmelt$y)
@@
 \begin{figure}[hbtp]\centering
-\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfil
+%\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfil
+%   Swap Back at later date (shrinks size of pdf to managable)
+\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
 \subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}

### 2013-04-07 Paper.Rnw
couple of comments

@@
 This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points.
+\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
 A modern alternative to open circles is alpha-blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.
@@
 
+\hh{discussion of sufficient statistics for graphics could go here.
+Idea: the problems with large data in scatterplots all arise from overplotting, which is a form of implicit data aggregation.
+In order to keep track of overplotting, we switch to a weighted form of displays and explicitly keep track of overplotting.
+The aggregated data forms the sufficient statistics for a scatterplot, i.e.
+provide all of the information necessary to render it.
+The scatterplot therefore does not display more than that information.
+Making the data aggregation explicit allows us to calculate the loss we experience.}
+
 One approach to reduce the graphical complexity is to bin the displayed data.
@@
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The binned scatterplot is less prone to the Hermann-grid illusion than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
+The binned scatterplot is less prone to the Hermann-grid illusion \hh{need a reference} than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
 
@@
 \caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.
-(b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
+\hh{could we make (a) square again?
+the squished panel looks funky.}  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
 	\end{figure}
@@
 %   Swap Back at later date (shrinks size of pdf to managable)
-\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
+\subfloat[Million-point sample of Airline Data \hh{this is the bad practice - our counter-example of what not to do.
+we need to emphasize this a bit more}]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
 \subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}

### 2013-04-22 Paper.Rnw
edits to baseball example section
We still need to discuss the baseball binned scatterplot construction
and whether the issue was actually solved by using odd integer binwidths
(or if the scatterplot for (5,25) looks better only because the bin
dividers were no longer on integer values which is what i suspect)

@@
  As the visual loss is the difference between the numerical center and visual center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm).
-We would not expect to notice the difference in the  visual loss in either algorithm, so long as the bin width is constant, because the expected value of the random process that results in shifting numerical bin centers is zero.
+We expect a larger visual loss under the random binning algorithms than in the standard binning because the numerical bin centers are allowed to be drawn further away from the visual center under the random binning algorithm.
 
@@
 
-THIS SECTION UNDER CONSTRUCTION
-
 For this data example we will revisit the baseball data used earlier in this paper.
 The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}.
+We find that the visual and numerical loss are in fact higher while using the random binning algorithm, but that the two binning algorithms are almost indestiguishable in terms of frequency loss.
+
+Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.
+This is appearant in the order of colored lines based on the Strike Out binning size.
+This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as binwidth increases.
+Whereas the visual loss function sums loss over bin, thus as the binwidth increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.
 
@@
 
-\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
-\noindent
+%\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
+%\noindent
 %Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
@@
 We may also use this example to highlight the advantages of random binning.
-The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident: while the standard binning algorithm has rounding artifacts that are clearly visible for even bin widths, the random algorithm produces a graph that better represents the true shape of the data.
-This is because the In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.
+The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident.
+The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips.
+These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the binwidths are integer multiples of the data increments.
+This reinforces the arguement that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
+
+\textbf{Are we sure that this is because of an odd binwidths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
 
@@
 }
-\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chose bin width is not a multiple of the data resolution.}
+\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chosen bin width is not a multiple of the data resolution.}
 \end{figure}

### 2014-10-19 Paper.Rnw
update to new dbData package and new database on mysql2

@@
 
-connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-                   password="R R0cks", port=3306, dbname="baseball", 
-                   host="headnode.stat.iastate.edu")
+connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
+                     port=3306, user="dbaccess", dbname="baseball")
 pitch <- new("dataDB", co=connect, table="Pitching")
@@
 d1.hex$Freq <-  d1.hex.obj@count
-qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
-  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
+  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
   ggsave("./figure/BinScatter.pdf", width=5, height=5)

### 2015-02-09 Paper.Rnw
No changes, pulling and pushing to ensure sync

--- 2013-04-22 Paper.Rnw
@@
 
+\newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
 \newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike
-\newcommand{\svp}[1]{{\color{red}#1}} %green: Susan (It's more fun that way - comment roulette)
+\newcommand{\svp}[1]{{\color{yellow}#1}} %yellow: Susan (It's more fun that way - comment roulette)
 
@@
 \title{Binning Strategies and Related Loss for Large Data}
-\author{Susan VanderPlas$^1$, Karsten Maurer$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
+\author{Karsten Maurer$^1$, Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
 
@@
 @
+
 <<scatterplots, echo=F,include=F>>=
@@
 d1.hex$Freq <-  d1.hex.obj@count
-qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
-  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + scale_size_area("Frequency", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
   ggsave("./figure/BinScatter.pdf", width=5, height=5)
@@
 This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points.
-\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
+%\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
 A modern alternative to open circles is alpha-blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.
@@
 
-\hh{discussion of sufficient statistics for graphics could go here.
+%\hh{discussion of sufficient statistics for graphics could go here.
 Idea: the problems with large data in scatterplots all arise from overplotting, which is a form of implicit data aggregation.
@@
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The binned scatterplot is less prone to the Hermann-grid illusion \hh{need a reference} than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
+The binned scatterplot is less prone to the Hermann-grid illusion %\hh{need a reference}
+than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
 
@@
 
-\item {\it Visual Loss} is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
-This visual center will generally different from the numerical center, or the mean of the data corresponding to that bin, which can be thought of as the center of mass.
-Thus visual loss is the difference between the visual center $(x^\ast, y^\ast)$ and the numerical center $(\bar{x}, \bar{y})$.
-This causes some bias in the perception.
-
-\item {\it Numerical Loss} occurs when points in the data set are reduced to a single point at the mean, i.e.
-this is the collective difference between all data points $(x_i,y_i)$ and their corresponding numerical centers $(\bar{x}, \bar{y})$.
+\item {\it Numerical Loss}, $L^\text{N}$, occurs when points in the data set are reduced to a single points at the bin means.
+Allocating points to the numerical center, or the mean of the data corresponding to that bin, causes a loss of information about the exact location of the points.
+
+\item {\it Visual Loss}, $L^\text{V}$, is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
+This visual center will generally different from the numerical mean, causing some bias in the perception.
+Thus visual loss for a bin, $L_i^\text{N}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
+The visual loss is the sum of these bin distances.
 
@@
 \subsection{Numerical and Visual Loss}
-Reducing data points $\mathbf{x}$ to the visual bin centers $x^\ast$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
-Let us denote the difference between point $x_i$ from its bin center as $S_i = \left|x_i-b(x_i)\right|$ for each point in the data set $i=1, ..., n$.
-The total loss from binning an entire data set is defined as
-
-\[L_s(\mathbf{x}) = 1/S^2_{\emptyset} \cdot \sum_{i=1}^{n} S^2_i\] 
-where $S_\emptyset^2$ is the loss which results when the entire data set is reduced to a single bin,
-$S^2_{\emptyset} = \sum_i (x_i-\tilde{x})^2$ for the single bin center $\tilde{x}$.
-%The loss depends on the choice of bin center: candidates for bin centers are the mean of the points within the bin or the visual center of the bin. 
+Reducing data points $\mathbf{x}$ to the visual bin centers $\mathbf{x}^\ast$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
 If the visual bin centers are the same as the numerical centers for each bin, then any loss could be strictly attributed to difference between points and the bin centers. 
 The problem is that the visual center and the numerical center for bins will rarely match, which introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of the symbol representing the data.
+
 Therefore we distinguish two contributions to the total loss. 
-Numerical loss results from the binning process as points are reduced to the numerical center point and is thus the sum of numerical losses for individual points ($\sum_{i=1}^n L_i^\text{N}$).
-Visual loss is due to the difference between the numerical centers and the visual centers and is thus the sum of visual losses for each bin ( $\sum_{j=1}^{N_x} L_j^\text{V}$).
+Numerical loss results from the binning process as points are reduced to the numerical center point.
+The loss for an individual point, $L_i^\text{N}$, is the euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$.
+The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
+Visual loss is due to the distance between the numerical centers and the visual centers.
+The visual loss for each bin, $L_j^\text{V}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
+The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$).
 This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.
-
-Fortunately we can show that the total loss perceived is partitioned by these two components of loss, and it holds (see appendix \ref{proof:partition} for a proof):
-\begin{eqnarray}\label{eqn:partition}
-L_s(\mathbf{x}) =  1/S^2_{\emptyset} \cdot \left(\sum_i^n L^N_i + \sum_j^{N_x} L^V_j\right)
-\end{eqnarray}
+This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions napply to either approach to binning.
+
+The magnitude of the numerical and visual losses are clearly dictated by the scale on which the data was recorded, thus and the loss values have no direct interpretation.
+However, we can examine both the numerical and the visual loss from the perspective of proportion of the greatest possible loss.
+To do this for numerical loss we consider the binning scenario that would always lead to greatest possible loss; allocating all data points into a single bin.
+Thus we can define the greatest possible numerical loss as
+
+\begin{center}
+$ L_0^\text{N} = \sum_{i=1}^{n} \sqrt{(x_i-\bar{x})^2 + (y_i-\bar{y})^2}  $, 
+\end{center}
+
+where $\bar{x}$ and $\bar{y}$ are the sample averages for x and y respectively.
+After we select a binning strategy and binwidth we can then assess the proportion of possible numerical loss incurred due to binning by computing $L^\text{N} / L_0^\text{N}$.
+
+%-------
+% Cut "total loss" stuff, because this was all a relic of "spatial loss" part of paper
+%Let us denote the difference between point $x_i$ from its bin center as $S_i = \left|x_i-b(x_i)\right|$ for each point in the data set $i=1, ..., n$.
+%We may then combine these losses from the entire data set to be defined as 
+%\[L_s(\mathbf{x}) = 1/S^2_{\emptyset} \cdot \sum_{i=1}^{n} S^2_i\] 
+%where $S_\emptyset^2$ is the loss which results when the entire data set is reduced to a single bin,
+%$S^2_{\emptyset} = \sum_i (x_i-\tilde{x})^2$ for the single bin center $\tilde{x}$.
+%The loss depends on the choice of bin center: candidates for bin centers are the mean of the points within the bin or the visual center of the bin. 
+%Fortunately we can show that the total loss perceived is partitioned by these two components of loss, %and it holds (see appendix \ref{proof:partition} for a proof):
+%\begin{eqnarray}\label{eqn:partition}
+%L_s(\mathbf{x}) =  1/S^2_{\emptyset} \cdot \left(\sum_i^n L^N_i + \sum_j^{N_x} L^V_j\right)
+%\end{eqnarray}
+%-------
+
 <<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
@@
 \noindent
-Note that the discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms.
-The decomposition also holds for the random binning algorithm as well as the standard binning algorithm.
+
+%The decomposition also holds for the random binning algorithm as well as the standard binning algorithm. 
 %The expected total loss for either algorithm depends on the distribution the underlying data. 
@@
 
-It is worth noting that the loss function is a monotonically increasing function of binwidth.
+%---
+% I believe we showed that this is only true when "rebinning" previously binned data and not deterministic when randomly binning at doubled bins from original data, thus not practically useful information so commented out
+%It is worth noting that the loss function is a monotonically increasing function of binwidth.
 An interesting feature of the random binning algorithm is that the increase in loss that we experience by doubling the bin width is deterministic, even though the assignment of bins is not.
@@
 Doubling bin widths is, particularly in the framework of visualizations, a very natural step, making this an important property of the random binning.
-
-The loss as a function of bin width was computed for both algorithms for the baseball data and is shown in Figure~\ref{fig:LossComparison1}.
+%---
 
@@
     \centering
-		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
-    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
+		\subfloat[Numerical and Visual Loss]{\includegraphics[keepaspectratio=true, width=.46\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
+    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/FreqLoss.pdf}}
 \caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.
-\hh{could we make (a) square again?
-the squished panel looks funky.}  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
+(b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
 	\end{figure}
@@
   lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
-  datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
-    qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
-        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
+  datplot <- rbind(cbind(unique(lRandom$NewData), Algorithm="Random"), cbind(unique(lMean$NewData), Algorithm="Standard"))
+  datplot$xbin <- round(datplot[,1]/binning[1],0)*binning[1]
+  datplot$ybin <- round(datplot[,2]/binning[2],0)*binning[2]
+  #datplot2 <- ddply(datplot, .(G, SO,Algorithm), summarize, count=length(G))
+    qplot(xbin, ybin, geom="tile", data = datplot, fill = fsum , 
+          ,main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
         scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
-    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
+    ggsave(paste("./RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
   return(rbind(
@@
 %   Swap Back at later date (shrinks size of pdf to managable)
-\subfloat[Million-point sample of Airline Data \hh{this is the bad practice - our counter-example of what not to do.
-we need to emphasize this a bit more}]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
+\subfloat[Million-point sample of Airline Data.
+\ktm{This plot is not compiling and needs to be reconstructed (my apologies)}]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
 \subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}

### 2015-02-17 Paper.Rnw
Push minor change, will it let me?

@@
 d1.hex$Freq <-  d1.hex.obj@count
+#<<<<<<< HEAD
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + scale_size_area("Frequency", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+#=======
+qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
+  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+#>>>>>>> origin/master
   ggsave("./figure/BinScatter.pdf", width=5, height=5)

### 2015-02-17 Paper.Rnw
Attempt to resolve merge conflict

@@
 d1.hex$Freq <-  d1.hex.obj@count
-#<<<<<<< HEAD
+
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + scale_size_area("Frequency", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-#=======
+
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
   scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-#>>>>>>> origin/master
+
   ggsave("./figure/BinScatter.pdf", width=5, height=5)

### 2015-02-17 Paper.Rnw
small edits to section 2, unify "alpha blending" term and fix bubble plot label in fig 1

@@
 \section{Scatterplots for Large Data Sets}
-In the case of medium sized data scatterplots are great tools for showing relationships in two dimensions.
+In the case of medium sized data, scatterplots are great tools for showing relationships in two dimensions.
 For large data, scatterplots suffer from over-plotting -- i.e.
@@
 The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
-The number of games played in a season is plotted versus number of strikeouts a pitchr threw over the course of a season.
+The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season.
 While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
@@
 %\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
-A modern alternative to open circles is alpha-blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.
+A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides  more frequency information, allowing a distinction of higher-resolution density.
 
@@
 
-  \subfloat[Hexagonal binning]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{HexBinning.pdf}}
-  \subfloat[Scatterplot of Binned Data]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{BinScatter.pdf}}
-
-	\caption{\label{fig:scatter-alpha} Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha-blending ($\alpha$=0.05).
+  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{HexBinning.pdf}}
+  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{BinScatter.pdf}}
+
+	\caption{\label{fig:scatter-alpha} Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05).
 Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
@@
 Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers.
-The data set is large enough that neither alpha-blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.
+The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.
 
@@
 When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}.
-Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or $\alpha$-blended scatterplots in a more smooth, continuous fashion.
+Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion.
 These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.

### 2015-02-17 Paper.Rnw
Standard Binning Algorithm rewrite

@@
 \section{Binning Data}
-We will only consider binning in two dimensions, $X$ and $Y$.
+We will only consider binning in up to two dimensions, $X$ and $Y$.
 The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.
@@
 \subsection{Standard Rectangular Binning Algorithm}
+We will first define a binning algorithm that results in observed values being allocated to equal sized bins in a deterministic manner.
 Let $x_{(i)}$ correspond to the ordered values of $X$, where $i=1, ..., n$ and $n$ is the number of total observations.
@@
 
-We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le n_X$, where $n_X$ is the overall number of bins in dimension $X$.
-Binning can either be determined by the number of bins, $n_X$, or, equivalently, by the bin width, $b_X$.
+We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.
+Binning can be determined by the number of bins, $N_X$, which then dictates the bin width, $\omega_X$.
 For standard binning the relationship between these two parameters is  given as
 \[
-b_X = (x_{(n)} - x_{(1)})/n_X
+\omega_X = (x_{(n)} - x_{(1)})/N_X
 \]
 
-Mathematically, binning corresponds to a function $b(.)$ that assigns to a value $x$ the closest bin center $x^\ast \in \{ x_j^\ast \}, j \in \{1, ..., n_X\}$, i.e.
-the binning function   $b(.) : x \rightarrow x^\ast$ is defined as
-\begin{eqnarray*}\label{stdbin}
-x^\ast = b(x) =  b_X \cdot \left\lfloor (x-x_1^\ast)/ b_X + 1/2\right\rfloor + x_1^\ast
-\end{eqnarray*}
-where $x_{(1)}$ is the center of the first bin and $\lfloor .
+Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_i \in \{ x_j^\ast \}, j \in \{1, ..., N_X\}$, i.e.
+the binning function   $b_X(.) : x_i \rightarrow x^\ast_i$ is defined as
+% \begin{center}
+% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
+% \vspace{.03}
+% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
+% \end{center}
+\[ 
+ x^\ast_i = b_X(x_i) =
+  \begin{cases}
+    x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor & \quad \forall x_i < x_{(n)} \\
+    x_{(n)} - \omega_x/2 & \quad \forall x_i = x_{(n)}
+  \end{cases}
+\]
+
+where $x_{(1)}^\ast = x_{(1)} + \omega_x/2$ is the center of the first bin and $\lfloor .
 \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.
-An intuitive understanding of the formula is developed by noting that $(x-x_1^\ast)/ b_X$ is the number of bin widths that $x$ is above $x_1^\ast$ and that $\frac{b_X/2}{b_X}=\frac{1}{2}$ is the number of bin widths that $x_1^\ast$ is above $x_{(1)}$.
-The value $\lfloor (x-x_1^\ast)/ b_X + 1/2\rfloor$ then represents the number of bins centers that lie between $x_{(1)}$ and $x$.
-Thus, the binning function is simply finding the location of the bin center for $x$ by adding the distance between the bin centers for $x_{(1)}$ and $x$ to the known bin center for $x_{(1)}$.
-%to allow for bins of width $b_X$ centered at any sequence $O_X, O_X + b_X, ..., O_X + n b_X$.
-
-Alternate binning parameterizations include dividing up the range of data by a specified number of bins of equal width, but even these strategies fit into the framework of assigning values to the closest bin center.
-
-% , with smallest bin
-% $b_X   \cdot \lfloor (x_{(1)}-O_X)/b_X + 0.5\rfloor + O_X$ and largest bin 
-% $b_X   \cdot \lfloor (x_{(n)}-O_X)/b_X + 0.5\rfloor + O_X$.
+An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ represents the number of bins centers that lie between $x_{(1)}$ and $x_i$.
+Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between the bin centers for $x_{(1)}$ and $x_i$ to the known bin center for $x_{(1)}$.\\
+
+Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
+A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
+This strategy would still fit into the framework of assigning values to the closest bin center.
 

### 2015-02-17 Paper.Rnw
fixed?

@@
 
-qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + scale_size_area("Frequency", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
+  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
 
@@
         scale_alpha(range=c(.5, 1), guide=FALSE) +
-        scale_colour_identity(guide=FALSE) + 
-        scale_shape(solid=T,guide=FALSE) +
+      #  scale_colour_identity(guide=FALSE) + 
+      #  scale_shape(solid=T,guide=FALSE) +
         theme_bw() +
@@
         annotate("segment", x=temp$x, y=temp$y, 
-                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
+                 xend=temp$xend, yend=temp$yend, group=temp$group) # + ylim(c(0, 50))
       ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
@@
   \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
-  {./figure/RandomBinningDataPoints1.pdf}}
+  {./figure/RandomBinningDataPoints2.pdf}}
   \caption{Points are assigned to bins centered at (5, 25) an (15,25).
@@
   #datplot2 <- ddply(datplot, .(G, SO,Algorithm), summarize, count=length(G))
-    qplot(xbin, ybin, geom="tile", data = datplot, fill = fsum , 
+    qplot(xbin, ybin, geom="tile", data = datplot, fill = Freq , 
           ,main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
@@
 
-<<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
-loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
-loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
-loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
-loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
-
-loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
-
-table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
-                               c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
-                               c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
-                               c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
-                               c(binwidth=1, loss1[5:7]*100, loss1[8])))
-names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
-print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
-@
+\hh{the  code following this comment in the Rnw file is, I think, broken beyond repair - but we should include a table like the one it produced in the paper.
+- we will have different losses to report, though}
+
+% 
+% <<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
+% loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
+% loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
+% loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
+% loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
+% 
+% loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% 
+% table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
+%                                c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
+%                                c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
+%                                c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
+%                                c(binwidth=1, loss1[5:7]*100, loss1[8])))
+% names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
+% print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
+% @
 
@@
 \textbf{Are we sure that this is because of an odd binwidths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
+
+\hh{This is a valid concern, but we have to fix the notation at the beginning first, before we can address it.}
 
@@
 \begin{figure}[hbtp]\centering
-%\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfil
+\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfill
 %   Swap Back at later date (shrinks size of pdf to managable)
-\subfloat[Million-point sample of Airline Data.
+%\subfloat[Million-point sample of Airline Data.
 \ktm{This plot is not compiling and needs to be reconstructed (my apologies)}]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
@@
 setwd("./data/")
+if (!file.exists("airline.csv.gz")) {
+  # HH: I'm not sure, which files these should be ... 
   # airline <- do.call("rbind", lapply(list.files(), function(i) read.csv(i)))
+
   # write.csv(airline, "airline.csv", row.names=FALSE)
-airline <- read.csv("airline.csv")
+}
+airline <- read.csv("airline.csv.gz")
 na.departure <- (is.na(airline[,8]) + is.na(airline[,9]))> 0
@@
 
+if (!file.exists("ReducedAirlineData.csv")) {
   airline.freq <- ddply(airline, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=nrow(i)))
   write.csv(airline.freq, "ReducedAirlineData.csv", row.names=FALSE)
+}
 airline.freq <- read.csv("ReducedAirlineData.csv")
@@
 
+if (!file.exists("airlineBin5Std.csv")) {
   air.binned5 <- binStd(airline.freq, c(5, 5))
@@
   write.csv(air.binned5, "airlineBin5Std.csv", row.names=FALSE)
-
+}
 # air.binned5.rdm <- binRdm(cbind(airline, Freq=1), c(5, 5))

### 2015-02-18 Paper.Rnw
Add equation number to standard binning algorithm and edit notation in random binning section

@@
 library(RColorBrewer)
+library(hexbin)
 
@@
 % \end{center}
-\[ 
- x^\ast_i = b_X(x_i) =
-  \begin{cases}
+\begin{eqnarray}\label{standbin}
+ x^\ast_i = b_X(x_i) = \left\{\begin{array}{ll} 
     x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor & \quad \forall x_i < x_{(n)} \\
     x_{(n)} - \omega_x/2 & \quad \forall x_i = x_{(n)}
-  \end{cases}
-\]
+  \end{array}\right.
+\end{eqnarray}
 
@@
 An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
-A random binning algorithm utilizes a non-deterministic bin function $b^r(\cdot)$ which maps value $x$ to one of several bin centers $x^\ast$ with probability $p$.
-In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
-The bin function assigns $x$ to bin center $x^\ast$ with a probability depending on the distance to that bin center;  the closer a value is to a bin center, the higher the probability that the value is assigned to this bin center.
+A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
+In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
+The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
 More formally,
+%
 \begin{eqnarray}\label{randbin}
-b^r(x) = \left\{\begin{array}{ll} 
-  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x)}/b_X \\
-  x^\ast_{j+1} &\text{ with probability } 1-p = {(x - x^\ast_{j})}/b_X
+b^r_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
+  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
   \end{array}\right.
@@
 % In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
-This method is easily extensible to also map $x$ into one of more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
+This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
 

### 2015-02-18 Paper.Rnw
Add quantile binning section to draft

@@
 % linear binning in Gold book - random algorithm as a less computational method of linear binning
+
+\subsection{Quantile Binning Algorithm}
+
+We could also consider a quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
+This is a fundimentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
+This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.
+
+Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
+The number of quantile based bins, $N_q ,= J$ , must be defined in order to assign values to bin centers.
+The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as
+%
+\begin{eqnarray}\label{quantbin}
+c^\ast_{(j)} = b^q_C(c_{(j)}) = c_{(\eta)} \text{, where } \eta = \left\lfloor j/ \left\lceil J/N_c \right\rceil \right\rfloor \left\lceil J/N_c \right\rceil ,
+\end{eqnarray}
+%
+\noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.
+To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus the function assigns the $j^{th}$ value to the next lower ordered value that is a multiple of $n_q$.
 

### 2015-02-18 Paper.Rnw
Spell check and minor grammar repair

@@
 Histograms are an example of plots which use binned variables.
-To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tesselated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile.
+To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile.
 A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.
@@
   \caption{Points are assigned to bins centered at (5, 25) an (15,25).
-Points in the standard algorithm are contained within the defined bin region, where points from the random algorith may be assigned to bins from outside the defined bin regions.}
+Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to bins from outside the defined bin regions.}
   \label{fig:BinningTutorial}
@@
 We could also consider a quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
-This is a fundimentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
+This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
 This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.
@@
 \noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.
-To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus the function assigns the $j^{th}$ value to the next lower ordered value that is a multiple of $n_q$.
+To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus $\eta$ is the index for the ordered value where the next lower quantile based bin will be centered.
 
@@
 This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.
-This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions napply to either approach to binning.
+This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions apply to either approach to binning.
 
@@
 
-Frequency loss is argueably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary.
+Frequency loss is arguably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary.
 The standard and random algorithms are very similar when we compare frequency loss.
-The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticable than the increase seen in numerical loss.
+The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticeable than the increase seen in numerical loss.
 %~\ref{fig:LossComparison}
@@
 Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues.
-Random binning is especially advantagous when discrete data values are recorded at a consistant increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths.
-In this senario, we see that the non-synchronous data increments and bin widths will cause bins to contain unconsistant numbers of possible data value.
+Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths.
+In this scenario, we see that the non-synchronous data increments and bin widths will cause bins to contain inconstant numbers of possible data value.
 Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency.
-Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artifical frequency discrepancy that manifests in a striped pattern in the visualization.
+Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artificial frequency discrepancy that manifests in a striped pattern in the visualization.
 The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.
@@
 The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}.
-We find that the visual and numerical loss are in fact higher while using the random binning algorithm, but that the two binning algorithms are almost indestiguishable in terms of frequency loss.
+We find that the visual and numerical loss are in fact higher while using the random binning algorithm, but that the two binning algorithms are almost indistinguishable in terms of frequency loss.
 
 Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.
-This is appearant in the order of colored lines based on the Strike Out binning size.
+This is apparent in the order of colored lines based on the Strike Out binning size.
 This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as binwidth increases.
@@
 These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the binwidths are integer multiples of the data increments.
-This reinforces the arguement that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
+This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
 

### 2015-02-18 Paper.Rnw
Standard binning reorder: define general case with bins specified by first center and binwidths, not data range special case

@@
 We will first define a binning algorithm that results in observed values being allocated to equal sized bins in a deterministic manner.
-Let $x_{(i)}$ correspond to the ordered values of $X$, where $i=1, ..., n$ and $n$ is the number of total observations.
-Then $x_{(1)}$ and $x_{(n)}$ then denote the observed minimum and the maximum of $X$, respectively.
-
 We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.
-Binning can be determined by the number of bins, $N_X$, which then dictates the bin width, $\omega_X$.
-For standard binning the relationship between these two parameters is  given as
-\[
-\omega_X = (x_{(n)} - x_{(1)})/N_X
-\]
-
-Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_i \in \{ x_j^\ast \}, j \in \{1, ..., N_X\}$, i.e.
-the binning function   $b_X(.) : x_i \rightarrow x^\ast_i$ is defined as
+To define the set of bin centers we need only specify the center of the first bin, $x^\ast_1$ and the binwidths, $\omega_X$.
+Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
+the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as
 % \begin{center}
@@
 \begin{eqnarray}\label{standbin}
- x^\ast_i = b_X(x_i) = \left\{\begin{array}{ll} 
-    x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor & \quad \forall x_i < x_{(n)} \\
-    x_{(n)} - \omega_x/2 & \quad \forall x_i = x_{(n)}
+b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
+\end{eqnarray}
+
+where $x_1^\ast$ is the center of the first bin and $\lfloor .
+\rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.
+An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
+Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.
+
+A special case of standard binning can be used to divide the data range into $N_X$ equal sized bins.
+To do so we use the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
+The first bin would then be centered at $x_{(1)} + \omega_X/2$.
+
+%Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
+A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
+This strategy would still fit into the framework of assigning values to the closest bin center.
+
+\subsection{Random Binning Algorithm}
+
+An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
+A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
+In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
+The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
+More formally,
+%
+\begin{eqnarray}\label{randbin}
+b^r_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
+  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
   \end{array}\right.
@@
 
-where $x_{(1)}^\ast = x_{(1)} + \omega_x/2$ is the center of the first bin and $\lfloor .
-\rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.
-An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ represents the number of bins centers that lie between $x_{(1)}$ and $x_i$.
-Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between the bin centers for $x_{(1)}$ and $x_i$ to the known bin center for $x_{(1)}$.\\
-
-Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
-A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
-This strategy would still fit into the framework of assigning values to the closest bin center.
+% In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
+This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
+
+%To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
+%Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
+%1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
+
+%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
+
+The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
+``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
+With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
+Specifically, the expectation of the random binning algorithm is the same as for linear binning.
+%However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
+In the next section, we examine the loss of visual and numerical information due to binning.
+% linear binning in Gold book - random algorithm as a less computational method of linear binning
+
+\subsection{Extensions to Rectangular Binning}
 
@@
 
-\begin{figure}[hbtp]
-  \centering
-  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
-%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
-%   {./figure/RandomBinningDataPoints1}}
-  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
-  {./figure/RandomBinningDataPoints2.pdf}}
-  \caption{Points are assigned to bins centered at (5, 25) an (15,25).
-Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to bins from outside the defined bin regions.}
-  \label{fig:BinningTutorial}
-\end{figure}
-
-\subsection{Random Binning Algorithm}
 <<RandomBinDemo,echo=F,include=FALSE>>=
@@
 @
-An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
-A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
-In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
-The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
-More formally,
-%
-\begin{eqnarray}\label{randbin}
-b^r_X(x_i) = \left\{\begin{array}{ll} 
-  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
-  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
-  \end{array}\right.
-\end{eqnarray}
-
-% In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
-This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
-
-%To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
-%Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
-%1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
-
-%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
-
-The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
-``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
-With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
-Specifically, the expectation of the random binning algorithm is the same as for linear binning.
-%However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
-In the next section, we examine the loss of visual and numerical information due to binning.
-% linear binning in Gold book - random algorithm as a less computational method of linear binning
+
+\begin{figure}[hbtp]
+  \centering
+  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
+%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
+%   {./figure/RandomBinningDataPoints1}}
+  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
+  {./figure/RandomBinningDataPoints2.pdf}}
+  \caption{Points are assigned to bins centered at (5, 25) an (15,25).
+Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to bins from outside the defined bin regions.}
+  \label{fig:BinningTutorial}
+\end{figure}
 

### 2015-02-18 Paper.Rnw
first draft of subsection for 2d extension for rectangular binning

@@
 \subsection{Extensions to Rectangular Binning}
+
+The standard and random binning algorithms are easily extendible to higher dimension.
+For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
+In this case we wish to assign data tuples $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins in $\mathbb{R}^2$.
+The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
+%
+\begin{eqnarray}\label{standrecbin}
+b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
+\end{eqnarray}
+%
+where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
+The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
+%
+\begin{eqnarray}\label{randrecbin}
+b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
+\end{eqnarray}
+%
+where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
 
@@
 
-We could also consider a quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
+We could also consider a empirical quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
 This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
@@
 Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
-The number of quantile based bins, $N_q ,= J$ , must be defined in order to assign values to bin centers.
+The number of quantile based bins, $N_q$ , must be defined in order to assign values to bin centers.
 The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as

### 2015-02-18 Paper.Rnw
Small edits to rectangular binning section

@@
 
-Binning is a two-step procedure: we first assign each data tuple $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+Binning is a two-step procedure: we first assign each data pairs $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
 
@@
 For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
-In this case we wish to assign data tuples $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins in $\mathbb{R}^2$.
+In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins in $\mathbb{R}^2$.
 The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
@@
 where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
-
-Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process extended to a two dimensional situation.
+Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
+
 <<binningDemo,echo=F,include=F,cache=T>>=
@@
   \centering
-  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
+  \subfloat[Standard Binning]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
 %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
 %   {./figure/RandomBinningDataPoints1}}
-  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
+  \subfloat[Random Binning]{\includegraphics[keepaspectratio=TRUE, width=3in]
   {./figure/RandomBinningDataPoints2.pdf}}
-  \caption{Points are assigned to bins centered at (5, 25) an (15,25).
+  \caption{Visualization of how data pairs are assigned to bins centered at (5, 25) and (15,25) under the two rectangular binning algorithms.
 Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to bins from outside the defined bin regions.}

### 2015-02-18 Paper.Rnw
quantile binning edits

@@
 
-\begin{figure}[hbtp]
+\begin{figure}[h]
   \centering
@@
 Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
-The number of quantile based bins, $N_q$ , must be defined in order to assign values to bin centers.
-The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as
+Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers.
+It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function.
+\ktm{Note this is the quantile function as defined in R; find related literature source.} The bin centers are defined as the midpoint of each consecutive pair of bin boundaries $\{c^\ast_k = (Q_{k-1/K} + Q_{k/K})/2  \text{ for } k=0,\dots,K \}$.
+The quantile binning function $b^q_C(\cdot) : (c_j) \rightarrow c^\ast_k$ is defined as
 %
 \begin{eqnarray}\label{quantbin}
-c^\ast_{(j)} = b^q_C(c_{(j)}) = c_{(\eta)} \text{, where } \eta = \left\lfloor j/ \left\lceil J/N_c \right\rceil \right\rfloor \left\lceil J/N_c \right\rceil ,
+b^q_C(c_j) = (Q_{k-1/K} + Q_{k/K})/2 \text{ where } Q_{k-1/K} < c_j \le Q_{k/K}.
 \end{eqnarray}
 %
-\noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.
-To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus $\eta$ is the index for the ordered value where the next lower quantile based bin will be centered.
+The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in stictly the univariate setting for this manuscript.
 

### 2015-02-24 Paper.Rnw
couple of edits

@@
 \section{Introduction}
+
+\hh{
+\begin{itemize}
+\item Additional literature: binning papers: standard binning, e.g.: Karl Pearson (1895) ``Contributions to the Mathematical Theory of Evolution. II. Skew Variation in Homogeneous Material"\\
+David W. Scott (1979). "On optimal and data-based histograms". Biometrika 66 (3): 605–610
+
+quantile binning: there's a lot of implementations (e.g.
+cut2 in package Hmisc, the rattle package, the package extracut.
+It would be interesting to check, if they do the same things (probably not) and mention that there is disagreement in the implementations).
+
+As a reference I would check in 
+Scott, D. W. 1992.
+Multivariate Density Estimation: Theory, Practice, and Visualization. New York: Wiley.
+
+\item make sure to verify that the binning algorithms agree with the literature (and check that the code is doing the same thing as described in the paper).
+You can include a bit of discussion on the problem that histograms are over-parameterized - it's a known problem, but I can't find any paper on it at the moment.
+
+\end{itemize}
+}
 
@@
 
-\section{Binning Data}
+\section{Binning \hh{Algorithms}}
+\hh{XXX Motivation first - starting with the limitiations too  negative.}
+
 We will only consider binning in up to two dimensions, $X$ and $Y$.
@@
 
-Binning is a two-step procedure: we first assign each data pairs $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
-
- We will proceed with rectangular bins for simplicity, but other binning schemes, such as those which utilize hexagonal bins are also common \citep{scatterplots}.
-Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two dimensions.
+Binning is a two-step procedure: we first assign each \hh{observation} $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center \hh{``frequency of each bin center" ...
+that needs a few more words, because there is only one bin center.
+But what you mean is that each observation assigned to the bin center now counts as one value for the bin center.
+XXX The next sentence explains that, so combine the two sentences a bit more}, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+
+ We will proceed with rectangular bins for simplicity, but other binning schemes, such as \sout{those which utilize} hexagonal bins are also common \citep{scatterplots} \hh{XXX don't put citations at the end of the sentence, put them right after the most relevant phrase.
+in this case 'hexagonal binning'.}.
+Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two \hh{or more} dimensions.
+
+\hh{There are a lot of commonalities between the different univariate binning algorithms - it would make sense to write a single, more general section, and give a list of specific parameterisations for the individual binning approaches.
+}
 
 \subsection{Standard Rectangular Binning Algorithm}
-We will first define a binning algorithm that results in observed values being allocated to equal sized bins in a deterministic manner.
+\hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
+
+\hh{Define what $X$ is, what is $x_i$}
+
+\hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }
+
+We \sout{will} first define a binning algorithm that \hh{allocates}  observed values to equal sized bins in a deterministic manner.
 We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.
 To define the set of bin centers we need only specify the center of the first bin, $x^\ast_1$ and the binwidths, $\omega_X$.
+\hh{I am still not sure how bin centers are defined in the end. XXX}
+
+\hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}
+
 Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
@@
 
-where $x_1^\ast$ is the center of the first bin and $\lfloor .
-\rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.
+where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. 
+\hh{something went wrong with the next sentence}
 An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
@@
 
+\hh{get rid of the woulds, shoulds, cans, ...}
 A special case of standard binning can be used to divide the data range into $N_X$ equal sized bins.

### 2015-02-24 Paper.Rnw
consistency in sections 3.2 and 3.3

@@
 For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
-In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins in $\mathbb{R}^2$.
+In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins \sout{in $\mathbb{R}^2$} \hh{XXX J is an integer number, definitely not a real-valued vector}.
 The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
@@
 Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
+
+\hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation.
+$J$ is defined, as above as a multiple of $N_X$ and $N_Y$, this needs to be properly adjusted in the binning algorithm.
+It will be some linear relationship of the form $j_X\cdot N_X + j_Y$.
+}
 
@@
 
-Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
+Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ \hh{shouldn't you go back to the same notation as before?
+$N_X$, not $J$?
+what are the $c_j$ bin centers?
+before they were $c_j^ast$.} that we wish to divide into quantile based bins.
 Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers.
 It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function.
-\ktm{Note this is the quantile function as defined in R; find related literature source.} The bin centers are defined as the midpoint of each consecutive pair of bin boundaries $\{c^\ast_k = (Q_{k-1/K} + Q_{k/K})/2  \text{ for } k=0,\dots,K \}$.
+\ktm{Note this is the quantile function as defined in R; find related literature source.} \hh{which function exactly?
+There are multiple.} The bin centers are defined as the midpoint of each consecutive pair of bin boundaries $\{c^\ast_k = (Q_{k-1/K} + Q_{k/K})/2  \text{ for } k=0,\dots,K \}$.
 The quantile binning function $b^q_C(\cdot) : (c_j) \rightarrow c^\ast_k$ is defined as

### 2015-02-25 Paper.Rnw
Changes made to generally binning description

@@
 
-\section{Binning \hh{Algorithms}}
+\section{Binning Algorithms}
 \hh{XXX Motivation first - starting with the limitiations too  negative.}
@@
 
-Binning is a two-step procedure: we first assign each \hh{observation} $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center \hh{``frequency of each bin center" ...
-that needs a few more words, because there is only one bin center.
-But what you mean is that each observation assigned to the bin center now counts as one value for the bin center.
-XXX The next sentence explains that, so combine the two sentences a bit more}, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+\ktm{Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.}
 

### 2015-02-25 Paper.Rnw
Minor edits and note

@@
 
- We will proceed with rectangular bins for simplicity, but other binning schemes, such as \sout{those which utilize} hexagonal bins are also common \citep{scatterplots} \hh{XXX don't put citations at the end of the sentence, put them right after the most relevant phrase.
-in this case 'hexagonal binning'.}.
-Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two \hh{or more} dimensions.
+ We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
+Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two or more dimensions.
 
@@
 In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins \sout{in $\mathbb{R}^2$} \hh{XXX J is an integer number, definitely not a real-valued vector}.
-The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
+\ktm{The wording here was not clear.
+J does not exist in $\mathbb{R}^2$, but it is being used to index 2d bins that exist in that plane.} The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
 %

### 2015-02-26 Paper.Rnw
Notes about Pearson and Scott

@@
 }
+
+\ktm{I have read through the sections of Pearson (1895), Scott (1979) and Scott (1992) that pertained to binning.
+I will work on integrating them into the lit review next.
+Scott talked much about the optimal choice of bin width based on MSE but there was not much talk about what to do with points at the boundaries of bins (he effectively assumed infinite precision in the measurement process to avoid it).
+
+I will need to search more to find papers about the binning algorithms and the boundaries they set.
+The \texttt{cut} function allows for intervals that are closed on either the left or right and also allows the points on the outermost open bound to be included in the last bin.
+
+The \texttt{cut2} function allows the same type of quantile binning as discussed in Scott (1992).}
 

### 2015-02-26 Paper.Rnw
standard binning section edits

@@
 
-We \sout{will} first define a binning algorithm that \hh{allocates}  observed values to equal sized bins in a deterministic manner.
-We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.
-To define the set of bin centers we need only specify the center of the first bin, $x^\ast_1$ and the binwidths, $\omega_X$.
 \hh{I am still not sure how bin centers are defined in the end. XXX}
@@
 
-Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
+\ktm{  We define the process that allocates observed values, $x_i$ for $i \in \{1, ..., n\}$, to equal sized bins in a deterministic manner as a standard binning algorithm.
+The bins are controlled by two parameters: the center of the first bin, $x^\ast_1$ and the binwidth, $\omega_X$.
+This leads to the set of bin centers $\{ x_j^\ast =  x_1^\ast + (j-1)\omega_X \text{ for } j \in \{1, ..., N_X\} \}$,  where $N_X$ is the overall number of bins.
+}
+
+Mathematically, the standard binning algorithm corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
 the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as
@@
 where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
-\hh{something went wrong with the next sentence}
-An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
+\ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $. }
+%\hh{something went wrong with the next sentence}
+%An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
 Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.
 
-\hh{get rid of the woulds, shoulds, cans, ...}
-A special case of standard binning can be used to divide the data range into $N_X$ equal sized bins.
-To do so we use the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
+\ktm{ A special case of standard binning divides the data range into $N_X$ equal sized bins.
+This leads to the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
 The first bin would then be centered at $x_{(1)} + \omega_X/2$.
+}
 

### 2015-02-26 Paper.Rnw
more edits 2/25

@@
 The bins are controlled by two parameters: the center of the first bin, $x^\ast_1$ and the binwidth, $\omega_X$.
-This leads to the set of bin centers $\{ x_j^\ast =  x_1^\ast + (j-1)\omega_X \text{ for } j \in \{1, ..., N_X\} \}$,  where $N_X$ is the overall number of bins.
+This leads to bin centers $ x_j^\ast =  x_1^\ast + (j-1)\omega_X \text{ for } j \in \{1, ..., N_X\} $,  where $N_X$ is the overall number of bins.
 }
@@
 where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
-\ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $. }
+\ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $.
+After all observations have been allocated to bins, the bin count, $c_j$, is joined with the bin center to form the tuple $(x^\ast_1, c_j)$ }
 %\hh{something went wrong with the next sentence}
@@
 It will be some linear relationship of the form $j_X\cdot N_X + j_Y$.
+}  \ktm{ I am having trouble tring to reconcile the indexing of bins in 2 dimensions.
+If we use an pair of indeces it makes the extension of binning to 2 dimensions easy; but it makes the notation messier and loss functions more difficult to define.
 }

### 2015-02-26 Paper.Rnw
response note made in the quantile binning section

@@
 This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
-This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.
-
-Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ \hh{shouldn't you go back to the same notation as before?
+This binning will not be considered for binning the data pairs directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.
+
+Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
+Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers.
+\hh{shouldn't you go back to the same notation as before?
 $N_X$, not $J$?
 what are the $c_j$ bin centers?
-before they were $c_j^ast$.} that we wish to divide into quantile based bins.
-Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers.
-It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function.
+before they were $c_j^\ast$.} \ktm{I conciously decided to define quantile binning in a way consistent with how it will be used later in the paper for discretizing the color palettes for the $c_j$ counts.
+} It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function.
 \ktm{Note this is the quantile function as defined in R; find related literature source.} \hh{which function exactly?

### 2015-02-27 Paper.Rnw
Added discussion and references for the two generalized scatterplot papers

@@
 A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density. 
+\ktm{ Generalized scatterplots \citep{GenScat} have also been introduced, which display all individual observations but use distortion of the point locations and axes to avoid overlapping points.
+This approach preserves fine data structure and allows color to be reserved for a third variable instead of frequency; however, the distortion warps the spatial information for the primary locations variables.
+A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{EllipseScatter}.}
 

### 2015-02-27 Paper.Rnw
edits to generalized scatterplot discussion

--- 2015-02-26 Paper.Rnw
@@
 A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density. 
+\ktm{ Generalized scatterplots \citep{GenScat} have also been introduced, which display all individual observations but use distortion of the point locations to avoid overlapping points.
+This approach preserves fine data structure and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations wraps the visualization of the association between the two primary variables.
+A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{EllipseScatter}.}
 

### 2015-02-27 Paper.Rnw
move paragraph and add variable binned scatterplot citation

--- 2015-02-26 Paper.Rnw
@@
 Making the data aggregation explicit allows us to calculate the loss we experience.}
+
+\ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.
+Generalized scatterplots \citep{GeneralizedScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
+A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{EllipseScatter}.
+Variable binned scatterplots \citep{hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
+This variable binning fragments the continuity of the axes into many segments on different scales.
+Theses approachs make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}
 

### 2015-02-27 Paper.Rnw
add Hadley "bin-summarize-smooth" paper and Liu "imMens" paper citations to bib

@@
 \ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.
-Generalized scatterplots \citep{GeneralizedScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
-A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{EllipseScatter}.
-Variable binned scatterplots \citep{hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
+Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
+A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
+Variable binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
 This variable binning fragments the continuity of the axes into many segments on different scales.
@@
 This has the additional advantage of reducing the size of the stored data, as only the bin centers and the frequency of points which correspond to that bin must be stored.
+\ktm{ Add \citep{Wickham2013Bin}  \citep{Liu2013imMens} here to argue the merits of binning.
+}
 

### 2015-02-27 Paper.Rnw
comments in section 2

@@
 As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information.
-Tukey's open circles (b) help some, but are as prone to over-plotting as solid circles when the data set is large.
-Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers.
+Tukey's open circles (b) help \hh{to a degree, but are also prone to over-plotting  when the data set is very large.} Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers.
 The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.
@@
 \ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.
-Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
-A extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
+\hh{These approaches do not help in the situation of ties in the data, which are very common for large data sets due to the finite resolution of data collection (e.g.\ airline data is collected at minute intervals).}  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
+A\hh{n} extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
 Variable binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
 This variable binning fragments the continuity of the axes into many segments on different scales.
-Theses approachs make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}
+Theses approachs make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.} \hh{please check the spelling in the previous paragraph.
+}
 
@@
 \ktm{ Add \citep{Wickham2013Bin}  \citep{Liu2013imMens} here to argue the merits of binning.
-}
+} \hh{also add \ref{gold} for binning reference - it's called `weighting' there.}
 
 Histograms are an example of plots which use binned variables.
-To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile.
-A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.
+\hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information.
+Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
+%To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step
+\hh{A natural extensions of histograms to higher dimensions is} to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
+A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, \hh{similar to} a two-dimensional histogram viewed from above.
 
@@
 Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion.
-These estimates require parameter tuning and may hide gaps in the data by over-smoothing while simultaneously de-emphasizing outlying points.
-
-As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e).
+\hh{However,} these estimates require \hh{careful} parameter tuning, \hh{as over-smoothing} may hide gaps in the data while simultaneously de-emphasizes outlying points.
+
+As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) \hh{XXX the bubble plots already got mention above XXX} and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e).
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The binned scatterplot is less prone to the Hermann-grid illusion %\hh{need a reference}
+The binned scatterplot is less prone to the Hermann-grid illusion \hh{needs a reference}
 than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.

### 2015-02-27 Paper.Rnw
changes during the meeting

@@
 \ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.
-\hh{These approaches do not help in the situation of ties in the data, which are very common for large data sets due to the finite resolution of data collection (e.g.\ airline data is collected at minute intervals).}  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
+\hh{Address how ties in the data are dealt with, which are very common for large data sets due to the finite resolution of data collection (e.g.\ airline data is collected at minute intervals).
+jittering?}  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
 A\hh{n} extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.

### 2015-02-28 Paper.Rnw
Start to Section 3 Binning Total Rewrite

@@
 This variable binning fragments the continuity of the axes into many segments on different scales.
-Theses approachs make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.} \hh{please check the spelling in the previous paragraph.
+Theses approaches make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.} \hh{please check the spelling in the previous paragraph.
 }
@@
 \section{Binning Algorithms}
-\hh{XXX Motivation first - starting with the limitiations too  negative.}
-
-We will only consider binning in up to two dimensions, $X$ and $Y$.
+
+Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.
+Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+
+We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
+Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
+We will however only consider binning in up to two dimensions, $X$ and $Y$.
 The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.
-Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.
-
-\ktm{Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.}
-
- We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
-Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two or more dimensions.
-
-\hh{There are a lot of commonalities between the different univariate binning algorithms - it would make sense to write a single, more general section, and give a list of specific parameterisations for the individual binning approaches.
+
+\ktm{  For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, the binning requires a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
+Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
+Values that fall at exactly $\beta_0$ are grouped into the first bin to close the leftmost bound.
+Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs.
+}
+
+\ktm{  This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as }
+%
+\ktm{  \begin{eqnarray}\label{rectbin}
+b_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
+  x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
+  \end{array}\right.
+\end{eqnarray}  }
+%
+\ktm{Note that this binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
+These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}.
+These specifications were selected for this paper, but these choices are by no means consider universal for binning.
+For example, the \texttt{ggplot2} packages creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.
+}
+
+\ktm{ An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
+In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
+The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
+More formally, }
+%
+\ktm{ \begin{eqnarray}\label{randbin1}
+b^r_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
+  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
+  \end{array}\right.
+\end{eqnarray}  
+}
+%
+\ktm{  for all $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. }
+
+\ktm{
+\begin{table}[hbtp]
+\small
+\centering
+\begin{tabular}{cccc} \hline
+ Type & Function & Boundaries & Centers \\ 
+ \hline  
+General & $b_X(x_i)$  &  &  \\
+%-------------
+Standard & $b_X(x_i)$  &  &  \\
+%-------------
+Quantile & $b_X(x_i)$  &  &  \\
+%-------------
+Random & $b^r_X(x_i)$   &   &  \\
+\hline
+\end{tabular}
+\caption{Rectangular Binning Functions}
+\label{tab:rectbinning}
+\end{table}
+\normalsize
+%\hh{There are a lot of commonalities between the different univariate binning algorithms - it would make sense to write a single, more general section, and give a list of specific parameterisations for the individual binning approaches.
+}
 }
@@
 
-The standard and random binning algorithms are easily extendible to higher dimension.
+The standard and random binning algorithms are easily extendable to higher dimension.
 For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
@@
 It will be some linear relationship of the form $j_X\cdot N_X + j_Y$.
-}  \ktm{ I am having trouble tring to reconcile the indexing of bins in 2 dimensions.
-If we use an pair of indeces it makes the extension of binning to 2 dimensions easy; but it makes the notation messier and loss functions more difficult to define.
+}  \ktm{ I am having trouble trying to reconcile the indexing of bins in 2 dimensions.
+If we use an pair of indices it makes the extension of binning to 2 dimensions easy; but it makes the notation messier and loss functions more difficult to define.
 }
@@
 what are the $c_j$ bin centers?
-before they were $c_j^\ast$.} \ktm{I conciously decided to define quantile binning in a way consistent with how it will be used later in the paper for discretizing the color palettes for the $c_j$ counts.
+before they were $c_j^\ast$.} \ktm{I consciously decided to define quantile binning in a way consistent with how it will be used later in the paper for discretizing the color palettes for the $c_j$ counts.
 } It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function.
@@
 %
-The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in stictly the univariate setting for this manuscript.
+The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in strictly the univariate setting for this manuscript.
 

### 2015-02-28 Paper.Rnw
continued work on binning section rewrite

@@
 %
-\ktm{  \begin{eqnarray}\label{rectbin}
+\begin{eqnarray}\label{rectbin}
 b_X(x_i) = \left\{\begin{array}{ll} 
@@
   \end{array}\right.
-\end{eqnarray}  }
+\end{eqnarray}  
 %
@@
 }
+
+\ktm{  Paragraph: Discussion of types of deterministic rectangular binning.  Standard and Quantile. Table~\ref{tab:rectbinning} }
+
+\begin{table}[hbtp]
+\centering
+\begin{tabular}{lll} \hline
+ & Bin Boundaries & Bin Centers \\ 
+ \hline  
+General &  $ \{\beta_j \text{ }|\text{ } \beta_j > \beta_{j-1} \} $ & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = (\beta_{j-1}+ \beta_j)/2 \}$ \\
+%-------------
+Standard \hspace{0.5cm} & $ \{\beta_j \text{ }|\text{ } \beta_j = \beta_{j-1} + \omega_X \} $\hspace{0.5cm} & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = \beta_{j-1} + \omega_X/2 \}$ \\
+%-------------
+Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
+%-------------
+Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
+\hline
+\end{tabular}
+\caption{Rectangular Binning Specifications}
+\label{tab:rectbinning}
+\end{table}
 
@@
 %
-\ktm{  for all $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. }
-
-\ktm{
-\begin{table}[hbtp]
-\small
-\centering
-\begin{tabular}{cccc} \hline
- Type & Function & Boundaries & Centers \\ 
- \hline  
-General & $b_X(x_i)$  &  &  \\
-%-------------
-Standard & $b_X(x_i)$  &  &  \\
-%-------------
-Quantile & $b_X(x_i)$  &  &  \\
-%-------------
-Random & $b^r_X(x_i)$   &   &  \\
-\hline
-\end{tabular}
-\caption{Rectangular Binning Functions}
-\label{tab:rectbinning}
-\end{table}
-\normalsize
-%\hh{There are a lot of commonalities between the different univariate binning algorithms - it would make sense to write a single, more general section, and give a list of specific parameterisations for the individual binning approaches.
-}
+\ktm{  for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.
+This probability structure is inversely proportional to the distance from each bin center.
+This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes.
+}
+
+\ktm{  The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the bin center.
+``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning in which adjacent bins are assigned a weight depending on the distance from the point to that bin, where all weights sum to one.
+With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
+Specifically, the expectation of the random binning algorithm is the same as for linear binning.
 }

### 2015-02-28 Paper.Rnw
reference for Hermann grid illusion

@@
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The binned scatterplot is less prone to the Hermann-grid illusion \hh{needs a reference}
+The binned scatterplot is less prone to the Hermann-grid illusion \citep{hermann:1870}
 than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.

### 2015-02-28 Paper.Rnw
Continued (2) work on binning rewrite

--- 2015-02-28 Paper.Rnw
@@
 
-\ktm{  Paragraph: Discussion of types of deterministic rectangular binning.  Standard and Quantile. Table~\ref{tab:rectbinning} }
+\ktm{ \textit{Standard rectangular binning} is a special cases of this general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified.
+Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density.
+}
+
+\ktm{  \textit{Quantile binning} is another option that divides the range of the observations into bins each containing equal number of points.
+The $j^{th}$ bin intervals is $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
+Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernable groups in visualization using discrete color palettes.
+}
+
+\ktm{  The bin boundaries and centers for each type of rectangular binning algorithm discussed above can be found in Table~\ref{tab:rectbinning}.  }
 
@@
 \end{tabular}
-\caption{Rectangular Binning Specifications}
+\caption{Rectangular and Random Binning Specifications}
 \label{tab:rectbinning}
@@
 This probability structure is inversely proportional to the distance from each bin center.
-This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes.
+In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.
+This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.
 }

### 2015-03-01 Paper.Rnw
Edits to Section 2

@@
 \section{Scatterplots for Large Data Sets}
-In the case of medium sized data, scatterplots are great tools for showing relationships in two dimensions.
+In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions.
 For large data, scatterplots suffer from over-plotting -- i.e.
-more and more points are drawn in close-by places, thereby masking relevant structure.
+\ktm{a large number of observations drawn in such close proximity that their points overlap}, thereby masking relevant structure.
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
@@
 As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information.
-Tukey's open circles (b) help \hh{to a degree, but are also prone to over-plotting  when the data set is very large.} Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers.
+Tukey's open circles (b) help to a degree, but are also prone to over-plotting  when the data set is very large.
+Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers.
 The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.
@@
 \ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.
-\hh{Address how ties in the data are dealt with, which are very common for large data sets due to the finite resolution of data collection (e.g.\ airline data is collected at minute intervals).
-jittering?}  Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations to avoid overlapping points.
-A\hh{n} extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
-Variable binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
-This variable binning fragments the continuity of the axes into many segments on different scales.
-Theses approaches make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.} \hh{please check the spelling in the previous paragraph.
-}
+Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations by having points repel one another to avoid overlapping points.
+An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
+Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
+This variable binning fragments the continuity of the axes into many segments on different scales and also does not deal with points that are exactly overlapped due to the resolution of data collection.
+Generalized and variable-binned scatterplots make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}
 
@@
 
-Histograms are an example of plots which use binned variables.
+%Histograms are an example of plots which use binned variables.
 \hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information.
 Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
-%To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step
-\hh{A natural extensions of histograms to higher dimensions is} to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
-A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, \hh{similar to} a two-dimensional histogram viewed from above.
+\ktm{Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.}  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
+A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, similar to a two-dimensional histogram viewed from above.
 
@@
 Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion.
-\hh{However,} these estimates require \hh{careful} parameter tuning, \hh{as over-smoothing} may hide gaps in the data while simultaneously de-emphasizes outlying points.
+However, these estimates require careful parameter tuning, as over-smoothing may hide gaps in the data while simultaneously de-emphasizes outlying points.
 

### 2015-03-01 Paper.Rnw
Rough rearrangement of section 2 + edits -Start with simple adjustments to points (open, alpha blend) -Adjust point locations (generalized, ellipse, variable-binned) -binned aggregations (sunflower, kernal, binned scatterplots, bubble plots)

@@
 While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
-Figure~\ref{fig:scatter-alpha} shows a traditional scatterplot on the left.
-Each observation is drawn with a filled circle.
-A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts.
+
+\ktm{ Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle.
+A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable}.
 Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting.
@@
 
-\ktm{ A number of scatterplot adaptations have been introduced that avoid overplotting by manipulating point locations or location scales.
+\ktm{ Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales.
 Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations by having points repel one another to avoid overlapping points.
@@
 
+Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and modified scatterplots \citep{martin-gold}.
+Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin.
+Sunflower plots are particularly useful when the number of points in each bin remains reasonably small.
+Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion.
+However, these estimates require careful parameter tuning, as over-smoothing may hide gaps in the data while simultaneously de-emphasizes outlying points.
+
 %Histograms are an example of plots which use binned variables.
@@
 Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
-\ktm{Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.}  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
-A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, similar to a two-dimensional histogram viewed from above.
-
-Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, modified scatterplots, and kernel density smoothing of tonal variation \citep{martin-gold}.
-Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin.
-Sunflower plots are particularly useful when the number of points in each bin remains reasonably small.
-
-When the point area is scaled in proportion to frequency, scatterplots can be used to display the binned data as well.
-When points are filled circles, these plots are also known as ``bubble plots", which were first used by William Playfair \citep{playfair, playfair2}.
-Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion.
-However, these estimates require careful parameter tuning, as over-smoothing may hide gaps in the data while simultaneously de-emphasizes outlying points.
-
-As alternatives, Figure~\ref{fig:scatter-alpha} (d-e) contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) \hh{XXX the bubble plots already got mention above XXX} and a ``bubble plot" (referred to as a modified scatterplot) with frequency encoded as point size (e).
+\ktm{Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.
+} A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
+\ktm{ A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, similar to a two-dimensional histogram viewed from above.
+A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.
+Bubble plots were first used by William Playfair \citep{playfair, playfair2}.
+}
+
+Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (\ref{fig:scatter-alpha}d) and a bubble plot with frequency encoded as point size (\ref{fig:scatter-alpha}e).
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The binned scatterplot is less prone to the Hermann-grid illusion \citep{hermann:1870}
-than the bubble plot, particularly for binned data, where bin centers usually fall on a grid.
+The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870} especially for data binned on a grid; whereas the binned scatterplot will not on a tesselated grid.
 

### 2015-03-01 Paper.Rnw
Update hadley and liu et al paragraph introducing binning to reduce graphical complexity/computational benefits.

@@
 
-One approach to reduce the graphical complexity is to bin the displayed data.
-This has the additional advantage of reducing the size of the stored data, as only the bin centers and the frequency of points which correspond to that bin must be stored.
-\ktm{ Add \citep{Wickham2013Bin}  \citep{Liu2013imMens} here to argue the merits of binning.
-} \hh{also add \ref{gold} for binning reference - it's called `weighting' there.}
+\ktm{ Another approach is to reduce the graphical complexity by plotting binned aggregations, namely frequencies, of the data as opposed to plotting every observation as an individual point.} This has the additional advantage of reducing the size of the stored data \ktm{necessary for the construction of the plot}, as only the bin centers and the frequency of points which correspond to that bin must be stored.
+\ktm{ Wickham argues for a ``bin-summarize-smooth'' procedure to to be applied to the visualization of big data and notes that simple summary functions, such as counts, scale very well with the size of data sources \citep{Wickham2013Bin}.
+Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.
+} \hh{also add \citep{gold} for binning reference - it's called `weighting' there.}
 

### 2015-03-01 Paper.Rnw
Add Wickham and Scott references to binning sections. Also spell check.

@@
 
-\ktm{ Another approach is to reduce the graphical complexity by plotting binned aggregations, namely frequencies, of the data as opposed to plotting every observation as an individual point.} This has the additional advantage of reducing the size of the stored data \ktm{necessary for the construction of the plot}, as only the bin centers and the frequency of points which correspond to that bin must be stored.
-\ktm{ Wickham argues for a ``bin-summarize-smooth'' procedure to to be applied to the visualization of big data and notes that simple summary functions, such as counts, scale very well with the size of data sources \citep{Wickham2013Bin}.
+\ktm{ Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point.} This has the additional advantage of reducing the size of the stored data \ktm{necessary for the construction of the plot}, as only the bin centers and the frequency of observations which correspond to that bin must be stored.
+\ktm{ Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale very well with the size of data \citep{Wickham2013Bin}.
 Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.
@@
 The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870} especially for data binned on a grid; whereas the binned scatterplot will not on a tesselated grid.
+The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870} especially for data binned on a grid; whereas the binned scatterplot will not on a tessellated grid.
 
@@
 % \end{figure}
-Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: 
-the joint density consists two distinct ridges following two lines with very different slopes.
+Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: the joint density consists two distinct ridges following two lines with very different slopes.
 The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game.
@@
 
-As with histograms, the width of bins (or the number of bins) is an important factor in the detail of the binned data and the resulting plot: if the bin width is too small in comparison to the amount of data available, there is little advantage to binning, but if the bin width is too large, interesting features of the joint distribution may be obscured by over-smoothing.
+As with histograms, the width of bins (or the number of bins) is an important factor in the detail of the binned data and the resulting plot: if the bin width is too small in comparison to the amount of data available, there is little advantage to binning, but if the bin width is too large, interesting features of the joint distribution may be obscured by over-plotting.
 
@@
 We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
-Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
+\ktm{ While hexagonal binning has been shown to have slightly better graphical properties \citep{scott2009multivariate};} rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
 We will however only consider binning in up to two dimensions, $X$ and $Y$.
@@
 Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density.
+Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.
 }
@@
 The $j^{th}$ bin intervals is $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
-Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernable groups in visualization using discrete color palettes.
+Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernible groups in visualization using discrete color palettes.
 }

### 2015-03-01 Paper.Rnw
move 2 dimension extension discussion up and add description of linearization.  Note that I have adjusted the notation such that j indexes X, k indexes Y and l is the linearized index s.t. l=j+J(k-1)
oops new extension to 2d not saved before last push

@@
 
-\subsection{Standard Rectangular Binning Algorithm}
+\subsection{Extensions to Rectangular Binning}
+
+The standard and random binning algorithms are easily extendable to higher dimension.
+For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
+In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$, where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively.
+\ktm{The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $l = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
+With this linearized index for all bins we now have a set of bin centers of the form $(x_l^\ast,y_l^\ast)$, with $l \in \{1,\dots,L\}$, where $L=J\dotK$.}
+
+The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_l,y^\ast_l)$ is defined as 
+%
+\begin{eqnarray}\label{standrecbin}
+b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
+\end{eqnarray}
+%
+where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
+The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_l,y^\ast_l)$ can similarly be defined as
+%
+\begin{eqnarray}\label{randrecbin}
+b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
+\end{eqnarray}
+%
+where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
+Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
+
+\subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
 \hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
@@
 
-\subsection{Random Binning Algorithm}
+\subsection{Random Binning Algorithm (old-to be removed)}
 
@@
 
-\subsection{Extensions to Rectangular Binning}
+\subsection{Extensions to Rectangular Binning (old-to be removed)}
 

### 2015-03-02 Paper.Rnw
Fixing compiling problems due to repeated function labels and math notation error

@@
 %
-\ktm{ \begin{eqnarray}\label{randbin1}
+\begin{eqnarray}\label{randbin1}
 b^r_X(x_i) = \left\{\begin{array}{ll} 
@@
 \end{eqnarray}  
-}
 %
@@
 \ktm{The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $l = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
-With this linearized index for all bins we now have a set of bin centers of the form $(x_l^\ast,y_l^\ast)$, with $l \in \{1,\dots,L\}$, where $L=J\dotK$.}
+With this linearized index for all bins we now have a set of bin centers of the form $(x_l^\ast,y_l^\ast)$, with $l \in \{1,\dots,L\}$, where $L=J\dot K$.}
 
@@
 %
-\begin{eqnarray}\label{standrecbin}
+\begin{eqnarray}\label{standrecbin1}
 b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
@@
 %
-\begin{eqnarray}\label{randrecbin}
+\begin{eqnarray}\label{randrecbin1}
 b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))

### 2015-03-02 Paper.Rnw
Rescale the rectangular binning plot in figure 2 to better display the difference between standard and random rectangular binning

@@
 \ktm{The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $l = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
-With this linearized index for all bins we now have a set of bin centers of the form $(x_l^\ast,y_l^\ast)$, with $l \in \{1,\dots,L\}$, where $L=J\dot K$.}
+With this linearized index for all bins we now have a set of bin centers of the form $(x_l^\ast,y_l^\ast)$, with $l \in \{1,\dots,L\}$, where $L=J\cdot K$.}
 
@@
 
-\subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
-\hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
-
-\hh{Define what $X$ is, what is $x_i$}
-
-\hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }
-
-\hh{I am still not sure how bin centers are defined in the end. XXX}
-
-\hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}
-
-\ktm{  We define the process that allocates observed values, $x_i$ for $i \in \{1, ..., n\}$, to equal sized bins in a deterministic manner as a standard binning algorithm.
-The bins are controlled by two parameters: the center of the first bin, $x^\ast_1$ and the binwidth, $\omega_X$.
-This leads to bin centers $ x_j^\ast =  x_1^\ast + (j-1)\omega_X \text{ for } j \in \{1, ..., N_X\} $,  where $N_X$ is the overall number of bins.
-}
-
-Mathematically, the standard binning algorithm corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
-the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as
-% \begin{center}
-% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
-% \vspace{.03}
-% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
-% \end{center}
-\begin{eqnarray}\label{standbin}
-b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
-\end{eqnarray}
-
-where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
-\ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $.
-After all observations have been allocated to bins, the bin count, $c_j$, is joined with the bin center to form the tuple $(x^\ast_1, c_j)$ }
-%\hh{something went wrong with the next sentence}
-%An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
-Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.
-
-\ktm{ A special case of standard binning divides the data range into $N_X$ equal sized bins.
-This leads to the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
-The first bin would then be centered at $x_{(1)} + \omega_X/2$.
-}
-
-%Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
-A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
-This strategy would still fit into the framework of assigning values to the closest bin center.
-
-\subsection{Random Binning Algorithm (old-to be removed)}
-
-An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
-A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
-In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
-The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
-More formally,
-%
-\begin{eqnarray}\label{randbin}
-b^r_X(x_i) = \left\{\begin{array}{ll} 
-  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
-  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
-  \end{array}\right.
-\end{eqnarray}
-
-% In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
-This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
-
-%To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
-%Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
-%1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
-
-%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
-
-The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
-``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
-With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
-Specifically, the expectation of the random binning algorithm is the same as for linear binning.
-%However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
-In the next section, we examine the loss of visual and numerical information due to binning.
-% linear binning in Gold book - random algorithm as a less computational method of linear binning
-
-\subsection{Extensions to Rectangular Binning (old-to be removed)}
-
-The standard and random binning algorithms are easily extendable to higher dimension.
-For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
-In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins \sout{in $\mathbb{R}^2$} \hh{XXX J is an integer number, definitely not a real-valued vector}.
-\ktm{The wording here was not clear.
-J does not exist in $\mathbb{R}^2$, but it is being used to index 2d bins that exist in that plane.} The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
-%
-\begin{eqnarray}\label{standrecbin}
-b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
-\end{eqnarray}
-%
-where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
-The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
-%
-\begin{eqnarray}\label{randrecbin}
-b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
-\end{eqnarray}
-%
-where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
-Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
-
-\hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation.
-$J$ is defined, as above as a multiple of $N_X$ and $N_Y$, this needs to be properly adjusted in the binning algorithm.
-It will be some linear relationship of the form $j_X\cdot N_X + j_Y$.
-}  \ktm{ I am having trouble trying to reconcile the indexing of bins in 2 dimensions.
-If we use an pair of indices it makes the extension of binning to 2 dimensions easy; but it makes the notation messier and loss functions more difficult to define.
-}
-
 <<binningDemo,echo=F,include=F,cache=T>>=
       library(mvtnorm)
-      set.seed(33)
-      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
+      set.seed(12)
+      d <- rmvnorm(50, c(10, 25), matrix(c(50, 15, 15, 50), nrow=2))
       d <- data.frame(d)
@@
 temp <- data.frame(   
-                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
-             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
-                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
-             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
-            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
+          x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
+       xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
+          y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
+       yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
+      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
       qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
@@
         annotate("segment", x=temp$x, y=temp$y, 
-                 xend=temp$xend, yend=temp$yend, group=temp$group) # + ylim(c(0, 50))
+                 xend=temp$xend, yend=temp$yend, group=temp$group) + 
+        ylim(c(10, 40)) + xlim(c(-10,30))
       ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
@@
         l="Numerical Center", a=1, size=2, bin=TRUE))
-      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
-        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
-                 fill="blue", alpha=.10) + 
-        geom_point() +
-        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
-                                             "Binned Data" = 15, 
-                                             "Numerical Center" = 17, 
-                                             "Visual Center" = 18)) +
-        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
-                                              "Binned Data" = "blue",
-                                              "Numerical Center" = "blue4", 
-                                              "Visual Center" = "blue4"))+
-        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
-                                            "Numerical Center" = 5, 
-                                            "Visual Center" = 5))+
-        theme_bw() +
-        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
-        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
-                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
-                 colour="blue4") +
-#        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
-#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
-#                  y=d$y[which(d$l=="Numerical Center")], 
-#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
-        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
-        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
+#       ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
+#         annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
+#                  fill="blue", alpha=.10) + 
+#         geom_point() +
+#         scale_shape_manual(name="", values=c("Data outside bin" = 16, 
+#                                              "Binned Data" = 15, 
+#                                              "Numerical Center" = 17, 
+#                                              "Visual Center" = 18)) +
+#         scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
+#                                               "Binned Data" = "blue",
+#                                               "Numerical Center" = "blue4", 
+#                                               "Visual Center" = "blue4"))+
+#         scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
+#                                             "Numerical Center" = 5, 
+#                                             "Visual Center" = 5))+
+#         theme_bw() +
+#         annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
+#         annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
+#                  y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
+#                  colour="blue4") +
+# #        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
+# #         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
+# #                  y=d$y[which(d$l=="Numerical Center")], 
+# #                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
+#         scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
+#         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
+#       ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
 @
@@
 temp <- data.frame(   
-          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
-       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
-          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
-       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
-      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
+          x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
+       xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
+          y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
+       yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
+      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
 
@@
   theme_bw() +
-  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
-         y=rep(seq(-5, 55, 10), times=6), 
-         shape='+', size=8, colour="gray90") + 
+#  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
+#        y=rep(seq(-5, 55, 10), times=6), 
+#         shape='+', size=8, colour="gray90") + 
   geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
@@
   scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
-  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
+  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(10, 40)) + xlim(c(-10,30))
 ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
@@
 \end{figure}
+
+\subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
+\hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
+
+\hh{Define what $X$ is, what is $x_i$}
+
+\hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }
+
+\hh{I am still not sure how bin centers are defined in the end. XXX}
+
+\hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}
+
+\ktm{  We define the process that allocates observed values, $x_i$ for $i \in \{1, ..., n\}$, to equal sized bins in a deterministic manner as a standard binning algorithm.
+The bins are controlled by two parameters: the center of the first bin, $x^\ast_1$ and the binwidth, $\omega_X$.
+This leads to bin centers $ x_j^\ast =  x_1^\ast + (j-1)\omega_X \text{ for } j \in \{1, ..., N_X\} $,  where $N_X$ is the overall number of bins.
+}
+
+Mathematically, the standard binning algorithm corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
+the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as
+% \begin{center}
+% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
+% \vspace{.03}
+% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
+% \end{center}
+\begin{eqnarray}\label{standbin}
+b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
+\end{eqnarray}
+
+where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
+\ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $.
+After all observations have been allocated to bins, the bin count, $c_j$, is joined with the bin center to form the tuple $(x^\ast_1, c_j)$ }
+%\hh{something went wrong with the next sentence}
+%An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
+Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.
+
+\ktm{ A special case of standard binning divides the data range into $N_X$ equal sized bins.
+This leads to the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
+The first bin would then be centered at $x_{(1)} + \omega_X/2$.
+}
+
+%Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
+A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
+This strategy would still fit into the framework of assigning values to the closest bin center.
+
+\subsection{Random Binning Algorithm (old-to be removed)}
+
+An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
+A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
+In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
+The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
+More formally,
+%
+\begin{eqnarray}\label{randbin}
+b^r_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
+  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
+  \end{array}\right.
+\end{eqnarray}
+
+% In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
+This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
+
+%To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
+%Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
+%1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
+
+%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
+
+The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
+``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
+With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
+Specifically, the expectation of the random binning algorithm is the same as for linear binning.
+%However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
+In the next section, we examine the loss of visual and numerical information due to binning.
+% linear binning in Gold book - random algorithm as a less computational method of linear binning
+
+\subsection{Extensions to Rectangular Binning (old-to be removed)}
+
+The standard and random binning algorithms are easily extendable to higher dimension.
+For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
+In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins \sout{in $\mathbb{R}^2$} \hh{XXX J is an integer number, definitely not a real-valued vector}.
+\ktm{The wording here was not clear.
+J does not exist in $\mathbb{R}^2$, but it is being used to index 2d bins that exist in that plane.} The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
+%
+\begin{eqnarray}\label{standrecbin}
+b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
+\end{eqnarray}
+%
+where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
+The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
+%
+\begin{eqnarray}\label{randrecbin}
+b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
+\end{eqnarray}
+%
+where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
+Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
+
+\hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation.
+$J$ is defined, as above as a multiple of $N_X$ and $N_Y$, this needs to be properly adjusted in the binning algorithm.
+It will be some linear relationship of the form $j_X\cdot N_X + j_Y$.
+}  \ktm{ I am having trouble trying to reconcile the indexing of bins in 2 dimensions.
+If we use an pair of indices it makes the extension of binning to 2 dimensions easy; but it makes the notation messier and loss functions more difficult to define.
+}
 

### 2015-03-02 Paper.Rnw
More apt subsection title
Minor caption and label edits

@@
 
-\subsection{Extensions to Rectangular Binning}
+\subsection{Extension to Two Dimensional Binning}
 
@@
   \caption{Visualization of how data pairs are assigned to bins centered at (5, 25) and (15,25) under the two rectangular binning algorithms.
-Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to bins from outside the defined bin regions.}
+Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to any of the surrounding bin centers.}
   \label{fig:BinningTutorial}
@@
 
-\subsection{Quantile Binning Algorithm}
+\subsection{Quantile Binning Algorithm (old - to be removed)}
 

### 2015-03-02 Paper.Rnw
Changes to loss source description

@@
 
-Loss of information occurs on multiple levels during the binning and rendering process. We distinguish three sources:
+\ktm{Loss of information occurs during the binning and rendering process.
+For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.
+We distinguish two sources of loss in the construction of a binned scatterplot:
 \begin{itemize}
-
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) in the data set are reduced to a single points at the bin centers ($x_l^\ast$,$y_l^\ast$).
+By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
+
+\item {\it Frequency Loss}, $L^\text{F}$, occurs when frequency information is not rendered accurately.
+If a discrete color scale is used to display bin counts, then entire ranges of counts will be mapped to the same color; thus lossing precision in the display of frequencies.
+\end{itemize}
+
+Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points.
+}
+
+\ktm{The following \textit{numerical, visual and freq} loss descriptions are from the old list and will be removed: }
+\begin{itemize}
 \item {\it Numerical Loss}, $L^\text{N}$, occurs when points in the data set are reduced to a single points at the bin means.

### 2015-03-03 Paper.Rnw
Spatial loss subsection start

@@
 
+\subsection{Spatial Loss}
+
+\ktm{When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.
+This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.
+the bin centers).
+Note that other distance metrics could be used, but the euclidean distance has a desirable interpretability in $\mathbb{R}^2$.
+The spatial loss for the $i^{th}$ observation, $L_i^S$ and the total spatial loss, $L^S$ are defined as}
+%
+\begin{eqnarray}\label{indspatloss}
+L_i^S = \sqrt{ (x_i-b_X(x_i))^2 + (y_i-b_Y(y_i))^2 }
+\end{eqnarray}
+%
+\begin{eqnarray}\label{totspatloss}
+L^S = \sum_{i=1}^{n} L_i^S
+\end{eqnarray}  
+%
+
+\subsection{Old Loss Definitions (to be removed)}
 \ktm{The following \textit{numerical, visual and freq} loss descriptions are from the old list and will be removed: }

### 2015-03-03 Paper.Rnw
start of frequency loss subsection

@@
 %
+\ktm{Visual display of spatial loss still needed.}
+
+\subsection{Frequency Loss}
+
+\ktm{If the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
+While counts can certainly be modeled as continuous on the HSV color space (reference needed) discrete color breaks may be prefered for a few reasons.
+%The average person is able to  On the continuous scale, a viewer will likely not be able to exactly match a tile shading to the corresponding count on the color key in the legend of the plot without an aid such as interactive mouseover effects.
+}
 

### 2015-03-04 Paper.Rnw
Edits for Section 3 discussed in meeting 3/4/15.  Remove red, general binning function, l's to \ell's, comment out old.

@@
 We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
-\ktm{ While hexagonal binning has been shown to have slightly better graphical properties \citep{scott2009multivariate};} rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
+While hexagonal binning has been shown to have slightly better graphical properties \citep{scott2009multivariate}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
 We will however only consider binning in up to two dimensions, $X$ and $Y$.
@@
 
-\ktm{  For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, the binning requires a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
+\ktm{  In general, in the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a function that maps observations to a bin centers.
+}
+
+\ktm{ Rectangular binning accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
 Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
@@
 
-\ktm{  This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as }
+This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
 %
@@
 %
-\ktm{Note that this binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
+\textit{Standard rectangular binning} is a special cases of this general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified.
+Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density.
+Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.
+
+Note that this binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
 These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}.
@@
 For example, the \texttt{ggplot2} packages creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.
-}
-
-\ktm{ \textit{Standard rectangular binning} is a special cases of this general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified.
-Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density.
-Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.
-}
-
-\ktm{  \textit{Quantile binning} is another option that divides the range of the observations into bins each containing equal number of points.
+
+\textit{Quantile binning} is another option that divides the range of the observations into bins each containing equal number of points.
 The $j^{th}$ bin intervals is $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
 Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernible groups in visualization using discrete color palettes.
-}
-
-\ktm{  The bin boundaries and centers for each type of rectangular binning algorithm discussed above can be found in Table~\ref{tab:rectbinning}.  }
+
+The bin boundaries and centers for each type of rectangular binning algorithm discussed above can be found in Table~\ref{tab:rectbinning}.
 
@@
 \end{table}
-
-\ktm{ An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
+%
+An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
 The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
-More formally, }
+More formally,
 %
@@
 %
-\ktm{  for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.
+for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.
 This probability structure is inversely proportional to the distance from each bin center.
@@
 This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.
-}
-
-\ktm{  The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the bin center.
+
+The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the bin center.
 ``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning in which adjacent bins are assigned a weight depending on the distance from the point to that bin, where all weights sum to one.
@@
 Specifically, the expectation of the random binning algorithm is the same as for linear binning.
-}
 
@@
 In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$, where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively.
-\ktm{The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $l = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
-With this linearized index for all bins we now have a set of bin centers of the form $(x_l^\ast,y_l^\ast)$, with $l \in \{1,\dots,L\}$, where $L=J\cdot K$.}
-
-The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_l,y^\ast_l)$ is defined as 
+The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
+With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,L\}$, where $L=J\cdot K$.
+
+The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ is defined as 
 %
@@
 where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
-The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_l,y^\ast_l)$ can similarly be defined as
+The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ can similarly be defined as
 %
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \centering
@@
 
-\subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
-\hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
-
-\hh{Define what $X$ is, what is $x_i$}
-
-\hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }
-
-\hh{I am still not sure how bin centers are defined in the end. XXX}
-
-\hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}
-
-\ktm{  We define the process that allocates observed values, $x_i$ for $i \in \{1, ..., n\}$, to equal sized bins in a deterministic manner as a standard binning algorithm.
+% \subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
+% \hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
+% 
+% \hh{Define what $X$ is, what is $x_i$}
+% 
+% \hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }
+% 
+% \hh{I am still not sure how bin centers are defined in the end. XXX}
+% 
+% \hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}
+% 
+% \ktm{  We define the process that allocates observed values, $x_i$ for $i \in \{1, ..., n\}$, to equal sized bins in a deterministic manner as a standard binning algorithm.
 The bins are controlled by two parameters: the center of the first bin, $x^\ast_1$ and the binwidth, $\omega_X$.
@@
 }
-
-Mathematically, the standard binning algorithm corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
+% 
+% 
+% Mathematically, the standard binning algorithm corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
 the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as
-% \begin{center}
-% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
-% \vspace{.03}
-% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
-% \end{center}
-\begin{eqnarray}\label{standbin}
-b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
-\end{eqnarray}
-
-where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
-\ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $.
+% % \begin{center}
+% % $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
+% % \vspace{.03}
+% % $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
+% % \end{center}
+% \begin{eqnarray}\label{standbin}
+% b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
+% \end{eqnarray}
+% 
+% where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
+% \ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $.
 After all observations have been allocated to bins, the bin count, $c_j$, is joined with the bin center to form the tuple $(x^\ast_1, c_j)$ }
-%\hh{something went wrong with the next sentence}
-%An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
+% %\hh{something went wrong with the next sentence}
+% %An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
 Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.
-
-\ktm{ A special case of standard binning divides the data range into $N_X$ equal sized bins.
+% 
+% \ktm{ A special case of standard binning divides the data range into $N_X$ equal sized bins.
 This leads to the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
@@
 }
-
-%Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
+% 
+% 
+% 
+% %Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
 A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
 This strategy would still fit into the framework of assigning values to the closest bin center.
-
-\subsection{Random Binning Algorithm (old-to be removed)}
-
-An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
+% 
+% \subsection{Random Binning Algorithm (old-to be removed)}
+% 
+% An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
 A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
@@
 More formally,
-%
-\begin{eqnarray}\label{randbin}
-b^r_X(x_i) = \left\{\begin{array}{ll} 
-  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
-  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
-  \end{array}\right.
-\end{eqnarray}
-
-% In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
-This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm.
-
-%To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
-%Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
-%1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
-
-%Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
-
-The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
+% %
+% \begin{eqnarray}\label{randbin}
+% b^r_X(x_i) = \left\{\begin{array}{ll} 
+%   x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
+%   x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
+%   \end{array}\right.
+% \end{eqnarray}
+% 
+% % In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
+% This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm. 
+% 
+% %To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
+% %Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
+% %1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
+% 
+% %Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
+% 
+% The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
 ``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
@@
 Specifically, the expectation of the random binning algorithm is the same as for linear binning.
-%However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
-In the next section, we examine the loss of visual and numerical information due to binning.
-% linear binning in Gold book - random algorithm as a less computational method of linear binning
-
-\subsection{Extensions to Rectangular Binning (old-to be removed)}
-
-The standard and random binning algorithms are easily extendable to higher dimension.
+% %However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
+% In the next section, we examine the loss of visual and numerical information due to binning.
+% % linear binning in Gold book - random algorithm as a less computational method of linear binning
+% 
+% \subsection{Extensions to Rectangular Binning (old-to be removed)}
+% 
+% The standard and random binning algorithms are easily extendable to higher dimension.
 For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
@@
 J does not exist in $\mathbb{R}^2$, but it is being used to index 2d bins that exist in that plane.} The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
-%
-\begin{eqnarray}\label{standrecbin}
-b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
-\end{eqnarray}
-%
-where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
+% %
+% \begin{eqnarray}\label{standrecbin}
+% b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
+% \end{eqnarray}
+% %
+% where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
 The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
-%
-\begin{eqnarray}\label{randrecbin}
-b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
-\end{eqnarray}
-%
-where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
+% %
+% \begin{eqnarray}\label{randrecbin}
+% b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
+% \end{eqnarray}
+% %
+% where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
 Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
-
-\hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation.
+% 
+% \hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation.
 $J$ is defined, as above as a multiple of $N_X$ and $N_Y$, this needs to be properly adjusted in the binning algorithm.
@@
 }
-
-\subsection{Quantile Binning Algorithm (old - to be removed)}
-
-We could also consider a empirical quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
+% 
+% 
+% 
+% \subsection{Quantile Binning Algorithm (old - to be removed)}
+% 
+% We could also consider a empirical quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
 This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
 This binning will not be considered for binning the data pairs directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.
-
-Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
+% 
+% Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
 Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers.
@@
 The quantile binning function $b^q_C(\cdot) : (c_j) \rightarrow c^\ast_k$ is defined as
-%
-\begin{eqnarray}\label{quantbin}
-b^q_C(c_j) = (Q_{k-1/K} + Q_{k/K})/2 \text{ where } Q_{k-1/K} < c_j \le Q_{k/K}.
-\end{eqnarray}
-%
-The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in strictly the univariate setting for this manuscript.
+% %
+% \begin{eqnarray}\label{quantbin}
+% b^q_C(c_j) = (Q_{k-1/K} + Q_{k/K})/2 \text{ where } Q_{k-1/K} < c_j \le Q_{k/K}.
+% \end{eqnarray}
+% %
+% The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in strictly the univariate setting for this manuscript.
 

### 2015-03-04 Paper.Rnw
Spatial loss display plot addition to section 4.1

@@
 library(hexbin)
+library(MASS)
 
@@
         theme_bw() +
-        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
-               y=rep(seq(-5, 55, 10), times=6), 
-               shape='+', size=8, colour="gray90") + 
+      #  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
+      #         y=rep(seq(-5, 55, 10), times=6), 
+      #         shape='+', size=8, colour="gray90") + 
         geom_point(data=d[which(!is.na(d$xbin)),], 
@@
 \item {\it Frequency Loss}, $L^\text{F}$, occurs when frequency information is not rendered accurately.
-If a discrete color scale is used to display bin counts, then entire ranges of counts will be mapped to the same color; thus lossing precision in the display of frequencies.
+If a discrete color scale is used to display bin counts, then entire ranges of counts will be mapped to the same color; thus losing precision in the display of frequencies.
 \end{itemize}
@@
 %
-\ktm{Visual display of spatial loss still needed.}
+\ktm{Figure~\ref{fig:LossDisplay} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
+Observations $(x_i,y_i)$ and bin centers $(x_\ell^\ast,y_\ell^\ast)$ are displayed as circular and triangular points, respectively.
+The length of line segments connecting these represent, $L^S_i$ the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
+}
+
+<<SpatialLossDemo,echo=F,include=F,cache=T>>=
+#Try Multivariate Normal and Uniform for this plot
+
+#MVN plot
+n=15
+set.seed(2)
+Sigma <- matrix(c(16,12,12,14),2,2)
+dat <- mvrnorm(n = n, rep(0, 2), Sigma)
+x <- dat[,1]
+y <- dat[,2]
+xcent <- rep(-5,n) ; xcent[x>0] <- 5
+ycent <- rep(-5,n) ; ycent[y>0] <- 5
+temp <- data.frame(   
+  x=c( -10,   0,  10, -10, -10, -10), 
+  xend=c( -10,   0,  10,  10, 10, 10), 
+  y=c( -10, -10 , -10 , -10, 0 , 10), 
+  yend=c(10, 10 , 10, -10, 0, 10 ))
+qplot(xcent, ycent, geom="point", size=I(4), shape=I(2)) +
+  geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
+  geom_point(aes(x=x,y=y), size=I(3))+ 
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+  theme_bw() + xlab("") + ylab("")
+      ggsave("./figure/SpatialLossDisplayMVN.pdf", width=5, height=5, units="in")
+
+#uniform plot
+set.seed(2)
+x <- runif(n, -10,10)
+y <- runif(n, -10,10)
+xcent <- rep(-5,n) ; xcent[x>0] <- 5
+ycent <- rep(-5,n) ; ycent[y>0] <- 5
+temp <- data.frame(   
+  x=c( -10,   0,  10, -10, -10, -10), 
+  xend=c( -10,   0,  10,  10, 10, 10), 
+  y=c( -10, -10 , -10 , -10, 0 , 10), 
+  yend=c(10, 10 , 10, -10, 0, 10 ))
+qplot(xcent, ycent, geom="point", size=I(4), shape=I(2)) +
+  geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
+  geom_point(aes(x=x,y=y), size=I(3))+ 
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+  theme_bw() + xlab("") + ylab("")
+      ggsave("./figure/SpatialLossDisplayUnif.pdf", width=5, height=5, units="in")
+@
+
+\begin{figure}[h]
+  \centering
+  \subfloat[Spatial Loss MVN Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayMVN.pdf}}
+  \subfloat[Spatial Loss Uniform Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayUnif.pdf}}
+  \caption{Visualization of spatial loss. \ktm{Redundant plots using MVN and Uniform Data. Which makes a better plot?}}
+  \label{fig:LossDisplay}
+\end{figure}
+
+%------------------------------------------------
 
@@
 \ktm{If the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
-While counts can certainly be modeled as continuous on the HSV color space (reference needed) discrete color breaks may be prefered for a few reasons.
+While counts can certainly be modeled as continuous on the HSV color space (reference needed) discrete color breaks may be preferred for a few reasons.
 %The average person is able to  On the continuous scale, a viewer will likely not be able to exactly match a tile shading to the corresponding count on the color key in the legend of the plot without an aid such as interactive mouseover effects.

### 2015-03-05 Paper.Rnw
Add section labels and start work on freq loss section

@@
 \section{Introduction}
+\label{Intro}
 
@@
 \section{Scatterplots for Large Data Sets}
+\label{Scatter}
+
 In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions.
@@
 \section{Binning Algorithms}
+\label{GenBinning}
 
@@
 \subsection{Extension to Two Dimensional Binning}
+\label{TwoDimBin}
 
@@
 \section{Loss due to Binning}
+\label{LossIntro}
+
 While binning data has some notable advantages, it does result in loss of some information.
@@
 \subsection{Spatial Loss}
+\label{SpatialLoss}
 
@@
 \subsection{Frequency Loss}
-
-\ktm{If the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
-While counts can certainly be modeled as continuous on the HSV color space (reference needed) discrete color breaks may be preferred for a few reasons.
-%The average person is able to  On the continuous scale, a viewer will likely not be able to exactly match a tile shading to the corresponding count on the color key in the legend of the plot without an aid such as interactive mouseover effects.
+\label{FreqLoss}
+
+\ktm{ If the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
+This is effectively a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins  using any of the previously mentioned univariate binning algorithms in Section~\ref{GenBinning}.
+}
+
+\ktm{ While counts can certainly be modeled as continuous on the HSV color space (reference needed) discrete color breaks may be preferred for a few reasons.
+The average person is able to distinguish  On the continuous scale, a viewer will likely not be able to exactly match a tile shading to the corresponding count on the color key in the legend of the plot without an aid such as interactive mouseover effects.
 }

### 2015-03-05 Paper.Rnw
Work on frequency loss section

@@
 Note that other distance metrics could be used, but the euclidean distance has a desirable interpretability in $\mathbb{R}^2$.
-The spatial loss for the $i^{th}$ observation, $L_i^S$ and the total spatial loss, $L^S$ are defined as}
-%
-\begin{eqnarray}\label{indspatloss}
-L_i^S = \sqrt{ (x_i-b_X(x_i))^2 + (y_i-b_Y(y_i))^2 }
+The \textit{total spatial loss}, $L^S$, is defined as }
+\begin{eqnarray}\label{totspatloss}
+L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ (x_i-b_X(x_i))^2 + (y_i-b_Y(y_i))^2 }
 \end{eqnarray} 
 %
-\begin{eqnarray}\label{totspatloss}
-L^S = \sum_{i=1}^{n} L_i^S
-\end{eqnarray}  
+\ktm{where $L_i^S$ is the $i^{th}$ observation. }
 %
@@
 Observations $(x_i,y_i)$ and bin centers $(x_\ell^\ast,y_\ell^\ast)$ are displayed as circular and triangular points, respectively.
-The length of line segments connecting these represent, $L^S_i$ the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
+The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
 }
@@
 \ktm{ If the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
-This is effectively a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins  using any of the previously mentioned univariate binning algorithms in Section~\ref{GenBinning}.
-}
-
-\ktm{ While counts can certainly be modeled as continuous on the HSV color space (reference needed) discrete color breaks may be preferred for a few reasons.
-The average person is able to distinguish  On the continuous scale, a viewer will likely not be able to exactly match a tile shading to the corresponding count on the color key in the legend of the plot without an aid such as interactive mouseover effects.
-}
+This is effectively a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins  using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
+}
+
+\ktm{ Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus have no frequency loss.
+While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
+Color perception of shade is extremely context sensative (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
+}
+
+\ktm{The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned freqencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
+The \textit{total frequency loss}, $L^F$ are defined as}
+%
+\begin{eqnarray}\label{totfreqloss}
+L^F = \sum_{i=1}^{n} L_i^F = \sum_{i=1}^{n} (c_\ell-b_c(c_\ell))^2
+\end{eqnarray}  
+%
+\ktm{where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.}
 

### 2015-03-05 Paper.Rnw
continued work on frequency binning section.  Still incomplete, needs visual display.

@@
 While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
-Color perception of shade is extremely context sensative (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
-}
-
-\ktm{The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned freqencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
-The \textit{total frequency loss}, $L^F$ are defined as}
+Color perception of shade is extremely context sensitive (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
+}
+
+\ktm{The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
+The \textit{total frequency loss}, $L^F$, is defined as}
 %
 \begin{eqnarray}\label{totfreqloss}
-L^F = \sum_{i=1}^{n} L_i^F = \sum_{i=1}^{n} (c_\ell-b_c(c_\ell))^2
+L^F = \sum_{\ell=1}^{L} L_\ell^F = \sum_{\ell=1}^{L} (c_\ell-b_c(c_\ell))^2
 \end{eqnarray}  
@@
 \ktm{where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.}
+
+Frequency data consists of counts, which most commonly exhibit skew densities, i.e.
+there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
+A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution.
+This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}.
+Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph.
+The \textit{log frequency loss}, $L^{\log F}$, is defined as
+%
+\begin{eqnarray}\label{logfrequency}
+L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
+\end{eqnarray}
+%
+Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.
+
+\ktm{Plot needed here to demonstrate shading scales under binning.}
+
+This numerical assessment of frequency loss does not account for limitations in human perceptual ability.
+Research suggests that under optimal conditions, we can effectively compare about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
+As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
 
@@
 
-Frequency data consists of counts, which most commonly exhibit skew densities, i.e. there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
-
+Frequency data consists of counts, which most commonly exhibit skew densities, i.e.
+there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
 A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution.
@@
 This suggests that frequency loss be calculated as
+%
 \begin{eqnarray}\label{eqn:logfrequency}
@@
 \end{eqnarray}
-
+%
 Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

### 2015-03-09 Paper.Rnw
proofing

@@
 In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions.
-For large data, scatterplots suffer from over-plotting -- i.e.
-\ktm{a large number of observations drawn in such close proximity that their points overlap}, thereby masking relevant structure.
+For large data, scatterplots suffer from over-plotting, %-- i.e.\ a number of observations drawn in such close proximity that their points overlap,
+thereby masking relevant structure.
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
@@
 
-\ktm{ Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle.
-A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable}.
-Tukey \citep{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting.
+Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle.
+A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable.
+\citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting.
 Open circles make points visible that are close together.
@@
 
-\ktm{ Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales.
-Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations but use distortion of the point locations by having points repel one another to avoid overlapping points.
+Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales.
+Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations (including ties) but use distortion of the point locations by having points repel one another to avoid overlapping points.
 An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
 Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
-This variable binning fragments the continuity of the axes into many segments on different scales and also does not deal with points that are exactly overlapped due to the resolution of data collection.
-Generalized and variable-binned scatterplots make fine data structure more visible and allows color to be reserved for a third variable instead of frequency; however, the distortion of the point locations or the axes wrap the visual display of the association between the two primary variables.}
-
-\ktm{ Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point.} This has the additional advantage of reducing the size of the stored data \ktm{necessary for the construction of the plot}, as only the bin centers and the frequency of observations which correspond to that bin must be stored.
-\ktm{ Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale very well with the size of data \citep{Wickham2013Bin}.
+This variable binning fragments the continuity of the axes into  segments on different scales and also does not deal with ties.
+Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.
+
+Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point.
+This has the additional advantage of reducing the size of the stored data necessary for the construction of the plot, as only the bin centers and the frequency of observations which correspond to that bin must be stored.
+Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale  well with the size of data \citep{Wickham2013Bin}.
 Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.
-} \hh{also add \citep{gold} for binning reference - it's called `weighting' there.}
+\hh{also add \citep{gold} for binning reference - it's called `weighting' there.}
 
@@
 Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
-\ktm{Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.
-} A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
-\ktm{ A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, similar to a two-dimensional histogram viewed from above.
+Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.
+A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
+A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.
 A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.
 Bubble plots were first used by William Playfair \citep{playfair, playfair2}.
-}
 
@@
 
-\ktm{  In general, in the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a function that maps observations to a bin centers.
-}
-
-\ktm{ Rectangular binning accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
+In general, for the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.
+
+{\textit Rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
 Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
-Values that fall at exactly $\beta_0$ are grouped into the first bin to close the leftmost bound.
+Values $x_i$ exactly equal to the lowest bin boundar $\beta_0$ are grouped into the first bin to close the leftmost bound.
 Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs.
-}
-
-This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
-%
-\begin{eqnarray}\label{rectbin}
-b_X(x_i) = \left\{\begin{array}{ll} 
-  x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
-  x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
-  \end{array}\right.
-\end{eqnarray}  
-%
-\textit{Standard rectangular binning} is a special cases of this general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified.
-Standard rectangular binning is necessarily used in the construction of all histograms and is ideal for visualizing density because the consistency of binwidth makes the display of frequency proportional to density.
+
+% This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
+% %
+% \begin{eqnarray}\label{rectbin}
+% b_X(x_i) = \left\{\begin{array}{ll} 
+%   x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
+%   x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
+%   \end{array}\right.
+% \end{eqnarray}  
+% %
+
+\textit{Standard rectangular binning} is a special cases of general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified.
+Standard rectangular binning is necessarily used in the construction of all histograms;  the consistent binwidth makes the display of frequency proportional to density.
 Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.
 
-Note that this binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
+Note that this \hh{XXX which binning procedure are you referring to by 'this'?} binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
 These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}.
 These specifications were selected for this paper, but these choices are by no means consider universal for binning.
-For example, the \texttt{ggplot2} packages creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.
+For example, the \texttt{ggplot2} package creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.
 
@@
 
-As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \ref{eqn:partition}). We can make use of this and
+As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
+). We can make use of this and
  recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations.

### 2015-03-09 Paper.Rnw
Add figure to frequency loss section

@@
 \ktm{Plot needed here to demonstrate shading scales under binning.}
+
+<<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold'>>= 
+library(ggplot2)
+library(gridExtra)
+J = 0.01
+J2 = 200
+bound = 1000
+id <- seq(0+J/2,bound-J/2,by=J )
+fill <- seq(0+J/2,bound-J/2,by=J )
+id2 <- seq(0+J2/2,bound-J2/2, by=J2)
+fill2 <- seq(J2/2,bound-(J2/2), by=J2)
+p1 <- qplot(id, 2, geom="tile", fill=-fill) +
+  theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
+        legend.position = "none",
+        axis.ticks=element_blank(),
+        axis.text.y=element_blank(),
+        panel.background=element_blank()
+        ) + 
+  ylab("Continuous") + xlab("")
+p2 <- qplot(id2, 2, geom="tile", fill=-fill2) + 
+  theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
+        legend.position = "none",
+        axis.ticks=element_blank(),
+        axis.text.y=element_blank(),
+        panel.background=element_blank()) + 
+  ylab("Discrete") + xlab("Count")
+grid.arrange(p1,p2,nrow=2)
+#ggsave("./figure/FreqBin.pdf",width=5, height=1.5)
+@
 

### 2015-03-10 Paper.Rnw
mini changes

@@
 
-\ktm{Loss of information occurs during the binning and rendering process.
+Loss of information occurs during the binning and rendering process.
 For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.
@@
 \begin{itemize}
-\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) in the data set are reduced to a single points at the bin centers ($x_l^\ast$,$y_l^\ast$).
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) \hh{XXX define $i$, $l$} in the data set are reduced to a single points at the bin centers ($x_l^\ast$,$y_l^\ast$).
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
@@
 Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points.
-}
 
@@
 \begin{eqnarray}\label{totspatloss}
-L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ (x_i-b_X(x_i))^2 + (y_i-b_Y(y_i))^2 }
+L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ \left(x_i-b_X(x_i)\right)^2 + \left(y_i-b_Y(y_i)\right)^2 }
 \end{eqnarray} 
 %
-\ktm{where $L_i^S$ is the $i^{th}$ observation. }
+\ktm{where $L_i^S$ is the \hh{loss in the } $i^{th}$ observation. }
 %
 \ktm{Figure~\ref{fig:LossDisplay} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
-Observations $(x_i,y_i)$ and bin centers $(x_\ell^\ast,y_\ell^\ast)$ are displayed as circular and triangular points, respectively.
+Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as circular and triangular points, respectively.
 The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.

### 2015-03-11 Paper.Rnw
comments, color context

@@
 
-\item {\it Frequency Loss}, $L^\text{F}$, occurs when frequency information is not rendered accurately.
-If a discrete color scale is used to display bin counts, then entire ranges of counts will be mapped to the same color; thus losing precision in the display of frequencies.
+\item {\it Frequency Loss}, $L^\text{F}$, occurs when frequency information is not rendered, such as in standard scatterplots,  or not rendered accurately, e.g.\ when a discrete color scale is used to display bin counts, the entire ranges of counts is mapped to the same color; thus losing precision in the display of frequencies.
 \end{itemize}
@@
 
-\ktm{When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.
-This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.
-the bin centers).
+When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.
+This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers).
 Note that other distance metrics could be used, but the euclidean distance has a desirable interpretability in $\mathbb{R}^2$.
-The \textit{total spatial loss}, $L^S$, is defined as }
+The \textit{total spatial loss}, $L^S$, is defined as
 \begin{eqnarray}\label{totspatloss}
@@
 %
-\ktm{where $L_i^S$ is the \hh{loss in the } $i^{th}$ observation. }
-%
-\ktm{Figure~\ref{fig:LossDisplay} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
+where $L_i^S$ is the loss in the  $i^{th}$ observation. 
+%
+Figure~\ref{fig:LossDisplay} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
 Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as circular and triangular points, respectively.
 The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
-}
 
@@
   \subfloat[Spatial Loss Uniform Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayUnif.pdf}}
-  \caption{Visualization of spatial loss. \ktm{Redundant plots using MVN and Uniform Data. Which makes a better plot?}}
+  \caption{Visualization of spatial loss.
+The distance between points and bin centers adds up to the spatial loss due to binning.
+}%\ktm{Redundant plots using MVN and Uniform Data.
+Which makes a better plot?}}
   \label{fig:LossDisplay}
@@
 \label{FreqLoss}
+
+\hh{XXX I think we need a bit of discussion here, why it is not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
+Figure~\ref{fig:context} shows an example of the context sensitivity of colors.
+Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each of the points clouds.
+}
+
+\begin{figure}[ht]
+\centering
+<<context, echo=false, fig.width=4, fig.height=4, out.width='.5\\textwidth'>>=
+col <- c(rnorm(200, mean=0, sd=10), rnorm(10, mean=50), rnorm(200, mean=100, sd=10), rnorm(10, mean=50))
+x <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=1, sd=0.075))
+y <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=1, sd=0.075))
+
+require(ggplot2)
+qplot(x, y, colour=col, size=I(5)) + theme_bw()
+@
+\caption{\label{fig:context}Assess the colors in this scatterplot. Are the darkest spots in the cloud of lighter points as dark as the lightest spots in the cloud of darker points? Darker? Lighter? }
+\end{figure}
 

### 2015-03-11 Paper.Rnw
aspect ratio

@@
 \centering
-<<context, echo=false, fig.width=4, fig.height=4, out.width='.5\\textwidth'>>=
+<<context, echo=F, fig.width=6, fig.height=4.5, out.width='.5\\textwidth'>>=
 col <- c(rnorm(200, mean=0, sd=10), rnorm(10, mean=50), rnorm(200, mean=100, sd=10), rnorm(10, mean=50))
@@
 require(ggplot2)
-qplot(x, y, colour=col, size=I(5)) + theme_bw()
+qplot(x, y, colour=col, size=I(5)) + theme_bw() + theme(aspect.ratio=1)
 @
@@
 
-Frequency data consists of counts, which most commonly exhibit skew densities, i.e.
-there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
+Frequency data consists of counts, which most commonly exhibit skew densities, i.e.\ there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
 A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution.

### 2015-03-11 Paper.Rnw
Frequency loss section re-structured

@@
 
-\hh{XXX I think we need a bit of discussion here, why it is not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
+Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus theoretically have no frequency loss.
+While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
+Color perception of shade is extremely context sensitive (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
+
+ Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
+
+\hh{It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
 Figure~\ref{fig:context} shows an example of the context sensitivity of colors.
@@
 
-\ktm{ If the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
-This is effectively a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins  using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
-}
-
-\ktm{ Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus have no frequency loss.
-While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
-Color perception of shade is extremely context sensitive (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
-}
-
-\ktm{The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
-The \textit{total frequency loss}, $L^F$, is defined as}
+As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
+
+ Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
+We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins  using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
+
+While this numerical assessment of frequency loss does not exactly account for limitations in human perceptual ability, it does provide a more realistic model for the loss in perception that does occur.
+
+The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
+The \textit{total frequency loss}, $L^F$, is defined as
 %
@@
 %
-\ktm{where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.}
+where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.
 
@@
 @
-
-This numerical assessment of frequency loss does not account for limitations in human perceptual ability.
-Research suggests that under optimal conditions, we can effectively compare about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
-As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
 

### 2015-03-13 BinningLossFunctions.R
code for spatial and frequency binning/loss functions

@@ -1,0 +1,194 @@
+### Preliminaries 
+
+library(ggplot2)
+library(mvtnorm)
+library(plyr)
+library(Hmisc)
+
+#----------------------------------------------------
+### Binning Functions
+#----------------------------------------------------
+
+## Standard 1d Rectangular Binning
+StandRectBin1d <- function(xs, origin, width){
+  binxs <- origin + width*( ceiling((xs-origin)/width) - 0.5 )
+  binxs[xs == origin] <- origin + width/2
+  return(binxs)
+}
+
+## Random 1d Rectangular Binning
+# Establish bounding bin centers for each observation
+# Then use Unif(0,1) draws compared to assignment probs to allocate
+# Reassignment for values below first center (bx1) or
+#     above highest center (bxJ)
+RandRectBin1d <- function(xs, origin, width){
+  bx1 <- origin + width/2
+  bxJ <- width*(floor((max(xs)-bx1)/width) + .5)
+  lbs <- bx1 + width*floor((xs-bx1)/width)
+  ubs <- bx1 + width*ceiling((xs-bx1)/width)
+  # initially assign all values to upper bound
+  binxs <- ubs
+  # then use runif to reassign based on distance from 
+  plower <- (ubs - xs)/width
+  lowerindex <- (plower > runif(length(xs), 0, 1))
+  binxs[lowerindex] <- lbs[lowerindex]
+  binxs[xs < bx1] <- bx1
+  binxs[xs > bxJ] <- bxJ
+  return(binxs)
+}
+
+
+## Quantile 1d Binning
+# used for binning the counts values by quantile
+# define vector of counts and number of bin
+QuantBin1d <- function(cs, nbin){
+  bincenters <- quantile(cs, seq(1/(2*nbin) , 1- 1/(2*nbin), by=1/nbin))
+  binbounds <- quantile(cs, seq(0 , 1, by=1/nbin))
+  bincs <- rep(bincenters[1],length(cs))
+   for (i in 2:length(bincenters)){
+     bincs[(binbounds[i] < cs & cs <= binbounds[i+1])] <- bincenters[i]
+   }   
+  return(bincs)
+}
+
+
+## 2d Rectangular Binning (for either Standard or Random)
+RectBin2d <- function(xs,ys, originx, originy, widthx, widthy, type="standard"){
+  if(type=="standard"){
+  tempdat <- data.frame(xs = xs, ys=ys,
[…]

### 2015-03-16 Paper.Rnw
minor edits

@@
 Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
-Values $x_i$ exactly equal to the lowest bin boundar $\beta_0$ are grouped into the first bin to close the leftmost bound.
+Values $x_i$ exactly equal to the lowest bin boundary $\beta_0$ are grouped into the first bin to close the leftmost bound.
 Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs.
@@
 
-Note that this \hh{XXX which binning procedure are you referring to by 'this'?} binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
+Note that this standard rectangular binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
 These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}.
@@
 \begin{itemize}
-\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) \hh{XXX define $i$, $l$} in the data set are reduced to a single points at the bin centers ($x_l^\ast$,$y_l^\ast$).
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) \hh{XXX define $i$, $l$} in the data set are reduced to a single points at the bin centers ($x_\ell^\ast$,$y_\ell^\ast$).
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.

### 2015-03-18 BinningLossFunctions.R
Tricky post processing for 2d random binning, will add explaination to description of spatial loss in binning paper writeup

@@ -53,21 +53,62 @@ QuantBin1d <- function(cs, nbin){
 
 
 ## 2d Rectangular Binning (for either Standard or Random)
+# standard is straight forward extension of 1d binning
+# random binning needs post processing to calculate minimum spatial loss
 RectBin2d <- function(xs,ys, originx, originy, widthx, widthy, type="standard"){
-  if(type=="standard"){
   tempdat <- data.frame(xs = xs, ys=ys,
                         binxs = StandRectBin1d(xs,originx,widthx),
                         binys = StandRectBin1d(ys,originy,widthy))
-  }
+  
   if(type=="random"){
-  tempdat <- data.frame(xs = xs, ys=ys,
-                          binxs = RandRectBin1d(xs,originx,widthx),
-                          binys = RandRectBin1d(ys,originy,widthy))
+    tempdat<- data.frame( tempdat,
+                          randbinxs = RandRectBin1d(xs,originx,widthx),
+                          randbinys = RandRectBin1d(ys,originy,widthy))
+    tempdat$binname <- paste(tempdat$binxs,tempdat$binys)
+    tempdat$randbinname <- paste(tempdat$randbinxs,tempdat$randbinys)
+    tempdat$randdist <- with(tempdat,sqrt((xs-randbinxs)^2 + (ys-randbinys)^2))
+    tempdat$index <- 1:length(xs)
+    # points where mismatch between standard and random binning
+    mismatchindex <- which(tempdat$binxs != tempdat$randbinxs | tempdat$binys != tempdat$randbinys )
+    mmdat <- tempdat[mismatchindex,]
+    # identify which need to be swapped to standard for net spatial loss
+    
+    #loop over all neighboring bin pairs (shift all xs over by 1 then all )
+    xbins <- seq(min(c(tempdat$binxs,tempdat$randbinxs)) , max(c(tempdat$binxs,tempdat$randbinxs)), by=widthx)
+    ybins <- seq(min(c(tempdat$binys,tempdat$randbinys)) , max(c(tempdat$binys,tempdat$randbinys)), by=widthy)
+    nbrs <- data.frame(binxs = rep(rep(xbins,length(ybins)),2) ,
+                       binys = rep(rep(ybins,each=length(xbins)),2),
+                       nbrsxs = c(rep(xbins+widthx,length(ybins)),rep(xbins,length(ybins))),
[…]

### 2015-03-18 BinningLossFunctions.R
found bug in random binning function when working with origin not equal to 0. This fixes problem.

@@ -23,7 +23,7 @@ StandRectBin1d <- function(xs, origin, width){
 #     above highest center (bxJ)
 RandRectBin1d <- function(xs, origin, width){
   bx1 <- origin + width/2
-  bxJ <- width*(floor((max(xs)-bx1)/width) + .5)
+  bxJ <- origin + width*(floor((max(xs)-bx1)/width) + .5)
   lbs <- bx1 + width*floor((xs-bx1)/width)
   ubs <- bx1 + width*ceiling((xs-bx1)/width)
   # initially assign all values to upper bound

### 2015-03-18 Paper.Rnw
New spatial loss visualization and description of total vs. net spatial loss for random binning

@@
 %
-Figure~\ref{fig:LossDisplay} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
-Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as circular and triangular points, respectively.
+Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
+Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively.
 The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
+
+For random assignment the total spatial loss can be calculated by simply replacing $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.
+This total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
+Special consideration should be payed to the two points that with spatial losses highlighted in red; in standard binning these points are in neighboring bins and have been allocated to the opposite bin during random binning.
+We see that the binned scatterplot remains identical if these points are reallocated back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
+
+The minimum total spatial loss from all binning combinations that result in the same reduced data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}.
+The net spatial loss for standard binned data will always be equivalent to the total spatial loss due to the deterministic bin allocations.
+For random binning algorithms the net spatial loss is achieved by exchanging bin assignments for pairs of points in neighboring bins that are exist further from their own bin center that their partner's bin center.
 
@@
 #Try Multivariate Normal and Uniform for this plot
-
+library(MASS)
 #MVN plot
@@
 
-\begin{figure}[h]
-  \centering
-  \subfloat[Spatial Loss MVN Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayMVN.pdf}}
-  \subfloat[Spatial Loss Uniform Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayUnif.pdf}}
-  \caption{Visualization of spatial loss.
+% \begin{figure}[h]
+%   \centering
+%   \subfloat[Spatial Loss MVN Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayMVN.pdf}}
+%   \subfloat[Spatial Loss Uniform Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayUnif.pdf}}
+%   \caption{Visualization of spatial loss.
 The distance between points and bin centers adds up to the spatial loss due to binning.
@@
 Which makes a better plot?}}
-  \label{fig:LossDisplay}
-\end{figure}
+%   \label{fig:LossDisplay}
+% \end{figure}
+
+<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms.
+Note that the total spatial loss for the standard algorithm is smaller than by random binning.
+The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>=
+
+RandRectBin1d <- function(xs, origin, width){
+  bx1 <- origin + width/2
+  bxJ <- origin + width*(floor((max(xs)-bx1)/width) + .5)
+  lbs <- bx1 + width*floor((xs-bx1)/width)
+  ubs <- bx1 + width*ceiling((xs-bx1)/width)
+  # initially assign all values to upper bound
+  binxs <- ubs
+  # then use runif to reassign based on distance from 
+  plower <- (ubs - xs)/width
+  lowerindex <- (plower > runif(length(xs), 0, 1))
+  binxs[lowerindex] <- lbs[lowerindex]
+  binxs[xs < bx1] <- bx1
+  binxs[xs > bxJ] <- bxJ
+  return(binxs)
+}
+#---------------------------------------------------
+n=12
+param1 = 1
+param2 = 1.6
+set.seed(52)
+x <- rbeta(n,param1, param2)*20-10
+y <- rbeta(n,param1, param2)*20-10
+# standard bin centers
+xcent <- rep(-5,n) ; xcent[x>0] <- 5
+ycent <- rep(-5,n) ; ycent[y>0] <- 5
+# standard bin gridlines
+temp <- data.frame(   
+  x=c( -10,   0,  10, -10, -10, -10), 
+  xend=c( -10,   0,  10,  10, 10, 10), 
+  y=c( -10, -10 , -10 , -10, 0 , 10), 
+  yend=c(10, 10 , 10, -10, 0, 10 ))
+
+p1 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+      geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+      geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
+      geom_point(aes(x=x,y=y), size=I(3))+ 
+      theme_bw() + xlab("") + ylab("") +
+      ggtitle("Standard Binning") +
+      theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+            legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+
+# random bin centers
+set.seed(14)
+rxcent <- RandRectBin1d(x, -10, 10)
+rycent <- RandRectBin1d(y, -10, 10)
+# point 1 and 11 are special 
+linecolor <- rep("blue", 12)
+linecolor[c(1,11)] <- "red"
+
+p2 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor)) + 
+  scale_colour_manual(values = c("darkblue","red"))+
+  geom_point(aes(x=x,y=y), size=I(3))+ 
+  theme_bw() + xlab("") + ylab("") +
+  ggtitle("Random Binning") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+
+# post process random bin centers to visually 
+# equivalent allocation with minimum spatial loss
+rxcentpost <- rxcent
+rycentpost <- rycent
+rxcentpost[1] <- rxcent[11]
+rycentpost[1] <- rycent[11]
+rxcentpost[11] <- rxcent[1]
+rycentpost[11] <- rycent[1]
+# point 1 and 11 are special 
+linecolor <- rep("blue", 12)
+linecolor[c(1,11)] <- "red"
+
+p3 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor)) + 
+  scale_colour_manual(values = c("darkblue","red"))+
+  geom_point(aes(x=x,y=y), size=I(3))+ 
+  theme_bw() + xlab("") + ylab("") +
+  ggtitle("Post-Processed Random Binning") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+
+allbin <- data.frame(type=rep(c("Standard","Random","Post-Processed Random"),each=n),
+                     binxs = c(xcent,rxcent,rxcentpost), 
+                     binys = c(ycent,rycent,rycentpost))
+allbinfreq <- ddply(allbin, .(type,binxs,binys),summarise,
+                     freq = length(binxs))
+
+p4 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Standard")) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
+  theme_bw() + xlab("") + ylab("")  +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        plot.margin = unit(c(0,1,0, 1),"cm"))+
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
+
+p5 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Random")) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
+  theme_bw() + xlab("") + ylab("") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        plot.margin = unit(c(0,1,0, 1),"cm"))+
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
+
+p6 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Post-Processed Random")) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
+  theme_bw() + xlab("") + ylab("") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        plot.margin = unit(c(0,1,0, 1),"cm"))+
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
+
+grid.arrange(p1,p2,p3,p4,p5,p6,nrow=2)
+@
 

### 2015-03-19 Paper.Rnw
source in binning functions and clean up code chunks for figure 3

@@
 library(MASS)
+#source in binning functions
+source("BinningLossFunctions.R")
 
@@
 y <- round(rnorm(200), digits=2)
-library(ggplot2)
 X <- data.frame(x,y)
-library(plyr)
 @
@@
 binwidth=0.1
-X$xbin <- round(X$x/binwidth,0)*binwidth
-X$ybin <- round(X$y/binwidth,0)*binwidth
-Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
-qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey90") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey90") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
+binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+qplot(x,y, asp=1, geom="blank") +
+  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  theme_bw()+ theme(legend.position="bottom") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+                      trans="log", breaks=c(1, 2, 5, 10, 20))+
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())
 @
@@
 binwidth=0.25
-X$xbin <- round(X$x/binwidth,0)*binwidth
-X$ybin <- round(X$y/binwidth,0)*binwidth
-Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
-qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey80") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey80")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
+binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+qplot(x,y, asp=1, geom="blank") +
+  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  theme_bw()+ theme(legend.position="bottom") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+                      trans="log", breaks=c(1, 2, 5, 10, 20))+
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())
 @
@@
 binwidth=0.5
-X$xbin <- round(X$x/binwidth,0)*binwidth
-X$ybin <- round(X$y/binwidth,0)*binwidth
-Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
-qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey70") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey70")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
+binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+qplot(x,y, asp=1, geom="blank") +
+  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  theme_bw()+ theme(legend.position="bottom") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+                      trans="log", breaks=c(1, 2, 5, 10, 20))+
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())
 @
@@
 binwidth=1
-X$xbin <- round(X$x/binwidth,0)*binwidth
-X$ybin <- round(X$y/binwidth,0)*binwidth
-Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
-qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey60") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey60")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))
+binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+qplot(x,y, asp=1, geom="blank") +
+  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+                                by=binwidth)+binwidth/2), colour="grey90") + 
+  theme_bw()+ theme(legend.position="bottom") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+                      trans="log", breaks=c(1, 2, 5, 10, 20))+
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())
 @
@@
 The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>=
-
-RandRectBin1d <- function(xs, origin, width){
-  bx1 <- origin + width/2
-  bxJ <- origin + width*(floor((max(xs)-bx1)/width) + .5)
-  lbs <- bx1 + width*floor((xs-bx1)/width)
-  ubs <- bx1 + width*ceiling((xs-bx1)/width)
-  # initially assign all values to upper bound
-  binxs <- ubs
-  # then use runif to reassign based on distance from 
-  plower <- (ubs - xs)/width
-  lowerindex <- (plower > runif(length(xs), 0, 1))
-  binxs[lowerindex] <- lbs[lowerindex]
-  binxs[xs < bx1] <- bx1
-  binxs[xs > bxJ] <- bxJ
-  return(binxs)
-}
-#---------------------------------------------------
 n=12

### 2015-03-19 BinningLossFunctions.R

@@ -156,78 +156,78 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
   return(binout)
 }
 
-#----------------------------------------------------
-#### Function Tests
-#----------------------------------------------------
-
-## Try 1d binning functions
-xs <- runif(100,0,100)
-cs <- rexp(100,.1)
-
-# standard binning 
-standbinnedxs <- StandRectBin1d(xs, origin=0, width=10)
-qplot(xs,  standbinnedxs)
-# random binning
-randbinnedxs <- RandRectBin1d(xs, origin=0, width=10)
-qplot(xs,  randbinnedxs)
-# quantile binning 
-quantbinnedcs <- QuantBin1d(cs, 7)
-qplot(cs,  quantbinnedcs)
-
-
-### Try 2d rectangular binning with MVN data
-n=50000
-mu <- 50
-sig <- 15
-xnorm <- rnorm(n, mean=mu, sd=sig)
-ynorm <- rnorm(n, mean=mu, sd=sig)
-qplot(xnorm, ynorm)
-# create binned data for standard and random 
-binout1 <- RectBin2d(xnorm,ynorm,0,0,10,10,type="standard")
-binout2 <- RectBin2d(xnorm,ynorm,0,0,10,10,type="random")
-# take a peek at binned data format
-head(binout1[[1]])
-head(binout2[[2]])
-# compare spatial losses
-binout1[[2]]
-binout2[[2]]
-#compare visualizations
-qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]]) +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend")
-qplot(binxs, binys, geom="tile", fill=binfreq, data=binout2[[1]]) +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend")
-#compare with shade by log counts
-qplot(binxs, binys, geom="tile", fill=log(binfreq+1), data=binout1[[1]]) +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend")
-qplot(binxs, binys, geom="tile", fill=log(binfreq+1), data=binout2[[1]]) +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend")
-
-
-
-### Try out frequency binning on the same data using standard/quantile/log-count binning
[…]

### 2015-03-19 Paper.Rnw
Minor edits and commenting out old sections (to be removed)

@@
                                 by=binwidth)+binwidth/2), colour="grey90") +
-  theme_bw()+ theme(legend.position="bottom") +
+  theme_bw() + theme(panel.grid.major= element_blank(),
+                     panel.grid.minor= element_blank(), legend.position="bottom") +
   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      trans="log", breaks=c(1, 2, 5, 10, 20))+
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())
+                      trans="log", breaks=c(1, 2, 5, 10, 20))
+
 @
@@
                                 by=binwidth)+binwidth/2), colour="grey90") +
-  theme_bw()+ theme(legend.position="bottom") +
+  theme_bw() + theme(panel.grid.major= element_blank(),
+                     panel.grid.minor= element_blank(), legend.position="bottom") +
   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      trans="log", breaks=c(1, 2, 5, 10, 20))+
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())
+                      trans="log", breaks=c(1, 2, 5, 10, 20))
 @
@@
   theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        plot.margin = unit(c(0,1,0, 1),"cm"))+
+        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
   geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
@@
 
-\ktm{Plot needed here to demonstrate shading scales under binning.}
+\ktm{Can't figure out how to get rid of the ``washed-out'' look in the continuous color scale figure.
+It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}
 
 <<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold'>>= 
-library(ggplot2)
-library(gridExtra)
-J = 0.01
+J= 1
 J2 = 200
@@
         legend.position = "none",
-        axis.ticks=element_blank(),
-        axis.text.y=element_blank(),
+        axis.ticks=element_blank(),axis.text.y=element_blank(),
+        panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
         panel.background=element_blank()
@@
   ylab("Discrete") + xlab("Count")
+
 grid.arrange(p1,p2,nrow=2)
@@
 
-\subsection{Old Loss Definitions (to be removed)}
-\ktm{The following \textit{numerical, visual and freq} loss descriptions are from the old list and will be removed: }
-\begin{itemize}
-\item {\it Numerical Loss}, $L^\text{N}$, occurs when points in the data set are reduced to a single points at the bin means.
+% \subsection{Old Loss Definitions (to be removed)}
+% \ktm{The following \textit{numerical, visual and freq} loss descriptions are from the old list and will be removed: }
+% \begin{itemize}
+% \item {\it Numerical Loss}, $L^\text{N}$, occurs when points in the data set are reduced to a single points at the bin means.
 Allocating points to the numerical center, or the mean of the data corresponding to that bin, causes a loss of information about the exact location of the points.
-
-\item {\it Visual Loss}, $L^\text{V}$, is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
+% 
+% \item {\it Visual Loss}, $L^\text{V}$, is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
 This visual center will generally different from the numerical mean, causing some bias in the perception.
@@
 The visual loss is the sum of these bin distances.
-
-\item {\it Frequency Loss} results from our inability to render the frequency information accurately.
+% 
+% \item {\it Frequency Loss} results from our inability to render the frequency information accurately.
 Different rendering methods for the frequency information will lead to different losses.
@@
 Note that even though this loss will turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where frequency information is masked in large data situations due to over-plotting of points.
-
-% 
-% , so we must balance perceptual limits of frequency information with the limits of overplotting single points.
-
-% which may not be the center of the data within the bin, causing bias and numerical loss.
-% 
-% Visual loss occurs in the graphical display of the data, as individual points are represented as tiles, and tiles are perceived according to their visual center, rather than according to the center of the data.
-% 
-%  
-\end{itemize}
-
-We explore the different types of loss in the remainder of this section in more detail.
+% 
+% % 
+% % , so we must balance perceptual limits of frequency information with the limits of overplotting single points. 
+% 
+% 
+% % which may not be the center of the data within the bin, causing bias and numerical loss.
+% % 
+% % Visual loss occurs in the graphical display of the data, as individual points are represented as tiles, and tiles are perceived according to their visual center, rather than according to the center of the data.
+% % 
+% %  
+% \end{itemize}
+% 
+% We explore the different types of loss in the remainder of this section in more detail.
 For clarity we will again be specifying loss using notation for a one dimensional binning process.
 This notation can then be extended to higher dimensions in applications.
-
-\subsection{Numerical and Visual Loss}
-Reducing data points $\mathbf{x}$ to the visual bin centers $\mathbf{x}^\ast$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
-If the visual bin centers are the same as the numerical centers for each bin, then any loss could be strictly attributed to difference between points and the bin centers. 
-The problem is that the visual center and the numerical center for bins will rarely match, which introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of the symbol representing the data.
-
-Therefore we distinguish two contributions to the total loss. 
-Numerical loss results from the binning process as points are reduced to the numerical center point.
+% 
+% \subsection{Numerical and Visual Loss}
+% Reducing data points $\mathbf{x}$ to the visual bin centers $\mathbf{x}^\ast$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
+% If the visual bin centers are the same as the numerical centers for each bin, then any loss could be strictly attributed to difference between points and the bin centers. 
+% The problem is that the visual center and the numerical center for bins will rarely match, which introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of the symbol representing the data.
+% 
+% Therefore we distinguish two contributions to the total loss. 
+% Numerical loss results from the binning process as points are reduced to the numerical center point.
 The loss for an individual point, $L_i^\text{N}$, is the euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$.
 The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
-Visual loss is due to the distance between the numerical centers and the visual centers.
+% Visual loss is due to the distance between the numerical centers and the visual centers.
 The visual loss for each bin, $L_j^\text{V}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
 The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$).
-This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.
-This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions apply to either approach to binning.
-
-The magnitude of the numerical and visual losses are clearly dictated by the scale on which the data was recorded, thus and the loss values have no direct interpretation.
+% This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.
+% This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions apply to either approach to binning.
+% 
+% The magnitude of the numerical and visual losses are clearly dictated by the scale on which the data was recorded, thus and the loss values have no direct interpretation.
 However, we can examine both the numerical and the visual loss from the perspective of proportion of the greatest possible loss.
-To do this for numerical loss we consider the binning scenario that would always lead to greatest possible loss; allocating all data points into a single bin.
+% To do this for numerical loss we consider the binning scenario that would always lead to greatest possible loss; allocating all data points into a single bin.
 Thus we can define the greatest possible numerical loss as
-
-\begin{center}
-$ L_0^\text{N} = \sum_{i=1}^{n} \sqrt{(x_i-\bar{x})^2 + (y_i-\bar{y})^2}  $, 
-\end{center}
-
-where $\bar{x}$ and $\bar{y}$ are the sample averages for x and y respectively.
+% 
+% \begin{center}
+% $ L_0^\text{N} = \sum_{i=1}^{n} \sqrt{(x_i-\bar{x})^2 + (y_i-\bar{y})^2}  $, 
+% \end{center}
+% 
+% where $\bar{x}$ and $\bar{y}$ are the sample averages for x and y respectively.
 After we select a binning strategy and binwidth we can then assess the proportion of possible numerical loss incurred due to binning by computing $L^\text{N} / L_0^\text{N}$.
@@
 %-------
-
 <<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
@@
 @
-
-\noindent
 
@@
 
-\begin{figure}[hbtp]
-    \centering
-		\subfloat[Numerical and Visual Loss]{\includegraphics[keepaspectratio=true, width=.46\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
-    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/FreqLoss.pdf}}
-\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.
+% \begin{figure}[hbtp]
+%     \centering
+% 		\subfloat[Numerical and Visual Loss]{\includegraphics[keepaspectratio=true, width=.46\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
+%     \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/FreqLoss.pdf}}
+% \caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.
 (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
-	\end{figure}
+% 	\end{figure}
 

### 2015-03-19 Paper.Rnw
move old "information recovery" subsection into new subsection in binning section. Discuss binned data reduction (one row per point -> one row per bin with freq column added).  Reduced binned data is concept necessary to be referenced in loss section.

@@
 
+\subsection{Binned Data Reduction}
+
+\ktm{Section to describe how data is stored post binning.
+Section will discuss difference between binned data (bin centers added to original data, still n rows) and "reduced binned data" (aggregated into frequencies), sufficiency for constructing binned scatterplot and storage size.
+Want to reference these ideas in later sections.}
+
+% As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
+% ). We can make use of this and
+%  recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations.
+Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.
+
+<<DataReduction,echo=FALSE,eval=FALSE,results='asis'>>=
+n=12
+param1 = 1
+param2 = 1.6
+set.seed(52)
+x <- rbeta(n,param1, param2)*20-10
+y <- rbeta(n,param1, param2)*20-10
+# standard bin centers
+xcent <- rep(-5,n) ; xcent[x>0] <- 5
+ycent <- rep(-5,n) ; ycent[y>0] <- 5
+idx <- order(xcent,ycent)
+d <- data.frame(x,y,xcent,ycent)[idx,]
+dbin <- ddply(d, .(xcent,ycent), summarise,
+              c=length(x))
+dbin <- dbin[order(dbin$xcent,dbin$ycent),]
+
+print(xtable(d[1:2], digits=c(0, 4, 4), caption=paste('Original Data\\newline', nrow(d), 'rows')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)
+
+print(xtable(d[3:4], digits=c(0, 0, 0), caption=paste('Binned Data\\newline', nrow(d), 'rows')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE)
+
+print(xtable(dbin, digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')
+
+@
+
+\begin{table}[H]
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 12 rows]{
+\begin{tabular}{rr}\hline
+       $x$ &  $y$  \\ \hline
+-7.7325 & -9.6340 \\ 
+  -8.1176 & -1.4529 \\ 
+  -5.8996 & -3.2033 \\ 
+  -7.0375 & -5.5563 \\ 
+  -3.6354 & -3.9315 \\ 
+  -8.7639 & 0.9874 \\ 
+  -2.9781 & 8.6802 \\ 
+  0.8210 & -8.6118 \\ 
+  5.4477 & -8.4555 \\ 
+  4.6849 & -5.6620 \\ 
+  9.4785 & 1.1133 \\ 
+  1.7579 & 5.3759 \\  \hline
+\end{tabular}}
+\end{minipage}\hfil
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Centers, 12 rows]{
+\begin{tabularx}{.7\textwidth}{rr} \hline
+  $b_X(x)$ & \hspace{.4cm} $b_Y(y)$  \\ \hline
+  -5 & -5 \\ 
+  -5 & -5 \\ 
+  -5 & -5 \\ 
+  -5 & -5 \\ 
+  -5 & -5 \\ 
+  -5 & 5 \\ 
+  -5 & 5 \\ 
+  5 & -5 \\ 
+  5 & -5 \\ 
+  5 & -5 \\ 
+  5 & 5 \\ 
+  5 & 5 \\  \hline
+\end{tabularx}}
+\end{minipage}\hfil
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Reduced Binned Data, 4 rows]{
+\begin{tabularx}{.7\textwidth}{rrr}\hline
+ \hspace{.2cm} $x^\ast$   &   \hspace{.2cm} $y^\ast$ & \hspace{.4cm}c \\ \hline
+-5 & -5 & 5 \\ 
+  -5 & 5 & 2 \\ 
+  5 & -5 & 3 \\ 
+  5 & 5 & 2 \\  \hline
+\end{tabularx}}
+\end{minipage}
+\label{DataReductionExample}
+\caption{Original, Binned and Reduced Binned Data Tables, with data storage sizes.  Binned using standard rectangular approach with origin at (-10,-10) and binwidths $\omega_x = \omega_y = 10$. }
+\end{table}
+
+Information recovery comes with a two-fold cost: the additional storage of numeric bin centers triples the amount of information stored, and the computational cost to compute numeric bin centers is O($nm^2$), where $n$ is the number of data points and $m^2$, the total number of bins.
+O($nm^2$) $\approx$ O($n$), as $m$ is generally negligible compared to $n$.
+
+In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable.
+Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning.
+
 % \subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
@@
 
-\subsection{Recovery of Numerical Information}
-% The total loss from reducing a value $x$ to bin center $x^\ast$ is at least as large as the loss from reducing points $x_i$ to $\overline x_i$, the numerical center of the binned data.
-% % that's only true for the expected value - not for all individuals, but you don't need - we just use the partition.
-
-As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
-). We can make use of this and
- recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations.
-Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.
-
-<<recover,echo=FALSE,eval=FALSE,results='asis'>>=
-set.seed(90)
-x <- rnorm(100, 5, 2)
-y <- rnorm(100, 5, 2)
-d <- data.frame(x=x, y=y, c=1)
-d <- d[order(d$x, d$y),]
-print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
-dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
-dbin <- dbin[order(dbin$x, dbin$y),]
-print(xtable(head(dbin), digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')
-names(d) <- c("x", "y", "Freq")
-drecover <- unique(lossCalc(d, c(1,1), "standard", newData=TRUE)$NewData)
-names(drecover) <- c("x", "y", "c")
-drecover <- drecover[order(drecover$x, drecover$y),]
-print(xtable(head(drecover), digits=c(0, 4, 4, 0), caption=paste('Recovered Data\\newline', nrow(drecover), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
-@
-
-\begin{table}[hbtp]
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 100 rows]{
-\begin{tabular}{rrr}\hline
-       $x$ &      $y$ & c \\ \hline
- -0.2582 & 1.5497 & 1 \\ 
- -0.3843 & 5.6508 & 1 \\ 
- 1.0597 & 2.4308 & 1 \\ 
- 2.2680 & 1.5877 & 1 \\ 
- 2.4992 & 4.6260 & 1 \\ 
- 1.6788 & 5.0260 & 1 \\ 
- 1.6054 & 6.4125 & 1 \\ 
- 2.4180 & 6.4438 & 1 \\ 
- 2.4345 & 6.4462 & 1 \\ 
- 2.3620 & 10.3119 & 1 \\  \hline
-\end{tabular}}
-\end{minipage}\hfil
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Visual Centers, 49 rows]{
-\begin{tabularx}{.7\textwidth}{XXc} \hline
-  $x^\ast$ & $y^\ast$ & c \\ \hline
-  0 & 2 & 1 \\ 
-  0 & 6 & 1 \\ 
-  1 & 2 & 1 \\
-  2 & 2 & 1 \\
-  2 & 5 & 2 \\ 
-  2 & 6 & 3 \\ 
-  2 & 10 & 1 \\ 
-   \hline
-\end{tabularx}}
-\end{minipage}\hfil
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Numerical Centers, 49 rows]{
-\begin{tabular}{rrc}\hline
-  $\bar{x}$   &    $\bar{y}$ & c \\ \hline
- -0.3843 & 5.6508 & 1 \\ 
- -0.2582 & 1.5497 & 1 \\ 
-  1.0597 & 2.4308 & 1 \\ 
-  2.2680 & 1.5877 & 1 \\  
-  2.0890 & 4.8260 & 2 \\ 
-  2.1526 & 6.4341 & 3 \\ 
-  2.2680 & 1.5877 & 1 \\ \hline
-\end{tabular}}
-\end{minipage}
-\caption{\label{RecoveryComparison}ten rows from Original and Binned Data Tables, with data storage sizes}
-\end{table}
-
-Information recovery comes with a two-fold cost: the additional storage of numeric bin centers triples the amount of information stored, and the computational cost to compute numeric bin centers is O($nm^2$), where $n$ is the number of data points and $m^2$, the total number of bins.
-O($nm^2$) $\approx$ O($n$), as $m$ is generally negligible compared to $n$.
-
-In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable.
-Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning.
-%but creates an additional complexity in that the geometric center of the tile (the visual bin center) does not correspond to the stored, binned data.
-% The random binning algorithm serves as a less computationally intensive, unbiased alternative for 
-% data sets which are too large to partition the loss in this manner.
-% \hh{is that true, or does it require the expected value to be the visual center?}
+%-----------------------------------------------------------------------------
 

### 2015-03-19 Paper.Rnw
fixing reference to DataReductionExample

@@
 \end{minipage}
-\label{DataReductionExample}
-\caption{Original, Binned and Reduced Binned Data Tables, with data storage sizes.  Binned using standard rectangular approach with origin at (-10,-10) and binwidths $\omega_x = \omega_y = 10$. }
+\caption{\label{DataReductionExample}Original, Binned and Reduced Binned Data Tables, with data storage sizes.
+Binned using standard rectangular approach with origin at (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
+}
 \end{table}
@@
 %
-Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for observations binned using a standard rectangular binning algorithm.
+Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning.
 Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively.

### 2015-03-20 Paper.Rnw
lots of minor edits to sections 3 and 4: moving plots, cleaning plots, wording reworks.

@@
 In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions.
-For large data, scatterplots suffer from over-plotting, %-- i.e.\ a number of observations drawn in such close proximity that their points overlap,
-thereby masking relevant structure.
+With large data, scatterplots suffer from over-plotting of points, %-- i.e.\ a number of observations drawn in such close proximity that their points overla
+which masks relevant structure.
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
@@
 \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting.
-Open circles make points visible that are close together.
-This technique is  not suitable to determine the magnitude, but gives more information than is available with filled points.
-%\hh{We need to decide on one way for referring to $\alpha$-blending in one way - we've got about four different ways of writing it right now. I'm ok with any of them, but we need to stick to it.}
-A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.
+Open circles make points that are close together more visually distinct; thus allowing for the perception of more density information than with filled points.
+A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c).
+Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.
 
@@
 
-%\hh{discussion of sufficient statistics for graphics could go here.
+Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales.
+Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations (including ties) but use distortion of the point locations by having points repel one another to avoid overlapping points.
+An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
+Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
+This variable binning fragments the continuity of the axes into  segments on different scales and also does not deal with ties.
+Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.
+
+\hh{discussion of sufficient statistics for graphics could go here.
 Idea: the problems with large data in scatterplots all arise from overplotting, which is a form of implicit data aggregation.
@@
 Making the data aggregation explicit allows us to calculate the loss we experience.}
-
-Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales.
-Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations (including ties) but use distortion of the point locations by having points repel one another to avoid overlapping points.
-An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
-Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
-This variable binning fragments the continuity of the axes into  segments on different scales and also does not deal with ties.
-Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.
 
@@
 
-In general, for the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.
-
-{\textit Rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
+For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.
+
+\textit{Rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
 Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
@@
 
-% This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
-% %
-% \begin{eqnarray}\label{rectbin}
-% b_X(x_i) = \left\{\begin{array}{ll} 
-%   x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
-%   x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
-%   \end{array}\right.
-% \end{eqnarray}  
-% %
+This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
+%
+\begin{eqnarray}\label{rectbin}
+b_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
+  x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
+  \end{array}\right.
+\end{eqnarray}  
+%
 
@@
 Note that this standard rectangular binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
-These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R}.
+These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R} \citep{R}.
 These specifications were selected for this paper, but these choices are by no means consider universal for binning.
-For example, the \texttt{ggplot2} package creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin.
+For example, the \texttt{ggplot2} package creates histograms with intervals open to the right and does not close the outer bound of the rightmost bin \citep{ggplot2}.
 
@@
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
-The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
+The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
 More formally,
@@
 \end{figure}
+
+%--------------------------------------------------------------
 
@@
 
+%--------------------------------------------------------------
+
 \section{Loss due to Binning}
@@
 The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
+
+<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms.
+Note that the total spatial loss for the standard algorithm is smaller than by random binning.
+The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>=
+n=12
+param1 = 1
+param2 = 1.6
+set.seed(52)
+x <- rbeta(n,param1, param2)*20-10
+y <- rbeta(n,param1, param2)*20-10
+# standard bin centers
+xcent <- rep(-5,n) ; xcent[x>0] <- 5
+ycent <- rep(-5,n) ; ycent[y>0] <- 5
+# standard bin gridlines
+temp <- data.frame(   
+  x=c( -10,   0,  10, -10, -10, -10), 
+  xend=c( -10,   0,  10,  10, 10, 10), 
+  y=c( -10, -10 , -10 , -10, 0 , 10), 
+  yend=c(10, 10 , 10, -10, 0, 10 ))
+
+p1 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+      geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+      geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
+      geom_point(aes(x=x,y=y), size=I(3))+ 
+      theme_bw() + xlab("") + ylab("") +
+      ggtitle("Standard Binning") +
+      theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+            legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+
+# random bin centers
+set.seed(14)
+rxcent <- RandRectBin1d(x, -10, 10)
+rycent <- RandRectBin1d(y, -10, 10)
+# point 1 and 11 are special 
+linecolor <- rep("blue", 12)
+linecolor[c(1,11)] <- "red"
+
+p2 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor)) + 
+  scale_colour_manual(values = c("darkblue","red"))+
+  geom_point(aes(x=x,y=y), size=I(3))+ 
+  theme_bw() + xlab("") + ylab("") +
+  ggtitle("Random Binning") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+
+# post process random bin centers to visually 
+# equivalent allocation with minimum spatial loss
+rxcentpost <- rxcent
+rycentpost <- rycent
+rxcentpost[1] <- rxcent[11]
+rycentpost[1] <- rycent[11]
+rxcentpost[11] <- rxcent[1]
+rycentpost[11] <- rycent[1]
+# point 1 and 11 are special 
+linecolor <- rep("blue", 12)
+linecolor[c(1,11)] <- "red"
+
+p3 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
+  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor)) + 
+  scale_colour_manual(values = c("darkblue","red"))+
+  geom_point(aes(x=x,y=y), size=I(3))+ 
+  theme_bw() + xlab("") + ylab("") +
+  ggtitle("Post-Processed Random Binning") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+
+allbin <- data.frame(type=rep(c("Standard","Random","Post-Processed Random"),each=n),
+                     binxs = c(xcent,rxcent,rxcentpost), 
+                     binys = c(ycent,rycent,rycentpost))
+allbinfreq <- ddply(allbin, .(type,binxs,binys),summarise,
+                     freq = length(binxs))
+
+p4 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Standard")) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
+  theme_bw() + xlab("") + ylab("")  +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
+
+p5 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Random")) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
+  theme_bw() + xlab("") + ylab("") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
+
+p6 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
+  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Post-Processed Random")) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
+  theme_bw() + xlab("") + ylab("") +
+  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
+        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
+
+grid.arrange(p1,p2,p3,p4,p5,p6,nrow=2)
+@
 
@@
 % \end{figure}
-
-<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms.
-Note that the total spatial loss for the standard algorithm is smaller than by random binning.
-The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>=
-n=12
-param1 = 1
-param2 = 1.6
-set.seed(52)
-x <- rbeta(n,param1, param2)*20-10
-y <- rbeta(n,param1, param2)*20-10
-# standard bin centers
-xcent <- rep(-5,n) ; xcent[x>0] <- 5
-ycent <- rep(-5,n) ; ycent[y>0] <- 5
-# standard bin gridlines
-temp <- data.frame(   
-  x=c( -10,   0,  10, -10, -10, -10), 
-  xend=c( -10,   0,  10,  10, 10, 10), 
-  y=c( -10, -10 , -10 , -10, 0 , 10), 
-  yend=c(10, 10 , 10, -10, 0, 10 ))
-
-p1 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
-      geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-      geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
-      geom_point(aes(x=x,y=y), size=I(3))+ 
-      theme_bw() + xlab("") + ylab("") +
-      ggtitle("Standard Binning") +
-      theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-            legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
-
-# random bin centers
-set.seed(14)
-rxcent <- RandRectBin1d(x, -10, 10)
-rycent <- RandRectBin1d(y, -10, 10)
-# point 1 and 11 are special 
-linecolor <- rep("blue", 12)
-linecolor[c(1,11)] <- "red"
-
-p2 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor)) + 
-  scale_colour_manual(values = c("darkblue","red"))+
-  geom_point(aes(x=x,y=y), size=I(3))+ 
-  theme_bw() + xlab("") + ylab("") +
-  ggtitle("Random Binning") +
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
-
-# post process random bin centers to visually 
-# equivalent allocation with minimum spatial loss
-rxcentpost <- rxcent
-rycentpost <- rycent
-rxcentpost[1] <- rxcent[11]
-rycentpost[1] <- rycent[11]
-rxcentpost[11] <- rxcent[1]
-rycentpost[11] <- rycent[1]
-# point 1 and 11 are special 
-linecolor <- rep("blue", 12)
-linecolor[c(1,11)] <- "red"
-
-p3 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor)) + 
-  scale_colour_manual(values = c("darkblue","red"))+
-  geom_point(aes(x=x,y=y), size=I(3))+ 
-  theme_bw() + xlab("") + ylab("") +
-  ggtitle("Post-Processed Random Binning") +
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
-
-allbin <- data.frame(type=rep(c("Standard","Random","Post-Processed Random"),each=n),
-                     binxs = c(xcent,rxcent,rxcentpost), 
-                     binys = c(ycent,rycent,rycentpost))
-allbinfreq <- ddply(allbin, .(type,binxs,binys),summarise,
-                     freq = length(binxs))
-
-p4 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
-  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Standard")) + 
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
-  theme_bw() + xlab("") + ylab("")  +
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
-
-p5 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
-  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Random")) + 
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
-  theme_bw() + xlab("") + ylab("") +
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
-
-p6 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
-  geom_tile(aes(x=binxs, y=binys, fill=freq),data=subset(allbinfreq,type=="Post-Processed Random")) + 
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend") + 
-  theme_bw() + xlab("") + ylab("") +
-  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
-
-grid.arrange(p1,p2,p3,p4,p5,p6,nrow=2)
-@
-
 %------------------------------------------------
@@
 
- Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
-
 \hh{It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
@@
 <<context, echo=F, fig.width=6, fig.height=4.5, out.width='.5\\textwidth'>>=
-col <- c(rnorm(200, mean=0, sd=10), rnorm(10, mean=50), rnorm(200, mean=100, sd=10), rnorm(10, mean=50))
-x <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=1, sd=0.075))
-y <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=1, sd=0.075))
+#col <- c(rnorm(200, mean=0, sd=10), rnorm(10, mean=50), rnorm(200, mean=100, sd=10), rnorm(10, mean=50))
+Shade <- c(rbeta(200, 1,10),rep(.5,10), rbeta(200,10,1), rep(.5,10))
+x <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=.5, sd=0.075))
+y <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=.5, sd=0.075))
+Shape <- c(rep("circle",200),rep("triangle",10),rep("circle",200),rep("triangle",10))
 
 require(ggplot2)
-qplot(x, y, colour=col, size=I(5)) + theme_bw() + theme(aspect.ratio=1)
-@
-\caption{\label{fig:context}Assess the colors in this scatterplot. Are the darkest spots in the cloud of lighter points as dark as the lightest spots in the cloud of darker points? Darker? Lighter? }
+qplot(x, y, colour=Shade, size=I(5),shape=Shape) + theme_bw() + 
+  xlab("") + ylab("")+
+  theme(aspect.ratio=1, axis.ticks=element_blank(), axis.text=element_blank()) +
+  scale_colour_gradient(limits=c(0,1), breaks=c(0,.5,1))
+@
+\caption{\label{fig:context}Assess the colors in this scatterplot.
+Are the triangles in the lighter cloud of points as dark as the triangles in the darker cloud of points?
+The answer is that all the triangles have the exact same shade!}
 \end{figure}
 
-As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
+Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
 

### 2015-03-20 Paper.Rnw
Binned Data Reduction subsection

@@
   \centering
-  \subfloat[Standard Binning]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
+  \subfloat[Standard Binning]{\includegraphics[keepaspectratio=true, width=2.5in]{./figure/Binning-DataAndBin.pdf}}
 %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
 %   {./figure/RandomBinningDataPoints1}}
-  \subfloat[Random Binning]{\includegraphics[keepaspectratio=TRUE, width=3in]
+  \subfloat[Random Binning]{\includegraphics[keepaspectratio=TRUE, width=2.5in]
   {./figure/RandomBinningDataPoints2.pdf}}
@@
 
-\ktm{Section to describe how data is stored post binning.
-Section will discuss difference between binned data (bin centers added to original data, still n rows) and "reduced binned data" (aggregated into frequencies), sufficiency for constructing binned scatterplot and storage size.
-Want to reference these ideas in later sections.}
-
-% As shown before, the overall loss partitions naturally into visual and numerical loss (see equation \hh{XXX need new equation reference} %\ref{eqn:partition}
-% ). We can make use of this and
+\ktm{ As previously mentioned second stage of binning requires that we count the number of observations assigned to each unique bin center; forming data triples, $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+Table~\ref{DataReductionExample} uses a small set of simulated data to show the progression from the original data (a), to the binned data (b), to the reduced binned data (c).
+The reduced binned data is sufficient for constructing the binned scatterplot, as only bin centers and frequencies are necessary.
+As such the storage size for the information and the computation time needed to construct the plot is greatly reduced.
+}
+
+\ktm{ The data reduction comes at the expense of spatial information of any of the individual points.
+After the binned data is aggregated, there is no way to recover the original spatial locations.
+The loss of information incurred from binning will be explored in following sections.
+}
+
 % recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations.
@@
 
-Information recovery comes with a two-fold cost: the additional storage of numeric bin centers triples the amount of information stored, and the computational cost to compute numeric bin centers is O($nm^2$), where $n$ is the number of data points and $m^2$, the total number of bins.
+%Information recovery comes with a two-fold cost: the additional storage of numeric bin centers triples the amount of information stored, and the computational cost to compute numeric bin centers is O($nm^2$), where $n$ is the number of data points and $m^2$, the total number of bins.
 O($nm^2$) $\approx$ O($n$), as $m$ is generally negligible compared to $n$.
 
-In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable.
+% In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable.
 Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning.

### 2015-03-20 Paper.Rnw
editing section 4.1

@@
 In the first set of binned data, the bins and the points on the scatterplot are nearly identical, but the scatterplot contains information about overlapping points.
-The second and third sets of binned data, with bin width=0.25 and 0.50 respectively, show higher-level summaries of the data that contain some numerical and visual loss but which may also provide more visually accessible information about the shape of the two-dimensional density between $x$ and $y$.
+The second and third sets of binned data, with bin width=0.25 and 0.50 respectively, show higher-level summaries of the data but which may also provide more visually accessible information about the shape of the two-dimensional density between $x$ and $y$.
 The fourth set of binned data with bin width = 1.0 is nearly unrecognizable, because the bins are large enough that they provide very little additional information.
@@
 \begin{itemize}
-\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) \hh{XXX define $i$, $l$} in the data set are reduced to a single points at the bin centers ($x_\ell^\ast$,$y_\ell^\ast$).
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers ($x_\ell^\ast$,$y_\ell^\ast$) for bins $\ell \in \{1,\dots,L\}$.
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
@@
 For random assignment the total spatial loss can be calculated by simply replacing $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.
-This total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
+The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
 Special consideration should be payed to the two points that with spatial losses highlighted in red; in standard binning these points are in neighboring bins and have been allocated to the opposite bin during random binning.
@@
 
-The minimum total spatial loss from all binning combinations that result in the same reduced data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}.
+The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}.
 The net spatial loss for standard binned data will always be equivalent to the total spatial loss due to the deterministic bin allocations.
-For random binning algorithms the net spatial loss is achieved by exchanging bin assignments for pairs of points in neighboring bins that are exist further from their own bin center that their partner's bin center.
+For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then finding the total spatial loss from this processed binned data.
 

### 2015-03-20 Paper.Rnw
Edits to section 4.2

@@
 While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
-Color perception of shade is extremely context sensitive (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
-
+Color perception of shade is extremely context sensitive allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
 \hh{It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
@@
 
-Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
+Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
+As a result, a binning width which produces seven of fewer frequency categories is preferable.%, while minimizing numerical loss within that constraint.
 
@@
 We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
-
-While this numerical assessment of frequency loss does not exactly account for limitations in human perceptual ability, it does provide a more realistic model for the loss in perception that does occur.
-
-The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
-The \textit{total frequency loss}, $L^F$, is defined as
-%
-\begin{eqnarray}\label{totfreqloss}
-L^F = \sum_{\ell=1}^{L} L_\ell^F = \sum_{\ell=1}^{L} (c_\ell-b_c(c_\ell))^2
-\end{eqnarray}  
-%
-where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.
-
-Frequency data consists of counts, which most commonly exhibit skew densities, i.e.\ there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
-A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution.
-This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}.
-Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph.
-The \textit{log frequency loss}, $L^{\log F}$, is defined as
-%
-\begin{eqnarray}\label{logfrequency}
-L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
-\end{eqnarray}
-%
-Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.
-
-\ktm{Can't figure out how to get rid of the ``washed-out'' look in the continuous color scale figure.
-It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}
-
-<<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold'>>= 
+Figure~\ref{fig:FreqLossDemo} provides a visual example of a discrete color palette resulting from frequency binning.
+
+<<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Continuous and discrete color scales for counts in the range 0 to 1000.
+Frequency binning done using standard rectangular binning with origin $\\beta_0 = 0$ and binwidth $\\omega_X = 200$.
+\\ktm{Can't figure out how to get rid of the ``washed-out'' look in the continuous color scale figure.
+It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}">>=
 J= 1
@@
 
+The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
+The \textit{total frequency loss}, $L^F$, is defined as
+%
+\begin{eqnarray}\label{totfreqloss}
+L^F = \sum_{\ell=1}^{L} L_\ell^F = \sum_{\ell=1}^{L} (c_\ell-b_c(c_\ell))^2
+\end{eqnarray}  
+%
+where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.
+
+While this numerical assessment of frequency loss does not exactly account for limitations in human perceptual ability, it does provide a more realistic model for the loss in perception that does occur.
+
+Frequency data consists of counts, which most commonly exhibit skew densities, i.e.\ there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
+The use of quantile binning is promising in the case of frequency binning because it seeks to place the same number of values in each frequency bin, thus balance the number of tiles at each shade.
+
+Another alternative is to use a log transformation which produces a more symmetric distribution of frequency information, increasing perceptual resolution.
+This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}.
+Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph.
+The \textit{log frequency loss}, $L^{\log F}$, is defined as
+%
+\begin{eqnarray}\label{logfrequency}
+L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
+\end{eqnarray}
+%
+Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.
+
 % \subsection{Old Loss Definitions (to be removed)}
@@
 
-\subsection{Loss of Frequency Information} 
+\subsection{Loss of Frequency Information (old writeup - to be reworked or removed)} 
 Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently.
@@
 
-\paragraph{Traditional Scatterplots and Frequency Information:}
+\paragraph{Traditional Scatterplots and Frequency Information: (old writeup - to be reworked or removed)}
 

### 2015-03-20 Paper.Rnw
Outline of New Section 5 for loss properties (explored analytically and through simulation)

@@
 
+\section{Exploring Properties of Loss}
+
+\begin{itemize}
+\item Binning and frequency binning involve choices (standard/random/quantile, origin, binwidth, number of colors) want to make optimal choices given goals of binning.
+\item 3 simulated data sets: from bivariate Exponential, Normal, Uniform
+\item Since most data recorded to rough resolution we also assess loss when each dataset is rounded to nearest even value.  
+\end{itemize}
+
+%-----------------------------------------------------------------------------
+
+\subsection{Spatial binning choices}
+
+\begin{itemize}
+\item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytical proof)
+\item Origin: 
+  \begin{itemize}
+    \item Doesn't much matter for fine resolution data but it does for coarse data.
+    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytical proof then demo)
+  \end{itemize} 
+\item Binwidth: 
+  \begin{itemize}
+    \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
+    \item for fine data any binwidth is fine
+    \item for coarse data need binwidths $\omega \ge \alpha$ (show plot with whitespace gaps)
+    \item for coarse data binwidths should be scalar multiple of $\alpha$ (show plot with strips)
+    \item Does the distribution type matter? 
+  \end{itemize}
+\end{itemize}
+
+%-----------------------------------------------------------------------------
+
+\subsection{Frequency binning choices}
+
+\begin{itemize}
+\item Freq Binning Type
+  \begin{itemize}
+    \item Comparison of standard, log and quantile binning
+  \end{itemize}
+\end{itemize}
+
+\begin{itemize}
+\item Number of Colors:
+  \begin{itemize}
+    \item Perception driven limitation to 7 or less color bins
+    \item As number colors increases, freq loss goes up?
+  \end{itemize}
+\end{itemize}
+
+%-----------------------------------------------------------------------------
+
 \section{Discussion and Examples}

### 2015-03-22 Paper.Rnw
Appendix for proof that if data recorded to resolution alpha and bin width is scalar multiple of alpha then it is optimal to shift the origin to alpha/2 below the minimum data value.

@@
 
+%----------------------------------------------------------------------------
+
 \begin{appendix}
 \section{Mathematical Considerations}
+
+%----------------------------------------------------------------------------
+\subsection{Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}\label{proof:offset}
+
+The following is proof that when univariate data is uniformly distributed at resolution $\alpha$ and standard rectangular binning is used with binwidths that are a scalar multiple of the data resolution (i.e.
+$\omega = k\alpha$ for some $k \in \{1,2,\dots\})$, then spatial loss is minimized by setting the binning origin to $\alpha/2$ units below the minimum data value; set $\beta = x_{(1)} - \alpha/2$.\\
+
+\noindent Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha$ for some constant $\alpha \in \mathbb{R}$. Thus $x_j = x_1 + (j-1)\alpha$.\\
+
+\noindent Suppose then that we bin the data using standard rectangular binning with origin, $\beta = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data.
+Thus $b(x_j) = \beta + \omega/2 = (x_1 - \theta) + (k\alpha/2)$\\
+
+\noindent Spatial Loss, $L^S = \sum_{i=1}^{k} ||x_i-b_(x_i)|| $ is definitional minimized when $b_(x_i)$ is the \textit{geometric median}.
+The geometric median for $x_1, \dots, x_k = Q_x(.5) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2$ , where $Q_x(\cdot)$ is the empirical quantile function.
+\\
+
+\noindent Thus the optimal offset is the $\theta$ such that\\
+
+$ b(x_i) = Q_x(.5) $ \\
+$ \Rightarrow (x_1 - \theta) + (k\alpha/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
+$ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
+$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
+$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha ) $ \\
+$ \Rightarrow -2\theta = -\alpha $ \\
+$ \Rightarrow \theta = \alpha/2 $ \\
+
+Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha/2$.
+This result holds for data that is symmetrically distributed within the bin since the median will not change.
+It extends to multiple contiguous bins with resolution $\alpha$ data that has symmetrically distributed data withing each bin.
+\\
+
+If the same conditions are extended to the two dimensional case, then the origin for minimal spatial loss is at $(x_{(1)}-\alpha_x/2, y_{(1)}-\alpha_y/2)$ where $\alpha_x$ and $\alpha_y$  are the data resolution for each dimesion, respectively.
+
+%----------------------------------------------------------------------------
+
 \subsection{Partition of Spatial Loss}\label{proof:partition}

### 2015-03-22 Paper.Rnw
reference to appendix
Intro to Section 5 and also coloring sections worked on over break in red

@@
 
-For random assignment the total spatial loss can be calculated by simply replacing $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.
+\ktm{ For random assignment the total spatial loss can be calculated by simply replacing $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.
 The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
@@
 We see that the binned scatterplot remains identical if these points are reallocated back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
-
-The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}.
+}
+
+\ktm{ The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}.
 The net spatial loss for standard binned data will always be equivalent to the total spatial loss due to the deterministic bin allocations.
 For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then finding the total spatial loss from this processed binned data.
+}
 
@@
 
+\ktm{ Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm and also the parameters of the chosen binning algorithm.
+This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
+Some results are derived optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.
+}
+
+\ktm{ Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
+Bivariate datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
+These distributions were selected for their variety in terms of shape, center and spread.
+100,000 observation pairs were simulated for each set of simulated data, and the x and y values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.
+The parameters were selected to have data values roughly span the region $[0,100]^2$.
+}
+
+\ktm{ These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places.
+Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles.
+To observe binning loss under more realistic conditions, three corresponding data sets were also created by rounding the values from the original simulated data sets to the nearest even number.
+By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.
+}
+
+\ktm{
 \begin{itemize}
@@
 \end{itemize}
+}
 
@@
 
+\ktm{
 \begin{itemize}
@@
     \item Doesn't much matter for fine resolution data but it does for coarse data.
-    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytical proof then demo)
+    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytical proof then demo) Proof in Appendix~\ref{proof:offset}.
   \end{itemize} 
@@
 
+\ktm{
 The following is proof that when univariate data is uniformly distributed at resolution $\alpha$ and standard rectangular binning is used with binwidths that are a scalar multiple of the data resolution (i.e.
@@
 
+} %end ktm color
 %----------------------------------------------------------------------------

### 2015-03-25 Paper.Rnw
proofing through sections 4

@@
-\documentclass[letterpaper,12pt]{article}
+\documentclass[letterpaper,10pt]{article}
 \usepackage[utf8]{inputenc}
@@
 
-\ktm{ As previously mentioned second stage of binning requires that we count the number of observations assigned to each unique bin center; forming data triples, $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
-Table~\ref{DataReductionExample} uses a small set of simulated data to show the progression from the original data (a), to the binned data (b), to the reduced binned data (c).
-The reduced binned data is sufficient for constructing the binned scatterplot, as only bin centers and frequencies are necessary.
-As such the storage size for the information and the computation time needed to construct the plot is greatly reduced.
-}
-
-\ktm{ The data reduction comes at the expense of spatial information of any of the individual points.
-After the binned data is aggregated, there is no way to recover the original spatial locations.
-The loss of information incurred from binning will be explored in following sections.
-}
+The second stage of binning requires a frequency breakdown of the number of observations associated with each bin center, forming data triples, $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+Table~\ref{DataReductionExample} makes use of a small set of simulated data to show the progression from the original data (a), to the binned data (b), to the reduced binned data (c).
+The reduced binned data is sufficient for constructing the binned scatterplot. %, as only bin centers and frequencies are necessary.  
+Binning greatly reduces the storage size for the information and the computation time needed to construct the plot.
+\hh{XXX this is not always true - when the bins have only 1s the binned data set is actually bigger than the original. }
+
+Data reduction comes at the expense of spatial information of any of the individual points.
+After aggregation the original spatial locations cannot be recovered. The loss of information incurred from binning will be explored in following sections.
 
@@
 \end{figure}
+
+\hh{XXX do not use symbols in the text; write out = as 'equal to'}
 In the first set of binned data, the bins and the points on the scatterplot are nearly identical, but the scatterplot contains information about overlapping points.
@@
 \begin{itemize}
-\item {\it Spatial Loss}, $L^\text{S}$, occurs when points ($x_i$,$y_i$) for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers ($x_\ell^\ast$,$y_\ell^\ast$) for bins $\ell \in \{1,\dots,L\}$.
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,L\}$.
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
@@
 %
-where $L_i^S$ is the loss in the  $i^{th}$ observation. 
-%
-Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning.
+where $L_i^S$ is the loss in the  $i$th observation. 
+%
+Figure~\ref{fig:spatlossdemo2} \hh{XXX in the figure, make the tiles at the bottom squares, so they match with the scatter plot above.
+This is a neat picture!
+XXX} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning.
 Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively.

### 2015-03-25 Paper.Rnw
continuous scaling for color bar

@@
 It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}">>=
-J= 1
+#J= 1
 J2 = 200
 bound = 1000
-id <- seq(0+J/2,bound-J/2,by=J )
-fill <- seq(0+J/2,bound-J/2,by=J )
+#id <- seq(0+J/2,bound-J/2,by=J )
+#fill <- seq(0+J/2,bound-J/2,by=J )
 id2 <- seq(0+J2/2,bound-J2/2, by=J2)
 fill2 <- seq(J2/2,bound-(J2/2), by=J2)
-p1 <- qplot(id, 2, geom="tile", fill=-fill) +
+
+low = "#132B43"
+high = "#56B1F7"
+library(scales)
+blueGradient <- matrix(seq_gradient_pal(high, low, "Lab")(0:4400/4400), byrow=TRUE,
+       nrow=1000, ncol=4401)
+g <- rasterGrob(blueGradient, interpolate=TRUE)
+
+p1 <- qplot(c(0,1000), 2, geom="blank") +   # just to set up the dimensions
+  annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +  
   theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
@@
         panel.background=element_blank()
-        ) + 
-  ylab("Continuous") + xlab("")
+        ) + scale_x_continuous("", breaks=c(0,250,500,750,1000), labels=c(0,250,500,750,1000)) + 
+  ylab("Continuous") 
 p2 <- qplot(id2, 2, geom="tile", fill=-fill2) + 

### 2015-03-29 Paper.Rnw
Unneeded ")" removed
Small edits made

@@
 
-\hh{XXX do not use symbols in the text; write out = as 'equal to'}
 In the first set of binned data, the bins and the points on the scatterplot are nearly identical, but the scatterplot contains information about overlapping points.
-The second and third sets of binned data, with bin width=0.25 and 0.50 respectively, show higher-level summaries of the data but which may also provide more visually accessible information about the shape of the two-dimensional density between $x$ and $y$.
-The fourth set of binned data with bin width = 1.0 is nearly unrecognizable, because the bins are large enough that they provide very little additional information.
+The second and third sets of binned data, with bin width equal to 0.25 and 0.50 respectively, show higher-level summaries of the data but which may also provide more visually accessible information about the shape of the two-dimensional density between $x$ and $y$.
+The fourth set of binned data with bin width equal to 1.0 is nearly unrecognizable, because the bins are large enough that they provide very little additional information.
 
@@
 Frequency binning done using standard rectangular binning with origin $\\beta_0 = 0$ and binwidth $\\omega_X = 200$.
-\\ktm{Can't figure out how to get rid of the ``washed-out'' look in the continuous color scale figure.
-It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}">>=
+">>=
 #J= 1
@@
 $ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
-$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
-$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha ) $ \\
+$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha  $ \\
+$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha  $ \\
 $ \Rightarrow -2\theta = -\alpha $ \\

### 2015-03-29 BinningLossFunctions.R
Change center of bin for maximum potential loss to (max-min)/2 instead of mean so that it represents the use of the smallest possible tile that covers the data

@@ -107,8 +107,8 @@ RectBin2d <- function(xs,ys, originx, originy, widthx, widthy, type="standard"){
   outdat <- ddply(tempdat, .(binxs,binys), summarise,
                   binfreq = length(xs),
                   binspatialloss = sum(sqrt((xs-binxs)^2+(ys-binys)^2)) )
-  #centerofdata <- c((max(xs)-min(xs))/2,(max(ys)-min(ys))/2)
-  centerofdata <- c(mean(xs),mean(ys))
+  centerofdata <- c((max(xs)-min(xs))/2,(max(ys)-min(ys))/2)
+  #centerofdata <- c(mean(xs),mean(ys))
   summarydata <- data.frame( originx = originx, originy = originy, 
                              widthx = widthx, widthy = widthy,
                              totalSpatialLoss = sum(outdat$binspatialloss) ,

### 2015-03-29 Paper.Rnw
Discussion on proportion of potential loss under single bin added to section 4.1.  I don't love the notation, but there is a lot of overlapping things that it is hard to get a coherent labeling theme to work well.

@@
 
-\ktm{ The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}.
+\ktm{ The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}, $L^S_{net}$.
 The net spatial loss for standard binned data will always be equivalent to the total spatial loss due to the deterministic bin allocations.
 For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then finding the total spatial loss from this processed binned data.
+}
+
+\ktm{ The net spatial loss is perhaps more interpretable if it is viewed in comparison with the spatial loss from a ``worst-case scenario''.
+The largest displacement of points will occur if we place all points into a single bin.
+While the total spatial loss for single bin could be made arbitrarily large if the bin center is placed far away from the data values, we will consider a reasonable specification of a bin with boundaries that span only the range of the data in each dimension; thus centered at  $( (x_{(1)}-x_{(n)})/2$ , $(y_{(1)}-y_{(n)})/2 )$.
+The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as }
+\begin{eqnarray}\label{potspatloss}
+L^S_p = \sum_{i=1}^{n} \sqrt{ \left(x_i-(x_{(1)}-x_{(n)})/2\right)^2 + \left(y_i-(y_{(1)}-y_{(n)})/2\right)^2 }
+\end{eqnarray} 
+%
+\ktm{ Thus the net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$, will be used as the primary metric for assessing spatial loss incurred from binning.}
+
+%\ktm{ We recognize that the spatial loss is inherently influenced by the scale and units of the data.
+Standardizing the data pairs with respect to each axis will ensure that one of the dimensions does not have an inordinate effect on the spatial loss value.
 }

### 2015-03-29 Paper.Rnw
Add potential freq loss to section 4.2

@@
 
-\ktm{ The net spatial loss is perhaps more interpretable if it is viewed in comparison with the spatial loss from a ``worst-case scenario''.
+\ktm{ The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable.
+We will use the spatial loss from a ``worst-case'' binning scenario as a reference value.
 The largest displacement of points will occur if we place all points into a single bin.
@@
 %
-\ktm{ Thus the net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$, will be used as the primary metric for assessing spatial loss incurred from binning.}
+\ktm{ Thus the net spatial loss, as a proportion of potential spatial loss, $L^S_{net}/L^S_p$, will be used as the primary metric for assessing spatial loss incurred from binning.}
 
@@
 
+\ktm{ As with spatial binning, there will be a ``worst-case'' scenario for frequency binning that will provide context to the magnitude of the total frequency loss.
+If we elect to use a single frequency bin then all tiles of the binned scatterplot will be rendered with the same shade; thus the shade would represent the average bin density, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.
+We define the total frequency loss in the single shade scenario as the \textit{protential frequency loss}, $L^F_p$,  where
+%
+\begin{eqnarray}\label{potfreqloss}
+L^F_p = \sum_{\ell=1}^{L} (c_\ell-\bar{c})^2.
+\end{eqnarray}  
+%
+Thus total frequency loss after binning will be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.
+Similarly we define the \textit{protential log frequency loss}, $L^{\log F}_p$, as
+%
+\begin{eqnarray}\label{potlogfreqloss}
+L^{\log F}_p = \sum_{\ell=1}^{L} (\log(c_\ell+1)-\log(\bar{c}+1))^2,
+\end{eqnarray}  
+% 
+which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted. } %end ktm{} red
+
 % \subsection{Old Loss Definitions (to be removed)}

### 2015-03-29 BinningLossFunctions.R

@@ -128,7 +128,7 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
   # add frequency bin centers and bin labels of the form "(a,b]" to binout
   # binning depends on type / log counts
   if (logCount)  cs <- log(binout[[1]]$binfreq+1)
-    if(binType=="standard"){
+  if(binType=="standard"){
       width <- ceiling((max(cs)-min(cs))/ncolor)
       binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),2)
       binout[[1]]$freqlabel <- paste("(",round(binout[[1]]$freqgroup - width/2,2),
@@ -137,7 +137,7 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
       closeidx <- binout[[1]]$freqlabel == min(binout[[1]]$freqlabel)
       binout[[1]]$freqlabel[closeidx] <- paste("[",round(min(binout[[1]]$freqgroup) - width/2,2),
                                      ",",round(min(binout[[1]]$freqgroup) + width/2,2),"]",sep="")
-    } 
+  } 
   if(binType=="quantile"){
     binout[[1]]$freqgroup <- round(QuantBin1d(cs, ncolor),2)
     quantbounds <- unique(quantile(cs, (0:ncolor)/ncolor))

### 2015-03-29 Paper.Rnw
Complete intro to section 5 with scatterplot of simulated data

@@
 The parameters were selected to have data values roughly span the region $[0,100]^2$.
-}
+The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
+}
+
+<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=6, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.">>=
+### Try 2d binning with MVN data
+n=100000
+set.seed(31415)
+unifdat <- data.frame(x= runif(n, min=0, max=100),
+                      y= runif(n, min=0, max=100),
+                      type="Uniform",
+                      record = "Fine")
+normdat <- data.frame(x= rnorm(n, mean=50, sd=11),
+                     y= rnorm(n, mean=50, sd=11),
+                     type="Normal",
+                     record = "Fine")
+expdat <- data.frame(x= rexp(n, .11),
+                     y= rexp(n, .11),
+                     type="Exponential",
+                     record = "Fine")
+datlist <- list(unifdat,normdat,expdat)
+### round all data to resolution alpha
+alpha <- 2
+for(i in 1:3){
+  datlist[[i]]$roundx <- ((datlist[[i]]$x + 1)%/%alpha)*alpha
+  datlist[[i]]$roundy <- ((datlist[[i]]$y + 1)%/%alpha)*alpha
+}
+
+# plot the original data
+# make rounded data find overlapping points to make plot rendering faster
+dat <- rbind(datlist[[1]],datlist[[2]],datlist[[3]])
+smallround <- ddply(dat,.(type,record,roundx,roundy),summarise,
+      overlap = length(roundx))
+smallround <-cbind(data.frame(x=smallround$roundx,y=smallround$roundy), 
+                   smallround[,1:4])
+smallround$record <- "Coarse"
+alldat <- rbind(dat,smallround)
+
+qplot(x,y,geom="point",data=alldat,facets=record~type, size=I(1)) + 
+  theme_bw() + 
+  #ggtitle("Scatterplots of Simulated Data") +
+  xlab("X") + ylab("Y") + 
+  scale_x_continuous(breaks=c(0,25,50,75,100)) + 
+  scale_y_continuous(breaks=c(0,25,50,75,100))
+@
 
@@
 To observe binning loss under more realistic conditions, three corresponding data sets were also created by rounding the values from the original simulated data sets to the nearest even number.
+This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
 By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.
-}
-
-\ktm{
-\begin{itemize}
-\item Binning and frequency binning involve choices (standard/random/quantile, origin, binwidth, number of colors) want to make optimal choices given goals of binning.
-\item 3 simulated data sets: from bivariate Exponential, Normal, Uniform
-\item Since most data recorded to rough resolution we also assess loss when each dataset is rounded to nearest even value.  
-\end{itemize}
 }

### 2015-03-30 Paper.Rnw
Comment out old materials from frequency loss and beginning of new section 5.1

@@
 % 	\end{figure}
-
-\subsection{Loss of Frequency Information (old writeup - to be reworked or removed)} 
-Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently.
+% 
+% \subsection{Loss of Frequency Information (old writeup - to be reworked or removed)} 
+% Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently.
 By employing a color scheme for filling tiles according to the observed frequencies corresponds to a binning of the frequencies into a few different levels, with losses stemming from the same source as the previously discussed numerical loss.
-We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_k\}$ with $1 \le k \le n_C$.
+% We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_k\}$ with $1 \le k \le n_C$.
 An observed frequency $c_j$ with $1 \le j \le N_x \times N_y$, where $N_x$ and $N_y$ are the number of bin divisions in the X and Y directions respectively,  is assigned to the closest frequency level $c^\ast_k = b(c_j)$.
-The \textit{unscaled} frequency loss is then defined as
-\begin{eqnarray}\label{eqn:frequency}
-L_\text{Freq} = \sum_{j=1}^{N_x N_y} \left(c_j - b(c_j)\right)^2
-\end{eqnarray}
-
-% When we consider loss of frequency information, we must consider both dimensions simultaneously, as frequency information is based on the combination of binning in $x$ and  $y$ direction.
+% The \textit{unscaled} frequency loss is then defined as
+% \begin{eqnarray}\label{eqn:frequency}
+% L_\text{Freq} = \sum_{j=1}^{N_x N_y} \left(c_j - b(c_j)\right)^2
+% \end{eqnarray}
+% 
+% % When we consider loss of frequency information, we must consider both dimensions simultaneously, as frequency information is based on the combination of binning in $x$ and  $y$ direction.
 We define loss of frequency information as: \hh{what's that in words?}
-% $$ L_\text{Freq} = \sum_j \left(c_j - \frac{1}{n_X n_Y}\sum_j c_j\right)^2,$$ 
-% where $n_X, n_Y$ are the number of bins in $x$ and $y$, respectively.
-
-The loss of frequency information due to binning results from tiling over larger areas, thus results from {\it smoothing} the frequency variation (see Figure~\ref{fig:stdBinTutorial2}).
-
-Frequency data consists of counts, which most commonly exhibit skew densities, i.e.
+% % $$ L_\text{Freq} = \sum_j \left(c_j - \frac{1}{n_X n_Y}\sum_j c_j\right)^2,$$ 
+% % where $n_X, n_Y$ are the number of bins in $x$ and $y$, respectively. 
+% 
+% The loss of frequency information due to binning results from tiling over larger areas, thus results from {\it smoothing} the frequency variation (see Figure~\ref{fig:stdBinTutorial2}). 
+% 
+% Frequency data consists of counts, which most commonly exhibit skew densities, i.e.
 there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
@@
 This suggests that frequency loss be calculated as
-%
-\begin{eqnarray}\label{eqn:logfrequency}
-L_{\log \text{Freq}} = \sum_{j=1}^{N_x N_y} \left(\log(c_j+1) - \log(b(c_j)+1)\right)^2
-\end{eqnarray}
-%
-Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.
-
-This numerical assessment of frequency loss does not account for limitations in human perceptual ability.
+% %
+% \begin{eqnarray}\label{eqn:logfrequency}
+% L_{\log \text{Freq}} = \sum_{j=1}^{N_x N_y} \left(\log(c_j+1) - \log(b(c_j)+1)\right)^2
+% \end{eqnarray}
+% %
+% Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells. 
+% 
+% This numerical assessment of frequency loss does not account for limitations in human perceptual ability.
 Research suggests that under optimal conditions, we can effectively compare about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
-As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.
-
-To examine this further, we simulated points from three different distributions: Uniform, Normal, and Exponential to determine under what circumstances log frequency scaling is performing better than linear frequency scaling.
+% As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint. 
+% 
+% To examine this further, we simulated points from three different distributions: Uniform, Normal, and Exponential to determine under what circumstances log frequency scaling is performing better than linear frequency scaling.
 Figure~\ref{fig:logcolor} gives an overview of this simulation.
 In the top row binned scatterplots with a color scheme based on linear frequency scaling is used, the bottom row shows the same data based on log frequency scaled colors.
-
-For uniformly distributed data (right column) the choice of color scheme does not have a huge impact, but under both the normal and exponential distribution, log frequency scaling results in categories which account for more uniform proportions of the data (as can be seen by the color strips in the second and third row of Figure~\ref{fig:logcolor}).
+% 
+% For uniformly distributed data (right column) the choice of color scheme does not have a huge impact, but under both the normal and exponential distribution, log frequency scaling results in categories which account for more uniform proportions of the data (as can be seen by the color strips in the second and third row of Figure~\ref{fig:logcolor}).
 This is desirable because it provides more discernible frequency information.
-\begin{figure}[hbtp]
-\centering
-<<FrequencyCategories,echo=FALSE, results='hide', cache=TRUE>>=
-options(stringsAsFactors=FALSE)
-library(plyr)
-library(ggplot2)
-library(RColorBrewer)
-set.seed(5032)
-data <- rbind(data.frame(x=runif(1000, 0, 20), y=runif(1000, 0, 20), 
-                         Freq=1, Distribution="Uniform"), 
-              data.frame(x=rnorm(1000, 10, 3), y=rnorm(1000, 10, 3), 
-                         Freq=1, Distribution="Normal"),
-              data.frame(x=rexp(1000, .3), y=rexp(1000, .3), 
-                         Freq=1, Distribution="Exponential"))
-data <- data[which(abs(data$x-10)<10 & abs(data$y-10)<10),]
-# qplot(data=data, x=x, y=y, geom="point")+facet_wrap(~Distribution)
-
-data.bin <- data
-data.bin[,1:3] <- round(data[,1:3])
-data.bin <- ddply(data.bin, .(Distribution, x, y), summarise, Freq=sum(Freq))
-data.bin <- data.bin[order(data.bin$Distribution, data.bin$Freq),]
-# 
-# 
-# qplot(data=data.bin, x=Freq,geom="blank") + geom_density(aes(y=..scaled..),adjust=2) + facet_wrap(~Distribution, scales="free") + ylab("Density")
-# 
-# qplot(data=data.bin, x=Freq, geom="blank") + geom_density(aes(y=..scaled..), adjust=2) + facet_wrap(~Distribution,scales="free") + scale_x_continuous(trans="log", breaks=c(1, 5, 10, 20, 40, 60)) + ylab("Density")
-#
-
-get.upper <- function(x){
-  num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][2])
-  num <- sapply(num, function(i) substr(i, 1, nchar(i)-1))
-  return(as.numeric(num))
-}
-get.lower <- function(x){
-  num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][1])
-  num <- sapply(num, function(i) substr(i, 2, nchar(i)))
-  return(as.numeric(num))
-}
-
-labelsToValues <- function(x){
-  return(data.frame(lower=get.lower(x), upper=get.upper(x)))
-}
-
-linearbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(x$Freq, 7)), label=as.character(cut(x$Freq,7)), id=1:nrow(x))))[,2:4])
-linearbins$fill <- factor(linearbins$fill)
-
-x <- subset(linearbins, Distribution=="Uniform" & fill==1)
-linearbins2 <- ddply(linearbins, .(Distribution, fill),
-                     function(x){
-                       temp <- labelsToValues(x$label[1])
-                       return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
-                     })
-linearbins2 <- ddply(linearbins2, .(Distribution), transform, height=height/sum(height))
-
-logbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(log(x$Freq), 7)), label=as.character(cut(log(x$Freq),7)), id=1:nrow(x))))[,2:4])
-logbins$fill <- factor(logbins$fill)
-
-x <- subset(logbins, Distribution=="Uniform" & fill==1)
-logbins2 <- ddply(logbins, .(Distribution, fill),
-                  function(x){
-                    temp <- exp(labelsToValues(x$label[1]))
-                    newlabel <- paste(c("(", round(temp[1], 3), ",", round(temp[2], 3), "]"), sep="", collapse="")
-                    return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
-                  })
-
-logbins2 <- ddply(logbins2, .(Distribution), transform, height=height/sum(height))
-
-colors <- brewer.pal(9, "Blues")[-c(1:2)]
-
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, heights=NULL) {
-  require(grid)
-
-  # Make a list from the ... arguments and plotlist
-  plots <- c(list(...), plotlist)
-
-  numPlots = length(plots)
-
-  # If layout is NULL, then use 'cols' to determine layout
-  if (is.null(layout)) {
-    # Make the panel
-    # ncol: Number of columns of plots
-    # nrow: Number of rows needed, calculated from # of cols
-    nrow = ceiling(numPlots/cols)
-
-    if (is.null(heights)) heights = unit(rep(1, numPlots), "null")
-    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                     ncol = cols, nrow = nrow)
-  }
-
-  if (numPlots==1) {
-    print(plots[[1]])
-
-  } else {
-    # Set up the page
-    grid.newpage()
-    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout), heights=heights)))
-
-    # Make each plot, in the correct location
-    for (i in 1:numPlots) {
-      # Get the i,j matrix positions of the regions that contain this subplot
-      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-
-      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                      layout.pos.col = matchidx$col))
-    }
-  }
-}
-
-p1 <- qplot(data=linearbins, x=x, y=y, geom="tile", fill=fill)+facet_wrap(~Distribution) +
-  scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
-  ggtitle("(a) Joint Density: Linear Binning") +  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-
-X <- data.frame(fill=with(linearbins, rep(fill, Freq)), Distribution = with(linearbins, rep(Distribution, Freq)))
-X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
-p2 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(1)) + 
-  scale_fill_manual(values=colors, name="Frequency", guide="none") + 
-  facet_grid(.~Distribution, scales="free") +
-  xlab("Count") + theme_bw() + ylab("") +
-  ggtitle("(b) Color Frequency: Linear Binning") + 
-  theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
-  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
-# linear binning
-pdf(file="p2.pdf", width=8, height=3)
-p2
-dev.off()
-
-# p2 <- qplot(data=linearbins2, xmin=lower, xmax=upper, ymin=0, ymax=height*1000, fill=fill, geom="blank", asp=1.5) + 
-#   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
-#   scale_fill_manual("Tile Color", values=colors)+
-#   facet_wrap(~Distribution, scales="free") + 
-#   theme_bw() + scale_x_continuous("Bin Frequency") + 
-#   scale_y_continuous("") + ggtitle("(b) Distribution of Frequency: Linear Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-
-# p4 <- qplot(data=logbins2, xmin=lower, xmax=upper, ymin=0, ymax=1000*height, fill=fill,  geom="blank", asp=1.5) + 
-#   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
-#   scale_fill_manual("Tile Color", values=colors)+
-#   facet_wrap(~Distribution, scales="free") + 
-#   theme_bw() + scale_x_continuous("Bin Frequency") + 
-#   scale_y_continuous("") + ggtitle("(c) Distribution of Frequency: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-
-X <- data.frame(fill=with(logbins, rep(fill, Freq)), Distribution = with(logbins, rep(Distribution, Freq)))
-X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
-p4 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(NULL)) + 
-  scale_fill_manual(values=colors, name="Frequency", guide="none") + 
-  facet_grid(.~Distribution, scales="free") +
-  xlab("Count") + theme_bw() + ylab("") +
-  ggtitle("(c) Color Frequency: Log Binning") + 
-  theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
-  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
-
-p3 <- qplot(data=logbins, x=x, y=y, geom="tile", fill=fill, asp=1.0)+facet_wrap(~Distribution) +
-  scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
-  ggtitle("(d) Joint Density: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-
-multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
-#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
-@
-%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure.pdf}
-\caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
-\end{figure}
-
-Frequency loss is much more difficult to normalize, as we must balance the partial loss of frequency information due to over-plotting with the total loss of frequency information that results if we use only a single bin to categorize the data.
-For the purposes of numerical assessment of frequency loss, we calculate the baseline frequency loss as the loss corresponding to a minimal bin width given by the smallest non-zero absolute difference of successive values in the original data set, as shown in Figure~\ref{binning}.
+% \begin{figure}[hbtp]
+% \centering
+% <<FrequencyCategories,echo=FALSE, results='hide', cache=TRUE>>=
+% options(stringsAsFactors=FALSE)
+% library(plyr)
+% library(ggplot2)
+% library(RColorBrewer)
+% set.seed(5032)
+% data <- rbind(data.frame(x=runif(1000, 0, 20), y=runif(1000, 0, 20), 
+%                          Freq=1, Distribution="Uniform"), 
+%               data.frame(x=rnorm(1000, 10, 3), y=rnorm(1000, 10, 3), 
+%                          Freq=1, Distribution="Normal"),
+%               data.frame(x=rexp(1000, .3), y=rexp(1000, .3), 
+%                          Freq=1, Distribution="Exponential"))
+% data <- data[which(abs(data$x-10)<10 & abs(data$y-10)<10),]
+% # qplot(data=data, x=x, y=y, geom="point")+facet_wrap(~Distribution)
+% 
+% data.bin <- data
+% data.bin[,1:3] <- round(data[,1:3])
+% data.bin <- ddply(data.bin, .(Distribution, x, y), summarise, Freq=sum(Freq))
+% data.bin <- data.bin[order(data.bin$Distribution, data.bin$Freq),]
+% # 
+% # 
+% # qplot(data=data.bin, x=Freq,geom="blank") + geom_density(aes(y=..scaled..),adjust=2) + facet_wrap(~Distribution, scales="free") + ylab("Density")
+% # 
+% # qplot(data=data.bin, x=Freq, geom="blank") + geom_density(aes(y=..scaled..), adjust=2) + facet_wrap(~Distribution,scales="free") + scale_x_continuous(trans="log", breaks=c(1, 5, 10, 20, 40, 60)) + ylab("Density")
+% # 
+% 
+% get.upper <- function(x){
+%   num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][2])
+%   num <- sapply(num, function(i) substr(i, 1, nchar(i)-1))
+%   return(as.numeric(num))
+% }
+% get.lower <- function(x){
+%   num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][1])
+%   num <- sapply(num, function(i) substr(i, 2, nchar(i)))
+%   return(as.numeric(num))
+% }
+% 
+% labelsToValues <- function(x){
+%   return(data.frame(lower=get.lower(x), upper=get.upper(x)))
+% }
+% 
+% 
+% linearbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(x$Freq, 7)), label=as.character(cut(x$Freq,7)), id=1:nrow(x))))[,2:4])
+% linearbins$fill <- factor(linearbins$fill)
+% 
+% x <- subset(linearbins, Distribution=="Uniform" & fill==1)
+% linearbins2 <- ddply(linearbins, .(Distribution, fill),
+%                      function(x){
+%                        temp <- labelsToValues(x$label[1])
+%                        return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
+%                      })
+% linearbins2 <- ddply(linearbins2, .(Distribution), transform, height=height/sum(height))
+% 
+% 
+% 
+% logbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(log(x$Freq), 7)), label=as.character(cut(log(x$Freq),7)), id=1:nrow(x))))[,2:4])
+% logbins$fill <- factor(logbins$fill)
+% 
+% x <- subset(logbins, Distribution=="Uniform" & fill==1)
+% logbins2 <- ddply(logbins, .(Distribution, fill),
+%                   function(x){
+%                     temp <- exp(labelsToValues(x$label[1]))
+%                     newlabel <- paste(c("(", round(temp[1], 3), ",", round(temp[2], 3), "]"), sep="", collapse="")
+%                     return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
+%                   })
+% 
+% logbins2 <- ddply(logbins2, .(Distribution), transform, height=height/sum(height))
+% 
+% colors <- brewer.pal(9, "Blues")[-c(1:2)]
+% 
+% multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, heights=NULL) {
+%   require(grid)
+%   
+%   # Make a list from the ... arguments and plotlist
+%   plots <- c(list(...), plotlist)
+%   
+%   numPlots = length(plots)
+%   
+%   # If layout is NULL, then use 'cols' to determine layout
+%   if (is.null(layout)) {
+%     # Make the panel
+%     # ncol: Number of columns of plots
+%     # nrow: Number of rows needed, calculated from # of cols
+%     nrow = ceiling(numPlots/cols)
+%     
+%     if (is.null(heights)) heights = unit(rep(1, numPlots), "null")
+%     layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+%                      ncol = cols, nrow = nrow)
+%   }
+%   
+%   if (numPlots==1) {
+%     print(plots[[1]])
+%     
+%   } else {
+%     # Set up the page
+%     grid.newpage()
+%     pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout), heights=heights)))
+%     
+%     # Make each plot, in the correct location
+%     for (i in 1:numPlots) {
+%       # Get the i,j matrix positions of the regions that contain this subplot
+%       matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+%       
+%       print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+%                                       layout.pos.col = matchidx$col))
+%     }
+%   }
+% }
+% 
+% p1 <- qplot(data=linearbins, x=x, y=y, geom="tile", fill=fill)+facet_wrap(~Distribution) +
+%   scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
+%   ggtitle("(a) Joint Density: Linear Binning") +  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
+% 
+% X <- data.frame(fill=with(linearbins, rep(fill, Freq)), Distribution = with(linearbins, rep(Distribution, Freq)))
+% X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
+% p2 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(1)) + 
+%   scale_fill_manual(values=colors, name="Frequency", guide="none") + 
+%   facet_grid(.~Distribution, scales="free") +
+%   xlab("Count") + theme_bw() + ylab("") +
+%   ggtitle("(b) Color Frequency: Linear Binning") + 
+%   theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
+%   theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
+% # linear binning
+% pdf(file="p2.pdf", width=8, height=3)
+% p2
+% dev.off()
+% 
+% # p2 <- qplot(data=linearbins2, xmin=lower, xmax=upper, ymin=0, ymax=height*1000, fill=fill, geom="blank", asp=1.5) + 
+% #   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
+% #   scale_fill_manual("Tile Color", values=colors)+
+% #   facet_wrap(~Distribution, scales="free") + 
+% #   theme_bw() + scale_x_continuous("Bin Frequency") + 
+% #   scale_y_continuous("") + ggtitle("(b) Distribution of Frequency: Linear Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
+% 
+% # p4 <- qplot(data=logbins2, xmin=lower, xmax=upper, ymin=0, ymax=1000*height, fill=fill,  geom="blank", asp=1.5) + 
+% #   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
+% #   scale_fill_manual("Tile Color", values=colors)+
+% #   facet_wrap(~Distribution, scales="free") + 
+% #   theme_bw() + scale_x_continuous("Bin Frequency") + 
+% #   scale_y_continuous("") + ggtitle("(c) Distribution of Frequency: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
+% 
+% X <- data.frame(fill=with(logbins, rep(fill, Freq)), Distribution = with(logbins, rep(Distribution, Freq)))
+% X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
+% p4 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(NULL)) + 
+%   scale_fill_manual(values=colors, name="Frequency", guide="none") + 
+%   facet_grid(.~Distribution, scales="free") +
+%   xlab("Count") + theme_bw() + ylab("") +
+%   ggtitle("(c) Color Frequency: Log Binning") + 
+%   theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
+%   theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
+% 
+% 
+% p3 <- qplot(data=logbins, x=x, y=y, geom="tile", fill=fill, asp=1.0)+facet_wrap(~Distribution) +
+%   scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
+%   ggtitle("(d) Joint Density: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
+% 
+% multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
+% #ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
+% @
+% %\includegraphics[width=\textwidth]{figure/MultipleDensityFigure.pdf}
+% \caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
+% \end{figure}
+% 
+% Frequency loss is much more difficult to normalize, as we must balance the partial loss of frequency information due to over-plotting with the total loss of frequency information that results if we use only a single bin to categorize the data.
+% For the purposes of numerical assessment of frequency loss, we calculate the baseline frequency loss as the loss corresponding to a minimal bin width given by the smallest non-zero absolute difference of successive values in the original data set, as shown in Figure~\ref{binning}.
 
@@
 
-\subsection{Spatial binning choices}
+\subsection{Binning Specification and Spatial Loss}
+
+To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning.
+If equivalent bin widths are used, the net spatial loss under standard binning will always be less than or equal to the net spatial loss under random binning.
+This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.
+
+Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins ranging from 2 units to 20 units.
+It is clear to see that the spatial loss for random binning is always higher than for standard binning and that the loss grows as binwidth increases.
+Also, the proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.
+
+Another attribute that stands out in Figure~\ref{fig:SpatialLossCompare} is that proportion of potential spatial loss is much lower for the exponential data than for the normal or uniform data; this is because the potential spatial loss is much higher for skewed data than symmetric.
+This is not surprising when we consider that if a single rectangular tile were placed over the bivariate range of the exponential data, the center of the tile would be a fair distance away from the center of mass for the points.
+The plot is reflecting that the ``worst-case'' binning scenario for skewed data is worse than for symmetric data.
+
+<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for proportion of potential spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
+spatlossdat <- read.csv("SpatialLossData.csv",header=TRUE)
+# use all the square bins to plot bin width against proportion of max spataial loss
+spatlossdatSquareBins <- spatlossdat[spatlossdat$widthx==spatlossdat$widthy,]
+qplot(widthx,propSpatialLoss,geom="path",colour=bintype, linetype=bintype,
+      data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
+  theme_bw() + ylab("Prop. of Potential Spatial Loss") +
+  xlab("Width/Height of Square Bins")
+@
 

### 2015-03-30 Paper.Rnw
more work on section 5.1

@@
 
-\subsection{Binning Specification and Spatial Loss}
-
+\subsection{Rectangular Binning Specifications and Spatial Loss}
+
+For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension.
 To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning.
@@
 
-Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins ranging from 2 units to 20 units.
+Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
 It is clear to see that the spatial loss for random binning is always higher than for standard binning and that the loss grows as binwidth increases.
-Also, the proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.
+The proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.
 
@@
 @
+
+All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin.
+Smaller binwidths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats.
+With smaller bins comes more computation requirements; a non-negligable consideration in settings with truly massive data.
+Additionally, the binwidth should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.
+
+Consider the simulated uniform data, we have the fine version and the coarse version, which rounded to nearest even integer.
+Figure~\ref{fig:strippedplots} displays the binned scatterplots under several scenarios.
 

### 2015-03-31 Paper.Rnw
read-through

@@
 
-\ktm{ For random assignment the total spatial loss can be calculated by simply replacing $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.
-The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
-Special consideration should be payed to the two points that with spatial losses highlighted in red; in standard binning these points are in neighboring bins and have been allocated to the opposite bin during random binning.
-We see that the binned scatterplot remains identical if these points are reallocated back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
-}
-
-\ktm{ The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}, $L^S_{net}$.
-The net spatial loss for standard binned data will always be equivalent to the total spatial loss due to the deterministic bin allocations.
+\ktm{ For random assignment the total spatial loss can be calculated by simply replacing  \hh{call these guys by name} $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.}  The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
+Special consideration should be paid to the two points highlighted in red in the panel for random binning: in standard binning these points are in neighboring bins but have been allocated to the opposite bin during random binning.
+We see that the binned scatterplot remains identical if these points are re-allocated back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
+
+ The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}, $L^S_{net}$.
+The net spatial loss for standard binned data is always  equivalent to the total spatial loss due to the deterministic bin allocations.
 For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then finding the total spatial loss from this processed binned data.
-}
-
-\ktm{ The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable.
-We will use the spatial loss from a ``worst-case'' binning scenario as a reference value.
-The largest displacement of points will occur if we place all points into a single bin.
-While the total spatial loss for single bin could be made arbitrarily large if the bin center is placed far away from the data values, we will consider a reasonable specification of a bin with boundaries that span only the range of the data in each dimension; thus centered at  $( (x_{(1)}-x_{(n)})/2$ , $(y_{(1)}-y_{(n)})/2 )$.
-The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as }
+
+The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable.
+We  use the spatial loss from a ``worst-case'' binning scenario as a reference value.
+The largest displacement of points  occurs if we place all points into a single bin.
+While the total spatial loss for single bin could be made arbitrarily large if the bin center is placed far away from the data values, we will consider a reasonable specification of a bin with boundaries that span only the range of the data in each dimension centered at  $( (x_{(1)}-x_{(n)})/2$ , $(y_{(1)}-y_{(n)})/2 )$.
+The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as
 \begin{eqnarray}\label{potspatloss}
@@
 %
-\ktm{ Thus the net spatial loss, as a proportion of potential spatial loss, $L^S_{net}/L^S_p$, will be used as the primary metric for assessing spatial loss incurred from binning.}
+We use net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$,  as the primary metric for assessing spatial loss incurred from binning.
 
@@
 
-\ktm{ As with spatial binning, there will be a ``worst-case'' scenario for frequency binning that will provide context to the magnitude of the total frequency loss.
-If we elect to use a single frequency bin then all tiles of the binned scatterplot will be rendered with the same shade; thus the shade would represent the average bin density, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.
-We define the total frequency loss in the single shade scenario as the \textit{protential frequency loss}, $L^F_p$,  where
+ Similar to spatial binning, \hh{we define} a worst-case scenario for frequency binning that provides context to the magnitude of the total frequency loss.
+If we use a single frequency bin then all tiles of the binned scatterplot are rendered with the same shade; thus the shade  represents the average bin density \hh{XXX this is different from the spatial loss - visually the shade would be half way between lowest and highest frequency, wouldn't it?}, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.
+We define the total frequency loss in the single shade scenario as the \textit{potential frequency loss}, $L^F_p$,  where
 %
@@
 %
-Thus total frequency loss after binning will be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.
-Similarly we define the \textit{protential log frequency loss}, $L^{\log F}_p$, as
+Thus total frequency loss after binning can be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.
+Similarly we define the \textit{potential log frequency loss}, $L^{\log F}_p$, as
 %
@@
 % 
-which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted. } %end ktm{} red
+which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted.
+\hh{XXX could you give a couple more details what you mean by cases where frequency binning was conducted?}
 
@@
 
-\ktm{ Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm and also the parameters of the chosen binning algorithm.
+Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with the  binning algorithm.
 This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
-Some results are derived optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.
-}
-
-\ktm{ Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
-Bivariate datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
+\ktm Some results are derived \hh{XXX check the sentence} optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.
+}
+
+Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
+Datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
 These distributions were selected for their variety in terms of shape, center and spread.
-100,000 observation pairs were simulated for each set of simulated data, and the x and y values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.
+100,000 observation pairs were simulated for each set of simulated data, and the $x$ and $y$ values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.
 The parameters were selected to have data values roughly span the region $[0,100]^2$.
 The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
-}
 

### 2015-03-31 Paper.Rnw
Add stripped plot to section 5.1 showing what happens when binwidth is not integer multiple of data resolution.

--- 2015-03-30 Paper.Rnw
@@
 Consider the simulated uniform data, we have the fine version and the coarse version, which rounded to nearest even integer.
-Figure~\ref{fig:strippedplots} displays the binned scatterplots under several scenarios.
+Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
+
+<<UniformStripes, echo=F,include=T,eval=T,fig.width=5, fig.height=5, out.width='.7\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned Scatterplots of Uniform Data with 5X5 Bins">>=
+### Binned Scatterplots for Uniform Data to Show stripes
+# create binned data for standard and random with binwidth 5
+binoutfine <- RectBin2d(datlist[[1]]$x,datlist[[1]]$y,0,0,5,5,type="standard")
+binoutcoarse <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")
+binoutfineRand <- RectBin2d(datlist[[1]]$x,datlist[[1]]$y,0,0,5,5,type="random")
+binoutcoarseRand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")
+
+allbindat <- rbind(binoutfine[[1]],binoutcoarse[[1]],
+                   binoutfineRand[[1]],binoutcoarseRand[[1]])
+allbindat$SimDataType <- factor(rep(c("Fine","Coarse","Fine","Coarse"), 
+                             each=nrow(binoutfine[[1]])),
+                             levels=c("Fine","Coarse"))
+allbindat$BinningType <- factor(rep(c("Standard","Random"),
+                             each=2*nrow(binoutfine[[1]])),
+                             levels=c("Standard","Random"))
+
+qplot(binxs, binys, geom="tile", fill=binfreq, data=allbindat) +
+  facet_grid(BinningType ~ SimDataType)+
+  ggtitle(" ") +  
+  scale_fill_gradient(low="#56B1F7", high="#132B43",
+                      limits=c(0, 450),name="Bin Count") + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1) +
+  xlab("X") + ylab("Y")
+@
 

### 2015-03-31 Paper.Rnw
more proofreading

@@
 This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
-\ktm Some results are derived \hh{XXX check the sentence} optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.
+\ktm{Some results are derived \hh{XXX check the sentence} optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.
 }
@@
 Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
+
+\hh{Just list the three different scenarios:
+\begin{itemize}
+\item {\bf Scenario I:} $x \sim \text{Exp}_\lambda, y \sim \text{Exp}_\lambda$ with $\lambda = 11$
+\end{itemize}}
 Datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
@@
 
-\ktm{ These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places.
+These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places.
 Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles.
-To observe binning loss under more realistic conditions, three corresponding data sets were also created by rounding the values from the original simulated data sets to the nearest even number.
+To observe binning loss under more realistic conditions, we created three  data sets by rounding the values from the originally simulated data sets to the nearest even number.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
 By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.
-}
 
@@
 For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension.
-To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning.
-If equivalent bin widths are used, the net spatial loss under standard binning will always be less than or equal to the net spatial loss under random binning.
+To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning: For equal bin widths, the net spatial loss under standard binning is always less than or equal to the net spatial loss under random binning.
 This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.
@@
 Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
-It is clear to see that the spatial loss for random binning is always higher than for standard binning and that the loss grows as binwidth increases.
-The proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.
+Obviously, the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
+The proportion of potential spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.
 
@@
 All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin.
-Smaller binwidths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats.
-With smaller bins comes more computation requirements; a non-negligable consideration in settings with truly massive data.
-Additionally, the binwidth should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.
+Smaller bin widths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats.
+With smaller bins comes more computation requirements; a non-negligible consideration in settings with truly massive data.
+Additionally, the bin width should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.
 
@@
 \begin{itemize}
-\item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytical proof)
+\item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytic proof)
 \item Origin: 
@@
     \item Doesn't much matter for fine resolution data but it does for coarse data.
-    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytical proof then demo) Proof in Appendix~\ref{proof:offset}.
+    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
   \end{itemize} 
-\item Binwidth: 
+\item Bin width: 
   \begin{itemize}
     \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
-    \item for fine data any binwidth is fine
-    \item for coarse data need binwidths $\omega \ge \alpha$ (show plot with whitespace gaps)
-    \item for coarse data binwidths should be scalar multiple of $\alpha$ (show plot with strips)
+    \item for fine data any bin width is fine
+    \item for coarse data need bin widths $\omega \ge \alpha$ (show plot with white space gaps)
+    \item for coarse data bin widths should be scalar multiple of $\alpha$ (show plot with strips)
     \item Does the distribution type matter? 
@@
 Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues.
-Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths.
+Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a bin width that is a non-integer multiple of the incremental widths.
 In this scenario, we see that the non-synchronous data increments and bin widths will cause bins to contain inconstant numbers of possible data value.
@@
 This is apparent in the order of colored lines based on the Strike Out binning size.
-This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as binwidth increases.
-Whereas the visual loss function sums loss over bin, thus as the binwidth increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.
+This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as bin width increases.
+Whereas the visual loss function sums loss over bin, thus as the bin width increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.
 
@@
 The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips.
-These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the binwidths are integer multiples of the data increments.
+These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the bin widths are integer multiples of the data increments.
 This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
 
-\textbf{Are we sure that this is because of an odd binwidths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
+\textbf{Are we sure that this is because of an odd bin widths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
 

### 2015-03-31 Paper.Rnw
Removing proportion of potential loss. Scaling by worst-case scenario was more of a headache than it was worth.  Comparisons of net spatial loss are sufficient.  Figure 8 much more interpretable now.

@@
 
-The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable.
+% The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable.
 We  use the spatial loss from a ``worst-case'' binning scenario as a reference value.
@@
 The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as
-\begin{eqnarray}\label{potspatloss}
-L^S_p = \sum_{i=1}^{n} \sqrt{ \left(x_i-(x_{(1)}-x_{(n)})/2\right)^2 + \left(y_i-(y_{(1)}-y_{(n)})/2\right)^2 }
-\end{eqnarray} 
-%
-We use net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$,  as the primary metric for assessing spatial loss incurred from binning.
+% \begin{eqnarray}\label{potspatloss}
+% L^S_p = \sum_{i=1}^{n} \sqrt{ \left(x_i-(x_{(1)}-x_{(n)})/2\right)^2 + \left(y_i-(y_{(1)}-y_{(n)})/2\right)^2 }
+% \end{eqnarray} 
+% %
+% We use net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$,  as the primary metric for assessing spatial loss incurred from binning.
 
@@
 
- Similar to spatial binning, \hh{we define} a worst-case scenario for frequency binning that provides context to the magnitude of the total frequency loss.
+%  Similar to spatial binning, \hh{we define} a worst-case scenario for frequency binning that provides context to the magnitude of the total frequency loss.
 If we use a single frequency bin then all tiles of the binned scatterplot are rendered with the same shade; thus the shade  represents the average bin density \hh{XXX this is different from the spatial loss - visually the shade would be half way between lowest and highest frequency, wouldn't it?}, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.
 We define the total frequency loss in the single shade scenario as the \textit{potential frequency loss}, $L^F_p$,  where
-%
-\begin{eqnarray}\label{potfreqloss}
-L^F_p = \sum_{\ell=1}^{L} (c_\ell-\bar{c})^2.
-\end{eqnarray}  
-%
-Thus total frequency loss after binning can be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.
+% %
+% \begin{eqnarray}\label{potfreqloss}
+% L^F_p = \sum_{\ell=1}^{L} (c_\ell-\bar{c})^2.
+% \end{eqnarray}  
+% %
+% Thus total frequency loss after binning can be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.
 Similarly we define the \textit{potential log frequency loss}, $L^{\log F}_p$, as
-%
-\begin{eqnarray}\label{potlogfreqloss}
-L^{\log F}_p = \sum_{\ell=1}^{L} (\log(c_\ell+1)-\log(\bar{c}+1))^2,
-\end{eqnarray}  
-% 
-which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted.
+% %
+% \begin{eqnarray}\label{potlogfreqloss}
+% L^{\log F}_p = \sum_{\ell=1}^{L} (\log(c_\ell+1)-\log(\bar{c}+1))^2,
+% \end{eqnarray}  
+% % 
+% which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted.
 \hh{XXX could you give a couple more details what you mean by cases where frequency binning was conducted?}
-
+% 
 % \subsection{Old Loss Definitions (to be removed)}
@@
 Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles.
-To observe binning loss under more realistic conditions, we created three  data sets by rounding the values from the originally simulated data sets to the nearest even number.
+The data resolution can be described by increment between data values.
+To observe binning loss under more realistic conditions, we created three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
@@
 
-Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
-Obviously, the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
-The proportion of potential spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.
+Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
+It is easily apparent that  the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
+The net spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.
 
@@
 
-<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for proportion of potential spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
+<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
 spatlossdat <- read.csv("SpatialLossData.csv",header=TRUE)
@@
 spatlossdatSquareBins <- spatlossdat[spatlossdat$widthx==spatlossdat$widthy,]
-qplot(widthx,propSpatialLoss,geom="path",colour=bintype, linetype=bintype,
+qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
-  theme_bw() + ylab("Prop. of Potential Spatial Loss") +
+  theme_bw() + ylab("Net Spatial Loss") +
   xlab("Width/Height of Square Bins")

### 2015-04-01 Paper.Rnw
Attempt to work around tile-size faceting issue in plot to display stripes. Takes too long to compile data on the fly and large for pushing to repo. Go to "UniformStripes" code chunk, uncomment, construct/save data, then recomment.

@@
 
-Another attribute that stands out in Figure~\ref{fig:SpatialLossCompare} is that proportion of potential spatial loss is much lower for the exponential data than for the normal or uniform data; this is because the potential spatial loss is much higher for skewed data than symmetric.
-This is not surprising when we consider that if a single rectangular tile were placed over the bivariate range of the exponential data, the center of the tile would be a fair distance away from the center of mass for the points.
-The plot is reflecting that the ``worst-case'' binning scenario for skewed data is worse than for symmetric data.
-
 <<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
@@
 All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin.
-Smaller bin widths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats.
-With smaller bins comes more computation requirements; a non-negligible consideration in settings with truly massive data.
-Additionally, the bin width should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.
-
-Consider the simulated uniform data, we have the fine version and the coarse version, which rounded to nearest even integer.
+While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a non-negligible consideration in settings with truly massive data.
+
+While bin widths can be set to any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.
+If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.
+These white-space gaps are undesireable because they create artificial discontinuity that interfers with interpretation of bivariate density.
+If bin dimensions are not integer multiples of the bivariate data resolution there is also a distruptive visual consequence of ``stripping'', where bins have systematically different numbers of possible data values that can cause artifical density stripes to appear in the binned scatterplot.
+
+To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of 2 units.
 Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
-
-<<UniformStripes, echo=F,include=T,eval=T,fig.width=5, fig.height=5, out.width='.7\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned Scatterplots of Uniform Data with 5X5 Bins">>=
-### Binned Scatterplots for Uniform Data to Show stripes
-# create binned data for standard and random with binwidth 5
-binoutfine <- RectBin2d(datlist[[1]]$x,datlist[[1]]$y,0,0,5,5,type="standard")
-binoutcoarse <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")
-binoutfineRand <- RectBin2d(datlist[[1]]$x,datlist[[1]]$y,0,0,5,5,type="random")
-binoutcoarseRand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")
-
-allbindat <- rbind(binoutfine[[1]],binoutcoarse[[1]],
-                   binoutfineRand[[1]],binoutcoarseRand[[1]])
-allbindat$SimDataType <- factor(rep(c("Fine","Coarse","Fine","Coarse"), 
-                             each=nrow(binoutfine[[1]])),
-                             levels=c("Fine","Coarse"))
-allbindat$BinningType <- factor(rep(c("Standard","Random"),
-                             each=2*nrow(binoutfine[[1]])),
-                             levels=c("Standard","Random"))
-
-qplot(binxs, binys, geom="tile", fill=binfreq, data=allbindat) +
-  facet_grid(BinningType ~ SimDataType)+
-  ggtitle(" ") +  
+Under standard binning we see the white-space gaps with 1 X 1 unit bins, stripes with 5 X 5 unit bins and the correct even spread density with 4X4 unit bins.
+Note that here the random binning is effective at eliminating the misleading visual effects when non-integer multiples of the data resolution are selected.
+
+<<UniformStripes, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data with square bins of size 1, 4 and 5">>=
+# Commented out code for data generation: run once then recomment
+### Create data to demonstrate stripes from improper binwidths
+# binoutcoarse1 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="standard")[[1]]
+# binoutcoarse4 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="standard")[[1]]
+# binoutcoarse5 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")[[1]]
+# binoutcoarse1Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="random")[[1]]
+# binoutcoarse4Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="random")[[1]]
+# binoutcoarse5Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")[[1]]
+# 
+# binoutcoarse1$BinSize <- "1 X 1 Unit" ; binoutcoarse1$BinningType <- "Standard" 
+# binoutcoarse4$BinSize <- "4 X 4 Unit" ; binoutcoarse4$BinningType <- "Standard" 
+# binoutcoarse5$BinSize <- "5 X 5 Unit" ; binoutcoarse5$BinningType <- "Standard" 
+# binoutcoarse1Rand$BinSize <- "1 X 1 Unit" ; binoutcoarse1Rand$BinningType <- "Random" 
+# binoutcoarse4Rand$BinSize <- "4 X 4 Unit" ; binoutcoarse4Rand$BinningType <- "Random" 
+# binoutcoarse5Rand$BinSize <- "5 X 5 Unit" ; binoutcoarse5Rand$BinningType <- "Random" 
+# 
+# allbindat <- rbind(binoutcoarse1,binoutcoarse4,binoutcoarse5,
+#                    binoutcoarse1Rand,binoutcoarse4Rand,binoutcoarse5Rand)
+# write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
+allbindat <- read.csv("StripeBinData.csv",header=TRUE)
+allbindat$BinningType <- factor(allbindat$BinningType, levels=c("Standard","Random"))
+
+p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="1 X 1 Unit")) +
+  facet_grid(BinningType~BinSize)+
+  ggtitle(" ") +   xlab("") + ylab("") +
   scale_fill_gradient(low="#56B1F7", high="#132B43",
                       limits=c(0, 450),name="Bin Count") + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1) +
-  xlab("X") + ylab("Y")
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+
+p2 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="4 X 4 Unit")) +
+  facet_grid(BinningType~BinSize)+
+  ggtitle(" ") +   xlab("") + ylab("") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43",
+                      limits=c(0, 450),name="Bin Count") + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+
+p3 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="5 X 5 Unit")) +
+  facet_grid(BinningType~BinSize)+
+  ggtitle(" ") +   xlab("") + ylab("") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43",
+                      limits=c(0, 450),name="Bin Count") + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+
+grid.arrange(p1,p2,p3, nrow=1)
+
 @

### 2015-04-01 Paper.Rnw
Work on Section 5.1

@@
 \end{itemize}}
-Datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
+Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
 These distributions were selected for their variety in terms of shape, center and spread.
-100,000 observation pairs were simulated for each set of simulated data, and the $x$ and $y$ values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.
+100,000 observation pairs were simulated for each set of simulated data, and the $x$ and $y$ values were drawn independently from univariate distributions; Exponential(11), Normal(50,$11^2$) and Uniform(0,100), respectively.
 The parameters were selected to have data values roughly span the region $[0,100]^2$.
@@
 
-<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=6, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.">>=
+<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.">>=
 ### Try 2d binning with MVN data
@@
 
-These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places.
+These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places.
 Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles.
@@
 
-<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
+<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
 spatlossdat <- read.csv("SpatialLossData.csv",header=TRUE)
@@
 If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.
-These white-space gaps are undesireable because they create artificial discontinuity that interfers with interpretation of bivariate density.
-If bin dimensions are not integer multiples of the bivariate data resolution there is also a distruptive visual consequence of ``stripping'', where bins have systematically different numbers of possible data values that can cause artifical density stripes to appear in the binned scatterplot.
+These white-space gaps are undesirable because they create artificial discontinuity that interferes with interpretation of bivariate density.
+If bin dimensions are not integer multiples of the bivariate data resolution there is also a disruptive visual consequence of ``stripping'', where bins have systematically different numbers of possible data values that can cause artificial density stripes to appear in the binned scatterplot.
 
@@
 Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
-Under standard binning we see the white-space gaps with 1 X 1 unit bins, stripes with 5 X 5 unit bins and the correct even spread density with 4X4 unit bins.
-Note that here the random binning is effective at eliminating the misleading visual effects when non-integer multiples of the data resolution are selected.
+Under standard binning we see the white-space gaps with 1 X 1 unit bins, vertical and horizontal stripes with 5 X 5 unit bins and the appropriate view of an evenly spread density with 4X4 unit bins.
+Note that random binning is effective at smoothing out the stripped patterns when non-integer multiples of the data resolution are selected.
 
@@
 
+The binning origin can also influence the spatial loss in the binned scatterplot.
+For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density in outermost bins.
+Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of 10 by 10 unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9).
+The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are empty in the negative regions.
+The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about 7\% lower than for the (-9,-9) origin.
+For data with very fine resolution, it is a good idea to set the origin at the minimum value for each dimension.
+
+<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
+These binnings have net spatial losses of 391067 units and 420458 units, respectively">>=
+
+#create binned data for fine exponential with 10X10 bins at good and bad origin
+binoutExp5 <- RectBin2d(datlist[[3]]$x,datlist[[3]]$y,0,0,10,10,type="standard")
+binoutExp5offset <- RectBin2d(datlist[[3]]$x,datlist[[3]]$y,-9,-9,10,10,type="standard")
+
+p1 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5[[1]]) +  
+  ggtitle("Origin at (0,0)") +
+  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+  theme_bw() + theme(legend.position="none",
+                     aspect.ratio=1, panel.grid= element_blank()) +
+  xlab("") + ylab("")
+p2 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5offset[[1]]) +  
+  ggtitle("Origin at (-9,-9)") + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+  theme_bw() + theme(legend.position="none",
+                     aspect.ratio=1, panel.grid= element_blank()) +
+  xlab("") + ylab("")
+
+grid.arrange(p1,p2,nrow=1)
+
+## loss comparison calculation for paragraph above:  7% lower
+# binoutExp5[[2]][5]/binoutExp5offset[[2]][5]
+
+@
+
 \ktm{
@@
 \item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytic proof)
-\item Origin: 
-  \begin{itemize}
-    \item Doesn't much matter for fine resolution data but it does for coarse data.
-    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
-  \end{itemize} 
 \item Bin width: 
@@
   \end{itemize}
+\item Origin: 
+  \begin{itemize}
+    \item Doesn't much matter for fine resolution data but it does for coarse data.
+    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
+  \end{itemize}
+
 \end{itemize}
@@
 
-\noindent Spatial Loss, $L^S = \sum_{i=1}^{k} ||x_i-b_(x_i)|| $ is definitional minimized when $b_(x_i)$ is the \textit{geometric median}.
+\noindent Spatial Loss, $L^S = \sum_{i=1}^{k} ||x_i-b_(x_i)|| $ is definitionally minimized when $b_(x_i)$ is the \textit{geometric median}.
 The geometric median for $x_1, \dots, x_k = Q_x(.5) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2$ , where $Q_x(\cdot)$ is the empirical quantile function.
@@
 
-If the same conditions are extended to the two dimensional case, then the origin for minimal spatial loss is at $(x_{(1)}-\alpha_x/2, y_{(1)}-\alpha_y/2)$ where $\alpha_x$ and $\alpha_y$  are the data resolution for each dimesion, respectively.
+If the same conditions are extended to the two dimensional case, then the origin for minimal spatial loss is at $(x_{(1)}-\alpha_x/2, y_{(1)}-\alpha_y/2)$ where $\alpha_x$ and $\alpha_y$  are the data resolution for each dimension, respectively.
 

### 2015-04-02 Paper.Rnw
Paragraph on the optimal alpha/2 offset proof in appendix

@@
 
+For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
+It can be shown analytically that for data with a resolution of $\alpha$ units that is spread symmetrically within bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum values (see proof in Appendix~\ref{proof:offset}).
+
 <<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
@@
 %----------------------------------------------------------------------------
-\subsection{Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}\label{proof:offset}
+\subsection{Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}
+\label{proof:offset}
 

### 2015-04-02 Paper.Rnw
New plot that shows the recommended origin offset (alpha/2 below minimum value) for the simulated data at 5 different binwiths.  Also push data for constructing this plot.  Next step will be paragraph describing the local minimums in loss achieved at that offset.

@@
 
-For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
-It can be shown analytically that for data with a resolution of $\alpha$ units that is spread symmetrically within bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum values (see proof in Appendix~\ref{proof:offset}).
-
 <<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
@@
 # binoutExp5[[2]][5]/binoutExp5offset[[2]][5]
+
+@
+
+For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
+It can be shown analytically that for data with a resolution of $\alpha$ units that is spread symmetrically within bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum values (see proof in Appendix~\ref{proof:offset}).
+
+<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a resolution of 2 units at various sized (2,4,6,8,10) square bins over the range of possible origin offsets.">>=
+
+##  Check out the spatial loss for the data that is recorded to resolution = alpha at various origin offsets
+# standard and random binning with bin sizes that are scalar multiples of the data resolution
+# save spatial loss information
+# spatlossdatRound <- data.frame(NULL)
+# index=0
+# distindex = 0
+# distrib <- c("unif", "norm", "exp")
+# widths <- seq(alpha,5*alpha,by=alpha)
+# for (dat in 1:3){
+#   distindex <- distindex + 1
+#   for (width in widths){
+#     for (origin in seq(-width/2,0, by=width/60)){
+#       index <- index+1
+#       binout <- RectBin2d(datlist[[dat]]$roundx,datlist[[dat]]$roundy,origin,origin,width,width,type="standard")
+#       binout[[2]]$bintype <- "standard"
+#       binout[[2]]$distrib <- distrib[distindex]
+#       spatlossdatRound <- rbind(spatlossdatRound, binout[[2]])
+#       index <- index+1
+#       binout <- RectBin2d(datlist[[dat]]$roundx,datlist[[dat]]$roundy,origin,origin,width,width,type="random")
+#       binout[[2]]$bintype <- "random"
+#       binout[[2]]$distrib <- distrib[distindex]
+#       spatlossdatRound <- rbind(spatlossdatRound, binout[[2]])
+#       print(c(distrib[distindex],width,origin))
+#     }
+#   }
+# }
+# write.csv(spatlossdatRound,"LossesByOriginBinwidth.csv", row.names=FALSE)
+lossOrigin <- read.csv("LossesByOriginBinwidth.csv",header=TRUE)
+
+qplot(originx,totalSpatialLoss,geom="path",group=distrib, colour=distrib,
+      data=subset(lossOrigin,bintype=="standard"), size=I(1))+ 
+  facet_grid(distrib~widthx) + theme_bw() + 
+  theme(panel.grid= element_blank())+
+  geom_vline(xintercept = -1, colour=I("darkgray")) +
+  ylab("Net Spatial Loss") + xlab("Origin Offset")
 

### 2015-04-02 Paper.Rnw
Description of Figure 11 added to writeup

@@
 Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of 10 by 10 unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9).
-The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are empty in the negative regions.
+The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions.
 The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about 7\% lower than for the (-9,-9) origin.
@@
 For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
-It can be shown analytically that for data with a resolution of $\alpha$ units that is spread symmetrically within bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum values (see proof in Appendix~\ref{proof:offset}).
-
-<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a resolution of 2 units at various sized (2,4,6,8,10) square bins over the range of possible origin offsets.">>=
+It can be shown analytically that for data with a resolution of $\alpha$ units that is spread symmetrically within bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum value (see proof in Appendix~\ref{proof:offset}).
+
+In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimesions that are integer multiples of $\alpha$.
+Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to half the binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit below the minimum data value was found to minimize the net spatial loss for each considered binsize (2, 4, 6, 8, and 10).
+For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the larger intervals.
+
+<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a resolution of 2 units at various sized (2,4,6,8,10) square bins over the range of possible origin offsets.
+The vertical gray line in each facet indicates where the origin is offset 1 unit below the minimum data value.">>=
 
@@
   facet_grid(distrib~widthx) + theme_bw() + 
-  theme(panel.grid= element_blank())+
+  theme(panel.grid= element_blank(), legend.position="none")+
   geom_vline(xintercept = -1, colour=I("darkgray")) +

### 2015-04-02 Paper.Rnw
Editing the intro to section 5

@@
 This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
-\ktm{Some results are derived \hh{XXX check the sentence} optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.
-}
-
+Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning.
 Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
 
-\hh{Just list the three different scenarios:
-\begin{itemize}
-\item {\bf Scenario I:} $x \sim \text{Exp}_\lambda, y \sim \text{Exp}_\lambda$ with $\lambda = 11$
-\end{itemize}}
 Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
 These distributions were selected for their variety in terms of shape, center and spread.
-100,000 observation pairs were simulated for each set of simulated data, and the $x$ and $y$ values were drawn independently from univariate distributions; Exponential(11), Normal(50,$11^2$) and Uniform(0,100), respectively.
-The parameters were selected to have data values roughly span the region $[0,100]^2$.
-The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
+100,000 observation pairs were simulated for each data set from the following distributions:
+
+\begin{itemize}
+\item Set I: $x_i \sim \text{iid Exp}(\lambda), y_i \sim \text{iid Exp}(\lambda)$ with $\lambda = 11$
+
+\item Set II: $x_i \sim \text{iid Normal}(\mu,\sigma^2), y_i \sim \text{iid Normal}(\mu,\sigma^2)$ with $\mu = 50$ and $\sigma=11$
+
+\item Set III: $x_i \sim \text{iid Uniform}(a,b), y_i \sim \text{iid Uniform}(a,b)$ with $a =0$ and $b=100$
+\end{itemize}
+%
+The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
 
@@
 
-\ktm{
-\begin{itemize}
-\item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytic proof)
-\item Bin width: 
-  \begin{itemize}
-    \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
-    \item for fine data any bin width is fine
-    \item for coarse data need bin widths $\omega \ge \alpha$ (show plot with white space gaps)
-    \item for coarse data bin widths should be scalar multiple of $\alpha$ (show plot with strips)
-    \item Does the distribution type matter? 
-  \end{itemize}
-\item Origin: 
-  \begin{itemize}
-    \item Doesn't much matter for fine resolution data but it does for coarse data.
-    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
-  \end{itemize}
-
-\end{itemize}
-}
+% 
+% \ktm{
+% \begin{itemize}
+% \item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytic proof)
+% \item Bin width: 
+%   \begin{itemize}
+%     \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
+%     \item for fine data any bin width is fine
+%     \item for coarse data need bin widths $\omega \ge \alpha$ (show plot with white space gaps)
+%     \item for coarse data bin widths should be scalar multiple of $\alpha$ (show plot with strips)
+%     \item Does the distribution type matter? 
+%   \end{itemize}
+% \item Origin: 
+%   \begin{itemize}
+%     \item Doesn't much matter for fine resolution data but it does for coarse data.
+%     \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
+%   \end{itemize} 
+% 
+% \end{itemize}
+% }
 

### 2015-04-02 Paper.Rnw
Minor edits

@@
 
-In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimesions that are integer multiples of $\alpha$.
+In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimensions that are integer multiples of $\alpha$.
 Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to half the binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
-For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit below the minimum data value was found to minimize the net spatial loss for each considered binsize (2, 4, 6, 8, and 10).
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit below the minimum data value was found to minimize the net spatial loss for each considered bin size (2, 4, 6, 8, and 10 units).
 For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the larger intervals.
@@
 
-\subsection{Frequency binning choices}
+\subsection{Frequency Binning Specifications and Frequency Loss}
+
+The loss of frequency information in frequency binning is highly subject to the selection of a discrete color mapping.
 
 \ktm{
+\begin{itemize}
+\item Number of Colors:
+  \begin{itemize}
+    \item Perception driven limitation to 7 or less color bins
+    \item As number colors increases, freq loss goes down
+  \end{itemize}
+\end{itemize}
+
 \begin{itemize}
@@
     \item Comparison of standard, log and quantile binning
-  \end{itemize}
-\end{itemize}
-
-\begin{itemize}
-\item Number of Colors:
-  \begin{itemize}
-    \item Perception driven limitation to 7 or less color bins
-    \item As number colors increases, freq loss goes up?
   \end{itemize}

### 2015-04-06 Paper.Rnw
couple of comments

@@
 The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}. 
-
-<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.">>=
+\hh{XXX what other purpose does this scatterplot have besides showing the distributions?}
+
+<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
+\\hh{include a finding here- e.g.
+how big is the loss under each scenario?}">>=
 ### Try 2d binning with MVN data
@@
 
-These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places.
-Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles.
-The data resolution can be described by increment between data values.
-To observe binning loss under more realistic conditions, we created three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
+These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places \hh{XXX how many exactly?
+.
+Real data is recorded to only the number of digits that the measurements allow, \hh{and} in many cases rounded \hh{even further.
+}% to the nearest whole dollar or five minutes or two miles.
+\hh{XXX this is a definition of data resolution, make that clear.} The data resolution can be described by increment between data values.
+To observe \hh{loss from} binning  under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
@@
 Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
-It is easily apparent that  the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
+It is  apparent that  the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
 The net spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.

### 2015-04-06 Paper.Rnw
ylim

@@
   theme_bw() + ylab("Net Spatial Loss") +
-  xlab("Width/Height of Square Bins")
+  xlab("Width/Height of Square Bins") + ylim(c(0, 1100000))
 @

### 2015-04-07 Paper.Rnw
Edits to sections 3 and 4 based on comments

@@
 Table~\ref{DataReductionExample} makes use of a small set of simulated data to show the progression from the original data (a), to the binned data (b), to the reduced binned data (c).
-The reduced binned data is sufficient for constructing the binned scatterplot. %, as only bin centers and frequencies are necessary.  
-Binning greatly reduces the storage size for the information and the computation time needed to construct the plot.
-\hh{XXX this is not always true - when the bins have only 1s the binned data set is actually bigger than the original. }
+The reduced binned data is sufficient for constructing the binned scatterplot.
+In cases of large data, binning greatly reduces the storage size for the information and the computation time needed to construct the plot.
 
 Data reduction comes at the expense of spatial information of any of the individual points.
-After aggregation the original spatial locations cannot be recovered. The loss of information incurred from binning will be explored in following sections.
+After aggregation the original spatial locations cannot be recovered.
+The loss of information incurred from binning will be explored in following sections.
 
@@
 %
-Figure~\ref{fig:spatlossdemo2} \hh{XXX in the figure, make the tiles at the bottom squares, so they match with the scatter plot above.
-This is a neat picture!
-XXX} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning.
+Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning.
 Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively.
@@
       theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-            legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+            legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"),
+            aspect.ratio=1)
 
@@
   theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"))
+        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"),
+            aspect.ratio=1)
 
@@
   theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
-        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom")+
+        plot.margin = unit(c(0,1,0, 1),"cm"),legend.position="bottom",
+            aspect.ratio=1)+
   geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp)
@@
 
-\ktm{ For random assignment the total spatial loss can be calculated by simply replacing  \hh{call these guys by name} $b_X(x_i)$ with $b^r_X(x_i)$ in Equation~\ref{totspatloss}.}  The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
+For random assignment the total spatial loss can be calculated by simply replacing the standard binning function with the random binning function in Equation~\ref{totspatloss}.
+The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
 Special consideration should be paid to the two points highlighted in red in the panel for random binning: in standard binning these points are in neighboring bins but have been allocated to the opposite bin during random binning.
@@
 Color perception of shade is extremely context sensitive allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
-\hh{It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
+It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
 Figure~\ref{fig:context} shows an example of the context sensitivity of colors.
 Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each of the points clouds.
-}
 
@@
 
-These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places \hh{XXX how many exactly?
-.
-Real data is recorded to only the number of digits that the measurements allow, \hh{and} in many cases rounded \hh{even further.
-}% to the nearest whole dollar or five minutes or two miles.
-\hh{XXX this is a definition of data resolution, make that clear.} The data resolution can be described by increment between data values.
-To observe \hh{loss from} binning  under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
+These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data.
+Real data is recorded to only the number of digits that the measurements allow, and in many cases rounded even further.
+
+The \textit{data resolution} is defined as the increment between data values.
+To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.

### 2015-04-07 Paper.Rnw
Add frequency loss comparison plot for standard and quantile frequency binning.

@@
 
-<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
+<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
 These binnings have net spatial losses of 391067 units and 420458 units, respectively">>=
@@
 
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss from standard and quantile frequency binning of reduced binned data from each bivariate distribution.
+Spatial binning to produce the reduced binned data done with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).">>=
+### Loop over ncolors 1 to 7 and also over data distribution
+# find freq loss from standard, quantile, logCount
+freqLossSimDat <- NULL
+distrib <- c("Uniform", "Normal", "Exponential")
+for (distidx in 1:3){
+  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
+  for (ncol in 1:7){
+    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
+                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
+    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
+    binoutFreq$DistType <- distrib[distidx]
+    binoutFreq$ncolor <- ncol 
+    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
+  }
+}
+
+freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Exponential"),]
+p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Normal"),]
+p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Uniform"),]
+p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+
+grid.arrange(p1,p2,p3, nrow=1)
+@
+
 \ktm{

### 2015-04-08 Paper.Rnw
New section 5.2 (including a few plots that I don't think will survive much longer than tomorrow's meeting)

@@
 \subsection{Frequency Binning Specifications and Frequency Loss}
-
-The loss of frequency information in frequency binning is highly subject to the selection of a discrete color mapping.
+\ktm{
+The reduced binned data from spatial binning contains the center and count information for all bins.
+The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.
+The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified.
+To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned using standard rectangular binning to produce three sets of reduced binned data; then standard, quantile and log frequency binning algorithms were applied with various numbers of frequency bins.
+As discussed in Section~\ref{FreqLoss}, we will limit the considered options to using between one and seven frequency bins due to the upper limits of human perceptual abilities \citep{colorperception}.
+
+Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and seven frequency bins, from each set of simulated data.
+We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.
+We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms, thus we should consider using the maximum number of bin shades that human perception is able to simultaneously discern.
+Lastly, the frequency loss with the standard frequency binning may be higher than for the quantile-based algorithm when a minimal number of bins are used; but in all three simulated data sets, this is reversed when seven frequency bins are used.
+Thus we recommend using the standard algorithm over the quantile-based algorithm for frequency binning because it had smaller frequency loss at the recommended seven bins, and additionally carries the benefit that equally dividing the bin count range into equally spaced bins creates a simpler shading interpretion than dividing by quantiles.
+
+Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
+Frequency binning using the log counts did however behave very similarly, where log frequency loss decreased as the number of frequency bins increased.
+Thus the same advise to use seven frequency bins holds for shading based on log counts as well.
+}
+
+<<FreqLossDisplay,echo=F,include=F,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning.
+If so, do the colors between freq bins need to be made more distinguishable?
+i.e.
+in the quantile binning the bins starting at 1, 8 and 36.5 are hard to differentiate because the shade is determined by the breaks.
+">>=
+binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
+binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
+binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
+binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)
+
+# Plot with Orginal Frequencies
+p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binoutFreqGroups[[1]],
+            main="No Freq Binning") +
+  xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+# Plot with Binned Frequencies
+p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
+            main="Standard Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+                      breaks=unique(binoutFreqGroups[[1]]$freqgroup)) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+# Plot with Quantile Binned Frequencies
+p3 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutQuantFreqGroups[[1]],
+            main="Quantile Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutQuantFreqGroups[[1]]$freqgroup)) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+# Plot with LogCount Binned Frequencies
+p4 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]],
+            main="Log Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutLogFreqGroups[[1]]$freqgroup)) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+
+grid.arrange(p1,p2,p3,p4, nrow=1)
+@
 
@@
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Freq Loss") + 
@@
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+
+grid.arrange(p1,p2,p3, nrow=1)
+@
+
+<<FreqLossCompare2,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total log frequency loss from log frequency binning of reduced binned data from each bivariate distribution.
+I don't think this plot will make the final cut, but it is interesting that after logging the bin counts, the log freq loss for the uniform data does not improve with more bins (might as well use one shade for all bins)">>=
+### Loop over ncolors 1 to 7 and also over data distribution
+# find freq loss from standard, quantile, logCount
+freqLossSimDat <- NULL
+distrib <- c("Uniform", "Normal", "Exponential")
+for (distidx in 1:3){
+  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
+  for (ncol in 1:7){
+    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
+                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
+    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
+    binoutFreq$DistType <- distrib[distidx]
+    binoutFreq$ncolor <- ncol 
+    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
+  }
+}
+
+freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Exponential"),]
+p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = I(3), size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom")
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Normal"),]
+p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = I(3), size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom")
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Uniform"),]
+p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = I(3), size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
   theme(legend.position="bottom")

### 2015-04-08 Paper.Rnw
mini change

--- 2015-04-07 Paper.Rnw
@@
 
-The \textit{data resolution} is defined as the increment between data values.
+The \textit{data resolution} is defined as the \hh{smallest} increment between \hh{successive} data values.
 To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
@@
 
-<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
-spatlossdat <- read.csv("SpatialLossData.csv",header=TRUE)
-# use all the square bins to plot bin width against proportion of max spataial loss
-spatlossdatSquareBins <- spatlossdat[spatlossdat$widthx==spatlossdat$widthy,]
-qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
+<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=6, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
+## create binned data for standard and random using variety of bin sizes (from 2 to 20, by 2)
+##save spatial loss information
+# spatlossdat <- data.frame(NULL)
+# index=0
+# distindex = 0
+# distrib <- c("Uniform", "Normal", "Exponential")
+# for (dat in 1:3){
+#   distindex <- distindex + 1
+#   for (width in seq(2,20,by=2)){
+#       index <- index+1
+#       timeStart <- Sys.time()
+#       binout <- RectBin2d(datlist[[dat]]$x,datlist[[dat]]$y,0,0,width,width,type="standard")
+#       binout[[2]]$binTime <- Sys.time() - timeStart
+#       binout[[2]]$bintype <- "standard"
+#       binout[[2]]$distrib <- distrib[distindex]
+#       spatlossdat <- rbind(spatlossdat, binout[[2]])
+#       index <- index+1
+#       timeStart <- Sys.time()
+#       binout <- RectBin2d(datlist[[dat]]$x,datlist[[dat]]$y,0,0,width,width,type="random")
+#       binout[[2]]$binTime <- Sys.time() - timeStart
+#       binout[[2]]$bintype <- "random"
+#       binout[[2]]$distrib <- distrib[distindex]
+#       spatlossdat <- rbind(spatlossdat, binout[[2]])
+#       print(c(distrib[distindex],width))
+#   }
+# }
+# # write.csv(spatlossdat,"SpatialLossData.csv", row.names=FALSE)
+spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE)
+spatlossdatSquareBins$bintype <- factor(spatlossdatSquareBins$bintype, levels=c("standard","random"))
+p1 <- qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
@@
   xlab("Width/Height of Square Bins") + ylim(c(0, 1100000))
+
+p2 <- qplot(widthx,as.numeric(binTime),geom="path",colour=bintype, linetype=bintype,
+      data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
+  theme_bw() + ylab("Binning Time (sec)") +
+  xlab("Width/Height of Square Bins")
+
+grid.arrange(p1,p2, nrow=2)
 @
@@
 All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin.
-While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a non-negligible consideration in settings with truly massive data.
+While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data.
+\ktm{ Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.
+Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
+Another pertinent argument for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}
 
@@
 In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimensions that are integer multiples of $\alpha$.
-Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to half the binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
+Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to one binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
 For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit below the minimum data value was found to minimize the net spatial loss for each considered bin size (2, 4, 6, 8, and 10 units).
-For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the larger intervals.
+For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals.
 
@@
 The vertical gray line in each facet indicates where the origin is offset 1 unit below the minimum data value.">>=
-
 ##  Check out the spatial loss for the data that is recorded to resolution = alpha at various origin offsets
@@
 # distindex = 0
-# distrib <- c("unif", "norm", "exp")
+# distrib <- c("Uniform", "Normal", "Exponential")
 # widths <- seq(alpha,5*alpha,by=alpha)
@@
 #   for (width in widths){
-#     for (origin in seq(-width/2,0, by=width/60)){
+#     for (origin in seq(-width,0, by=width/60)){
 #       index <- index+1
@@
       data=subset(lossOrigin,bintype=="standard"), size=I(1))+ 
-  facet_grid(distrib~widthx) + theme_bw() + 
+  facet_grid(distrib~widthx, scales="free_x") + theme_bw() + 
   theme(panel.grid= element_blank(), legend.position="none")+
@@
   ylab("Net Spatial Loss") + xlab("Origin Offset")
-
 @
@@
 \subsection{Frequency Binning Specifications and Frequency Loss}
-
-The loss of frequency information in frequency binning is highly subject to the selection of a discrete color mapping.
-
 \ktm{
-\begin{itemize}
-\item Number of Colors:
-  \begin{itemize}
-    \item Perception driven limitation to 7 or less color bins
-    \item As number colors increases, freq loss goes down
-  \end{itemize}
-\end{itemize}
-
-\begin{itemize}
-\item Freq Binning Type
-  \begin{itemize}
-    \item Comparison of standard, log and quantile binning
-  \end{itemize}
-\end{itemize}
+The reduced binned data from spatial binning contains the center and count information for all bins.
+The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.
+The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified.
+To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).
+Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
+As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
+
+The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
+We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.
+We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms.
+This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous.
+Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot.
+Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used.
+But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.
+
+Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
+The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data.
+Frequency binning using the log counts behaved very similarly, where log frequency loss decreased as the number of frequency bins increased.
+The primary difference uniform simulated data with bin counts that have low variability, are rendered virtually indistinguishable after a log transformation; thus changing the number of frequency bins does little to impact the log frequency loss.
+The same advise to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
 }
 
+<<FreqLossDisplay,echo=F,include=T,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning.
+If so, do the colors between freq bins need to be made more distinguishable?
+i.e.
+in the quantile binning the bins starting at 1, 8 and 36.5 are hard to differentiate because the shade is determined by the breaks.
+">>=
+binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
+binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
+binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
+binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)
+
+# Plot with Orginal Frequencies
+p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binoutFreqGroups[[1]],
+            main="No Freq Binning") +
+  xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+# Plot with Binned Frequencies
+p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
+            main="Standard Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+                      breaks=unique(binoutFreqGroups[[1]]$freqgroup)) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+# Plot with Quantile Binned Frequencies
+p3 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutQuantFreqGroups[[1]],
+            main="Quantile Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutQuantFreqGroups[[1]]$freqgroup)) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+# Plot with LogCount Binned Frequencies
+p4 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]],
+            main="Log Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutLogFreqGroups[[1]]$freqgroup)) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     panel.grid= element_blank())
+
+grid.arrange(p1,p2,p3,p4, nrow=1)
+@
+
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=8, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
+">>=
+### Loop over ncolors 1 to 10 and also over data distribution
+# find freq loss from standard, quantile, logCount
+freqLossSimDat <- NULL
+distrib <- c("Uniform", "Normal", "Exponential")
+for (distidx in 1:3){
+  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
+  for (ncol in 1:10){
+    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
+                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
+    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
+    binoutFreq$DistType <- distrib[distidx]
+    binoutFreq$ncolor <- ncol 
+    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
+  }
+}
+
+freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels= c("Standard", "Quantile", "Standard Log", "Quantile Log"))
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Exponential"),]
+p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Normal"),]
+p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Uniform"),]
+p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom")
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log")  & freqLossSimDat$DistType=="Exponential"),]
+p4 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color=FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom")
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Normal"),]
+p5 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom")
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Uniform"),]
+p6 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,40) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom")
+
+grid.arrange(p1,p2,p3,p4,p5,p6, nrow=2)
+@
+
 %-----------------------------------------------------------------------------
@@
 \section{Discussion and Examples}
-\subsection{Comparison of Loss in Binning Algorithms}
+
+\ktm{
+The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction.
+Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing (Would like to use an O(n)-type argument, but I don't rightly understand this comp.
+sci.
+notation) and does not highlight large scale density structure.
+Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear.
+Thus it is left to the plot designer to decide if they wish to highlight high level density structure with higher spatial loss or finer density structure with lower spatial loss.
+
+For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution, $\alpha$.
+However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution is not problematic so long as a binwidth that is an integer multiple of $\alpha$ is selected.
+It was also shown that in this scenario, we should use an origin that is offset to $\alpha / 2$ units below the minimum data value leads to minimize spatial loss.
+
+If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
+This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
+The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision falls to the interpretability of the resulting frequency bins.
+Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.
+
+These recommendations for binned scatterplot construction and loss properties will be demonstrated in the following examples using real bivariate data.
+First we will further explore the baseball pitching data.
+Additionally we will investigate the relationship between scheduled and actual departure times for a full year of American commercial airline data.
+}
+
+% \subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
 <<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=
@@
 
-In order to quantify the advantages of each algorithm, we first compare the three sources of loss at different bin widths.
+% In order to quantify the advantages of each algorithm, we first compare the three sources of loss at different bin widths.
 The numerical and visual losses inherent in creating a binned scatterplot are a result of our desire to reduce and summarize the data until it is more manageable in size, and hence is a factor in our decision to bin the data in the first place.
-%Figure~\ref{fig:binning-comparison} shows clearly that the standard algorithm can produce misleading graphical summaries. 
- The random algorithm has higher numeric and visual loss than the standard algorithm, but in many cases the binned scatterplots are visually indistinguishable.
-
- The numerical loss is a function of the distribution and bin width, and once the bin width is chosen, it is entirely fixed.
+% %Figure~\ref{fig:binning-comparison} shows clearly that the standard algorithm can produce misleading graphical summaries. 
+%  The random algorithm has higher numeric and visual loss than the standard algorithm, but in many cases the binned scatterplots are visually indistinguishable. 
+%  
+%  The numerical loss is a function of the distribution and bin width, and once the bin width is chosen, it is entirely fixed.
 The random binning algorithm will inherently have a larger numerical loss than the standard algorithm because it allows for random assignment of points to numeric bin centers further from the original data than does the standard algorithm.
-
- As the visual loss is the difference between the numerical center and visual center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm).
+%  
+%  As the visual loss is the difference between the numerical center and visual center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm).
 We expect a larger visual loss under the random binning algorithms than in the standard binning because the numerical bin centers are allowed to be drawn further away from the visual center under the random binning algorithm.
-
-Frequency loss is arguably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary.
+% 
+% Frequency loss is arguably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary.
 The standard and random algorithms are very similar when we compare frequency loss.
@@
 %~\ref{fig:LossComparison}
-
-Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues.
+% 
+% Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues.
 Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a bin width that is a non-integer multiple of the incremental widths.
@@
 
-\begin{figure}[hbtp]
-\centering
-<<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
-
-# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-#                      password="R R0cks", port=3306, dbname="baseball", 
-#                      host="headnode.stat.iastate.edu")
-# pitch <- new("dataDB", co=connect, table="Pitching")
-# d1 <- dbData(pitch, vars=c( "G", "SO"))
-# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
-# 
-# library(multicore)
-# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
-# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
-# 
-# lossesRdm <- cbind(bins, lossesRdm)
-# lossesStd <- cbind(bins, lossesStd)
-# 
-# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
-# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
-
-losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
-lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
-lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
-lossmelt$y <- as.factor(lossmelt$y)
-levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
-qplot(x=x, y=value,
-      group=interaction(y, Algorithm),
-      data=lossmelt, geom="line",
-      colour=y, size=I(1),
-      xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
-        theme_bw() + theme(legend.position="bottom") + 
-        guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-        facet_grid(variable ~ Algorithm, scales = "free")
-
-# qplot(x=x, y=value,
-#       group=interaction(y, Algorithm),
-#       data=lossmelt, geom="line",
-#       colour=y, size=I(1), linetype=Algorithm,
-#       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
-#         theme_bw() + theme(legend.position="bottom") + 
-#         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-#         facet_grid(. ~ variable, scales = "free")
-
-@
-\caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
-\end{figure}
+% \begin{figure}[hbtp]
+% \centering
+% <<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
+% 
+% # connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
+% #                      password="R R0cks", port=3306, dbname="baseball", 
+% #                      host="headnode.stat.iastate.edu")
+% # pitch <- new("dataDB", co=connect, table="Pitching")
+% # d1 <- dbData(pitch, vars=c( "G", "SO"))
+% # bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
+% # 
+% # library(multicore)
+% # lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
+% # lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
+% # 
+% # lossesRdm <- cbind(bins, lossesRdm)
+% # lossesStd <- cbind(bins, lossesStd)
+% # 
+% # losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
+% # write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
+% 
+% losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
+% lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
+% lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
+% lossmelt$y <- as.factor(lossmelt$y)
+% levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
+% qplot(x=x, y=value,
+%       group=interaction(y, Algorithm),
+%       data=lossmelt, geom="line",
+%       colour=y, size=I(1),
+%       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
+%         theme_bw() + theme(legend.position="bottom") + 
+%         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
+%         facet_grid(variable ~ Algorithm, scales = "free") 
+% 
+% # qplot(x=x, y=value,
+% #       group=interaction(y, Algorithm),
+% #       data=lossmelt, geom="line",
+% #       colour=y, size=I(1), linetype=Algorithm,
+% #       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
+% #         theme_bw() + theme(legend.position="bottom") + 
+% #         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
+% #         facet_grid(. ~ variable, scales = "free") 
+% 
+% @
+% \caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
+% \end{figure}
 
@@
 
-\ktm{
 The following is proof that when univariate data is uniformly distributed at resolution $\alpha$ and standard rectangular binning is used with binwidths that are a scalar multiple of the data resolution (i.e.
@@
 
-} %end ktm color
 %----------------------------------------------------------------------------
-
-\subsection{Partition of Spatial Loss}\label{proof:partition}
-We want to show that spatial loss can be written as a partition of the form 
-\[L_s(x) = 1/S^2_{\emptyset}\sum_i S^2_i = 1/S^2_{\emptyset}\left[\sum_i L^N_i + \sum_i L^V_i\right]\]
-
-For that let us consider the values $\{x_{i}: b(x_i) = x^\ast\}$ in a single bin with center $x^\ast$.
-The numerical center of these points is given as  $1/n \sum x_i = \overline x$.
-Then the total loss for each element $x_i$ in this bin is:
-
-\begin{align*}
-S_i^2 = \sum (x_i - x_i^\ast)^2 & = \sum (x_i - \overline x + \overline x - x_i^\ast)^2\\
-& = \sum(x_i-\overline x)^2 + 2\sum (x_i-\overline x)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
-& = \sum(x_i - \overline x)^2 + 2\sum(0)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
-& = \sum(x_i - \overline x)^2 + \sum(\overline x-x_i^\ast)^2\\
-& = L^N_i + L^V_i
-\end{align*}
-Adding over all elements in the bin and all bins gives the formula above.
-
-\subsection{Constancy of Loss under Binary Splits}\label{proof:constantloss}
-We want to show that under random binning the loss due to doubling of bin width is deterministic.
-
-Let us consider the contribution $L_i$ of a single point $x_i$ to the loss when starting with the minimal bin width, i.e. each unique point is in a separate bin:
-doubling the bin width leads to two  scenarios: 
-a point is either at  the bin center or directly between two bins.
-In the case that the point is at the bin center, its contribution to the total overall loss is zero,  $L_i = 0$, and the associated probability $p_i = 1$.
-
-Alternately, when the point is exactly half-way between two bin centers, the point is attributed with probability $p_i = \frac{1}{2}$ to either bin, leading to a loss of $L_i = \frac{1}{2} c_i$ regardless of which bin is assigned.
+% 
+% \subsection{Partition of Spatial Loss}\label{proof:partition}
+% We want to show that spatial loss can be written as a partition of the form 
+% \[L_s(x) = 1/S^2_{\emptyset}\sum_i S^2_i = 1/S^2_{\emptyset}\left[\sum_i L^N_i + \sum_i L^V_i\right]\] 
+% 
+% For that let us consider the values $\{x_{i}: b(x_i) = x^\ast\}$ in a single bin with center $x^\ast$.
+% The numerical center of these points is given as  $1/n \sum x_i = \overline x$.
+% Then the total loss for each element $x_i$ in this bin is:
+% 
+% \begin{align*}
+% S_i^2 = \sum (x_i - x_i^\ast)^2 & = \sum (x_i - \overline x + \overline x - x_i^\ast)^2\\
+% & = \sum(x_i-\overline x)^2 + 2\sum (x_i-\overline x)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
+% & = \sum(x_i - \overline x)^2 + 2\sum(0)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
+% & = \sum(x_i - \overline x)^2 + \sum(\overline x-x_i^\ast)^2\\
+% & = L^N_i + L^V_i
+% \end{align*}
+% Adding over all elements in the bin and all bins gives the formula above.
+% 
+% \subsection{Constancy of Loss under Binary Splits}\label{proof:constantloss}
+% We want to show that under random binning the loss due to doubling of bin width is deterministic.
+% 
+% Let us consider the contribution $L_i$ of a single point $x_i$ to the loss when starting with the minimal bin width, i.e. each unique point is in a separate bin:
+% doubling the bin width leads to two  scenarios: 
+% a point is either at  the bin center or directly between two bins.
+% In the case that the point is at the bin center, its contribution to the total overall loss is zero,  $L_i = 0$, and the associated probability $p_i = 1$. 
+% 
+% Alternately, when the point is exactly half-way between two bin centers, the point is attributed with probability $p_i = \frac{1}{2}$ to either bin, leading to a loss of $L_i = \frac{1}{2} c_i$ regardless of which bin is assigned.
 Hence, in this case, loss is also entirely independent of which bin is assigned.

### 2015-04-08 Paper.Rnw
Draft of new Section 6 introduction

--- 2015-04-08 Paper.Rnw
@@
 Lastly, the frequency loss with the standard frequency binning may be higher than for the quantile-based algorithm when a minimal number of bins are used; but in all three simulated data sets, this is reversed when seven frequency bins are used.
-Thus we recommend using the standard algorithm over the quantile-based algorithm for frequency binning because it had smaller frequency loss at the recommended seven bins, and additionally carries the benefit that equally dividing the bin count range into equally spaced bins creates a simpler shading interpretion than dividing by quantiles.
+Thus we recommend using the standard algorithm over the quantile-based algorithm for frequency binning because it had smaller frequency loss at the recommended seven bins, and additionally carries the benefit that equally dividing the bin count range into equally spaced bins creates a simpler shading interpretation than dividing by quantiles.
 
@@
 
+%-----------------------------------------------------------------------------
+
+\section{Discussion and Examples}
+
 \ktm{
-\begin{itemize}
-\item Number of Colors:
-  \begin{itemize}
-    \item Perception driven limitation to 7 or less color bins
-    \item As number colors increases, freq loss goes down
-  \end{itemize}
-\end{itemize}
-
-\begin{itemize}
-\item Freq Binning Type
-  \begin{itemize}
-    \item Comparison of standard, log and quantile binning
-  \end{itemize}
-\end{itemize}
+The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction.
+Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing (Would like to use an O(n)-type argument, but I don't rightly understand this comp.
+sci.
+notation) and does not highlight large scale density structure.
+Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear.
+Thus it is left to the plot designer to decide if they wish to highlight high level density structure with higher spatial loss or finer density structure with lower spatial loss.
+
+For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution, $\alpha$.
+However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution is not problematic so long as a binwidth that is an integer multiple of $\alpha$ is selected.
+It was also shown that in this scenario, we should use an origin that is offset to $\alpha / 2$ units below the minimum data value leads to minimize spatial loss.
+
+If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use seven distinct shades.
+This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
+Standard frequency binning is recommended over the quantile frequency binning because it displayed lower frequency loss for all three simulated binned data sets while using seven frequency bins.
+Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.
+
+These recommendations for binned scatterplot construction and loss properties will be demonstrated in the following examples using real bivariate data.
+First we will further explore the baseball pitching data.
+Additionally we will investigate the relationship between scheduled and actual departure times for a full year of American commercial airline data.
 }
 
-%-----------------------------------------------------------------------------
-
-\section{Discussion and Examples}
-\subsection{Comparison of Loss in Binning Algorithms}
+\subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
 <<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=

### 2015-04-08 Paper.Rnw
Commenting out old sections to clean up paper and speed up compiling

@@
 
-<<FreqLossDisplay,echo=F,include=F,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning.
+<<FreqLossDisplay,echo=F,include=T,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning.
 If so, do the colors between freq bins need to be made more distinguishable?
@@
 
-\subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
+% \subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
 <<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=

### 2015-04-08 Paper.Rnw
Edits to: Paper - add computation time plot for binning at various bin sizes
Binning Functions - remove proportion of max loss and also create binning function that does not calculate loss (for timing check)

Data for plot - add comp time values

@@
 All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin.
-While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a non-negligible consideration in settings with truly massive data.
+While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data.
+\ktm{ Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.
+Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
+Another pertinent arguement for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}
 

### 2015-04-08 BinningLossFunctions.R

@@ -102,18 +102,12 @@ RectBin2d <- function(xs,ys, originx, originy, widthx, widthy, type="standard"){
     tempdat$binxs[!(tempdat$index %in% swapindex)] <- tempdat$randbinxs[!(tempdat$index %in% swapindex)]
     tempdat$binys[!(tempdat$index %in% swapindex)] <- tempdat$randbinys[!(tempdat$index %in% swapindex)]
   }
-  
-  
   outdat <- ddply(tempdat, .(binxs,binys), summarise,
                   binfreq = length(xs),
                   binspatialloss = sum(sqrt((xs-binxs)^2+(ys-binys)^2)) )
-  centerofdata <- c((max(xs)-min(xs))/2,(max(ys)-min(ys))/2)
-  #centerofdata <- c(mean(xs),mean(ys))
   summarydata <- data.frame( originx = originx, originy = originy, 
                              widthx = widthx, widthy = widthy,
-                             totalSpatialLoss = sum(outdat$binspatialloss) ,
-                             maxSpatialLoss = sum(sqrt((xs-centerofdata[1])^2+(ys-centerofdata[2])^2)),
-                             propSpatialLoss = sum(outdat$binspatialloss) / sum(sqrt((xs-centerofdata[1])^2+(ys-centerofdata[2])^2)))
+                             totalSpatialLoss = sum(outdat$binspatialloss))
   templist <- list(bindat = outdat[,1:3],
                    summarydata)
   return(templist)
@@ -150,9 +144,9 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
   # Total Freq Loss
   binout[[2]]$totalFreqLoss <- sum((cs - binout[[1]]$freqgroup)^2)
   # Max Freq Loss based on shading at data average
-  binout[[2]]$maxFreqLoss <- sum((cs - mean(cs))^2)
+#  binout[[2]]$maxFreqLoss <- sum((cs - mean(cs))^2)
   # Proportion of Max Freq Loss
-  binout[[2]]$propFreqLoss <- sum((cs - binout[[1]]$freqgroup)^2) /  sum((cs - mean(cs))^2)
+#  binout[[2]]$propFreqLoss <- sum((cs - binout[[1]]$freqgroup)^2) /  sum((cs - mean(cs))^2)
   return(binout)
 }
 
@@ -232,4 +226,26 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
 
 
 
+
+### Create binning function that does not calculate loss
[…]

### 2015-04-08 Paper.Rnw
Frequency Binning edits: -before (1 to 7) now (1 to 10) bins -freq loss lineplot rebuild -add quantile binning of log counts -rewrite several paragraphs in section 5.2 -spell check

@@
 Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
-Another pertinent arguement for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}
+Another pertinent argument for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}
 
@@
 The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified.
-To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned using standard rectangular binning to produce three sets of reduced binned data; then standard, quantile and log frequency binning algorithms were applied with various numbers of frequency bins.
-As discussed in Section~\ref{FreqLoss}, we will limit the considered options to using between one and seven frequency bins due to the upper limits of human perceptual abilities \citep{colorperception}.
-
-Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and seven frequency bins, from each set of simulated data.
+To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).
+Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
+As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
+
+The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
 We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.
-We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms, thus we should consider using the maximum number of bin shades that human perception is able to simultaneously discern.
-Lastly, the frequency loss with the standard frequency binning may be higher than for the quantile-based algorithm when a minimal number of bins are used; but in all three simulated data sets, this is reversed when seven frequency bins are used.
-Thus we recommend using the standard algorithm over the quantile-based algorithm for frequency binning because it had smaller frequency loss at the recommended seven bins, and additionally carries the benefit that equally dividing the bin count range into equally spaced bins creates a simpler shading interpretation than dividing by quantiles.
+We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms.
+This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous.
+Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot.
+Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used.
+But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.
 
 Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
-Frequency binning using the log counts did however behave very similarly, where log frequency loss decreased as the number of frequency bins increased.
-Thus the same advise to use seven frequency bins holds for shading based on log counts as well.
+The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data.
+Frequency binning using the log counts behaved very similarly, where log frequency loss decreased as the number of frequency bins increased.
+The primary difference uniform simulated data with bin counts that have low variability, are rendered virtually indistinguishable after a log transformation; thus changing the number of frequency bins does little to impact the log frequency loss.
+The same advise to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
 }
@@
 
-<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss from standard and quantile frequency binning of reduced binned data from each bivariate distribution.
-Spatial binning to produce the reduced binned data done with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).">>=
-### Loop over ncolors 1 to 7 and also over data distribution
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=8, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
+">>=
+### Loop over ncolors 1 to 10 and also over data distribution
 # find freq loss from standard, quantile, logCount
@@
   binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
-  for (ncol in 1:7){
+  for (ncol in 1:10){
     binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                         freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
-                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
-    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
+                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
+    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
     binoutFreq$DistType <- distrib[distidx]
@@
 
-freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Exponential"),]
+freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels= c("Standard", "Quantile", "Standard Log", "Quantile Log"))
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Exponential"),]
 p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
@@
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Freq Loss") + 
   theme(legend.position="bottom")
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Normal"),]
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Normal"),]
 p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
@@
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Freq Loss") + 
   theme(legend.position="bottom")
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Uniform"),]
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Uniform"),]
 p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
@@
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Freq Loss") + 
@@
 
-grid.arrange(p1,p2,p3, nrow=1)
-@
-
-<<FreqLossCompare2,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total log frequency loss from log frequency binning of reduced binned data from each bivariate distribution.
-I don't think this plot will make the final cut, but it is interesting that after logging the bin counts, the log freq loss for the uniform data does not improve with more bins (might as well use one shade for all bins)">>=
-### Loop over ncolors 1 to 7 and also over data distribution
-# find freq loss from standard, quantile, logCount
-freqLossSimDat <- NULL
-distrib <- c("Uniform", "Normal", "Exponential")
-for (distidx in 1:3){
-  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
-  for (ncol in 1:7){
-    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
-                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
-                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
-    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
-    binoutFreq$DistType <- distrib[distidx]
-    binoutFreq$ncolor <- ncol 
-    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
-  }
-}
-
-freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Exponential"),]
-p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = I(3), size=I(1)) +
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log")  & freqLossSimDat$DistType=="Exponential"),]
+p4 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color=FreqBinType, size=I(1)) +
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
   theme(legend.position="bottom")
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Normal"),]
-p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = I(3), size=I(1)) +
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Normal"),]
+p5 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
   xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
   theme(legend.position="bottom")
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Uniform"),]
-p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = I(3), size=I(1)) +
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Uniform"),]
+p6 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
+  scale_x_continuous(breaks=1:10) + ylim(0,40) +
   xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
@@
 
-grid.arrange(p1,p2,p3, nrow=1)
+grid.arrange(p1,p2,p3,p4,p5,p6, nrow=2)
 @
@@
 
-If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use seven distinct shades.
+If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
 This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
-Standard frequency binning is recommended over the quantile frequency binning because it displayed lower frequency loss for all three simulated binned data sets while using seven frequency bins.
+The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision falls to the interpretability of the resulting frequency bins.
 Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.

### 2015-04-09 Paper.Rnw
FreqLossDisplay-1

@@
 # Plot with Binned Frequencies
+
 p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
@@
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      breaks=unique(binoutFreqGroups[[1]]$freqgroup)) + 
+  scale_fill_gradient(low="#56B1F7", high="#132B43", 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5),
+                      breaks=unique(binoutFreqGroups[[1]]$freqgroup),
+                      labels=unique(binoutFreqGroups[[1]]$freqlabel)) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(2, "cm"),
                      panel.grid= element_blank())
+
 # Plot with Quantile Binned Frequencies

### 2015-04-09 Paper.Rnw
quantile binning color

@@
 # Plot with Binned Frequencies
-
+labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
 p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
@@
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", 
+  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43", 
                       guide = guide_legend(label.position="bottom", label.hjust=0.5),
                       breaks=unique(binoutFreqGroups[[1]]$freqgroup),
-                      labels=unique(binoutFreqGroups[[1]]$freqlabel)) + 
+                      labels=labels) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
@@
 # Plot with Quantile Binned Frequencies
-p3 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutQuantFreqGroups[[1]],
+labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
+p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
             main="Quantile Freq Binning") +
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutQuantFreqGroups[[1]]$freqgroup)) + 
+  scale_fill_manual("Frequencies", values=cols, 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5),
+                      labels=labels) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(2, "cm"),
                      panel.grid= element_blank()) 

### 2015-04-09 BinningLossFunctions.R

@@ -136,10 +136,7 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
     binout[[1]]$freqgroup <- round(QuantBin1d(cs, ncolor),2)
     quantbounds <- unique(quantile(cs, (0:ncolor)/ncolor))
     if (length(quantbounds)-1 < ncolor) warning("two or more quantiles of the data have same value due to many bins with equivalent frequencies, color will be rendered equivalently for bins with matching quantiles")
-    binout[[1]]$freqlabel <- paste("[",round(quantbounds[1],2),",",round(quantbounds[2],2),"]",sep="")
-    for (i in 2:(length(quantbounds)-1)){
-      binout[[1]]$freqlabel <- paste("(",round(quantbounds[i],2),",",round(quantbounds[i+1],2),"]",sep="")
-    }
+    binout[[1]]$freqlabel <- cut(binout[[1]]$binfreq, breaks=quantbounds, include.lowest=TRUE)
   }
   # Total Freq Loss
   binout[[2]]$totalFreqLoss <- sum((cs - binout[[1]]$freqgroup)^2)

### 2015-04-09 Paper.Rnw
legend text at the top

--- 2015-04-09 Paper.Rnw
@@
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", 
-                      guide = guide_legend(label.position="bottom", label.hjust=0.5),
+  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43", 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                           title.position = "top"),
                       breaks=unique(binoutFreqGroups[[1]]$freqgroup),
-                      labels=unique(binoutFreqGroups[[1]]$freqlabel)) + 
+                      labels=labels) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(2, "cm"),
+                     legend.key.width = unit(1, "cm"),
                      panel.grid= element_blank())
@@
 # Plot with Quantile Binned Frequencies
-p3 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutQuantFreqGroups[[1]],
+labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
+p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
             main="Quantile Freq Binning") +
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutQuantFreqGroups[[1]]$freqgroup)) + 
+  scale_fill_manual("Frequencies", values=cols, 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                           title.position = "top"),
+                      labels=labels) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
                      panel.grid= element_blank()) 
@@
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutLogFreqGroups[[1]]$freqgroup)) + 
+  scale_fill_gradient("Log Frequency", low="#56B1F7", high="#132B43", guide = guide_legend(label.position="bottom", label.hjust=0.5,  title.position = "top"), breaks=unique(binoutLogFreqGroups[[1]]$freqgroup)) +
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
                      panel.grid= element_blank())

### 2015-04-10 Paper.Rnw
Frequency binning work.  Figure 12 changed, quantile binning function fix, add scales library to setup

--- 2015-04-09 Paper.Rnw
@@
 library(MASS)
+library(scales)
 #source in binning functions
@@
 
-% which challenges?
 Data sets with millions of records and thousands of variables are not uncommon. 
@@
 
-% binning as a way out
 Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots that utilize raw data. 
@@
 In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions.
-With large data, scatterplots suffer from over-plotting of points, %-- i.e.\ a number of observations drawn in such close proximity that their points overla
-which masks relevant structure.
+With large data, scatterplots suffer from over-plotting of points, which masks relevant structure.
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
@@
 
-%Histograms are an example of plots which use binned variables.
-\hh{XXX No, histograms don't use binned variables in their standard definition, histograms bin variables to render the information.
-Weighted histograms (again, \citet{gold}) make use of binned data, because we can think of the tabular information as the sufficient statistics for histograms.}
 Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.
@@
 
-% \begin{figure}[hbtp]
-% <<conclusions, out.width='\\textwidth', fig.width=5, fig.height=3>>=
-% library(dbData)
-% connect <- dbConnect(dbDriver("MySQL"), user="2009Expo",
-% password="R R0cks", port=3306, dbname="baseball",
-% host="headnode.stat.iastate.edu")
-% pitch <- new("dataDB", co=connect, table="Pitching")
-% pitch.stats <- dbData(vars=list("G", "SO", "yearID"), pitch)
-% qplot(yearID, SO/G, alpha=I(0.2), weight=Freq, data=pitch.stats, colour=G) + xlab("season")
-% @
-% \caption{\caption{strikeout-rate}Scatterplot of rate of strikeouts per game over time.
-Each dot corresponds to a pitcher's seasonal average.
-Coloring indicates number of games played, alpha-blending is used additionally to emphasize high-density regions.
-There's clear break points in the strikeout rates over time.
-In 1902 XXXX}
-% \end{figure}
 Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: the joint density consists two distinct ridges following two lines with very different slopes.
@@
 Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and  bats.
-
-% two centers - almost figure~\ref{strikeout-rate} is numerically loss-free (binning is to integers, reduces the data by 16\%)
 
@@
 \subsection{Frequency Binning Specifications and Frequency Loss}
-\ktm{
+
 The reduced binned data from spatial binning contains the center and count information for all bins.
@@
 Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
-As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
+\ktm{ Figure~\ref{freqbinning} shows the binned scatterplots resulting from each type of frequency using five frequency bins.} As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
 
@@
 The same advise to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
-}
-
-<<FreqLossDisplay,echo=F,include=T,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning.
-If so, do the colors between freq bins need to be made more distinguishable?
-i.e.
-in the quantile binning the bins starting at 1, 8 and 36.5 are hard to differentiate because the shade is determined by the breaks.
-">>=
+
+\begin{figure}[hbtp]
+  \begin{minipage}{.3\linewidth}
+<<noFreqBin, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
 binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
-binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
-binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
-binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)
-
-# Plot with Orginal Frequencies
-p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binoutFreqGroups[[1]],
+qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
             main="No Freq Binning") +
@@
                      panel.grid= element_blank())
+@
+  \end{minipage}
+\hfil\begin{minipage}{.695\linewidth}
+<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=10>>=
+binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
+binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
+binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)
+binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=TRUE)
+
 # Plot with Binned Frequencies
-
+labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
 p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
@@
 # Plot with Quantile Binned Frequencies
-p3 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutQuantFreqGroups[[1]],
+labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
+p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
             main="Quantile Freq Binning") +
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutQuantFreqGroups[[1]]$freqgroup)) + 
+  scale_fill_manual("Frequencies", values=cols, 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                           title.position = "top"),
+                      labels=labels) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
                      panel.grid= element_blank()) 
-# Plot with LogCount Binned Frequencies
+# Plot with standard binned Log Frequencies
+labels = paste(unique(binoutLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutLogFreqGroups[[1]]$freqgroup), ")", sep="")
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
 p4 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]],
-            main="Log Freq Binning") +
+            main="Standard Log Freq Binning") +
     xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutLogFreqGroups[[1]]$freqgroup)) + 
+  scale_fill_gradient("Log Frequencies", low="#56B1F7", high="#132B43", 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                           title.position = "top"),
+                      breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
+                      labels=labels) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
                      panel.grid= element_blank())
 
-grid.arrange(p1,p2,p3,p4, nrow=1)
-@
+# Plot with quantile binned Log Frequencies
+labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
+p5 <- qplot(binxs, binys, geom="tile", fill=freqgroup,
+            data=binoutQuantLogFreqGroups[[1]],
+            main="Quantile Log Freq Binning") +
+    xlab("") + ylab("")+
+  scale_fill_gradient("Log Frequencies", low="#56B1F7", high="#132B43", 
+                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                           title.position = "top"),
+                      breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
+                      labels=labels) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
+                     panel.grid= element_blank())
+
+grid.arrange(p2,p3,p4,p5, nrow=2)
+@
+\end{minipage}
+\caption{\label{freqbinning}Simulated bivariate normal data spatially binned using standard algorithm with 5X5 unit bins and origin at (0,0).
+Binned Scatterplot with raw frequencies (left) and binned scatterplots with five frequency bins (right)."}
+\end{figure}
 

### 2015-04-10 BinningLossFunctions.R

@@ -123,8 +123,9 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
   # binning depends on type / log counts
   if (logCount)  cs <- log(binout[[1]]$binfreq+1)
   if(binType=="standard"){
-      width <- ceiling((max(cs)-min(cs))/ncolor)
-      binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),2)
+      #width <- ceiling((max(cs)-min(cs))/ncolor)
+      width <- (max(cs)-min(cs))/ncolor 
+      binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),1)
       binout[[1]]$freqlabel <- paste("(",round(binout[[1]]$freqgroup - width/2,2),
                                      ",",round(binout[[1]]$freqgroup + width/2,2),"]",sep="")
       #close interval for smallest counts
@@ -133,10 +134,10 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
                                      ",",round(min(binout[[1]]$freqgroup) + width/2,2),"]",sep="")
   } 
   if(binType=="quantile"){
-    binout[[1]]$freqgroup <- round(QuantBin1d(cs, ncolor),2)
+    binout[[1]]$freqgroup <- as.numeric(round(QuantBin1d(cs, ncolor),2))
     quantbounds <- unique(quantile(cs, (0:ncolor)/ncolor))
     if (length(quantbounds)-1 < ncolor) warning("two or more quantiles of the data have same value due to many bins with equivalent frequencies, color will be rendered equivalently for bins with matching quantiles")
-    binout[[1]]$freqlabel <- cut(binout[[1]]$binfreq, breaks=quantbounds, include.lowest=TRUE)
+    binout[[1]]$freqlabel <- cut(cs, breaks=quantbounds, include.lowest=TRUE)
   }
   # Total Freq Loss
   binout[[2]]$totalFreqLoss <- sum((cs - binout[[1]]$freqgroup)^2)

### 2015-04-10 Paper.Rnw
rescale y axis on origin offset plot
edits on freq loss plot and it's description writeup

@@
 
-<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a resolution of 2 units at various sized (2,4,6,8,10) square bins over the range of possible origin offsets.
+<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=4, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a resolution of 2 units at various sized (2,4,6,8,10) square bins over the range of possible origin offsets.
 The vertical gray line in each facet indicates where the origin is offset 1 unit below the minimum data value.">>=
@@
 lossOrigin <- read.csv("LossesByOriginBinwidth.csv",header=TRUE)
-
-qplot(originx,totalSpatialLoss,geom="path",group=distrib, colour=distrib,
+qplot(originx,totalSpatialLoss/1000,geom="path",group=distrib, colour=distrib,
       data=subset(lossOrigin,bintype=="standard"), size=I(1))+ 
@@
   geom_vline(xintercept = -1, colour=I("darkgray")) +
-  ylab("Net Spatial Loss") + xlab("Origin Offset")
+  ylab("Net Spatial Loss (in thousands of units)") + xlab("Origin Offset")
 @
@@
 Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
-\ktm{ Figure~\ref{freqbinning} shows the binned scatterplots resulting from each type of frequency using five frequency bins.} As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
+As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
+
+\ktm{ Figure~\ref{freqbinning} shows the binned scatterplots resulting from standard and quantile frequency binning algorithms using five frequency bins.
+Note that for standard binning, the shade is to be interpreted as an ordinal indicator of the groups based on bin frequencies, whereas for quantile binning the shade denotes groups based on frequency quantiles.
+This distinction explains why the shading of the binned scatterplot is invariant to the log transformation for quantile frequency binning but not for standard; because a monotone transformation will not alter groupings based on order, but will change groupings based on value magnitude.}
 
@@
 Frequency binning using the log counts behaved very similarly, where log frequency loss decreased as the number of frequency bins increased.
-The primary difference uniform simulated data with bin counts that have low variability, are rendered virtually indistinguishable after a log transformation; thus changing the number of frequency bins does little to impact the log frequency loss.
-The same advise to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
+The same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
 
 \begin{figure}[hbtp]
-  \begin{minipage}{.3\linewidth}
+  \begin{minipage}{.31\linewidth}
 <<noFreqBin, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
@@
   xlab("") + ylab("")+
-  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
+                      guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
+                                           title.position = "top")) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     panel.grid= element_blank())
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(.1,.1,.1,.1),"cm") )
 @
   \end{minipage}
-\hfil\begin{minipage}{.695\linewidth}
+\hfil\begin{minipage}{.685\linewidth}
 <<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=10>>=
@@
                      legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank())
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     legend.text=element_text(size=12) )
 
@@
                      legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank()) 
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     legend.text=element_text(size=12) ) 
 # Plot with standard binned Log Frequencies
@@
                      legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank())
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
+                     legend.text=element_text(size=12))
 
@@
                      legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank())
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     legend.text=element_text(size=12))
 
@@
 \caption{\label{freqbinning}Simulated bivariate normal data spatially binned using standard algorithm with 5X5 unit bins and origin at (0,0).
-Binned Scatterplot with raw frequencies (left) and binned scatterplots with five frequency bins (right)."}
+Binned Scatterplot with raw frequencies (left) and binned scatterplots with five frequency bins (right).
+Bin shades labeled by frequency range and average.
+}
 \end{figure}
 
-<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=8, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
 ">>=
@@
   xlab("Freq Bins") + ylab("Total Freq Loss") + 
-  theme(legend.position="bottom")
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
 
@@
   xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
-  theme(legend.position="bottom")
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
 
@@
   facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,40) +
+  scale_x_continuous(breaks=1:10) + ylim(0,2) +
   xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
-  theme(legend.position="bottom")
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
 

### 2015-04-10 BinningLossFunctions.R

@@ -125,16 +125,16 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
   if(binType=="standard"){
       #width <- ceiling((max(cs)-min(cs))/ncolor)
       width <- (max(cs)-min(cs))/ncolor 
-      binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),1)
-      binout[[1]]$freqlabel <- paste("(",round(binout[[1]]$freqgroup - width/2,2),
-                                     ",",round(binout[[1]]$freqgroup + width/2,2),"]",sep="")
+      binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),2)
+      binout[[1]]$freqlabel <- paste("(",round(binout[[1]]$freqgroup - width/2,1),
+                                     ",",round(binout[[1]]$freqgroup + width/2,1),"]",sep="")
       #close interval for smallest counts
       closeidx <- binout[[1]]$freqlabel == min(binout[[1]]$freqlabel)
-      binout[[1]]$freqlabel[closeidx] <- paste("[",round(min(binout[[1]]$freqgroup) - width/2,2),
-                                     ",",round(min(binout[[1]]$freqgroup) + width/2,2),"]",sep="")
+      binout[[1]]$freqlabel[closeidx] <- paste("[",round(min(binout[[1]]$freqgroup,1) - width/2),
+                                     ",",round(min(binout[[1]]$freqgroup) + width/2,1),"]",sep="")
   } 
   if(binType=="quantile"){
-    binout[[1]]$freqgroup <- as.numeric(round(QuantBin1d(cs, ncolor),2))
+    binout[[1]]$freqgroup <- as.numeric(round(QuantBin1d(cs, ncolor),1))
     quantbounds <- unique(quantile(cs, (0:ncolor)/ncolor))
     if (length(quantbounds)-1 < ncolor) warning("two or more quantiles of the data have same value due to many bins with equivalent frequencies, color will be rendered equivalently for bins with matching quantiles")
     binout[[1]]$freqlabel <- cut(cs, breaks=quantbounds, include.lowest=TRUE)

### 2015-04-10 Paper.Rnw
edits to section 6 intro

@@
 
-\ktm{
 The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction.
-Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing (Would like to use an O(n)-type argument, but I don't rightly understand this comp.
-sci.
-notation) and does not highlight large scale density structure.
+Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
 Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear.
@@
 For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution, $\alpha$.
-However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution is not problematic so long as a binwidth that is an integer multiple of $\alpha$ is selected.
-It was also shown that in this scenario, we should use an origin that is offset to $\alpha / 2$ units below the minimum data value leads to minimize spatial loss.
+However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution of $\alpha$ units can avoid the stripe illusion if a binwidth that is an integer multiple of $\alpha$ is selected.
+It was also shown that in this scenario, a reasonable default for the binning origin is offset to $\alpha / 2$ units below the minimum data value because it will minimize spatial loss for symmetric data and generally preform well even for non-symmetric data.
 
@@
 This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
-The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision falls to the interpretability of the resulting frequency bins.
-Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.
+The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision becomes one of how we wish the shade of a tile to be interpreted.
+Standard log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.
+We also discovered that if we elect to use quantile binning, the log transformation will have no effect on the binned scatterplot shading due to the identical quantile-based mapping to ordinal frequency groupings for all monotone transformations on bin counts.
 
@@
 Additionally we will investigate the relationship between scheduled and actual departure times for a full year of American commercial airline data.
-}
 

### 2015-04-10 Paper.Rnw
"L" used to both denote loss and index 2d bins.  change to uppercase script L to match \ell notation.  Also get rid of "+1" to bin counts in log freq loss since we can simply change notation to sum over only, non-empty bins.  This in turn makes the log frequency transformation for binning more simple

@@
 \usepackage{hyperref}
+\usepackage{mathrsfs}
 
@@
 The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
-With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,L\}$, where $L=J\cdot K$.
+With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,\mathscr{L}\}$, where $\mathscr{L}=J\cdot K$.
 
@@
 \begin{itemize}
-\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,L\}$.
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$.
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
@@
  Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
-We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
+We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, \mathscr{L} \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
 Figure~\ref{fig:FreqLossDemo} provides a visual example of a discrete color palette resulting from frequency binning.
@@
 \begin{eqnarray}\label{totfreqloss}
-L^F = \sum_{\ell=1}^{L} L_\ell^F = \sum_{\ell=1}^{L} (c_\ell-b_c(c_\ell))^2
+L^F = \sum_{\ell=1}^{\mathscr{L}} L_\ell^F = \sum_{\ell=1}^{\mathscr{L}} (c_\ell-b_c(c_\ell))^2
 \end{eqnarray}  
@@
 \begin{eqnarray}\label{logfrequency}
-L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
+L^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} L_\ell^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} (\log(c_\ell) - \log(b_c(c_\ell)))^2
 \end{eqnarray}
 %
-Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.
+where $\mathscr{L}^\ast$ is the index set for all non-empty bins, which is done to avoid asymptotic problems from log transforming bin counts of zero.
 
@@
 <<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=10>>=
-binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
-binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
-binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)
-binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=TRUE)
+freqgrps = 5
+binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=FALSE)
+binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=FALSE)
+binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=TRUE)
+binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=TRUE)
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:freqgrps)/freqgrps)
 
@@
 labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
-p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
+p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutFreqGroups[[1]],
             main="Standard Freq Binning") +
     xlab("") + ylab("")+
-  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43", 
+  scale_fill_manual("Frequencies", values=cols, 
                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
@@
 labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
-cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
 p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
@@
 labels = paste(unique(binoutLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutLogFreqGroups[[1]]$freqgroup), ")", sep="")
-cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
-p4 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]],
+p4 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutLogFreqGroups[[1]],
             main="Standard Log Freq Binning") +
     xlab("") + ylab("")+
-  scale_fill_gradient("Log Frequencies", low="#56B1F7", high="#132B43", 
+  scale_fill_manual("Log Frequencies", values=cols, 
                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
@@
 labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
-p5 <- qplot(binxs, binys, geom="tile", fill=freqgroup,
+p5 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup),
             data=binoutQuantLogFreqGroups[[1]],
@@
     xlab("") + ylab("")+
-  scale_fill_gradient("Log Frequencies", low="#56B1F7", high="#132B43", 
+  scale_fill_manual("Log Frequencies", values=cols, 
                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  

### 2015-04-10 BinningLossFunctions.R

@@ -121,20 +121,20 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
   cs <- binout[[1]]$binfreq
   # add frequency bin centers and bin labels of the form "(a,b]" to binout
   # binning depends on type / log counts
-  if (logCount)  cs <- log(binout[[1]]$binfreq+1)
+  if (logCount)  cs <- log(binout[[1]]$binfreq)
   if(binType=="standard"){
       #width <- ceiling((max(cs)-min(cs))/ncolor)
       width <- (max(cs)-min(cs))/ncolor 
-      binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),2)
+      binout[[1]]$freqgroup <- round(StandRectBin1d(cs, min(cs) , width),5)
       binout[[1]]$freqlabel <- paste("(",round(binout[[1]]$freqgroup - width/2,1),
                                      ",",round(binout[[1]]$freqgroup + width/2,1),"]",sep="")
       #close interval for smallest counts
       closeidx <- binout[[1]]$freqlabel == min(binout[[1]]$freqlabel)
-      binout[[1]]$freqlabel[closeidx] <- paste("[",round(min(binout[[1]]$freqgroup,1) - width/2),
+      binout[[1]]$freqlabel[closeidx] <- paste("[",round(min(binout[[1]]$freqgroup)- width/2,1),
                                      ",",round(min(binout[[1]]$freqgroup) + width/2,1),"]",sep="")
   } 
   if(binType=="quantile"){
-    binout[[1]]$freqgroup <- as.numeric(round(QuantBin1d(cs, ncolor),1))
+    binout[[1]]$freqgroup <- as.numeric(round(QuantBin1d(cs, ncolor),5))
     quantbounds <- unique(quantile(cs, (0:ncolor)/ncolor))
     if (length(quantbounds)-1 < ncolor) warning("two or more quantiles of the data have same value due to many bins with equivalent frequencies, color will be rendered equivalently for bins with matching quantiles")
     binout[[1]]$freqlabel <- cut(cs, breaks=quantbounds, include.lowest=TRUE)

### 2015-04-11 Paper.Rnw
Write-up of 6.1 and edits to 6.2

@@
 d1.ext <- xtd.data(d1)
+dbDisconnect(connect)
 @
@@
 
-For this data example we will revisit the baseball data used earlier in this paper.
-The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}.
-We find that the visual and numerical loss are in fact higher while using the random binning algorithm, but that the two binning algorithms are almost indistinguishable in terms of frequency loss.
-
-Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.
+<<BaseballBinningExamplePrep, echo=FALSE, eval=TRUE, include=FALSE, cache=TRUE>>=
+connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
+                     port=3306, user="dbaccess", dbname="baseball")
+pitchdat <- dbGetQuery(connect,  sprintf("select G, SO from Pitching"))
+dbDisconnect(connect)
+
+# Standard binning ~ 50 in each direction
+# so range(x)/50 = 100/50 = 2 , range(y)/50 = 500/50 = 10 
+# origin at (min(pitchdat$G)-0.5, min(pitchdat$SO)-0.5) = (0.5, -0.5)
+binout1 <- RectBin2d(pitchdat$G,pitchdat$SO,.5,-.5,2,10,type="standard")
+binout1[[2]]$totalSpatialLoss / nrow(pitchdat)
+# spatial loss per observation ~ 2.598362 (units?)
+
+# how much better is loss under origin at recommended offset?
+# compare loss on standardize scale with equivalent binning on standardized variables
+standG <- (pitchdat$G - mean(pitchdat$G))/sd(pitchdat$G)
+standSO <- (pitchdat$SO - mean(pitchdat$SO))/sd(pitchdat$SO)
+binoutSimpOrigin <- RectBin2d(standG,standSO,min(standG),min(standSO),2/mean(pitchdat$G),10/mean(pitchdat$SO),type="standard")
+binoutRecOrigin <- RectBin2d(standG,standSO,min(standG)-.5/mean(pitchdat$G),min(standSO)-.5/mean(pitchdat$SO),2/mean(pitchdat$G),10/mean(pitchdat$SO),type="standard")
+binoutRecOrigin[[2]]$totalSpatialLoss / binoutSimpOrigin[[2]]$totalSpatialLoss
+# spatial loss down 3% 
+standBaseballLoss2X10 <- binoutRecOrigin[[2]]$totalSpatialLoss
+standBaseballLoss2X10PerObs <- binoutRecOrigin[[2]]$totalSpatialLoss / nrow(pitchdat)
+@
+
+\ktm{For this data example we will revisit the baseball data used earlier in this paper.
+To construct the binned scatterplot to display the relationship between strikeouts and game counts, we begin by specifying the bin dimensions and origin for standard rectangular binning.
+Since the data are counts, each variable is recorded to an integer resolution so integer bin widths in each dimension of spatial bins will be used to avoid artificial density stripes.
+The data has only only 42583 observations, so we can bin finely without taking an inordinate amount of computation time.
+We elect to use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are 2 games X 10 strikeouts.
+The origin will be set to a half unit below the minimum value in each dimension, at ($0.5$,$-0.5$).
+}
+
+\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot that with a continuous shading of the raw bin densities.
+We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
+In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts.
+Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities.
+Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins then visualizes density on the log scale.
+Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure.
+}
+
+<<BaseballBinnedScatters2X10,echo=F,include=T,eval=T,fig.width=12, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
+">>=
+ # raw frequencies plot
+p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
+      main="Standard Freq Binning") +
+  xlab("") + ylab("")+
+  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43") + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     legend.text=element_text(size=12) )
+# # bin frequencies distribution hugely skewed
+# qplot(binfreq, geom="density", data= binout1[[1]], color=I("blue")) + theme_bw()
+# qplot(log(binfreq), geom="density", data= binout1[[1]], color=I("blue")) + theme_bw()
+
+## quantile freq binning
+binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=4, logCount=FALSE)
+labels = sort(unique(binoutQuantFreqGroups[[1]]$freqlabel))
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
+p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqlabel), data=binoutQuantFreqGroups[[1]],
+            main="Quantile Freq Binning") +
+  xlab("") + ylab("")+
+  scale_fill_manual("Frequencies", values=cols, 
+                    guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                         title.position = "top"),
+                    labels=labels) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     legend.text=element_text(size=12) )
+
+## standard log freq binning
+binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=4, logCount=TRUE)
+binoutLogFreqGroups[[1]]$freqlabel <- factor(binoutLogFreqGroups[[1]]$freqlabel, 
+                                             levels = unique(binoutLogFreqGroups[[1]]$freqlabel)[c(3,5,2,4,1)])
+labels = levels(binoutLogFreqGroups[[1]]$freqlabel)
+cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
+
+p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqlabel), data=binoutLogFreqGroups[[1]],
+      main="Standard Log Freq Binning") +
+  xlab("") + ylab("")+
+  scale_fill_manual("Log Frequencies", values=cols, 
+                    guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                                         title.position = "top"),
+                    labels=labels) + 
+  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+                     legend.key.width = unit(1, "cm"),
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     legend.text=element_text(size=12) )
+
+# plot three binned scatterplots (raw, standard log freq, quantile freq)
+grid.arrange(p1,p2,p3, nrow=1)
+
+@
+
+\ktm{The interpretability of the net spatial loss in this scenario suffers because the units and scales of the two variables differ.
+To remedy this, the two variables can be standardized then binned equivalently by rescaling the bin dimension, where the net spatial loss is interpreted as a distance in units of standard deviations.
+The net spatial loss for the standardized data is \Sexpr{round(standBaseballLoss2X10,4)} standard deviations, an average of \Sexpr{round(standBaseballLoss2X10PerObs,4)} standard deviation per data point.
+We also find that we had approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification than had we naively set the binning origin at the minimum data values, ($1$,$0$).
+}
+
+% Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.
 This is apparent in the order of colored lines based on the Strike Out binning size.
@@
 Whereas the visual loss function sums loss over bin, thus as the bin width increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.
-
-% \begin{figure}[hbtp]
-% \centering
-% <<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
-% 
-% # connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-% #                      password="R R0cks", port=3306, dbname="baseball", 
-% #                      host="headnode.stat.iastate.edu")
-% # pitch <- new("dataDB", co=connect, table="Pitching")
-% # d1 <- dbData(pitch, vars=c( "G", "SO"))
-% # bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
-% # 
-% # library(multicore)
-% # lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
-% # lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
-% # 
-% # lossesRdm <- cbind(bins, lossesRdm)
-% # lossesStd <- cbind(bins, lossesStd)
-% # 
-% # losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
-% # write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
-% 
-% losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
-% lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
-% lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
-% lossmelt$y <- as.factor(lossmelt$y)
-% levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
-% qplot(x=x, y=value,
-%       group=interaction(y, Algorithm),
-%       data=lossmelt, geom="line",
-%       colour=y, size=I(1),
-%       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
-%         theme_bw() + theme(legend.position="bottom") + 
-%         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-%         facet_grid(variable ~ Algorithm, scales = "free") 
-% 
-% # qplot(x=x, y=value,
-% #       group=interaction(y, Algorithm),
-% #       data=lossmelt, geom="line",
-% #       colour=y, size=I(1), linetype=Algorithm,
-% #       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
-% #         theme_bw() + theme(legend.position="bottom") + 
-% #         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-% #         facet_grid(. ~ variable, scales = "free") 
-% 
-% @
-% \caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
-% \end{figure}
-
-%\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
-%\noindent
-%Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
-
-\hh{the  code following this comment in the Rnw file is, I think, broken beyond repair - but we should include a table like the one it produced in the paper.
+% 
+% 
+% % \begin{figure}[hbtp]
+% % \centering
+% % <<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
+% % 
+% % # connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
+% % #                      password="R R0cks", port=3306, dbname="baseball", 
+% % #                      host="headnode.stat.iastate.edu")
+% % # pitch <- new("dataDB", co=connect, table="Pitching")
+% % # d1 <- dbData(pitch, vars=c( "G", "SO"))
+% % # bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
+% % # 
+% % # library(multicore)
+% % # lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
+% % # lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
+% % # 
+% % # lossesRdm <- cbind(bins, lossesRdm)
+% % # lossesStd <- cbind(bins, lossesStd)
+% % # 
+% % # losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
+% % # write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
+% % 
+% % losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
+% % lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
+% % lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
+% % lossmelt$y <- as.factor(lossmelt$y)
+% % levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
+% % qplot(x=x, y=value,
+% %       group=interaction(y, Algorithm),
+% %       data=lossmelt, geom="line",
+% %       colour=y, size=I(1),
+% %       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
+% %         theme_bw() + theme(legend.position="bottom") + 
+% %         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
+% %         facet_grid(variable ~ Algorithm, scales = "free") 
+% % 
+% % # qplot(x=x, y=value,
+% % #       group=interaction(y, Algorithm),
+% % #       data=lossmelt, geom="line",
+% % #       colour=y, size=I(1), linetype=Algorithm,
+% % #       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
+% % #         theme_bw() + theme(legend.position="bottom") + 
+% % #         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
+% % #         facet_grid(. ~ variable, scales = "free") 
+% % 
+% % @
+% % \caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
+% % \end{figure}
+% 
+% 
+% %\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
+% %\noindent
+% %Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
+% 
+% \hh{the  code following this comment in the Rnw file is, I think, broken beyond repair - but we should include a table like the one it produced in the paper.
 - we will have different losses to report, though}
-
-% 
-% <<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
-% loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
-% loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
-% loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
-% loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
-% 
-% loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% 
-% table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
-%                                c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
-%                                c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
-%                                c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
-%                                c(binwidth=1, loss1[5:7]*100, loss1[8])))
-% names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
-% print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
-% @
-
-We may also use this example to highlight the advantages of random binning.
+% 
+% % 
+% % <<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
+% % loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
+% % loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
+% % loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
+% % loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
+% % 
+% % loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% % loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% % loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% % loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
+% % 
+% % table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
+% %                                c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
+% %                                c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
+% %                                c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
+% %                                c(binwidth=1, loss1[5:7]*100, loss1[8])))
+% % names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
+% % print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
+% % @
+% 
+% 
+% We may also use this example to highlight the advantages of random binning.
 The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident.
@@
 This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
-
-\textbf{Are we sure that this is because of an odd bin widths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
-
-\hh{This is a valid concern, but we have to fix the notation at the beginning first, before we can address it.}
-
-\begin{figure}[hbtp]
-\centering
-\subfloat[(2, 2) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
-}\\
-\subfloat[(5, 25) tiling of G, SO data]{
-\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
-}
-\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chosen bin width is not a multiple of the data resolution.}
-\end{figure}
+% 
+% 
+% \textbf{Are we sure that this is because of an odd bin widths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
+% 
+% \hh{This is a valid concern, but we have to fix the notation at the beginning first, before we can address it.}
+% 
+% \begin{figure}[hbtp]
+% \centering
+% \subfloat[(2, 2) tiling of G, SO data]{
+% \includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
+% }\\
+% \subfloat[(5, 25) tiling of G, SO data]{
+% \includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
+% }
+% \caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chosen bin width is not a multiple of the data resolution.}
+% \end{figure}
+% 
+%
 
@@
 The cloud of outliers halfway between the two main structures is potentially interesting, since no immediate explanation comes to mind, and would be worthy of a follow-up investigation.
+
 What is not apparent in the plot on the left, is some fine-level structure that the plot based on all of the data shows.
-A close inspection of the plot on the right hand side reveals  darker colored vertical lines at 30 minute intervals.
+\ktm{Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not lead to artificial density stripes.} A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
 It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.

### 2015-04-12 Paper.Rnw
misc. edits

@@
 Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.
-
-\hh{discussion of sufficient statistics for graphics could go here.
-Idea: the problems with large data in scatterplots all arise from overplotting, which is a form of implicit data aggregation.
-In order to keep track of overplotting, we switch to a weighted form of displays and explicitly keep track of overplotting.
-The aggregated data forms the sufficient statistics for a scatterplot, i.e.
-provide all of the information necessary to render it.
-The scatterplot therefore does not display more than that information.
-Making the data aggregation explicit allows us to calculate the loss we experience.}
 
@@
 
-While binning data has some notable advantages, it does result in loss of some information.
+The problems with large data in scatterplots arise from overplotting, which is a form of implicit data aggregation.
+In order to keep track of overplotting, we switch to a weighted form of displays and explicitly keep track of overplotting.
+The reduced binned data carries the sufficient information necessary to render the binned scatterplot.
+Making the data aggregation explicit allows us to calculate the loss we experience.
+
 Using a small number of bins removes much of the relevant information in the data, while an extremely large number of bins may not reduce the data size or complexity to a sufficient degree.
@@
 
-The \textit{data resolution} is defined as the \hh{smallest} increment between \hh{successive} data values.
+The \textit{data resolution} is defined as the smallest increment between successive data values.
 To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
@@
 To remedy this, the two variables can be standardized then binned equivalently by rescaling the bin dimension, where the net spatial loss is interpreted as a distance in units of standard deviations.
-The net spatial loss for the standardized data is \Sexpr{round(standBaseballLoss2X10,4)} standard deviations, an average of \Sexpr{round(standBaseballLoss2X10PerObs,4)} standard deviation per data point.
+The net spatial loss for the standardized data is \Sexpr{round(standBaseballLoss2X10,2)} standard deviations, an average of \Sexpr{round(standBaseballLoss2X10PerObs,4)} standard deviation per data point.
 We also find that we had approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification than had we naively set the binning origin at the minimum data values, ($1$,$0$).
@@
 
-Figure~\ref{airline-scatter} shows examples of two scatterplots of scheduled versus actual departure times.
-The plot on the left shows a sample of one million of those records.
+\ktm{Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
+The plot on the left shows the scatterplot from a sample of one million of those records.
 Even while using alpha blending this results in a severely over-plotted graph.
-On the right is a binned scatterplot of the complete data binned at 1-minute intervals.
-This does not have any spatial loss, as the times are recorded by minute.
-The 1-minute binning also reduces the data file to 176,384 individual records, less than 3\% of the original data.
-Both plots show the same big picture patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the line of identity.
-Scheduled departure times past 6 am in the morning are much more common than earlier flights. It is much more likely for a  flight to be delayed than to leave early,
-leading to the wash-out effect above the line, that is getting thinner with increasing delays.
+On the right is a binned scatterplot of the reduced data from standard binning at 1-minute intervals.
+If the origin is offset by a half minute for both departures and arrivals, this binning does not have any spatial loss, as all observations will be centered within bins.
+The 1-minute standard binning also reduces the data to 176,384 individual records, less than 3\% of the original data.
+}
+
+Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the line of identity.
+Scheduled departure times past 6 am in the morning are much more common than earlier flights.
+It is much more likely for a  flight to be delayed than to leave early, leading to the wash-out effect above the line, that is getting thinner with increasing delays.
 The range of delays on usual days starts at about one hour at 6 am and increases during the day to about 2 hours.
@@
 
-What is not apparent in the plot on the left, is some fine-level structure that the plot based on all of the data shows.
-\ktm{Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not lead to artificial density stripes.} A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
+What is not apparent in the alpha blended scatterplot, is some fine-level structure that the plot based on all of the data shows.
+\ktm{Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not cause any artificial density stripes.} A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
 It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.
@@
 The plot on the left is based on a sample of the data, the plot on the right shows all flights.
-The big patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}
+The large scale distributional patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}
 \end{figure}

### 2015-04-12 Paper.Rnw
Add brief discussion of unit interpretability of net spatial loss if scales and/or units differ between the two  binned dimensions.

@@
 
-% The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable.
+\ktm{The net spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
+If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
+However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard devations.
+}
+
+%but requires a reference scale on which the magnitude may be interpretable.
 We  use the spatial loss from a ``worst-case'' binning scenario as a reference value.

### 2015-04-13 Paper.Rnw
Edits to initial description of Spatial and Frequency Loss to make them more consistent with each other and the language in the rest of the paper.

@@
 \begin{itemize}
-\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$.
+\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$.
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
 
-\item {\it Frequency Loss}, $L^\text{F}$, occurs when frequency information is not rendered, such as in standard scatterplots,  or not rendered accurately, e.g.\ when a discrete color scale is used to display bin counts, the entire ranges of counts is mapped to the same color; thus losing precision in the display of frequencies.
+\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely.
+A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
+This is analogous to the fundimental problem for overplotted standard scatterplots where frequency is masked by other points.
 \end{itemize}

### 2015-04-13 Paper.Rnw
Move/resize airline plots and small edits to their descriptions.

@@
 
-Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the line of identity.
-Scheduled departure times past 6 am in the morning are much more common than earlier flights.
-It is much more likely for a  flight to be delayed than to leave early, leading to the wash-out effect above the line, that is getting thinner with increasing delays.
-The range of delays on usual days starts at about one hour at 6 am and increases during the day to about 2 hours.
-The few number of flights before 6 am are also visible in both plots.
-The triangle of observations on the bottom right visible in both plots is nothing but an artifact of the data collection consisting of flights that are scheduled before midnight, but are delayed to departures past midnight.
-The cloud of outliers halfway between the two main structures is potentially interesting, since no immediate explanation comes to mind, and would be worthy of a follow-up investigation.
-
-What is not apparent in the alpha blended scatterplot, is some fine-level structure that the plot based on all of the data shows.
-\ktm{Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not cause any artificial density stripes.} A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
-It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.
-
 \begin{figure}[hbtp]\centering
@@
 
+Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the line of identity.
+Scheduled departure times past 6 am in the morning are much more common than earlier flights.
+It is much more likely for a  flight to be delayed than to leave early, leading to the wash-out effect above the line, that is getting thinner with increasing delays.
+The range of delays on usual days starts at about one hour at 6 am and increases during the day to about 2 hours.
+The few number of flights before 6 am are also visible in both plots.
+The triangle of observations on the bottom right visible in both plots is nothing but an artifact of the data collection consisting of flights that are scheduled before midnight, but are delayed to departures past midnight.
+The cloud of outliers halfway between the two main structures is potentially interesting, since no immediate explanation comes to mind, and would be worthy of a follow-up investigation.
+
+What is not apparent in the alpha blended scatterplot, is some fine-level structure that the plot based on all of the data shows.
+\ktm{Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not cause any artificial density stripes.} A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
+It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.
+
 \begin{figure}[hbtp]\centering
-\includegraphics[keepaspectratio=true, height=.43\textheight]{images/AirlineStdBinning5mins.pdf}
-\includegraphics[keepaspectratio=true, height=.43\textheight]{images/AirlineStdBinning15mins.pdf}
+\includegraphics[width=.49\linewidth,keepaspectratio=true]{images/AirlineStdBinning5mins.pdf}
+\includegraphics[width=.49\linewidth,keepaspectratio=true]{images/AirlineStdBinning15mins.pdf}
 \caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
@@
 \end{figure}
+
+Figure~\ref{airline5mins-binning} displays the binned scatterplots for the flight departure data with larger bins.
 Binning data by five-minute intervals produces a more high-level summary of the relationship between actual and scheduled departure time, though it necessarily obscures some of the finer details.
@@
 @
-\section{Conclusion}
+
+%--------------------------------------------------------------------------------------
+
+\section{Conclusions and Future Work}
+
 Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
-The loss due to binning is the cost of reducing the data set to a more manageable size, but some of that loss can be recovered with additional computational investment.
-
-We have presented two algorithms for binning; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient.
-The choice of algorithm is a trade-off between speed and accuracy, and may be highly data set dependent.
-It is possible that there are computational modifications which could improve the accuracy of the random algorithm, but these modifications are likely to require additional computational operations.
-
-% \begin{subsection}{Numerical Comparison of Binning Algorithms}
-<<PrintTable,echo=F,include=FALSE, dependson='LossTableGraphs'>>=
-  print(tab, include.rownames=FALSE)
-@
-% \end{subsection}
+Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.
+
+We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient.
+The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid problematic visual features such as binning caused stripping.
+In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value.
 

### 2015-04-13 Paper.Rnw
Frequency loss conclusions

@@
 We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient.
-The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid problematic visual features such as binning caused stripping.
+The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid generating problematic visual features.
 In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value.
+
+We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading.
+It is recommended to use less than ten seperate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins.
+Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.
 

### 2015-04-15 Paper.Rnw
Move a paragraph and color new changes

@@
 
-Using a small number of bins removes much of the relevant information in the data, while an extremely large number of bins may not reduce the data size or complexity to a sufficient degree.
+\ktm{Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
+Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
+The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information.
+By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points.
+This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c.
+}
+
+\ktm{By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures.
+Using a small number of large bins may mask the real signal in the data, while an extremely large number of small bins may not sufficiently smooth over {\it noise} inherent in any real data set.
 Figure~\ref{binning} gives an overview of a data set and binned representations using different numbers of bins, demonstrating the loss of information with increasing bin size.
+}
 
@@
 We distinguish two sources of loss in the construction of a binned scatterplot:
+
+\ktm{
 \begin{itemize}
@@
 % For the purposes of numerical assessment of frequency loss, we calculate the baseline frequency loss as the loss corresponding to a minimal bin width given by the smallest non-zero absolute difference of successive values in the original data set, as shown in Figure~\ref{binning}.
-
-\paragraph{Traditional Scatterplots and Frequency Information: (old writeup - to be reworked or removed)}
-
-Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
-Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
-The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information.
-By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points.
-This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c.
-
-By additionally increasing the bin width, we  provide increasingly higher-level summaries of the data by  smoothing over local structures.
-For small bin widths these structures are likely to be  {\it noise} inherent in any real data set, while removing structures associated with  larger bin widths likely masks real signal in the data.
 
@@
 
-Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
+\ktm{Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
 Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.
-
-We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient.
+}
+
+\ktm{We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient.
 The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid generating problematic visual features.
 In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value.
-
-We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading.
+}
+
+\ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading.
 It is recommended to use less than ten seperate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins.
 Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.
+}
 
@@
 \begin{appendix}
-\section{Mathematical Considerations}
+\section{Appendix}
 

### 2015-04-15 Paper.Rnw
couple of comments in section 5.2

@@
 \newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike
-\newcommand{\svp}[1]{{\color{yellow}#1}} %yellow: Susan (It's more fun that way - comment roulette)
 
@@
 \label{Intro}
+\hh{General: write out any positive integer less than or equal to ten out in words, even if there is the word `unit' behind it. So it's one unit, and not 1 unit.}
 
@@
 While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data.
-\ktm{ Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.
+ Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.
 Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
-Another pertinent argument for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}
+ \ktm{Another pertinent argument for using larger bins is that it will smooth over  larger regions if visualizing the large scale density structure is of primary interest.}
+\hh{This is an argument along the lines of the bias vs variability tradeoff. Extend on it, don't put it here as a throwaway sentence. }
 
@@
 # write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
+
 allbindat <- read.csv("StripeBinData.csv",header=TRUE)
@@
 For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
-It can be shown analytically that for data with a resolution of $\alpha$ units that is spread symmetrically within bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum value (see proof in Appendix~\ref{proof:offset}).
+It can be shown analytically that for data with a resolution of $\alpha$ units with a  symmetric distribution within each of the bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum value (see proof in Appendix~\ref{proof:offset}).
+
+\hh{XXX which bin size are you talking about? add this, otherwise the statement is wrong XXX}
+
+\hh{`$\alpha/2$ origin offset' this an implicit naming.
+You have to make it either explicit or write this out.
+At the moment it is imprecise, because you mean this offset to be below the minimal value of the data.
+XXX }
 
 In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimensions that are integer multiples of $\alpha$.
+\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss.
+it's a bit confusing with the two minima floating around.
+It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
 Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to one binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
 For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit below the minimum data value was found to minimize the net spatial loss for each considered bin size (2, 4, 6, 8, and 10 units).
+\hh{XXX - no!
+the value at -1 is close, but the dots are in the actual minimal values.
+This is only a single example - the expected minimum might still be in -1.
+}
 For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals.
@@
 lossOrigin <- read.csv("LossesByOriginBinwidth.csv",header=TRUE)
+mins <- ddply(lossOrigin, .(widthx, bintype, distrib), summarise,
+              minx = max(originx[which(totalSpatialLoss == min(totalSpatialLoss))]),
+              minloss = min(totalSpatialLoss))
 qplot(originx,totalSpatialLoss/1000,geom="path",group=distrib, colour=distrib,
       data=subset(lossOrigin,bintype=="standard"), size=I(1))+ 
-  facet_grid(distrib~widthx, scales="free_x") + theme_bw() + 
+  facet_grid(distrib~widthx, scales="free_x", space="free_x") + theme_bw() + 
   theme(panel.grid= element_blank(), legend.position="none")+
   geom_vline(xintercept = -1, colour=I("darkgray")) +
-  ylab("Net Spatial Loss (in thousands of units)") + xlab("Origin Offset")
+  geom_point(aes(x=minx, y=minloss/1000), data=subset(mins, bintype=="standard"), size=3.5) +
+  ylab("Net Spatial Loss (in thousands of units)") + xlab("Origin Offset") + 
+  scale_x_continuous(breaks = seq(-10, 0, by = 1))
 @
@@
 Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
-As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.
-
-\ktm{ Figure~\ref{freqbinning} shows the binned scatterplots resulting from standard and quantile frequency binning algorithms using five frequency bins.
+As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities \citep{colorperception}.
+
+\ktm{ Figure~\ref{freqbinning} shows binned scatterplots resulting from standard and quantile frequency binning algorithms using five frequency bins.
 Note that for standard binning, the shade is to be interpreted as an ordinal indicator of the groups based on bin frequencies, whereas for quantile binning the shade denotes groups based on frequency quantiles.

### 2015-04-15 Paper.Rnw
Add "binning for distributional approximation" sources to beginning of Section 3.

--- 2015-04-15 Paper.Rnw
@@
 
+\ktm{Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}.
+More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations (\citealp{scott1979} and \citealp{scott1992}).
+}
+
 Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.
@@
 We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
-While hexagonal binning has been shown to have slightly better graphical properties \citep{scott2009multivariate}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
+While hexagonal binning has been shown to have slightly better graphical properties \citep{scott1992}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
 We will however only consider binning in up to two dimensions, $X$ and $Y$.

### 2015-04-15 Paper.Rnw
section 5.2 comments

@@
 The reduced binned data from spatial binning contains the center and count information for all bins. 
+\hh{This sounds like it is set in stone - it's not.
+We can define which information is kept for each of the bins.
+Make this an assumption rather than a statement and define what kind of center you are talking about - visual or numeric.
+}
+
 The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.
@@
 As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities \citep{colorperception}.
+
+\hh{XXX I don't think that figure 12 is doing its job - and for that matter, we should talk about what its job is.
+There are too many different things going on in it, and the figure might feed better into your argument when things are separated out.
+}
+
+\hh{XXX Let's try a first picture with a continuous scale and three discrete binned frequencies of four, seven, and ten bins.
+This should drive home the message that Healey is right, and we can't distinguish more than seven shades anyway, i.e.
+we don't see a difference between the continuously shaded plot and the one with discrete shades.
+This re-enforces our approach to treat color as a categorical variable, because we preceive it that way (otherwise why would we?)}
+
+\hh{XXX figure 13 is a plot with a screaming message: never ever use a raw data mapping to show color.
+use quantiles or log frequencies.
+It doesn't scream that at the moment, so emphasize it in the text.}
+
+\hh{XXX the rest of figure 12 should probably go into a figure 14 - to show that quantile binning and log frequency binning yield very similar results.
+We need to fix the legend.
+Right now they are atrocious (and I am aware that I am responsible for them, so I can say that :) )}
 

### 2015-04-15 Paper.Rnw
last (?) new paragraph: choosing the binwidth
add references to binning loss work by Scott in section 4

--- 2015-04-15 Paper.Rnw
@@
 Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points.
+\ktm{The idea of loss from one-dimensional binning was explored by Scott using mean squared error as a loss function where the number of bins could be optimized for the construction of histograms \citep{scott1979}.
+He later extended this discussion to two-dimensional binning, where he compared the mean squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
+We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section\ref{GenBinning} above.
+}
 

### 2015-04-15 Paper.Rnw

--- 2015-04-15 Paper.Rnw
@@
 % }
+
+%-----------------------------------------------------------------------------
+\subsection{Choice of bin width}
 

### 2015-04-15 Paper.Rnw
more accurate description of *near* minimum spatial loss for recommended origin offset

@@
 \label{Intro}
-\hh{General: write out any positive integer less than or equal to ten out in words, even if there is the word `unit' behind it. So it's one unit, and not 1 unit.}
-
-\hh{
-\begin{itemize}
-\item Additional literature: binning papers: standard binning, e.g.: Karl Pearson (1895) ``Contributions to the Mathematical Theory of Evolution. II. Skew Variation in Homogeneous Material"\\
-David W. Scott (1979). "On optimal and data-based histograms". Biometrika 66 (3): 605–610
-
-quantile binning: there's a lot of implementations (e.g.
+% \hh{General: write out any positive integer less than or equal to ten out in words, even if there is the word `unit' behind it. So it's one unit, and not 1 unit.}
+% 
+% \hh{
+% \begin{itemize}
+% \item Additional literature: binning papers: standard binning, e.g.: Karl Pearson (1895) ``Contributions to the Mathematical Theory of Evolution. II. Skew Variation in Homogeneous Material"\\
+% David W. Scott (1979). "On optimal and data-based histograms". Biometrika 66 (3): 605–610
+% 
+% quantile binning: there's a lot of implementations (e.g.
 cut2 in package Hmisc, the rattle package, the package extracut.
 It would be interesting to check, if they do the same things (probably not) and mention that there is disagreement in the implementations).
-
-As a reference I would check in 
-Scott, D. W. 1992.
-Multivariate Density Estimation: Theory, Practice, and Visualization. New York: Wiley.
-
-\item make sure to verify that the binning algorithms agree with the literature (and check that the code is doing the same thing as described in the paper).
-You can include a bit of discussion on the problem that histograms are over-parameterized - it's a known problem, but I can't find any paper on it at the moment.
-
-\end{itemize}
-}
-
-\ktm{I have read through the sections of Pearson (1895), Scott (1979) and Scott (1992) that pertained to binning.
+% 
+% As a reference I would check in 
+% Scott, D. W. 1992.
+% Multivariate Density Estimation: Theory, Practice, and Visualization. New York: Wiley.
+% 
+% \item make sure to verify that the binning algorithms agree with the literature (and check that the code is doing the same thing as described in the paper).
+% You can include a bit of discussion on the problem that histograms are over-parameterized - it's a known problem, but I can't find any paper on it at the moment.
+% 
+% \end{itemize}
+% }
+% 
+% \ktm{I have read through the sections of Pearson (1895), Scott (1979) and Scott (1992) that pertained to binning.
 I will work on integrating them into the lit review next.
 Scott talked much about the optimal choice of bin width based on MSE but there was not much talk about what to do with points at the boundaries of bins (he effectively assumed infinite precision in the measurement process to avoid it).
-
-I will need to search more to find papers about the binning algorithms and the boundaries they set.
+% 
+% I will need to search more to find papers about the binning algorithms and the boundaries they set.
 The \texttt{cut} function allows for intervals that are closed on either the left or right and also allows the points on the outermost open bound to be included in the last bin.
-
-The \texttt{cut2} function allows the same type of quantile binning as discussed in Scott (1992).}
+% 
+% The \texttt{cut2} function allows the same type of quantile binning as discussed in Scott (1992).}
 
@@
 Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to one binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
-For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit below the minimum data value was found to minimize the net spatial loss for each considered bin size (2, 4, 6, 8, and 10 units).
-\hh{XXX - no!
+\ktm{The round glyphs indicade the origin offset where the net loss reached an absolute minimum.
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within 0.5\% of the magnitude of the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
 the value at -1 is close, but the dots are in the actual minimal values.

### 2015-04-15 Paper.Rnw
add code (commented) to code chuck for calculating % increase from minimum loss values for offset at -1 .  Also edit wording in paragraph slightly. May still need phrasing edits.

@@
 \ktm{The round glyphs indicade the origin offset where the net loss reached an absolute minimum.
-For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within 0.5\% of the magnitude of the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
 the value at -1 is close, but the dots are in the actual minimal values.
@@
 }
-For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals.
+For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
 
@@
               minloss = min(totalSpatialLoss))
+
+# #Code to show percent of minimum loss at offset -1
+# spatlossatorigin1 <- data.frame(NULL)
+# index=0
+# distindex = 0
+# distrib <- c("Uniform", "Normal", "Exponential")
+# widths <- seq(alpha,5*alpha,by=alpha)
+# for (dat in 1:3){
+#   distindex <- distindex + 1
+#   for (width in widths){
+#       index <- index+1
+#       binout <- RectBin2d(datlist[[dat]]$roundx,datlist[[dat]]$roundy,-1,-1,width,width,type="standard")
+#       binout[[2]]$bintype <- "standard"
+#       binout[[2]]$distrib <- distrib[distindex]
+#       spatlossatorigin1 <- rbind(spatlossatorigin1, binout[[2]])
+#   }
+# }
+# offset1 <- ddply(spatlossatorigin1, .(widthx, bintype, distrib), summarise,
+#               lossat1 = totalSpatialLoss)
+# # Check percentage of minimum at -1
+# offset1$PercofMin <- (offset1$lossat1+0.000000000001) / (mins$minloss[which(mins$bintype =="standard")]+0.000000000001)
+# offset1
+
 qplot(originx,totalSpatialLoss/1000,geom="path",group=distrib, colour=distrib,

### 2015-04-15 Paper.Rnw
Attempt to clarify notation and wording for "origin offsets".  I think the problem was univariate language for a bivariate idea. Changes in red (blue comments not deleted, just hidden)

@@
 \caption{\label{DataReductionExample}Original, Binned and Reduced Binned Data Tables, with data storage sizes.
-Binned using standard rectangular approach with origin at (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
+Binned using standard rectangular approach with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
 }
@@
 The \textit{data resolution} is defined as the smallest increment between successive data values.
-To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units.
+To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
@@
 
-To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of 2 units.
+To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension.
 Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
-Under standard binning we see the white-space gaps with 1 X 1 unit bins, vertical and horizontal stripes with 5 X 5 unit bins and the appropriate view of an evenly spread density with 4X4 unit bins.
+Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.
 Note that random binning is effective at smoothing out the stripped patterns when non-integer multiples of the data resolution are selected.
@@
 For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density in outermost bins.
-Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of 10 by 10 unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9).
+Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9).
 The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions.
-The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about 7\% lower than for the (-9,-9) origin.
+The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about seven percent lower than for the (-9,-9) origin.
 For data with very fine resolution, it is a good idea to set the origin at the minimum value for each dimension.
@@
 For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
-It can be shown analytically that for data with a resolution of $\alpha$ units with a  symmetric distribution within each of the bins, the origin that minimizes net spatial loss is located $\alpha/2$ units below the minimum value (see proof in Appendix~\ref{proof:offset}).
-
-\hh{XXX which bin size are you talking about? add this, otherwise the statement is wrong XXX}
-
-\hh{`$\alpha/2$ origin offset' this an implicit naming.
+\ktm{We define the \textit{origin offset} for each dimension, ($o_x$, $o_y$), such that
+%
+\begin{eqnarray}\label{OriginOffset}
+(\beta_{0,x},\beta_{0,y}) = (x_{(1)}, y_{(1)}) - (o_x, o_y),
+\end{eqnarray}  
+%
+where $(\beta_{0,x},\beta_{0,y})$ is the bivariate bin origin, $x_{(1)}$ and $y_{(1)}$ are the minimum data value for each dimension, and $(o_x, o_y)$ is the origin offset.
+Thus the offset indicates the number of units in each dimension to shift the origin below the highest possible origin.
+}
+
+\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimesions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
+}
+
+%\hh{`$\alpha/2$ origin offset' this an implicit naming.
 You have to make it either explicit or write this out.
@@
 
-In practice the $\alpha/2$ origin offset is found to be a reasonable choice for minimizing spatial loss for symmetric bivariate data at resolution $\alpha$ while using bin dimensions that are integer multiples of $\alpha$.
-\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss.
+In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
+%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss.
 it's a bit confusing with the two minima floating around.
 It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
-Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin is slide from the minimum value to one binwidth below the minimum value using the coarse simulated data with a resolution of 2 units.
-\ktm{The round glyphs indicade the origin offset where the net loss reached an absolute minimum.
-For the two symmetrically distributed data sets, normal and uniform, the origin offset of 1 unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
+\ktm{Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm.
+Note that for simplicity, only origin offsets for with equal offset in each dimension are included in this simulation.
+Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees.
+The round glyphs indicade the origin offset where the net loss reached an absolute minimum.
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
 the value at -1 is close, but the dots are in the actual minimal values.

### 2015-04-17 Paper.Rnw
discussion of trade-off of spatial precision for frequecy information added to Section 4.2 (also, spell check)

@@
 A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
-This is analogous to the fundimental problem for overplotted standard scatterplots where frequency is masked by other points.
+This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points.
 \end{itemize}
@@
 If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
-However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard devations.
+However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations.
 }
@@
 \label{FreqLoss}
+
+\ktm{It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates.
+The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from overplotting, which can be immense for large data sets.
+When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location.
+This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points.
+In binned scatterplots, we have traded the exact locations for exact frequencies within bins.
+}
 
@@
 
-\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimesions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
+\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
 }
@@
 Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees.
-The round glyphs indicade the origin offset where the net loss reached an absolute minimum.
+The round glyphs indicate the origin offset where the net loss reached an absolute minimum.
 For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
@@
 we don't see a difference between the continuously shaded plot and the one with discrete shades.
-This re-enforces our approach to treat color as a categorical variable, because we preceive it that way (otherwise why would we?)}
+This re-enforces our approach to treat color as a categorical variable, because we perceive it that way (otherwise why would we?)}
 
@@
 \ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading.
-It is recommended to use less than ten seperate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins.
+It is recommended to use less than ten separate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins.
 Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.

### 2015-04-17 Paper.Rnw
Add description of how many hidden points exist in the traditional histogram for the coarse data

@@
 <<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
-\\hh{include a finding here- e.g.
-how big is the loss under each scenario?}">>=
+In the coarse data 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the uniform, normal and exponential data, respectively.">>=
 ### Try 2d binning with MVN data
@@
 }
+
+# # Find % hidden points in coarse data
+# unifPercHidden <- (100000 - nrow(unique(datlist[[1]][,c("roundx","roundy")]))) / 100000
+# normPercHidden <- (100000 - nrow(unique(datlist[[2]][,c("roundx","roundy")]))) / 100000
+# expoPercHidden <- (100000 - nrow(unique(datlist[[3]][,c("roundx","roundy")]))) / 100000
 

### 2015-04-17 Paper.Rnw
Paragraph added to Section 5.1 exploring properties of visual density estimation. Consistency and Unbiasness tradeoff. Efficiency.

@@
 <<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
-In the coarse data 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the uniform, normal and exponential data, respectively.">>=
+97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=
 ### Try 2d binning with MVN data
@@
 Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
- \ktm{Another pertinent argument for using larger bins is that it will smooth over  larger regions if visualizing the large scale density structure is of primary interest.}
-\hh{This is an argument along the lines of the bias vs variability tradeoff. Extend on it, don't put it here as a throwaway sentence. }
+Another pertinent argument for using larger bins is that it will smooth over larger regions if visualizing the large scale density structure is of primary interest.
+
+\ktm{If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasness, consistency and efficiency.
+As binsize becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but bin density have higher sampling variability due to the smaller number of observation per bin.
+Choosing bin size for a binned scatterplot carries a trade-off between unbiasness and consistency of the visual density estimate.
+For efficiency of density estimation, it may be tempting to claim that a minimally binning, with only one unique coordinate pair per bin, is an efficient estimator because it minimizes spatial loss.
+Minimally binned scatterplots are \textit{spatially} efficient, but density estimation requires the combination of spatial and frequency information, each of which have seperate loss functions.
+Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
+}
 

### 2015-04-17 Paper.Rnw
clarification about "reduced binned data" to acknowlege that other bin summaries could be recorded, however in our paper we only consider frequency because it is the only thing required to construct a binned scatterplot

@@
 
+Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
 Data reduction comes at the expense of spatial information of any of the individual points.
@@
 %-----------------------------------------------------------------------------
-\subsection{Choice of bin width}
+%\subsection{Choice of bin width}
 
@@
 The reduced binned data from spatial binning contains the center and count information for all bins. 
-\hh{This sounds like it is set in stone - it's not.
+%\hh{This sounds like it is set in stone - it's not.
 We can define which information is kept for each of the bins.
@@
 }
-
 The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.

### 2015-04-17 Paper.Rnw
edits to description of data resolution and origin offset in section 6

@@
 The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}. 
-\hh{XXX what other purpose does this scatterplot have besides showing the distributions?}
+%\hh{XXX what other purpose does this scatterplot have besides showing the distributions?}
 
 <<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
+Density changes are impossible to make out in the solid black areas of the plots.
 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=
@@
 
-For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution, $\alpha$.
-However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution of $\alpha$ units can avoid the stripe illusion if a binwidth that is an integer multiple of $\alpha$ is selected.
-It was also shown that in this scenario, a reasonable default for the binning origin is offset to $\alpha / 2$ units below the minimum data value because it will minimize spatial loss for symmetric data and generally preform well even for non-symmetric data.
+For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution.
+However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid the stripe illusion if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.
+It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for non-symmetric data.
 

### 2015-04-18 Paper.Rnw
Many changes to section 5.2 on frequency loss properties and recommendations

@@
 library(scales)
+library(plyr)
 #source in binning functions
@@
 \ktm{If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasness, consistency and efficiency.
-As binsize becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but bin density have higher sampling variability due to the smaller number of observation per bin.
+As bin size becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but bin density have higher sampling variability due to the smaller number of observation per bin.
 Choosing bin size for a binned scatterplot carries a trade-off between unbiasness and consistency of the visual density estimate.
 For efficiency of density estimation, it may be tempting to claim that a minimally binning, with only one unique coordinate pair per bin, is an efficient estimator because it minimizes spatial loss.
-Minimally binned scatterplots are \textit{spatially} efficient, but density estimation requires the combination of spatial and frequency information, each of which have seperate loss functions.
+Minimally binned scatterplots are \textit{spatially} efficient, but density estimation requires the combination of spatial and frequency information, each of which have separate loss functions.
 Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
@@
 The reduced binned data from spatial binning contains the center and count information for all bins.
-%\hh{This sounds like it is set in stone - it's not.
-We can define which information is kept for each of the bins.
-Make this an assumption rather than a statement and define what kind of center you are talking about - visual or numeric.
-}
 The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.
+\ktm{Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups.
+If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading.
+This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}.
+Our exploration of frequency loss will focus on frequency binning with at most ten bins.
+}
+
+<<StandardFreqBinning4710none,echo=F,include=T,eval=T,fig.width=10, fig.height=3, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Binned scatterplots for the simulated bivarate normal data with varying numbers of standard binned frequency groups.
+At 10 frequency bins the shades are difficult to distinguish, with little difference from a binned scatterplot with continuous frequency shading.">>=
+binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
+binoutFreqGroups4 <- freqBin(binout1, binType="standard", ncolor=4, logCount=FALSE)
+binoutFreqGroups7 <- freqBin(binout1, binType="standard", ncolor=7, logCount=FALSE)
+binoutFreqGroups10 <- freqBin(binout1, binType="standard", ncolor=10, logCount=FALSE)
+
+# Plot with 4 Binned Frequencies
+p4<- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups4[[1]],
+           main="4 Standard Freq Bins") +
+  xlab("") + ylab("")+
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43")+ 
+  theme_bw() + theme(legend.position="none",aspect.ratio=1,
+                     plot.margin = unit(c(0.1,.1,.1,.1),"cm") )
+
+# Plot with 7 Binned Frequencies
+p7<- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups7[[1]],
+            main="7 Standard Freq Bins") +
+  xlab("") + ylab("")+
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43")+ 
+  theme_bw() + theme(legend.position="none",aspect.ratio=1,
+                     plot.margin = unit(c(0.1,.1,.1,.1),"cm") )
+
+# Plot with 10 Binned Frequencies
+p10 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups10[[1]],
+           main="10 Standard Freq Bins") +
+  xlab("") + ylab("")+
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43")+ 
+  theme_bw() + theme(legend.position="none",aspect.ratio=1,
+                     plot.margin = unit(c(0.1,.1,.1,.1),"cm") )
+
+p0 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
+      main="No Freq Binning") +
+  xlab("") + ylab("")+
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
+                      guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
+                                              title.position = "top")) + 
+  theme_bw() + theme(legend.position="none",aspect.ratio=1,
+                     panel.grid= element_blank(),
+                     plot.margin = unit(c(.1,.1,.1,.1),"cm") )
+
+grid.arrange(p4,p7,p10,p0,nrow=1)
+@
+
 The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified.
-To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).
-Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
-As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities \citep{colorperception}.
-
-\hh{XXX I don't think that figure 12 is doing its job - and for that matter, we should talk about what its job is.
-There are too many different things going on in it, and the figure might feed better into your argument when things are separated out.
-}
-
-\hh{XXX Let's try a first picture with a continuous scale and three discrete binned frequencies of four, seven, and ten bins.
-This should drive home the message that Healey is right, and we can't distinguish more than seven shades anyway, i.e.
-we don't see a difference between the continuously shaded plot and the one with discrete shades.
-This re-enforces our approach to treat color as a categorical variable, because we perceive it that way (otherwise why would we?)}
-
-\hh{XXX figure 13 is a plot with a screaming message: never ever use a raw data mapping to show color.
-use quantiles or log frequencies.
-It doesn't scream that at the moment, so emphasize it in the text.}
-
-\hh{XXX the rest of figure 12 should probably go into a figure 14 - to show that quantile binning and log frequency binning yield very similar results.
-We need to fix the legend.
-Right now they are atrocious (and I am aware that I am responsible for them, so I can say that :) )}
-
-\ktm{ Figure~\ref{freqbinning} shows binned scatterplots resulting from standard and quantile frequency binning algorithms using five frequency bins.
-Note that for standard binning, the shade is to be interpreted as an ordinal indicator of the groups based on bin frequencies, whereas for quantile binning the shade denotes groups based on frequency quantiles.
-This distinction explains why the shading of the binned scatterplot is invariant to the log transformation for quantile frequency binning but not for standard; because a monotone transformation will not alter groupings based on order, but will change groupings based on value magnitude.}
-
 The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
@@
 But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.
+\ktm{Therefore, it is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies.}
 
@@
 
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
+">>=
+### Loop over ncolors 1 to 10 and also over data distribution
+# find freq loss from standard, quantile, logCount
+freqLossSimDat <- NULL
+distrib <- c("Uniform", "Normal", "Exponential")
+for (distidx in 1:3){
+  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
+  for (ncol in 1:10){
+    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
+                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
+                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
+    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
+    binoutFreq$DistType <- distrib[distidx]
+    binoutFreq$ncolor <- ncol 
+    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
+  }
+}
+
+freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels= c("Standard", "Quantile", "Standard Log", "Quantile Log"))
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Exponential"),]
+p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Normal"),]
+p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+
+plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Uniform"),]
+p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Freq Loss") + 
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log")  & freqLossSimDat$DistType=="Exponential"),]
+p4 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color=FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Normal"),]
+p5 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+
+plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Uniform"),]
+p6 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+      data=plotdat , color = FreqBinType, size=I(1)) +
+  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
+  scale_x_continuous(breaks=1:10) + ylim(0,2) +
+  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
+  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+
+grid.arrange(p1,p2,p3,p4,p5,p6, nrow=2)
+@
+
+%To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).
+Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
+As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities \citep{colorperception}.
+
+%\hh{XXX I don't think that figure 12 is doing its job - and for that matter, we should talk about what its job is.
+There are too many different things going on in it, and the figure might feed better into your argument when things are separated out.
+}
+
+%\hh{XXX Let's try a first picture with a continuous scale and three discrete binned frequencies of four, seven, and ten bins.
+This should drive home the message that Healey is right, and we can't distinguish more than seven shades anyway, i.e.
+we don't see a difference between the continuously shaded plot and the one with discrete shades.
+This re-enforces our approach to treat color as a categorical variable, because we perceive it that way (otherwise why would we?)}
+
+%\hh{XXX figure 13 is a plot with a screaming message: never ever use a raw data mapping to show color.
+use quantiles or log frequencies.
+It doesn't scream that at the moment, so emphasize it in the text.}
+
+\hh{XXX the rest of figure 12 should probably go into a figure 14 - to show that quantile binning and log frequency binning yield very similar results.
+We need to fix the legend.
+Right now they are atrocious (and I am aware that I am responsible for them, so I can say that :) )}
+
+\ktm{Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods of trying to fix the same problem: heavily right skewed bin counts.
+However, using both -- quantile log frequency binning -- is not a recommended approach.
+The frequency groups for quantile binning are invariant to the log transformation because a monotone one-to-one transformation will not alter groupings based on quantiles.
+In the common situation where bin frequencies are heavily skewed, the choice of frequency binning is between between quantile frequency binning and standard log frequency binning.}
+
+\ktm{Since the loss scales are not comparable, the choice is guided by desired interpretability.
+Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
+For standard log frequency binning, the shade is to be interpreted as an ordinal indicator of the groups based on equally spaced groupings of log bin frequencies, whereas for quantile frequency binning the shade denotes groups based on frequency quantiles.
+This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.}
+
+%We also see that the shading of the binned scatterplot is invariant to the log transformation for quantile frequency binning but not for standard; because a monotone one-to-one transformation will not alter groupings based on order, but will change groupings based on value magnitude.
+
 \begin{figure}[hbtp]
-  \begin{minipage}{.31\linewidth}
-<<noFreqBin, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
+%  \begin{minipage}{.31\linewidth}
+% <<noFreqBin, echo=FALSE, out.width='.8\\linewidth',fig.width=4, fig.height=4>>=
+% binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
+% qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
+%             main="No Freq Binning") +
+%   xlab("") + ylab("")+
+%   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
+%                       guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
+%                                            title.position = "top")) + 
+%   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+%                      panel.grid= element_blank(),
+%                      plot.margin = unit(c(.1,.1,.1,.1),"cm") )
+% @
+%  \end{minipage}
+%\hfil\begin{minipage}{.685\linewidth}
+<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=5>>=
 binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
-qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
-            main="No Freq Binning") +
-  xlab("") + ylab("")+
-  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
-                      guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
-                                           title.position = "top")) + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     panel.grid= element_blank(),
-                     plot.margin = unit(c(.1,.1,.1,.1),"cm") )
-@
-  \end{minipage}
-\hfil\begin{minipage}{.685\linewidth}
-<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=10>>=
 freqgrps = 5
@@
 
-# Plot with Binned Frequencies
+#Plot with Binned Frequencies
 labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
@@
 
-grid.arrange(p2,p3,p4,p5, nrow=2)
-@
-\end{minipage}
+grid.arrange(p3,p4, nrow=1)
+@
+%\end{minipage}
 \caption{\label{freqbinning}Simulated bivariate normal data spatially binned using standard algorithm with 5X5 unit bins and origin at (0,0).
-Binned Scatterplot with raw frequencies (left) and binned scatterplots with five frequency bins (right).
-Bin shades labeled by frequency range and average.
+Binned scatterplots with five frequency bins from quantile frequency binning (left) and standard log frequency binning (right).
 }
 \end{figure}
-
-<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
-">>=
-### Loop over ncolors 1 to 10 and also over data distribution
-# find freq loss from standard, quantile, logCount
-freqLossSimDat <- NULL
-distrib <- c("Uniform", "Normal", "Exponential")
-for (distidx in 1:3){
-  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
-  for (ncol in 1:10){
-    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
-                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
-                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
-                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
-    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
-    binoutFreq$DistType <- distrib[distidx]
-    binoutFreq$ncolor <- ncol 
-    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
-  }
-}
-
-freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels= c("Standard", "Quantile", "Standard Log", "Quantile Log"))
-
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Exponential"),]
-p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = FreqBinType, size=I(1)) +
-  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
-  xlab("Freq Bins") + ylab("Total Freq Loss") + 
-  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
-
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Normal"),]
-p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = FreqBinType, size=I(1)) +
-  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
-  xlab("Freq Bins") + ylab("Total Freq Loss") + 
-  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
-
-plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Uniform"),]
-p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = FreqBinType, size=I(1)) +
-  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
-  xlab("Freq Bins") + ylab("Total Freq Loss") + 
-  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
-
-plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log")  & freqLossSimDat$DistType=="Exponential"),]
-p4 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color=FreqBinType, size=I(1)) +
-  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
-  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
-  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
-
-plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Normal"),]
-p5 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = FreqBinType, size=I(1)) +
-  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
-  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
-  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
-
-plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Uniform"),]
-p6 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
-      data=plotdat , color = FreqBinType, size=I(1)) +
-  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
-  scale_x_continuous(breaks=1:10) + ylim(0,2) +
-  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
-  theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
-
-grid.arrange(p1,p2,p3,p4,p5,p6, nrow=2)
-@
 

### 2015-04-19 Paper.Rnw
edits to frequency loss section and commenting out some unused plots from R code

@@
 
-\ktm{Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods of trying to fix the same problem: heavily right skewed bin counts.
+\ktm{Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
 However, using both -- quantile log frequency binning -- is not a recommended approach.
 The frequency groups for quantile binning are invariant to the log transformation because a monotone one-to-one transformation will not alter groupings based on quantiles.
-In the common situation where bin frequencies are heavily skewed, the choice of frequency binning is between between quantile frequency binning and standard log frequency binning.}
+In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.}
 
@@
 
-#Plot with Binned Frequencies
-labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
-p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutFreqGroups[[1]],
-            main="Standard Freq Binning") +
-    xlab("") + ylab("")+
-  scale_fill_manual("Frequencies", values=cols, 
-                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-                                           title.position = "top"),
-                      breaks=unique(binoutFreqGroups[[1]]$freqgroup),
-                      labels=labels) + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank(),
-                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
-                     legend.text=element_text(size=12) )
+# #Plot with Binned Frequencies
+# labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
+# p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutFreqGroups[[1]],
+#             main="Standard Freq Binning") +
+#     xlab("") + ylab("")+
+#   scale_fill_manual("Frequencies", values=cols, 
+#                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+#                                            title.position = "top"),
+#                       breaks=unique(binoutFreqGroups[[1]]$freqgroup),
+#                       labels=labels) + 
+#   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+#                      legend.key.width = unit(1, "cm"),
+#                      panel.grid= element_blank(),
+#                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+#                      legend.text=element_text(size=12) )
 
@@
 
-# Plot with quantile binned Log Frequencies
-labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
-p5 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup),
-            data=binoutQuantLogFreqGroups[[1]],
-            main="Quantile Log Freq Binning") +
-    xlab("") + ylab("")+
-  scale_fill_manual("Log Frequencies", values=cols, 
-                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-                                           title.position = "top"),
-                      breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
-                      labels=labels) + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank(),
-                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
-                     legend.text=element_text(size=12))
+# # Plot with quantile binned Log Frequencies
+# labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
+# p5 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup),
+#             data=binoutQuantLogFreqGroups[[1]],
+#             main="Quantile Log Freq Binning") +
+#     xlab("") + ylab("")+
+#   scale_fill_manual("Log Frequencies", values=cols, 
+#                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+#                                            title.position = "top"),
+#                       breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
+#                       labels=labels) + 
+#   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+#                      legend.key.width = unit(1, "cm"),
+#                      panel.grid= element_blank(),
+#                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+#                      legend.text=element_text(size=12))
 

### 2015-04-19 Paper.Rnw
Minor edits to section 6 intro

@@
 However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid the stripe illusion if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.
-It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for non-symmetric data.
+It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.
 
@@
 The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision becomes one of how we wish the shade of a tile to be interpreted.
-Standard log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.
+Standard log frequency binning is an alternative if the bin counts are heavily right skewed and we wish to find distributional structure in bins with low density.
 We also discovered that if we elect to use quantile binning, the log transformation will have no effect on the binned scatterplot shading due to the identical quantile-based mapping to ordinal frequency groupings for all monotone transformations on bin counts.
@@
 First we will further explore the baseball pitching data.
-Additionally we will investigate the relationship between scheduled and actual departure times for a full year of American commercial airline data.
+Additionally we will investigate the relationship between scheduled and actual departure times for American commercial airline data.
 

### 2015-04-21 Paper.Rnw
Reorder distributions in scatterplot figure to match others

@@
 alldat <- rbind(dat,smallround)
-
-qplot(x,y,geom="point",data=alldat,facets=record~type, size=I(1)) + 
+alldat$type <- factor(alldat$type, levels=c("Exponential","Normal","Uniform"))
+
+qplot(x,y,geom="point",data=alldat,facets=record~type, size=I(1))+
   theme_bw() + 

### 2015-04-22 Paper.Rnw
Paper edits for problems that I noticed while making graphics working group presentation slides

@@
 \item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely.
-A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
+A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
 This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points.
@@
 \begin{eqnarray}\label{logfrequency}
-L^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} L_\ell^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} (\log(c_\ell) - \log(b_c(c_\ell)))^2
+L^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} L_\ell^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} (\log(c_\ell) - b_c(\log(c_\ell)))^2
 \end{eqnarray}
@@
 
-<<BaseballBinnedScatters2X10,echo=F,include=T,eval=T,fig.width=12, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
-">>=
+<<BaseballBinnedScatters2X10,echo=F,include=T,eval=T,fig.width=12, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Binned Scatterplots for Games vs Strikeouts">>=
  # raw frequencies plot
@@
       main="Standard Freq Binning") +
-  xlab("") + ylab("")+
+  xlab("Games (count)") + ylab("Strike Outs (count)")+
   scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43") + 
@@
                      panel.grid= element_blank(),
-                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+                     plot.margin = unit(c(0.5,-.5,.1,-.5),"cm"),
                      legend.text=element_text(size=12) )
@@
             main="Quantile Freq Binning") +
-  xlab("") + ylab("")+
+  xlab("Games (count)") + ylab("Strike Outs (count)")+
   scale_fill_manual("Frequencies", values=cols, 
@@
       main="Standard Log Freq Binning") +
-  xlab("") + ylab("")+
+  xlab("Games (count)") + ylab("Strike Outs (count)")+
   scale_fill_manual("Log Frequencies", values=cols, 

### 2015-04-24 Paper.Rnw
baseball - frequencies label top left

@@
   xlab("Games (count)") + ylab("Strike Outs (count)")+
-  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43") + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43",
+                      guide=guide_colourbar( 
                                         legend.key.width = unit(1, "cm"),
+                                             legend.text=element_text(size=12),
+                                            label.hjust=0.5,  
+                                         title.position = "top")) + 
+  theme_bw() + theme(aspect.ratio=1, 
                      panel.grid= element_blank(),
                      plot.margin = unit(c(0.5,-.5,.1,-.5),"cm"),
-                     legend.text=element_text(size=12) )
+                     legend.position="bottom")
 # # bin frequencies distribution hugely skewed

### 2015-04-24 Paper.Rnw
legend in colourbar. bit better, not sure that we can make it much better

--- 2015-04-22 Paper.Rnw
@@
   xlab("Games (count)") + ylab("Strike Outs (count)")+
-  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43") + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
+  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43",
+                      guide=guide_colourbar( 
+                                        legend.key.width = unit(1.5, "cm"),
+                                             legend.text=element_text(size=12),
+                                            label.hjust=0.5,  
+                                         title.position = "top")) + 
+  theme_bw() + theme(aspect.ratio=1, 
                      panel.grid= element_blank(),
                      plot.margin = unit(c(0.5,-.5,.1,-.5),"cm"),
-                     legend.text=element_text(size=12) )
+                     legend.position="bottom")
 # # bin frequencies distribution hugely skewed
@@
 labels = sort(unique(binoutQuantFreqGroups[[1]]$freqlabel))
+upper <- as.numeric(gsub(".*,([-+]?[0-9]*.?[0-9]+([eE][-+]?[0-9]+)?).*","\\1", as.character(labels) ))
 cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
@@
   xlab("Games (count)") + ylab("Strike Outs (count)")+
-  scale_fill_manual("Frequencies", values=cols, 
-                    guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-                                         title.position = "top"),
-                    labels=labels) + 
+  scale_fill_manual("Frequencies (Quartiles)", values=cols, labels=upper,
+#                    guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+#                                         title.position = "top"),
+                    guide = guide_legend(label.position="bottom", label.hjust=1.1,  
+                                         title.position = "top")) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
+                     legend.key.width = unit(1.5, "cm"),
                      panel.grid= element_blank(),
@@
 labels = levels(binoutLogFreqGroups[[1]]$freqlabel)
+upper <- as.numeric(gsub(".*,([-+]?[0-9]*.?[0-9]+([eE][-+]?[0-9]+)?).*","\\1", as.character(labels) ))
 cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
@@
   scale_fill_manual("Log Frequencies", values=cols, 
-                    guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+                    guide = guide_legend(label.position="bottom", label.hjust=1.1,  
                                          title.position = "top"),
-                    labels=labels) + 
+                    labels=upper) + 
   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
+                     legend.key.width = unit(1.5, "cm"),
                      panel.grid= element_blank(),

### 2015-04-24 Paper.Rnw
streamline the terminology about artificial stripes
restructured Discussion/Conclusion sections to remove redundancy

@@
 If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.
-These white-space gaps are undesirable because they create artificial discontinuity that interferes with interpretation of bivariate density.
-If bin dimensions are not integer multiples of the bivariate data resolution there is also a disruptive visual consequence of ``stripping'', where bins have systematically different numbers of possible data values that can cause artificial density stripes to appear in the binned scatterplot.
+These white-space gaps are undesirable because they create visual discontinuity that interferes with interpretation of bivariate density.
+There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial stripping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.
 
@@
 Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
-Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.
-Note that random binning is effective at smoothing out the stripped patterns when non-integer multiples of the data resolution are selected.
+Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.
+Note that random binning is effective at smoothing out the artificial stripping patterns when non-integer multiples of the data resolution are selected.
 
@@
 The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction.
-Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
-Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear.
-Thus it is left to the plot designer to decide if they wish to highlight high level density structure with higher spatial loss or finer density structure with lower spatial loss.
-
-For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution.
-However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid the stripe illusion if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.
-It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.
-
-If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
-This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
-The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision becomes one of how we wish the shade of a tile to be interpreted.
-Standard log frequency binning is an alternative if the bin counts are heavily right skewed and we wish to find distributional structure in bins with low density.
-We also discovered that if we elect to use quantile binning, the log transformation will have no effect on the binned scatterplot shading due to the identical quantile-based mapping to ordinal frequency groupings for all monotone transformations on bin counts.
-
-These recommendations for binned scatterplot construction and loss properties will be demonstrated in the following examples using real bivariate data.
-First we will further explore the baseball pitching data.
+These recommendations and loss properties will now be demonstrated through two examples using real bivariate data.
+First we further explore the baseball pitching data.
 Additionally we will investigate the relationship between scheduled and actual departure times for American commercial airline data.
@@
 
-\ktm{We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient.
-The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid generating problematic visual features.
-In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value.
-}
-
-\ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading.
-It is recommended to use less than ten separate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins.
+\ktm{We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
+The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution.
+However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid creating artificial stripes if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.
+It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.
+}
+
+\ktm{Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
+It is left to the plot designer to decide how much spatial information they are willing to sacrific in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
+}
+
+\ktm{If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
+This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
 Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.

### 2015-04-24 Paper.Rnw
Remove old plot comparing standard and random, now refered to new figure.
Plot caption improvements

@@
 where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
-The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ can similarly be defined as
+The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ is similarly be defined as
 %
@@
 where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
-Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
-
-<<binningDemo,echo=F,include=F,cache=T>>=
-      library(mvtnorm)
-      set.seed(12)
-      d <- rmvnorm(50, c(10, 25), matrix(c(50, 15, 15, 50), nrow=2))
-      d <- data.frame(d)
-      names(d) <- c("x", "y")
-      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
-      d.rand <- d
-      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
-      dnew$x <- dnew$x+5
-      dnew$y <- dnew$y+5
-      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
-      d$a <- .33 + .66*d$bin
-      d.orig <- d
-
-      d$xbin <- NA
-      d$ybin <- NA
-      idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
-      d$xbin[idx] <- dnew$x[idx]
-      d$ybin[idx] <- 25
-
-temp <- data.frame(   
-          x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
-       xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
-          y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
-       yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
-      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
-      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
-            xlim=c(-20, 40), ylim=c(0, 50)) +
-        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
-                 fill="blue", alpha=.25) +
-        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
-                 fill="green", alpha=.25) +
-        scale_alpha(range=c(.5, 1), guide=FALSE) +
-      #  scale_colour_identity(guide=FALSE) + 
-      #  scale_shape(solid=T,guide=FALSE) +
-        theme_bw() +
-      #  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
-      #         y=rep(seq(-5, 55, 10), times=6), 
-      #         shape='+', size=8, colour="gray90") + 
-        geom_point(data=d[which(!is.na(d$xbin)),], 
-                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
-                   size=I(3), aes.inherit=FALSE) + 
-        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
-                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
-        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
-                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
-        annotate("segment", x=temp$x, y=temp$y, 
-                 xend=temp$xend, yend=temp$yend, group=temp$group) + 
-        ylim(c(10, 40)) + xlim(c(-10,30))
-      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
-
-      d.orig$bin <- dnew$x==15 & dnew$y==25
-      d.orig$a <- .33+.66*d.orig$bin
-      d.orig$l <- "Data outside bin"
-      d.orig$l[d.orig$bin] <- "Binned Data"
-      d.orig$size <- 2
-      d.orig$size[d.orig$bin] <- 3
-      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE),
-        data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]),                                                                                    
-        l="Numerical Center", a=1, size=2, bin=TRUE))
-#       ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
-#         annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
-#                  fill="blue", alpha=.10) + 
-#         geom_point() +
-#         scale_shape_manual(name="", values=c("Data outside bin" = 16, 
-#                                              "Binned Data" = 15, 
-#                                              "Numerical Center" = 17, 
-#                                              "Visual Center" = 18)) +
-#         scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
-#                                               "Binned Data" = "blue",
-#                                               "Numerical Center" = "blue4", 
-#                                               "Visual Center" = "blue4"))+
-#         scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
-#                                             "Numerical Center" = 5, 
-#                                             "Visual Center" = 5))+
-#         theme_bw() +
-#         annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
-#         annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
-#                  y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
-#                  colour="blue4") +
-# #        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
-# #         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
-# #                  y=d$y[which(d$l=="Numerical Center")], 
-# #                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
-#         scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
-#         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-#       ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
-@
-
-<<RandomBinDemo,echo=F,include=FALSE>>=
-nl <- ncol(d.rand)
-dnew <- d.rand
-binning <- c(10, 10)
-for(x in 1:(nl)){
-  y <- d.rand[,x] - binning[x]/2
-dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
-  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
-}
-temp <- data.frame(   
-          x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
-       xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
-          y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
-       yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
-      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
-
-dnew$id <- interaction(dnew[,1:(nl)])
-idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
-d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
-d.rand$a <- .33 + .66*d.rand$bin
-d.rand$xbin <- NA
-d.rand$ybin <- NA
-d.rand$xbin[idx] <- dnew$x[idx]
-d.rand$ybin[idx] <- 25
-# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
-qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
-  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
-  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
-  scale_alpha(range=c(.5, 1), guide=FALSE) +
-  scale_colour_identity(guide=FALSE) + 
-  scale_shape(solid=T,guide=FALSE) +
-  theme_bw() +
-#  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
-#        y=rep(seq(-5, 55, 10), times=6), 
-#         shape='+', size=8, colour="gray90") + 
-  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
-             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
-  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
-  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
-  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(10, 40)) + xlim(c(-10,30))
-ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
-@
-
-\begin{figure}[h]
-  \centering
-  \subfloat[Standard Binning]{\includegraphics[keepaspectratio=true, width=2.5in]{./figure/Binning-DataAndBin.pdf}}
-%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
-%   {./figure/RandomBinningDataPoints1}}
-  \subfloat[Random Binning]{\includegraphics[keepaspectratio=TRUE, width=2.5in]
-  {./figure/RandomBinningDataPoints2.pdf}}
-  \caption{Visualization of how data pairs are assigned to bins centered at (5, 25) and (15,25) under the two rectangular binning algorithms.
+Figure~\ref{fig:spatlossdemo2} provides an illustration of each binning process extended to a two dimensional situation.
+
+% 
+% <<binningDemo,echo=F,include=F,cache=T>>=
+%       library(mvtnorm)
+%       set.seed(12)
+%       d <- rmvnorm(50, c(10, 25), matrix(c(50, 15, 15, 50), nrow=2))
+%       d <- data.frame(d)
+%       names(d) <- c("x", "y")
+%       d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
+%       d.rand <- d
+%       dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
+%       dnew$x <- dnew$x+5
+%       dnew$y <- dnew$y+5
+%       d$bin <- dnew$x%in%c(5,15) & dnew$y==25
+%       d$a <- .33 + .66*d$bin
+%       d.orig <- d
+% 
+%       d$xbin <- NA
+%       d$ybin <- NA
+%       idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
+%       d$xbin[idx] <- dnew$x[idx]
+%       d$ybin[idx] <- 25
+% 
+% temp <- data.frame(   
+%           x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
+%        xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
+%           y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
+%        yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
+%       group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
+%       qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
+%             xlim=c(-20, 40), ylim=c(0, 50)) +
+%         annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
+%                  fill="blue", alpha=.25) +
+%         annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
+%                  fill="green", alpha=.25) +
+%         scale_alpha(range=c(.5, 1), guide=FALSE) +
+%       #  scale_colour_identity(guide=FALSE) + 
+%       #  scale_shape(solid=T,guide=FALSE) +
+%         theme_bw() +
+%       #  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
+%       #         y=rep(seq(-5, 55, 10), times=6), 
+%       #         shape='+', size=8, colour="gray90") + 
+%         geom_point(data=d[which(!is.na(d$xbin)),], 
+%                    aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
+%                    size=I(3), aes.inherit=FALSE) + 
+%         scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
+%                            labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
+%         scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
+%                             labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
+%         annotate("segment", x=temp$x, y=temp$y, 
+%                  xend=temp$xend, yend=temp$yend, group=temp$group) + 
+%         ylim(c(10, 40)) + xlim(c(-10,30))
+%       ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
+%       
+%       d.orig$bin <- dnew$x==15 & dnew$y==25
+%       d.orig$a <- .33+.66*d.orig$bin
+%       d.orig$l <- "Data outside bin"
+%       d.orig$l[d.orig$bin] <- "Binned Data"
+%       d.orig$size <- 2
+%       d.orig$size[d.orig$bin] <- 3
+%       d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE),
+%         data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]),                                                                                    
+%         l="Numerical Center", a=1, size=2, bin=TRUE))
+% #       ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
+% #         annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
+% #                  fill="blue", alpha=.10) + 
+% #         geom_point() +
+% #         scale_shape_manual(name="", values=c("Data outside bin" = 16, 
+% #                                              "Binned Data" = 15, 
+% #                                              "Numerical Center" = 17, 
+% #                                              "Visual Center" = 18)) +
+% #         scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
+% #                                               "Binned Data" = "blue",
+% #                                               "Numerical Center" = "blue4", 
+% #                                               "Visual Center" = "blue4"))+
+% #         scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
+% #                                             "Numerical Center" = 5, 
+% #                                             "Visual Center" = 5))+
+% #         theme_bw() +
+% #         annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
+% #         annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
+% #                  y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
+% #                  colour="blue4") +
+% # #        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
+% # #         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
+% # #                  y=d$y[which(d$l=="Numerical Center")], 
+% # #                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
+% #         scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
+% #         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
+% #       ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
+% @
+% 
+% <<RandomBinDemo,echo=F,include=FALSE>>=
+% nl <- ncol(d.rand)
+% dnew <- d.rand
+% binning <- c(10, 10)
+% for(x in 1:(nl)){
+%   y <- d.rand[,x] - binning[x]/2
+% dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
+%   sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
+% }
+% temp <- data.frame(   
+%           x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
+%        xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
+%           y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
+%        yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
+%       group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
+% 
+% dnew$id <- interaction(dnew[,1:(nl)])
+% idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
+% d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
+% d.rand$a <- .33 + .66*d.rand$bin
+% d.rand$xbin <- NA
+% d.rand$ybin <- NA
+% d.rand$xbin[idx] <- dnew$x[idx]
+% d.rand$ybin[idx] <- 25
+% # res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
+% qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
+%   annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
+%   annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
+%   scale_alpha(range=c(.5, 1), guide=FALSE) +
+%   scale_colour_identity(guide=FALSE) + 
+%   scale_shape(solid=T,guide=FALSE) +
+%   theme_bw() +
+% #  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
+% #        y=rep(seq(-5, 55, 10), times=6), 
+% #         shape='+', size=8, colour="gray90") + 
+%   geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
+%              aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
+%   scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
+%   scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
+%   annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(10, 40)) + xlim(c(-10,30))
+% ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
+% @
+% 
+% \begin{figure}[h]
+%   \centering
+%   \subfloat[Standard Binning]{\includegraphics[keepaspectratio=true, width=2.5in]{./figure/Binning-DataAndBin.pdf}}
+% %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
+% %   {./figure/RandomBinningDataPoints1}}
+%   \subfloat[Random Binning]{\includegraphics[keepaspectratio=TRUE, width=2.5in]
+%   {./figure/RandomBinningDataPoints2.pdf}}
+%   \caption{Visualization of how data pairs are assigned to bins centered at (5, 25) and (15,25) under the two rectangular binning algorithms.
 Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to any of the surrounding bin centers.}
-  \label{fig:BinningTutorial}
-\end{figure}
+%   \label{fig:BinningTutorial}
+% \end{figure}
 
@@
 In cases of large data, binning greatly reduces the storage size for the information and the computation time needed to construct the plot.
-
-Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
-Data reduction comes at the expense of spatial information of any of the individual points.
-After aggregation the original spatial locations cannot be recovered.
-The loss of information incurred from binning will be explored in following sections.
-
-% recover some of the numerical information lost due to binning by storing the numerical center $(\bar{x},\bar{y})$ instead of the visual center $(x^\ast,y^\ast)$ for use in subsequent calculations.
-Storing the binned data as the arithmetic mean rather than the visual center reduces the total loss, and the visual center can be recovered from the bin width and  bin center $(O_X, O_Y)$, see table \ref{RecoveryComparison} for a numeric example for this.
 
@@
 
-%Information recovery comes with a two-fold cost: the additional storage of numeric bin centers triples the amount of information stored, and the computational cost to compute numeric bin centers is O($nm^2$), where $n$ is the number of data points and $m^2$, the total number of bins.
-O($nm^2$) $\approx$ O($n$), as $m$ is generally negligible compared to $n$.
-
-% In spite of the costs, information recovery is still useful for very large data, where the unbinned data set is  computationally intractable.
-Minimizing numerical loss in this manner provides for more accurate calculations from the data, while maintaining the storage space advantages of binning.
+Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
+Data reduction comes at the expense of spatial information of any of the individual points.
+After aggregation the original spatial locations cannot be recovered.
+The loss of information incurred from binning will be explored in following sections.
 
@@
 
-<<UniformStripes, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data with square bins of size 1, 4 and 5">>=
+<<UniformStripes, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data with 1X1, 4X4 and 5X5 square bins.
+For standard binning, the 1X1 bins leaves white-space gaps between bins, and 5X5 bins cause artificial density stripes; whereas random binning smoothes density over these poor choices in bin dimensions.">>=
 # Commented out code for data generation: run once then recomment
@@
 <<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
-These binnings have net spatial losses of 391067 units and 420458 units, respectively">>=
+The binning origin at (0,0) leads to 9\\% less net spatial loss than (-9,-9); with losses of 391067 units and 420458 units, respectively.">>=
 

### 2015-04-24 Paper.Rnw
Future work and spell check

@@
 
-\hh{XXX the rest of figure 12 should probably go into a figure 14 - to show that quantile binning and log frequency binning yield very similar results.
+%\hh{XXX the rest of figure 12 should probably go into a figure 14 - to show that quantile binning and log frequency binning yield very similar results.
 We need to fix the legend.
@@
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
-Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.
+Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation can come at the cost of losing precision in the spatial and frequency information.
 }
@@
 \ktm{Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
-It is left to the plot designer to decide how much spatial information they are willing to sacrific in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
+It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
 }
@@
 Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.
+}
+
+\ktm{Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
+The findings of this research provide suggestions for reasonable default settings for binning parameters that maintain of spatial and frequency information and have desirable visual properties.
 }

### 2015-04-24 Paper.Rnw
updating and expanding big data references

@@
 
-Technological advances have  facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online.
-According to estimates published by the IDC in 2009 the worldwide digital content  doubles about every 18 months -- in 2011 the digital amount was estimated to be at 1.8 zeta bytes (1 ZB = $2^{70}$ bytes $\approx 10^{21}$ bytes).
-This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: 
-\begin{quote}``It’s a revolution ...
-The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government.
-There is no area that is going to be untouched"\end{quote} (Gary King, Harvard Institute).
+\ktm{Technological advances have facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online.
+According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes" (\url{http://www.emc.com/about/news/press/2014/20140409-01.htm}).
+This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: ``It's a revolution $\dots$ The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government.
+There is no area that is going to be untouched"-- Gary King, Harvard Institute.
 
 Data sets with millions of records and thousands of variables are not uncommon.
-Such data sets are often too large to fit into a single computer's working memory, and  single-user machines cannot easily deal with them.
-We can employ database tools to extract relevant variables, thereby vertically subsetting the data.
-However, with millions of records, it can be very difficult to work with data even at that level.
-
 \citet{Friedman97} proposed in his paper on data mining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it".
-The same holds for visualizations.
+\citet{jacobs2009} echoed the sentiment, stating that ``big data should be defined at any point in time as \textit{data whose size forces us to look beyond the tried-and-true methods that are prevalent at that time}".
+The same holds true for visualizations.
 With a 100-1000 fold increase in the amount of data, the utility of some of our most used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.
@@
 In this paper we describe first the problem scatterplots are exposed to in large-data situations.
-We discuss two different binning methods and introduce the  {\it loss of information} inherent to binning.
-We conclude with a detailed example.
-
-% It is important to be able to describe the contents of the database in a quick and easily comprehensible manner, similar to the use of the mean and standard deviation to describe univariate data, or the mean, and variance matrix to describe multivariate data.
-Graphical summaries of the data contained within the database are likely to more fully describe large, complex data sets which may not originate from a single multivariate distribution than a set of numerical summaries (i.e.
-the mean, median, mode, and range used in introductory statistics classes).
+We discuss different binning algorithms use the construct binned scatterplots and the  {\it loss of information} inherent to binning.
+We will then explore the binning properties through simulation and real-data case studies.
+We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.
+}
 
@@
 %----------------------------------------------------------------------------
-
+\newpage
 \begin{appendix}

### 2015-04-24 Paper.Rnw
Adjust language about reduced binned data sizes for consistency  (data triples, not observations)

@@
 %\hfil\begin{minipage}{.685\linewidth}
-<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=5>>=
+<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=4.8>>=
 binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
@@
 Binning data by five-minute intervals produces a more high-level summary of the relationship between actual and scheduled departure time, though it necessarily obscures some of the finer details.
-In addition, binning data by five minute intervals reduces the size of the data set to a much more manageable 19,787 observations  which can be easily manipulated on probably any modern computer.
-Binning by 15-minute intervals reduces the data set to a nearly-trivial 3,575 observations, but the graphical summary becomes granular and less appealing at that resolution.
+In addition, binning data by five minute intervals reduces the size of the data set to a much more manageable 19,787 reduced binned data triples, which can be easily manipulated on probably any modern computer.
+Binning by 15-minute intervals reduces the data set to a nearly-trivial 3,575 reduced binned data triples, but the graphical summary becomes granular and less appealing at that resolution.
 

### 2015-04-26 Paper.Rnw
minor grammar edits

@@
 \item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely.
-A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
+A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$, for spatial bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
 This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points.

### 2015-04-29 Paper.Rnw
first couple of sections

@@
 
-\ktm{Technological advances have facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online.
+Technological advances have facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online.
 According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes" (\url{http://www.emc.com/about/news/press/2014/20140409-01.htm}).
@@
 We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.
-}
+
+\hh{XXX `binning properties' is too vague. Put in a bit more detail.}
 
@@
 
-\ktm{Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}.
+Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}.
 More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations (\citealp{scott1979} and \citealp{scott1992}).
-}
+\hh{XXX the multivariate book is going beyond one dimension}
 
@@
 
-\ktm{Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
+Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
 Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
-The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information.
+The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is usually reached well before the maximum frequency, truncating the perceivable frequency information.
 By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points.
 This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c.
-}
-
-\ktm{By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures.
+
+By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures.
 Using a small number of large bins may mask the real signal in the data, while an extremely large number of small bins may not sufficiently smooth over {\it noise} inherent in any real data set.
 Figure~\ref{binning} gives an overview of a data set and binned representations using different numbers of bins, demonstrating the loss of information with increasing bin size.
-}
 
@@
 Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points.
-\ktm{The idea of loss from one-dimensional binning was explored by Scott using mean squared error as a loss function where the number of bins could be optimized for the construction of histograms \citep{scott1979}.
+The idea of loss from one-dimensional binning was explored by Scott using mean squared error as a loss function where the number of bins could be optimized for the construction of histograms \citep{scott1979}.
 He later extended this discussion to two-dimensional binning, where he compared the mean squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
-We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section\ref{GenBinning} above.
-}
+We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section~\ref{GenBinning} above.
 

### 2015-04-29 Paper.Rnw
couple more comments

@@
 By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
+\hh{XXX the frequency information is not what causes the loss}
 
@@
 This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points.
+\hh{XXX ok, I don't quite agree with the description of frequency loss here.
+(a) the rendering on a continuous color scheme IS precise.
+but on a cognitive level we are not able to read the rendered values at the same precision.
+This effectively intorduces a second stage of binning ...
+then we are on the same page again.}
 \end{itemize}
@@
 
+\hh{XXX what are you referring to by `it' in the next sentence?}
 Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points.
-The idea of loss from one-dimensional binning was explored by Scott using mean squared error as a loss function where the number of bins could be optimized for the construction of histograms \citep{scott1979}.
+The idea of loss from one-dimensional binning was explored by Scott using mean squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}.
 He later extended this discussion to two-dimensional binning, where he compared the mean squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
@@
 \ktm{The net spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
-If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
+\hh{XXX isn't this true for spatial loss in general and not just the net spatial loss?
+}  If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
 However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations.
@@
 
-\ktm{It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates.
+It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates.
 The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from overplotting, which can be immense for large data sets.
@@
 In binned scatterplots, we have traded the exact locations for exact frequencies within bins.
-}
-
-Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus theoretically have no frequency loss.
+
+Bin counts can be displayed using a continuous shading scale in HSV color space \ktm{(reference needed)} and thus theoretically have no frequency loss.
 While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
@@
 
-\ktm{If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasness, consistency and efficiency.
-As bin size becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but bin density have higher sampling variability due to the smaller number of observation per bin.
-Choosing bin size for a binned scatterplot carries a trade-off between unbiasness and consistency of the visual density estimate.
+If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
+\ktm{As bin size becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but \hh{XXX what do you mean by bin density?} bin density have higher sampling variability due to the smaller number of observation per bin.} Choosing bin size for a binned scatterplot carries a trade-off between unbiasedness and consistency of the visual density estimate.
 For efficiency of density estimation, it may be tempting to claim that a minimally binning, with only one unique coordinate pair per bin, is an efficient estimator because it minimizes spatial loss.
@@
 Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
-}
 
@@
 These white-space gaps are undesirable because they create visual discontinuity that interferes with interpretation of bivariate density.
-There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial stripping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.
+There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial striping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.
 
@@
 Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.
-Note that random binning is effective at smoothing out the artificial stripping patterns when non-integer multiples of the data resolution are selected.
+Note that random binning is effective at smoothing out the artificial striping patterns when non-integer multiples of the data resolution are selected.
 

### 2015-04-30 Paper.Rnw
all but the last section

@@
 <<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
-The binning origin at (0,0) leads to 9\\% less net spatial loss than (-9,-9); with losses of 391067 units and 420458 units, respectively.">>=
+The binning origin at (0,0) leads to 9\\% less net spatial loss than (-9,-9); with losses of 391,067 units and 420,458 units, respectively.">>=
 
@@
 For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
-\ktm{We define the \textit{origin offset} for each dimension, ($o_x$, $o_y$), such that
+We define the \textit{origin offset} for each dimension, as the tuple, ($o_x$, $o_y$), by which we shift the bivariate bin origin $(\beta_{0,x},\beta_{0,y})$ from the minimal values in $x$ and $y$:
 %
@@
 %
-where $(\beta_{0,x},\beta_{0,y})$ is the bivariate bin origin, $x_{(1)}$ and $y_{(1)}$ are the minimum data value for each dimension, and $(o_x, o_y)$ is the origin offset.
+where $x_{(1)}$ and $y_{(1)}$ are the minimal data values in $x$ and $y$. %, and $(o_x, o_y)$ is the origin offset. 
 Thus the offset indicates the number of units in each dimension to shift the origin below the highest possible origin.
-}
-
-\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
-}
+
+It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units  minimizes the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension and $\alpha_y$ units in the $Y$ dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
 
@@
 It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
-\ktm{Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm.
-Note that for simplicity, only origin offsets for with equal offset in each dimension are included in this simulation.
-Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees.
-The round glyphs indicate the origin offset where the net loss reached an absolute minimum.
-For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no!
-the value at -1 is close, but the dots are in the actual minimal values.
-This is only a single example - the expected minimum might still be in -1.
-}
-For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
+Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm.
+Note that for simplicity, only origin offsets  with equal offsets in each dimension are included in this simulation.
+Since the data resolution  consists of one unit in each dimension, we  pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees.
+The round glyphs indicate the origin offset where the net loss reaches an absolute minimum in the simulation.
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.
+%\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
+For the bivariately skewed exponential data, the one unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
 
@@
 
-The reduced binned data from spatial binning contains the center and count information for all bins.
+\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception.
+We use the binning to reflect what happens in practice.
+The binning then allows us to measure what kind of loss occurs because of the limitations of the human cognitive system.}
+
+\ktm{The reduced binned data from spatial binning contains the center and count information for all bins.
 The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.
-\ktm{Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups.
+}
+Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups.
 If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading.
@@
 Our exploration of frequency loss will focus on frequency binning with at most ten bins.
-}
 
@@
 
-\ktm{Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
-However, using both -- quantile log frequency binning -- is not a recommended approach.
-The frequency groups for quantile binning are invariant to the log transformation because a monotone one-to-one transformation will not alter groupings based on quantiles.
-In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.}
-
-\ktm{Since the loss scales are not comparable, the choice is guided by desired interpretability.
+Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
+However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong.
+It actually does not help to do both, as can be seen in the figure}.
+The frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles.
+In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
+
+Since the loss scales are not comparable, the choice is guided by desired interpretation. 
 Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
 For standard log frequency binning, the shade is to be interpreted as an ordinal indicator of the groups based on equally spaced groupings of log bin frequencies, whereas for quantile frequency binning the shade denotes groups based on frequency quantiles.
-This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.}
+This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.
 

### 2015-04-30 Paper.Rnw
into the baseball example

@@
 
-\ktm{For this data example we will revisit the baseball data used earlier in this paper.
+For this data example we will revisit the baseball data used earlier.
 To construct the binned scatterplot to display the relationship between strikeouts and game counts, we begin by specifying the bin dimensions and origin for standard rectangular binning.
-Since the data are counts, each variable is recorded to an integer resolution so integer bin widths in each dimension of spatial bins will be used to avoid artificial density stripes.
-The data has only only 42583 observations, so we can bin finely without taking an inordinate amount of computation time.
-We elect to use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are 2 games X 10 strikeouts.
+The data are counts, each variable is therefore recorded as an integer, corresponding to a resolution of one.
+Integer bin widths in each dimension of spatial bins will be used to avoid artificial density stripes.
+The data has only 42,583 observations, so we can bin finely without taking an inordinate amount of computation time.
+We elect to use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high.
 The origin will be set to a half unit below the minimum value in each dimension, at ($0.5$,$-0.5$).
-}
-
-\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot that with a continuous shading of the raw bin densities.
-We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
+
+\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies.
+\hh{ XXX the legend needs fixing.
+the numbers are running into one another.
+XXX be careful with `density' of the bins - the bins show frequencies not densities.} We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
 In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts.

### 2015-05-02 Paper.Rnw
Lots of edits.  Most importantly an emphasis that we choose to bin frequencies to make frequency loss explicit. We experience this loss implicitly because of our inability to reverse the mapping of frequency to continuous shade through perception.

@@
 We discuss different binning algorithms use the construct binned scatterplots and the  {\it loss of information} inherent to binning.
-We will then explore the binning properties through simulation and real-data case studies.
-We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.
+\ktm{We will then explore the effects of binning specification on the properties of binned scatterplots through simulation and real-data case studies.} We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.
 
@@
 Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.
-A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.
+A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size (or 3D renderings, which we do not want to recommend at this point) to provide joint density information within each grid cell, known as a tile.
 A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.
@@
 Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}.
-More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations (\citealp{scott1979} and \citealp{scott1992}).
-\hh{XXX the multivariate book is going beyond one dimension}
+More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}.
+\citet{scott1992} extends to the properties of multivariate binning strategies.
 
@@
 
-\ktm{
 \begin{itemize}
 \item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$.
-By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.
-\hh{XXX the frequency information is not what causes the loss}
-
-\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely.
+By displaying frequency information using shaded tiles instead of individual points there is a loss of information about the exact location of the points.
+%\hh{XXX the frequency information is not what causes the loss}
+\ktm{
+\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not mapped to a continuous shading scale.
+While shade can be \textit{rendered} continuously in HSV color space, thus representing frequency exactly, this information is extracted via imperfect human perception.
 A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$, for spatial bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
-This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points.
+}
+\end{itemize}
+
 \hh{XXX ok, I don't quite agree with the description of frequency loss here.
@@
 This effectively intorduces a second stage of binning ...
-then we are on the same page again.}
-\end{itemize}
-}
-
-\hh{XXX what are you referring to by `it' in the next sentence?}
-Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points.
+then we are on the same page again.} \ktm{Is this new description any better?}
+
+Note that even though the losses from creating a binned scatterplot may turn out to be substantial there a huge gain with respect to the original scatterplot, where information can be masked in large data situations due to over-plotting of points.
 The idea of loss from one-dimensional binning was explored by Scott using mean squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}.
@@
 
-\ktm{The net spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
-\hh{XXX isn't this true for spatial loss in general and not just the net spatial loss?
-}  If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
+\ktm{The spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
+If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
 However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations.
@@
 If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
-\ktm{As bin size becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but \hh{XXX what do you mean by bin density?} bin density have higher sampling variability due to the smaller number of observation per bin.} Choosing bin size for a binned scatterplot carries a trade-off between unbiasedness and consistency of the visual density estimate.
-For efficiency of density estimation, it may be tempting to claim that a minimally binning, with only one unique coordinate pair per bin, is an efficient estimator because it minimizes spatial loss.
-Minimally binned scatterplots are \textit{spatially} efficient, but density estimation requires the combination of spatial and frequency information, each of which have separate loss functions.
-Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
+\ktm{Binned scatterplots are inherently bias displays of spatial information as they shift visual emphasis from true location of individual points to the geometric centers of tiles.
+However, as bin size becomes increasingly fine, the plot will place visual emphasis more precise spatial locations.
+When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate then spatially exact; thus making the binned scatterplot a consistent visual estimator.
+We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
+It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss.
+This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies to bin shades.
+%Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
+}
 
@@
 
-\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception.
+%\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception.
 We use the binning to reflect what happens in practice.
@@
 \ktm{The reduced binned data from spatial binning contains the center and count information for all bins.
-The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot.
-}
+The bin frequencies may be mapped continuously to a precisely rendered shade, however it is niave to believe that human perception will be able to perfectly extract that information.
+There is implicitly loss of frequency information occuring when the shade of a tile is visually mapped back to a frequency through the use of a shading scale index.
+Bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot, thus making the loss explicit.
+
 Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups.
@@
 This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}.
-Our exploration of frequency loss will focus on frequency binning with at most ten bins.
+Our exploration of frequency loss will focus on frequency binning with at most ten bins because above this we experience implicit frequency from perceptual bounds that are not well reflected in the explicitly defined frequency loss.
+}
 
@@
 It actually does not help to do both, as can be seen in the figure}.
-The frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles.
-In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
+\ktm{Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.} In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
 
@@
 \ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies.
-\hh{ XXX the legend needs fixing.
-the numbers are running into one another.
-XXX be careful with `density' of the bins - the bins show frequencies not densities.} We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
+We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
 In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts.

### 2015-05-06 Paper.Rnw
frequency loss definition

@@
 \item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not mapped to a continuous shading scale.
-While shade can be \textit{rendered} continuously in HSV color space, thus representing frequency exactly, this information is extracted via imperfect human perception.
-A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$, for spatial bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
+While shade can be \textit{rendered} continuously in HSV color space, thus representing frequency exactly, a human reader can not \textit{extract} this information at the same precision due to limitations of  human cognition.
+%A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively
+In order to model these limitations we 
+introduce a second stage of binning by using a discrete color scale for displaying binned frequencies, $b_C(c_\ell)$, $\ell \in \{1,\dots,\mathscr{L}\}$.
+%This loss in precision is then quantified as {\it Frequency loss}.
+%, thus losing precision in the display of the true frequencies.
 }
@@
 
-\hh{XXX ok, I don't quite agree with the description of frequency loss here.
+%\hh{XXX ok, I don't quite agree with the description of frequency loss here.
 (a) the rendering on a continuous color scheme IS precise.

### 2015-05-06 Paper.Rnw
Euclidean

@@
 %\hh{XXX the frequency information is not what causes the loss}
-\ktm{
+
 \item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not mapped to a continuous shading scale.
@@
 %, thus losing precision in the display of the true frequencies.
-}
+
 \end{itemize}
@@
 This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers).
-Note that other distance metrics could be used, but the euclidean distance has a desirable interpretability in $\mathbb{R}^2$.
+Note that other distance metrics could be used, but the Euclidean distance has a desirable interpretability in $\mathbb{R}^2$.
 The \textit{total spatial loss}, $L^S$, is defined as
@@
 
-\ktm{The spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
+\ktm{The spatial loss is a Euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
 If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
@@
 This visual center will generally different from the numerical mean, causing some bias in the perception.
-Thus visual loss for a bin, $L_i^\text{N}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
+Thus visual loss for a bin, $L_i^\text{N}$, is the Euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
 The visual loss is the sum of these bin distances.
@@
 % Numerical loss results from the binning process as points are reduced to the numerical center point.
-The loss for an individual point, $L_i^\text{N}$, is the euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$.
+The loss for an individual point, $L_i^\text{N}$, is the Euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$.
 The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
 % Visual loss is due to the distance between the numerical centers and the visual centers.
-The visual loss for each bin, $L_j^\text{V}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
+The visual loss for each bin, $L_j^\text{V}$, is the Euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
 The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$).

### 2015-05-06 Paper.Rnw
striping

@@
 
+\hh{OK, better - I still don't understand the 'but' at the top of the paragraph. There seems to be some implicit negation, which I can't see it. Could you make this explicit?}
+
 %but requires a reference scale on which the magnitude may be interpretable.
@@
 If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
-\ktm{Binned scatterplots are inherently bias displays of spatial information as they shift visual emphasis from true location of individual points to the geometric centers of tiles.
-However, as bin size becomes increasingly fine, the plot will place visual emphasis more precise spatial locations.
-When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate then spatially exact; thus making the binned scatterplot a consistent visual estimator.
+Binned scatterplots are inherently biased displays of spatial information as they shift visual emphasis from the true location of individual points to the geometric centers of tiles.
+However, as bin sizes become increasingly fine,  the bias decreases as the plot shows more precise spatial locations.
+\ktm{When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate then spatially exact; thus making the binned scatterplot a consistent visual estimator.
 We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.

### 2015-05-08 Paper.Rnw
added reference

@@
 However, as bin sizes become increasingly fine,  the bias decreases as the plot shows more precise spatial locations.
-\ktm{When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate then spatially exact; thus making the binned scatterplot a consistent visual estimator.
+When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate is spatially exact; thus making the binned scatterplot a consistent visual estimator.
 We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
 It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss.
-This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies to bin shades.
+This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies.
 %Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
-}
-
-While bin widths can be set to any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.
-If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.
-These white-space gaps are undesirable because they create visual discontinuity that interferes with interpretation of bivariate density.
+
+While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artefacts in the binned scatterplot.
+For example, if the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins in the binned scatterplot.
+These gaps of white space are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density.
+\hh{The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with an evaluation.}
+
 There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial striping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.

### 2015-05-12 Paper.Rnw
check proof in the back

@@
 %\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
-For the bivariately skewed exponential data, the one unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
+For the bivariately skewed exponential data, the one unit origin offset minimizes the net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
 
@@
 \label{proof:offset}
+\hh{you need to use the same conditions for the proof as you are stating when you refer to it.
+A uniform distribution was not mentioned before.
+Also, make sure that this section is 'stand-alone', i.e.
+explain every single one of the parameters that you are using.}
 

### 2015-05-12 Paper.Rnw
comments in section 5.2

@@
 
-\ktm{The reduced binned data from spatial binning contains the center and count information for all bins.
-The bin frequencies may be mapped continuously to a precisely rendered shade, however it is niave to believe that human perception will be able to perfectly extract that information.
-There is implicitly loss of frequency information occuring when the shade of a tile is visually mapped back to a frequency through the use of a shading scale index.
-Bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot, thus making the loss explicit.
-
-Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups.
-If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading.
-This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}.
-Our exploration of frequency loss will focus on frequency binning with at most ten bins because above this we experience implicit frequency from perceptual bounds that are not well reflected in the explicitly defined frequency loss.
-}
+Spatial binning reduces data to contain the center and frequency information for all bins.
+While bin frequencies could be mapped continuously to a precisely rendered shade, it is naive to believe that this information can be extracted at the same resolution due to limitations of our perception.
+Implicitly, we lose information  when (imperfectly) mapping the shade of a tile to its frequency.
+In order to make this loss explicit, we introduce a second stage of binning and discretize the color scale for frequencies in the spatially binned scatterplot.
+
+Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots.
+On the left frequency is mapped continuously to colour; on the right  there is a varying numbers of  standard binned frequencies.
+It is trivial to discern differences between similarly shaded tiles
+when there are only four shades.
+This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colours better than in the continuous shading.
+This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. We will therefore use an upper limit of ten frequency bins in our exploration of frequency loss.
 
@@
 The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
-We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.
-We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms.
+\hh{XXX the next sentence needs to be fixed.}
+\ktm{We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.} We also note that the frequency loss is a decreasing function in the number of frequency bins for both binning algorithms.
 This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous.

### 2015-05-12 Paper.Rnw
end of section 5.2

@@
 Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots.
-On the left frequency is mapped continuously to colour; on the right  there is a varying numbers of  standard binned frequencies.
+On the right frequency is mapped continuously to colour; on the left  there is a varying numbers of  standard binned frequencies.
 It is trivial to discern differences between similarly shaded tiles
@@
 The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
-\hh{XXX the next sentence needs to be fixed.}
-\ktm{We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.} We also note that the frequency loss is a decreasing function in the number of frequency bins for both binning algorithms.
-This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous.
-Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot.
-Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used.
-But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.
-\ktm{Therefore, it is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies.}
-
-Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
+
+There is a huge difference in the magnitude of frequency loss between the distributions: frequency losses are the highest in the scatterplot of exponentially distributed data, much lower for the normally distributed data and lowest in the case of the uniform data.
+For all distributions and both binning algorithms, frequency loss decreases as the number of frequency bins increases. This initially sharp decrease  largely flattens out at about four bins.
+Thus four and seven bins for color shades allow us to reduce frequency loss while allowing for easy distinction between frequency groups in the binned scatterplot.
+
+Lastly, for small number of bins, the frequency loss from standard frequency binning is higher than loss from quantile-based binning.
+However, in all three simulated data sets, the losses drop to similar levels for both algorithms when four or more frequency bins are used.
+
 The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data.
-Frequency binning using the log counts behaved very similarly, where log frequency loss decreased as the number of frequency bins increased.
-The same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
+Frequency binning using log counts exhibits a very similar behaviour to binning of untransformed frequencies.
+The difference in scales makes it impossible to directly compare frequency loss and log frequency loss. 
+But the same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
 
@@
 
-Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
-However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong.
-It actually does not help to do both, as can be seen in the figure}.
-\ktm{Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.} In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
-
-Since the loss scales are not comparable, the choice is guided by desired interpretation. 
-Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
+Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem in binned scatterplots: the distribution of bin frequencies is heavily right skewed.
+High frequency bins visually overshadow any structure in low frequency bins.
+
+%However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong. It actually does not help to do both, as can be seen in the figure}.
+
+Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.
+In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
+
+As losses from log-transformed and un-transformed frequencies are not directly comparable, the choice is guided by desired interpretation. 
 For standard log frequency binning, the shade is to be interpreted as an ordinal indicator of the groups based on equally spaced groupings of log bin frequencies, whereas for quantile frequency binning the shade denotes groups based on frequency quantiles.
 This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.
+
+\hh{XXX fig 13 starts a new thought: }
+Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning. 
+\hh{XXX What conclusion do you want to draw from figure 13?}
 

### 2015-05-12 Paper.Rnw
Fix appendix notation to match paper

@@
 
-While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artefacts in the binned scatterplot.
+While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artifacts in the binned scatterplot.
 For example, if the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins in the binned scatterplot.
@@
 Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots.
-On the right frequency is mapped continuously to colour; on the left  there is a varying numbers of  standard binned frequencies.
+On the right frequency is mapped continuously to color; on the left  there is a varying numbers of  standard binned frequencies.
 It is trivial to discern differences between similarly shaded tiles
 when there are only four shades.
-This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colours better than in the continuous shading.
+This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colors better than in the continuous shading.
 This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. We will therefore use an upper limit of ten frequency bins in our exploration of frequency loss.
@@
 The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data.
-Frequency binning using log counts exhibits a very similar behaviour to binning of untransformed frequencies.
+Frequency binning using log counts exhibits a very similar behavior to binning of untransformed frequencies.
 The difference in scales makes it impossible to directly compare frequency loss and log frequency loss. 
@@
 
-The following is proof that when univariate data is uniformly distributed at resolution $\alpha$ and standard rectangular binning is used with binwidths that are a scalar multiple of the data resolution (i.e.
-$\omega = k\alpha$ for some $k \in \{1,2,\dots\})$, then spatial loss is minimized by setting the binning origin to $\alpha/2$ units below the minimum data value; set $\beta = x_{(1)} - \alpha/2$.\\
-
-\noindent Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha$ for some constant $\alpha \in \mathbb{R}$. Thus $x_j = x_1 + (j-1)\alpha$.\\
-
-\noindent Suppose then that we bin the data using standard rectangular binning with origin, $\beta = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data.
-Thus $b(x_j) = \beta + \omega/2 = (x_1 - \theta) + (k\alpha/2)$\\
+\ktm{The following proof assumes three conditions for standard rectangular binning of a set univariate data hold: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension, (ii) points are symmetric distributed within standard rectangular bins, (iii) the bin width, $\omega_x$, is an integer multiples of $\alpha_x$ (i.e.
+$\omega_x = k\alpha_x$ for some $k \in \{1,2,\dots\}$).
+Under these conditions it will be shown that spatial loss is minimized by setting the binning origin, $\beta_x$, to $\alpha_x/2$ units below the minimum data value (i.e.
+$\beta_x = x_{(1)} - \alpha_x/2$).\\ }
+
+\noindent Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha_x$ for some constant $\alpha_x \in \mathbb{R}$.
+Thus $x_j = x_1 + (j-1)\alpha_x$.\\
+
+\noindent Suppose then that we bin the data using standard rectangular binning with origin, $\beta_x = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data.
+Thus $b(x_j) = \beta_x + \omega/2 = (x_1 - \theta) + (k\alpha_x/2)$\\
 
@@
 $ b(x_i) = Q_x(.5) $ \\
-$ \Rightarrow (x_1 - \theta) + (k\alpha/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
-$ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
-$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha  $ \\
-$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha  $ \\
-$ \Rightarrow -2\theta = -\alpha $ \\
-$ \Rightarrow \theta = \alpha/2 $ \\
-
-Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha/2$.
+$ \Rightarrow (x_1 - \theta) + (k\alpha_x/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
+$ \Rightarrow 2x_1 - 2\theta + k\alpha_x = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha_x ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha_x ) $ \\
+$ \Rightarrow -2\theta + k\alpha_x = (\lceil\frac{k+1}{2}\rceil - 1)\alpha_x  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha_x  $ \\
+$ \Rightarrow -2\theta + k\alpha_x = ((k+1) - 2)\alpha_x  $ \\
+$ \Rightarrow -2\theta = -\alpha_x $ \\
+$ \Rightarrow \theta = \alpha_x/2 $ \\
+
+Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha_x/2$.
 This result holds for data that is symmetrically distributed within the bin since the median will not change.
-It extends to multiple contiguous bins with resolution $\alpha$ data that has symmetrically distributed data withing each bin.
+It extends to multiple contiguous bins with resolution $\alpha_x$ data that has symmetrically distributed data withing each bin.
 \\

### 2015-05-12 Paper.Rnw
Completed the description of figure 13

@@
 
-%However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong. It actually does not help to do both, as can be seen in the figure}.
-
 Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.
@@
 
-\hh{XXX fig 13 starts a new thought: }
 Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
-\hh{XXX What conclusion do you want to draw from figure 13?}
+\ktm{Both shading schemes are effective at reducing the visual impact of the highest density at the center, allowing for the differences in the surrounding bin frequencies to be emphasized.
+It should be noted that the frequency binning algorithm produced binned scatterplots that visually very similar, but this will not always be the case.
+It occured in this scenario because the distribution of log bin frequencies from the bivariate normal data is nearly uniform, thus the standard log binning aligns closely with the quantile binning.}
 
@@
 
-\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies.
+The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies.
 We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
@@
 Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure.
-}
 
@@
 
-\ktm{The interpretability of the net spatial loss in this scenario suffers because the units and scales of the two variables differ.
+The interpretability of the net spatial loss in this scenario suffers because the units and scales of the two variables differ.
 To remedy this, the two variables can be standardized then binned equivalently by rescaling the bin dimension, where the net spatial loss is interpreted as a distance in units of standard deviations.
@@
 We also find that we had approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification than had we naively set the binning origin at the minimum data values, ($1$,$0$).
-}
 
@@
 
-\ktm{Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
+Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
 The plot on the left shows the scatterplot from a sample of one million of those records.
@@
 The 1-minute standard binning also reduces the data to 176,384 individual records, less than 3\% of the original data.
-}
 
@@
 
-\ktm{Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
+Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
 Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation can come at the cost of losing precision in the spatial and frequency information.
-}
-
-\ktm{We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
+
+We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
 The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution.
@@
 It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.
-}
-
-\ktm{Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
+
+Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
 It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
-}
-
-\ktm{If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
+
+If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
 This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
 Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.
-}
-
-\ktm{Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
+
+Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
 The findings of this research provide suggestions for reasonable default settings for binning parameters that maintain of spatial and frequency information and have desirable visual properties.
-}
 

### 2015-05-12 Paper.Rnw
section 6.1

@@
 
-The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction.
+The exploration of loss in spatial and frequency information in the process of creating a binned scatterplot yields a number of important properties and practical recommendations for their construction.
 These recommendations and loss properties will now be demonstrated through two examples using real bivariate data.
 First we further explore the baseball pitching data.
-Additionally we will investigate the relationship between scheduled and actual departure times for American commercial airline data.
+In a second example, we will investigate the relationship between scheduled and actual departure times for American commercial airline data.
 
@@
 
-For this data example we will revisit the baseball data used earlier.
+For this data example we revisit the baseball data used earlier.
 To construct the binned scatterplot to display the relationship between strikeouts and game counts, we begin by specifying the bin dimensions and origin for standard rectangular binning.
 The data are counts, each variable is therefore recorded as an integer, corresponding to a resolution of one.
-Integer bin widths in each dimension of spatial bins will be used to avoid artificial density stripes.
+We use integer bin widths in each dimension of spatial bins  to avoid artificial density stripes.
 The data has only 42,583 observations, so we can bin finely without taking an inordinate amount of computation time.
-We elect to use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high.
-The origin will be set to a half unit below the minimum value in each dimension, at ($0.5$,$-0.5$).
-
-The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies.
-We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts.
+We use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high.
+The origin will be set to a half unit below the minimum value in each dimension, at $(0.5,-0.5)$.
+
+The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of raw bin frequencies.
+The most striking feature in the frequency distribution is the dark spot in the bottom right of the plot representing a large number of pitchers that played very few games and had very few strikeouts.
+It is nearly impossible to differentiate the density structure across the majority of bins representing better pitchers who played many games and earned many strikeouts.
 In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts.
@@
                       guide=guide_colourbar( 
-                                        legend.key.width = unit(1.5, "cm"),
+                                        barwidth = unit(6, "cm"),
                                              legend.text=element_text(size=12),
@@
 The net spatial loss for the standardized data is \Sexpr{round(standBaseballLoss2X10,2)} standard deviations, an average of \Sexpr{round(standBaseballLoss2X10PerObs,4)} standard deviation per data point.
-We also find that we had approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification than had we naively set the binning origin at the minimum data values, ($1$,$0$).
+\hh{XXX for which of the two binnings is this loss?} We also find that we had approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification than had we naively set the binning origin at the minimum data values, $(1, 0)$.
 

### 2015-05-12 Paper.Rnw
done with the conclusions

@@
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
-Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation can come at the cost of losing precision in the spatial and frequency information.
+Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.
 
@@
 The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution.
-However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid creating artificial stripes if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.
-It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.
-
-Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
-It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
+However, standard binning algorithm is superior due to lower spatial loss.
+For data with resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.
+We were also able to show through simulation that  a reasonable default for the binning origin is an offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well even for skewed data.
+
+Spatial binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
+It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure.
+%; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
 

### 2015-05-12 Paper.Rnw
space?

--- 2015-05-12 Paper.Rnw
@@
 It should be noted that the frequency binning algorithm produced binned scatterplots that visually very similar, but this will not always be the case.
-It occured in this scenario because the distribution of log bin frequencies from the bivariate normal data is nearly uniform, thus the standard log binning aligns closely with the quantile binning.}
+It occurred in this scenario because the distribution of log bin frequencies from the bivariate normal data is nearly uniform, thus the standard log binning aligns closely with the quantile binning.}
 

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Version for jcgs

--- 2015-05-12 Paper.Rnw
@@
 \usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
-\graphicspath{{figure/}}
+%\graphicspath{{figure/}}
 %\usepackage{csquotes}
@@
 \title{Binning Strategies and Related Loss for Large Data}
-\author{Karsten Maurer$^1$, Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}
+\author{Karsten Maurer$^1$, Susan VanderPlas$^2$, Heike Hofmann$^{2,3}$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
 
@@
 The demands of large data require us to re-think strategies of visualizing data.
-Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much.
-This comes with the price of loss of information inherent to any binning scheme.
-In this paper we discuss properties of two commonly used binning algorithms.
-We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the framework of a simulation and two case studies.
+Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to be less affected by increases in data.
+This comes with the price of the loss of information inherent to any binning scheme.
+In this paper we discuss binning algorithms used in the construction of binned scatterplots.
+We define functions to quantify the loss of spatial and frequency information and discuss the effects of binning specification on loss in the framework of simulation and case studies.
+From this we provide several practical suggestions for binning strategies that lead to binned scatterplots with desirable visual properties.
 \end{abstract}
@@
 \label{Intro}
-% \hh{General: write out any positive integer less than or equal to ten out in words, even if there is the word `unit' behind it. So it's one unit, and not 1 unit.}
-% 
-% \hh{
-% \begin{itemize}
-% \item Additional literature: binning papers: standard binning, e.g.: Karl Pearson (1895) ``Contributions to the Mathematical Theory of Evolution. II. Skew Variation in Homogeneous Material"\\
-% David W. Scott (1979). "On optimal and data-based histograms". Biometrika 66 (3): 605–610
-% 
-% quantile binning: there's a lot of implementations (e.g.
-cut2 in package Hmisc, the rattle package, the package extracut.
-It would be interesting to check, if they do the same things (probably not) and mention that there is disagreement in the implementations).
-% 
-% As a reference I would check in 
-% Scott, D. W. 1992.
-% Multivariate Density Estimation: Theory, Practice, and Visualization. New York: Wiley.
-% 
-% \item make sure to verify that the binning algorithms agree with the literature (and check that the code is doing the same thing as described in the paper).
-% You can include a bit of discussion on the problem that histograms are over-parameterized - it's a known problem, but I can't find any paper on it at the moment.
-% 
-% \end{itemize}
-% }
-% 
-% \ktm{I have read through the sections of Pearson (1895), Scott (1979) and Scott (1992) that pertained to binning.
-I will work on integrating them into the lit review next.
-Scott talked much about the optimal choice of bin width based on MSE but there was not much talk about what to do with points at the boundaries of bins (he effectively assumed infinite precision in the measurement process to avoid it).
-% 
-% I will need to search more to find papers about the binning algorithms and the boundaries they set.
-The \texttt{cut} function allows for intervals that are closed on either the left or right and also allows the points on the outermost open bound to be included in the last bin.
-% 
-% The \texttt{cut2} function allows the same type of quantile binning as discussed in Scott (1992).}
-
-<<setup,echo=F,include=F,eval=T>>=
+
+<<setup4,echo=F,include=F,eval=T>>=
 library(ggplot2)
@@
 
-connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
-                     port=3306, user="dbaccess", dbname="baseball")
-pitch <- new("dataDB", co=connect, table="Pitching")
-
-d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs
-
-xtd.data <- function(x) {
-# replicate each line as often as given by the last entry
-  k <- dim(x)[2]
-
-  res <- data.frame(sapply(1:(k-1), function(var) {
-			return(rep(x[,var], x[,k]))
-		}
-	))
-	names(res) <- names(x)[1:(k-1)]
-	return(res)
-}
-d1.ext <- xtd.data(d1)
-dbDisconnect(connect)
+# connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
+#                      port=3306, user="dbaccess", dbname="baseball")
+# pitch <- new("dataDB", co=connect, table="Pitching")
+# 
+# d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs
+# 
+# xtd.data <- function(x) {
+# # replicate each line as often as given by the last entry
+#   k <- dim(x)[2]
+# 
+#   res <- data.frame(sapply(1:(k-1), function(var) {
+#   		return(rep(x[,var], x[,k]))
+# 		}
+# 	))
+# 	names(res) <- names(x)[1:(k-1)]
+# 	return(res)
+# }
+# d1.ext <- xtd.data(d1)
+# dbDisconnect(connect)
+# save(d1, d1.ext, file="pitching.Rdata")
+load("pitching.Rdata")
 @
@@
 
-Technological advances have facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online.
+Technological advances have facilitated collection and dissemination of large data as records are digitized and our lives are increasingly lived online.
 According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes" (\url{http://www.emc.com/about/news/press/2014/20140409-01.htm}).
@@
 The same holds true for visualizations.
-With a 100-1000 fold increase in the amount of data, the utility of some of our most used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.
-
-Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots that utilize raw data.
+With a 100-1000 fold increase in the amount of data, the utility of some of our most commonly used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.
+
+Area plots, such as histograms, do not tend to be as affected by increases in the amount of data because they display aggregations instead of raw data.
 By using binning strategies and the principles for displaying information in area plots, scatterplots can again become useful instruments for large data settings \citep{gold}.
 
-In this paper we describe first the problem scatterplots are exposed to in large-data situations.
-We discuss different binning algorithms use the construct binned scatterplots and the  {\it loss of information} inherent to binning.
-\ktm{We will then explore the effects of binning specification on the properties of binned scatterplots through simulation and real-data case studies.} We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.
-
-\hh{XXX `binning properties' is too vague. Put in a bit more detail.}
+In this paper we describe first the inadequacy of traditional scatterplots in large-data situations.
+We discuss different binning algorithms use in the construction of binned scatterplots and the \textit{loss of information} inherent to binning.
+We will then explore the effects of binning specification on the properties of binned scatterplots through simulation and real-data case studies.
+We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.
 
@@
 
-In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions.
+In the case of modestly sized data, scatterplots are great tools for showing bivariate data relationships.
 With large data, scatterplots suffer from over-plotting of points, which masks relevant structure.
@@
 The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
-The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season.
+The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season.
 While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
 
-Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle.
+Figure~\ref{fig:scatter-alpha}(a) shows a traditional scatterplot with each observation is drawn with a filled circle.
 A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable.
-\citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting.
+\citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}(b)) to mitigate the problem of over-plotting.
 Open circles make points that are close together more visually distinct; thus allowing for the perception of more density information than with filled points.
-A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c).
-Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.
+A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}(c)).
+Alpha blending renders points as semi-transparent to provides more visibility of underlying points.
 
 \begin{figure}[hbtp]
-  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{Overplotting.pdf}}
-  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingCircles.pdf}} 
-	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingAlpha.pdf}}
-
-  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{HexBinning.pdf}}
-  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{BinScatter.pdf}}
-
-	\caption{\label{fig:scatter-alpha} Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05).
-Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
+  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.pdf}}
+  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.pdf}} 
+	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.pdf}}
+
+  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.pdf}}
+  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.pdf}}
+
+	\caption[Traditional and adapted scatterplots for games vs.
+strikeouts data.]{\label{fig:scatter-alpha} Scatterplots of games versus strikeouts in major league baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05).
+Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively.}
 \end{figure}
@@
 
-Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales.
-Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations (including ties) but use distortion of the point locations by having points repel one another to avoid overlapping points.
+Other scatterplot adaptations have been introduced that avoid over-plotting by manipulating the display of the points by distorting the locations or the scales.
+Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations, including those sharing identical coordinates, and use distortion of the point locations by having points repel one another to avoid overlapping.
 An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
 Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
-This variable binning fragments the continuity of the axes into  segments on different scales and also does not deal with ties.
+This variable binning fragments the continuity of the axes into segments on different scales and also does not deal with points at identical coordinates.
 Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.
@@
 Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point.
-This has the additional advantage of reducing the size of the stored data necessary for the construction of the plot, as only the bin centers and the frequency of observations which correspond to that bin must be stored.
+This has the additional advantage of reducing the size of the stored data necessary for the construction of the plot, as only the bin centers and the bin frequencies must be stored.
 Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale  well with the size of data \citep{Wickham2013Bin}.
 Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.
-\hh{also add \citep{gold} for binning reference - it's called `weighting' there.}
-
-Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and modified scatterplots \citep{martin-gold}.
+
+Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and binned scatterplots \citep{martin-gold}.
 Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin.
@@
 Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.
-A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size (or 3D renderings, which we do not want to recommend at this point) to provide joint density information within each grid cell, known as a tile.
-A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.
+A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size or 3D renderings to provide joint density information within each grid cell, known as a tile.
 A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.
 Bubble plots were first used by William Playfair \citep{playfair, playfair2}.
-
-Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (\ref{fig:scatter-alpha}d) and a bubble plot with frequency encoded as point size (\ref{fig:scatter-alpha}e).
-The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c).
-The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870} especially for data binned on a grid; whereas the binned scatterplot will not on a tessellated grid.
-
-Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: the joint density consists two distinct ridges following two lines with very different slopes.
+A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.
+
+Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (\ref{fig:scatter-alpha}(d)) and a bubble plot with frequency encoded as point size (\ref{fig:scatter-alpha}(e)).
+The hexagonally binned scatterplot and the bubble plot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha}(a-c).
+The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870}, where the white spaces between circles on the evenly spaced grid appear shaded due to an optical illusion; whereas this will not occur for a binned scatterplot on a tessellated grid.
+
+The inner structure of the baseball data is only apparent in the binned scatterplot and the bubble plot.
+The joint density consists two distinct ridges following two lines with very different slopes.
 The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game.
@@
 
-%\hh{...
-This is the place to put some background literature: include the discussion of use of binning in the literature: Martin Theus's chapter in the GOLD book, Dan Carr's hexagonal binning, sunflower plots, ...
-}
-
 For extremely large data sets, binned scatterplots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be rendered separately.
@@
 As with histograms, the width of bins (or the number of bins) is an important factor in the detail of the binned data and the resulting plot: if the bin width is too small in comparison to the amount of data available, there is little advantage to binning, but if the bin width is too large, interesting features of the joint distribution may be obscured by over-plotting.
-
-%\hh{Main motivation: we want to investigate how much information we lose when we bin.}
-
-%\hh{I also would like to pick up the idea of sufficient statistics for scatterplots again - as a mathematical basis for measuring the amount of information that goes into a plot (since we can't expect to ever get more out of a plot).
-....
-I'm not sure, yet, where in the paper we should place this discussion.}
 
@@
 Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.
-Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
-
-We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.
+Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in reduced data triples of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+
+We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{Carr1987} are also common.
 While hexagonal binning has been shown to have slightly better graphical properties \citep{scott1992}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
@@
 
-This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
+This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ defined as 
 %
@@
 
-\textit{Quantile binning} is another option that divides the range of the observations into bins each containing equal number of points.
-The $j^{th}$ bin intervals is $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
-Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernible groups in visualization using discrete color palettes.
+\textit{Quantile binning} is another option that divides the range of the observations into bins each containing an equal number of points.
+The $j^{th}$ bin interval takes the form $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
+Note that this binning approach is \textit{not} desirable for spatially visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for binned scatterplots that employ a second stage of binning to create discrete shade scheme for displaying grouped bin frequencies, which will be discussed in Section~\ref{FreqLoss}.
 
@@
 \centering
+\caption{Rectangular and Random Binning Specifications}
 \begin{tabular}{lll} \hline
@@
 \end{tabular}
-\caption{Rectangular and Random Binning Specifications}
 \label{tab:rectbinning}
@@
 %
-An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
+An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
@@
 for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.
-This probability structure is inversely proportional to the distance from each bin center.
 In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.
@@
 For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
-In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$, where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively.
-The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
+In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$,where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively.
+The ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
 With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,\mathscr{L}\}$, where $\mathscr{L}=J\cdot K$.
@@
 where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
-The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ is similarly be defined as
+The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ is similarly defined as
 %
@@
 
-% 
-% <<binningDemo,echo=F,include=F,cache=T>>=
-%       library(mvtnorm)
-%       set.seed(12)
-%       d <- rmvnorm(50, c(10, 25), matrix(c(50, 15, 15, 50), nrow=2))
-%       d <- data.frame(d)
-%       names(d) <- c("x", "y")
-%       d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
-%       d.rand <- d
-%       dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
-%       dnew$x <- dnew$x+5
-%       dnew$y <- dnew$y+5
-%       d$bin <- dnew$x%in%c(5,15) & dnew$y==25
-%       d$a <- .33 + .66*d$bin
-%       d.orig <- d
-% 
-%       d$xbin <- NA
-%       d$ybin <- NA
-%       idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
-%       d$xbin[idx] <- dnew$x[idx]
-%       d$ybin[idx] <- 25
-% 
-% temp <- data.frame(   
-%           x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
-%        xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
-%           y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
-%        yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
-%       group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
-%       qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
-%             xlim=c(-20, 40), ylim=c(0, 50)) +
-%         annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
-%                  fill="blue", alpha=.25) +
-%         annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
-%                  fill="green", alpha=.25) +
-%         scale_alpha(range=c(.5, 1), guide=FALSE) +
-%       #  scale_colour_identity(guide=FALSE) + 
-%       #  scale_shape(solid=T,guide=FALSE) +
-%         theme_bw() +
-%       #  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
-%       #         y=rep(seq(-5, 55, 10), times=6), 
-%       #         shape='+', size=8, colour="gray90") + 
-%         geom_point(data=d[which(!is.na(d$xbin)),], 
-%                    aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
-%                    size=I(3), aes.inherit=FALSE) + 
-%         scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
-%                            labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
-%         scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
-%                             labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
-%         annotate("segment", x=temp$x, y=temp$y, 
-%                  xend=temp$xend, yend=temp$yend, group=temp$group) + 
-%         ylim(c(10, 40)) + xlim(c(-10,30))
-%       ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")
-%       
-%       d.orig$bin <- dnew$x==15 & dnew$y==25
-%       d.orig$a <- .33+.66*d.orig$bin
-%       d.orig$l <- "Data outside bin"
-%       d.orig$l[d.orig$bin] <- "Binned Data"
-%       d.orig$size <- 2
-%       d.orig$size[d.orig$bin] <- 3
-%       d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE),
-%         data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]),                                                                                    
-%         l="Numerical Center", a=1, size=2, bin=TRUE))
-% #       ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
-% #         annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
-% #                  fill="blue", alpha=.10) + 
-% #         geom_point() +
-% #         scale_shape_manual(name="", values=c("Data outside bin" = 16, 
-% #                                              "Binned Data" = 15, 
-% #                                              "Numerical Center" = 17, 
-% #                                              "Visual Center" = 18)) +
-% #         scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
-% #                                               "Binned Data" = "blue",
-% #                                               "Numerical Center" = "blue4", 
-% #                                               "Visual Center" = "blue4"))+
-% #         scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
-% #                                             "Numerical Center" = 5, 
-% #                                             "Visual Center" = 5))+
-% #         theme_bw() +
-% #         annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
-% #         annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
-% #                  y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
-% #                  colour="blue4") +
-% # #        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
-% # #         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
-% # #                  y=d$y[which(d$l=="Numerical Center")], 
-% # #                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
-% #         scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
-% #         scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
-% #       ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
-% @
-% 
-% <<RandomBinDemo,echo=F,include=FALSE>>=
-% nl <- ncol(d.rand)
-% dnew <- d.rand
-% binning <- c(10, 10)
-% for(x in 1:(nl)){
-%   y <- d.rand[,x] - binning[x]/2
-% dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
-%   sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
-% }
-% temp <- data.frame(   
-%           x=c( -10,   0,  10,  20,  30,  -10, -10, -10, -10), 
-%        xend=c( -10,   0,  10,  20,  30,   30,  30,  30,  30), 
-%           y=c( 10,   10,  10,  10,  10,   10,  20,  30,  40), 
-%        yend=c( 40,   40,  40,  40,  40,   10,  20,  30,  40), 
-%       group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9))
-% 
-% dnew$id <- interaction(dnew[,1:(nl)])
-% idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
-% d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
-% d.rand$a <- .33 + .66*d.rand$bin
-% d.rand$xbin <- NA
-% d.rand$ybin <- NA
-% d.rand$xbin[idx] <- dnew$x[idx]
-% d.rand$ybin[idx] <- 25
-% # res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
-% qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
-%   annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
-%   annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
-%   scale_alpha(range=c(.5, 1), guide=FALSE) +
-%   scale_colour_identity(guide=FALSE) + 
-%   scale_shape(solid=T,guide=FALSE) +
-%   theme_bw() +
-% #  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
-% #        y=rep(seq(-5, 55, 10), times=6), 
-% #         shape='+', size=8, colour="gray90") + 
-%   geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
-%              aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
-%   scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
-%   scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
-%   annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(10, 40)) + xlim(c(-10,30))
-% ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
-% @
-% 
-% \begin{figure}[h]
-%   \centering
-%   \subfloat[Standard Binning]{\includegraphics[keepaspectratio=true, width=2.5in]{./figure/Binning-DataAndBin.pdf}}
-% %   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
-% %   {./figure/RandomBinningDataPoints1}}
-%   \subfloat[Random Binning]{\includegraphics[keepaspectratio=TRUE, width=2.5in]
-%   {./figure/RandomBinningDataPoints2.pdf}}
-%   \caption{Visualization of how data pairs are assigned to bins centered at (5, 25) and (15,25) under the two rectangular binning algorithms.
-Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to any of the surrounding bin centers.}
-%   \label{fig:BinningTutorial}
-% \end{figure}
-
 %--------------------------------------------------------------
@@
 
-The second stage of binning requires a frequency breakdown of the number of observations associated with each bin center, forming data triples, $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
+The second stage of binning requires a frequency breakdown of the number of observations associated with each bin center, forming reduced data triples, $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.
 Table~\ref{DataReductionExample} makes use of a small set of simulated data to show the progression from the original data (a), to the binned data (b), to the reduced binned data (c).
 The reduced binned data is sufficient for constructing the binned scatterplot.
-In cases of large data, binning greatly reduces the storage size for the information and the computation time needed to construct the plot.
+In cases of large data, binning greatly reduces the storage size for the information and the computation time needed to construct a binned scatterplot.
 
@@
 @
-
-\begin{table}[H]
+%
+\begin{table}[hbtp]
+\centering
+\caption[Original, binned and reduced binned data table example.]{\label{DataReductionExample}Original, binned and reduced binned data tables, with data storage sizes.
+Binned using standard rectangular approach with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
+}
 \begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 12 rows]{
@@
 \end{minipage}
-\caption{\label{DataReductionExample}Original, Binned and Reduced Binned Data Tables, with data storage sizes.
-Binned using standard rectangular approach with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
-}
 \end{table}
-
+%
 Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
@@
 
-% \subsection{Standard Rectangular Binning Algorithm (old - to be removed)}
-% \hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}
-% 
-% \hh{Define what $X$ is, what is $x_i$}
-% 
-% \hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }
-% 
-% \hh{I am still not sure how bin centers are defined in the end. XXX}
-% 
-% \hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}
-% 
-% \ktm{  We define the process that allocates observed values, $x_i$ for $i \in \{1, ..., n\}$, to equal sized bins in a deterministic manner as a standard binning algorithm.
-The bins are controlled by two parameters: the center of the first bin, $x^\ast_1$ and the binwidth, $\omega_X$.
-This leads to bin centers $ x_j^\ast =  x_1^\ast + (j-1)\omega_X \text{ for } j \in \{1, ..., N_X\} $,  where $N_X$ is the overall number of bins.
-}
-% 
-% 
-% Mathematically, the standard binning algorithm corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e.
-the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as
-% % \begin{center}
-% % $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
-% % \vspace{.03}
-% % $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
-% % \end{center}
-% \begin{eqnarray}\label{standbin}
-% b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
-% \end{eqnarray}
-% 
-% where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$.  
-% \ktm{ This is equivalent to treating the $j^{th}$ bin as a half-open interval that contains all observations $x_i \in [x_j^\ast - \omega_X /2 , x_j^\ast + \omega_X /2  ) $.
-After all observations have been allocated to bins, the bin count, $c_j$, is joined with the bin center to form the tuple $(x^\ast_1, c_j)$ }
-% %\hh{something went wrong with the next sentence}
-% %An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.
-Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.
-% 
-% \ktm{ A special case of standard binning divides the data range into $N_X$ equal sized bins.
-This leads to the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively.
-The first bin would then be centered at $x_{(1)} + \omega_X/2$.
-}
-% 
-% 
-% 
-% %Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width.
-A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins.
-This strategy would still fit into the framework of assigning values to the closest bin center.
-% 
-% \subsection{Random Binning Algorithm (old-to be removed)}
-% 
-% An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties.
-A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins.
-In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
-The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
-More formally,
-% %
-% \begin{eqnarray}\label{randbin}
-% b^r_X(x_i) = \left\{\begin{array}{ll} 
-%   x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
-%   x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
-%   \end{array}\right.
-% \end{eqnarray}
-% 
-% % In this manner, the probability of assigning point $x$ to bin $x^\ast_j$ is inversely proportional to the distance between the bin center of bin $x^\ast_j$ and $x$, that is, the closer $x$ is to a bin center, the higher the probability that $x$ will be assigned to that bin.
-% This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform bin sizes, unlike the standard algorithm. 
-% 
-% %To take up the previous number example again, under random binning the values 1.35, 1.55, and 2.35 are assigned to bin centers $\{1, 2, 3\}$ non-deterministic. 
-% %Value 1.35 has a 35\% probability to be assigned to bin center 2, and accordingly is assigned to bin center 1 with probability 65\%. 
-% %1.55 is closer to the boundary between bin centers and has therefore a 45\% or 55\% chance to be assigned to bin centers 1 or 2, respectively.
-% 
-% %Figure~\ref{fig:BinningTutorial} illustrates one possible outcome of the random binning process.
-% 
-% The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the nearest bin.
-``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning  in which adjacent bins are assigned a weight  depending on the distance from the point to that bin, where all weights sum to one.
-With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
-Specifically, the expectation of the random binning algorithm is the same as for linear binning.
-% %However, linear binning requires evaluation of this expectation, whereas the random binning algorithm presented here uses a random sample from the underlying distribution, which is much more computationally tractable.
-% In the next section, we examine the loss of visual and numerical information due to binning.
-% % linear binning in Gold book - random algorithm as a less computational method of linear binning
-% 
-% \subsection{Extensions to Rectangular Binning (old-to be removed)}
-% 
-% The standard and random binning algorithms are easily extendable to higher dimension.
-For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.
-In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins \sout{in $\mathbb{R}^2$} \hh{XXX J is an integer number, definitely not a real-valued vector}.
-\ktm{The wording here was not clear.
-J does not exist in $\mathbb{R}^2$, but it is being used to index 2d bins that exist in that plane.} The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as
-% %
-% \begin{eqnarray}\label{standrecbin}
-% b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
-% \end{eqnarray}
-% %
-% where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively.
-The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
-% %
-% \begin{eqnarray}\label{randrecbin}
-% b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
-% \end{eqnarray}
-% %
-% where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension.
-Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.
-% 
-% \hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation.
-$J$ is defined, as above as a multiple of $N_X$ and $N_Y$, this needs to be properly adjusted in the binning algorithm.
-It will be some linear relationship of the form $j_X\cdot N_X + j_Y$.
-}  \ktm{ I am having trouble trying to reconcile the indexing of bins in 2 dimensions.
-If we use an pair of indices it makes the extension of binning to 2 dimensions easy; but it makes the notation messier and loss functions more difficult to define.
-}
-% 
-% 
-% 
-% \subsection{Quantile Binning Algorithm (old - to be removed)}
-% 
-% We could also consider a empirical quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values.
-This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density.
-This binning will not be considered for binning the data pairs directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.
-% 
-% Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins.
-Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers.
-\hh{shouldn't you go back to the same notation as before?
-$N_X$, not $J$?
-what are the $c_j$ bin centers?
-before they were $c_j^\ast$.} \ktm{I consciously decided to define quantile binning in a way consistent with how it will be used later in the paper for discretizing the color palettes for the $c_j$ counts.
-} It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function.
-\ktm{Note this is the quantile function as defined in R; find related literature source.} \hh{which function exactly?
-There are multiple.} The bin centers are defined as the midpoint of each consecutive pair of bin boundaries $\{c^\ast_k = (Q_{k-1/K} + Q_{k/K})/2  \text{ for } k=0,\dots,K \}$.
-The quantile binning function $b^q_C(\cdot) : (c_j) \rightarrow c^\ast_k$ is defined as
-% %
-% \begin{eqnarray}\label{quantbin}
-% b^q_C(c_j) = (Q_{k-1/K} + Q_{k/K})/2 \text{ where } Q_{k-1/K} < c_j \le Q_{k/K}.
-% \end{eqnarray}
-% %
-% The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in strictly the univariate setting for this manuscript.
-
 %--------------------------------------------------------------
@@
 
-The problems with large data in scatterplots arise from overplotting, which is a form of implicit data aggregation.
-In order to keep track of overplotting, we switch to a weighted form of displays and explicitly keep track of overplotting.
+Problems with large data in scatterplots arise from over-plotting, which is a form of implicit data aggregation.
+In order to keep track of the number of observations near a given location, we switch to a weighted visual display which explicitly aggregates the data.
 The reduced binned data carries the sufficient information necessary to render the binned scatterplot.
@@
 
-Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin.
-Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
+A traditional scatterplot is comparable to a \textit{minimally binned scatterplot} -- using small enough bins such that only a single unique coordinate pair exists within each bin --  however, the "bins" of a traditional scatterplot are shaded in a binary manner with no indication of overlapping observations.
+Alpha blending as used in Figure~\ref{fig:scatter-alpha}(c) extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
 The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is usually reached well before the maximum frequency, truncating the perceivable frequency information.
-By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points.
-This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c.
+By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which may represent more points.
+This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}(d) and (e) under the same framework as plots (a)-(c).
 
@@
 \end{minipage}
-\caption{\label{binning}Series of scatterplots showing the original data (scatterplot, left), and versions of the binned data for different bin widths.
+\caption[Series of binned scatterplots of decreasing bin dimensions.]{\label{binning}Series of scatterplots showing the original data (scatterplot, left), and versions of the binned data for different bin widths.
 The visual loss from binning at 0.1 is minimal, while a bin width of 1 gives a rough approximation.}
@@
 
-In the first set of binned data, the bins and the points on the scatterplot are nearly identical, but the scatterplot contains information about overlapping points.
-The second and third sets of binned data, with bin width equal to 0.25 and 0.50 respectively, show higher-level summaries of the data but which may also provide more visually accessible information about the shape of the two-dimensional density between $x$ and $y$.
-The fourth set of binned data with bin width equal to 1.0 is nearly unrecognizable, because the bins are large enough that they provide very little additional information.
+In Figure~\ref{binning} the minimally binned scatterplot, with bin width equal to 0.1, is visually very similar to the traditional scatterplot; but importantly the binned scatterplot contains information about overlapping points.
+The second and third binned scatterplots, with bin width equal to 0.25 and 0.50 respectively, show higher-level summaries of the data but which may also provide more visually accessible information about the shape of the two-dimensional density distribution between $x$ and $y$.
+The fourth binned scatterplot, with bin width equal to 1.0, provides only a rough bivariate density display due to over-smoothing from the large bins.
 
 Loss of information occurs during the binning and rendering process.
-For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.
+For the remainder of the paper we will assume that we are using shade in binned scatterplots to represent frequencies.
 We distinguish two sources of loss in the construction of a binned scatterplot:
@@
 %, thus losing precision in the display of the true frequencies.
-
 \end{itemize}
 
-%\hh{XXX ok, I don't quite agree with the description of frequency loss here.
-(a) the rendering on a continuous color scheme IS precise.
-but on a cognitive level we are not able to read the rendered values at the same precision.
-This effectively intorduces a second stage of binning ...
-then we are on the same page again.} \ktm{Is this new description any better?}
-
-Note that even though the losses from creating a binned scatterplot may turn out to be substantial there a huge gain with respect to the original scatterplot, where information can be masked in large data situations due to over-plotting of points.
-The idea of loss from one-dimensional binning was explored by Scott using mean squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}.
-He later extended this discussion to two-dimensional binning, where he compared the mean squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
+Note that even though the losses from creating a binned scatterplot may turn out to be substantial, there is a huge gain with respect to an traditional scatterplot, where information can be masked in large data situations due to over-plotting of points.
+The idea of loss from one-dimensional binning was explored by Scott using mean integrated squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}.
+He later extended this discussion to two-dimensional binning, where he compared the mean integrated squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
 We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section~\ref{GenBinning} above.
@@
 Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively.
-The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.
+The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all line segments represents the total spatial loss, $L^S$.
 
@@
 The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
-Special consideration should be paid to the two points highlighted in red in the panel for random binning: in standard binning these points are in neighboring bins but have been allocated to the opposite bin during random binning.
+Special consideration should be paid to the two points with line segments highlighted in red in the panel for random binning.
+In standard binning these points are in neighboring bins but have been allocated to the opposite bin during random binning.
 We see that the binned scatterplot remains identical if these points are re-allocated back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
@@
 The net spatial loss for standard binned data is always  equivalent to the total spatial loss due to the deterministic bin allocations.
-For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then finding the total spatial loss from this processed binned data.
-
-\ktm{The spatial loss is a Euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
+For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for all pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then calculating the total spatial loss from this processed binned data.
+
+The spatial loss is a Euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.
 If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.
-However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations.
-}
-
-\hh{OK, better - I still don't understand the 'but' at the top of the paragraph. There seems to be some implicit negation, which I can't see it. Could you make this explicit?}
-
-%but requires a reference scale on which the magnitude may be interpretable.
-We  use the spatial loss from a ``worst-case'' binning scenario as a reference value.
-The largest displacement of points  occurs if we place all points into a single bin.
-While the total spatial loss for single bin could be made arbitrarily large if the bin center is placed far away from the data values, we will consider a reasonable specification of a bin with boundaries that span only the range of the data in each dimension centered at  $( (x_{(1)}-x_{(n)})/2$ , $(y_{(1)}-y_{(n)})/2 )$.
-The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as
-% \begin{eqnarray}\label{potspatloss}
-% L^S_p = \sum_{i=1}^{n} \sqrt{ \left(x_i-(x_{(1)}-x_{(n)})/2\right)^2 + \left(y_i-(y_{(1)}-y_{(n)})/2\right)^2 }
-% \end{eqnarray} 
-% %
-% We use net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$,  as the primary metric for assessing spatial loss incurred from binning.
-
-%\ktm{ We recognize that the spatial loss is inherently influenced by the scale and units of the data.
-Standardizing the data pairs with respect to each axis will ensure that one of the dimensions does not have an inordinate effect on the spatial loss value.
-}
-
-<<SpatialLossDemo,echo=F,include=F,cache=T>>=
-#Try Multivariate Normal and Uniform for this plot
-library(MASS)
-#MVN plot
-n=15
-set.seed(2)
-Sigma <- matrix(c(16,12,12,14),2,2)
-dat <- mvrnorm(n = n, rep(0, 2), Sigma)
-x <- dat[,1]
-y <- dat[,2]
-xcent <- rep(-5,n) ; xcent[x>0] <- 5
-ycent <- rep(-5,n) ; ycent[y>0] <- 5
-temp <- data.frame(   
-  x=c( -10,   0,  10, -10, -10, -10), 
-  xend=c( -10,   0,  10,  10, 10, 10), 
-  y=c( -10, -10 , -10 , -10, 0 , 10), 
-  yend=c(10, 10 , 10, -10, 0, 10 ))
-qplot(xcent, ycent, geom="point", size=I(4), shape=I(2)) +
-  geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
-  geom_point(aes(x=x,y=y), size=I(3))+ 
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-  theme_bw() + xlab("") + ylab("")
-      ggsave("./figure/SpatialLossDisplayMVN.pdf", width=5, height=5, units="in")
-
-#uniform plot
-set.seed(2)
-x <- runif(n, -10,10)
-y <- runif(n, -10,10)
-xcent <- rep(-5,n) ; xcent[x>0] <- 5
-ycent <- rep(-5,n) ; ycent[y>0] <- 5
-temp <- data.frame(   
-  x=c( -10,   0,  10, -10, -10, -10), 
-  xend=c( -10,   0,  10,  10, 10, 10), 
-  y=c( -10, -10 , -10 , -10, 0 , 10), 
-  yend=c(10, 10 , 10, -10, 0, 10 ))
-qplot(xcent, ycent, geom="point", size=I(4), shape=I(2)) +
-  geom_segment(aes(x = x, y = y, xend = xcent, yend = ycent)) + 
-  geom_point(aes(x=x,y=y), size=I(3))+ 
-  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-  theme_bw() + xlab("") + ylab("")
-      ggsave("./figure/SpatialLossDisplayUnif.pdf", width=5, height=5, units="in")
-@
-
-% \begin{figure}[h]
-%   \centering
-%   \subfloat[Spatial Loss MVN Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayMVN.pdf}}
-%   \subfloat[Spatial Loss Uniform Data]{\includegraphics[keepaspectratio=true, width=3in]{./figure/SpatialLossDisplayUnif.pdf}}
-%   \caption{Visualization of spatial loss.
-The distance between points and bin centers adds up to the spatial loss due to binning.
-}%\ktm{Redundant plots using MVN and Uniform Data.
-Which makes a better plot?}}
-%   \label{fig:LossDisplay}
-% \end{figure}
+However, if the two variables do not share the same units or the same magnitude of values it is advisable to standardize the variables prior to binning, thus making the spatial loss more universally interpretable as a distance in units of standard deviations.
+
 %------------------------------------------------
@@
 It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates.
-The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from overplotting, which can be immense for large data sets.
+The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets.
 When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location.
@@
 
-Bin counts can be displayed using a continuous shading scale in HSV color space \ktm{(reference needed)} and thus theoretically have no frequency loss.
-While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.
-Color perception of shade is extremely context sensitive allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
-It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
+Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly.
+While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise.
+Color perception of shade is extremely context sensitive allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend.
+It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
 Figure~\ref{fig:context} shows an example of the context sensitivity of colors.
-Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each of the points clouds.
-
-\begin{figure}[ht]
+Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.
+
+\begin{figure}[hbtp]
 \centering
@@
 @
-\caption{\label{fig:context}Assess the colors in this scatterplot.
+\caption[Demonstration of the contextual sensitivity of shade perception.]{\label{fig:context}Assess the colors in this scatterplot.
 Are the triangles in the lighter cloud of points as dark as the triangles in the darker cloud of points?
@@
 
-Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
-As a result, a binning width which produces seven of fewer frequency categories is preferable.%, while minimizing numerical loss within that constraint.
-
  Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
@@
 
+Research suggests that even under optimal conditions, we can effectively compare only about seven color hues simultaneously, and that we are even more limited in terms of distinguishing shade \citep{colorperception}.
+This provides a physical upper limit on the amount of frequency variation we can perceive through color.
+As a result, a frequency binning which produces seven of fewer frequency categories is preferable.
+
 <<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Continuous and discrete color scales for counts in the range 0 to 1000.
-Frequency binning done using standard rectangular binning with origin $\\beta_0 = 0$ and binwidth $\\omega_X = 200$.
-">>=
+Frequency binning done using  standard rectangular binning with origin $\\beta_0 = 0$ and binwidth $\\omega_c = 200$.", fig.scap="Continuous and discrete color scales for counts.">>=
 #J= 1
@@
 
-The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
+The goal of binning the frequencies and using a ordinal shading scheme is to quantify the imprecision in visually extracting frequency information.
+It does so by using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$.
 The \textit{total frequency loss}, $L^F$, is defined as
@@
 
-Frequency data consists of counts, which most commonly exhibit skew densities, i.e.\ there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
-The use of quantile binning is promising in the case of frequency binning because it seeks to place the same number of values in each frequency bin, thus balance the number of tiles at each shade.
-
-Another alternative is to use a log transformation which produces a more symmetric distribution of frequency information, increasing perceptual resolution.
+Frequency data consists of counts, which commonly exhibit skew densities, i.e.\ there are usually a lot of bins with small bin counts and a few bins with extremely large frequencies.
+The use of quantile binning is promising in the case of frequency binning because it seeks to place the same number of bins in each shaded group.
+An alternative is to use a log transformation which produces a more symmetric distribution of frequency information, increasing perceptual resolution.
 This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}.
-Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph.
+Using a logarithmic mapping of frequency to the shade aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph.
 The \textit{log frequency loss}, $L^{\log F}$, is defined as
@@
 
-%  Similar to spatial binning, \hh{we define} a worst-case scenario for frequency binning that provides context to the magnitude of the total frequency loss.
-If we use a single frequency bin then all tiles of the binned scatterplot are rendered with the same shade; thus the shade  represents the average bin density \hh{XXX this is different from the spatial loss - visually the shade would be half way between lowest and highest frequency, wouldn't it?}, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.
-We define the total frequency loss in the single shade scenario as the \textit{potential frequency loss}, $L^F_p$,  where
-% %
-% \begin{eqnarray}\label{potfreqloss}
-% L^F_p = \sum_{\ell=1}^{L} (c_\ell-\bar{c})^2.
-% \end{eqnarray}  
-% %
-% Thus total frequency loss after binning can be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.
-Similarly we define the \textit{potential log frequency loss}, $L^{\log F}_p$, as
-% %
-% \begin{eqnarray}\label{potlogfreqloss}
-% L^{\log F}_p = \sum_{\ell=1}^{L} (\log(c_\ell+1)-\log(\bar{c}+1))^2,
-% \end{eqnarray}  
-% % 
-% which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted.
-\hh{XXX could you give a couple more details what you mean by cases where frequency binning was conducted?}
-% 
-% \subsection{Old Loss Definitions (to be removed)}
-% \ktm{The following \textit{numerical, visual and freq} loss descriptions are from the old list and will be removed: }
-% \begin{itemize}
-% \item {\it Numerical Loss}, $L^\text{N}$, occurs when points in the data set are reduced to a single points at the bin means.
-Allocating points to the numerical center, or the mean of the data corresponding to that bin, causes a loss of information about the exact location of the points.
-% 
-% \item {\it Visual Loss}, $L^\text{V}$, is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile.
-This visual center will generally different from the numerical mean, causing some bias in the perception.
-Thus visual loss for a bin, $L_i^\text{N}$, is the Euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
-The visual loss is the sum of these bin distances.
-% 
-% \item {\it Frequency Loss} results from our inability to render the frequency information accurately.
-Different rendering methods for the frequency information will lead to different losses.
-For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent this information.
-The number of colors that can be differentiated and mapped back to frequency information accurately is fairly limited.
-Note that even though this loss will turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where frequency information is masked in large data situations due to over-plotting of points.
-% 
-% % 
-% % , so we must balance perceptual limits of frequency information with the limits of overplotting single points. 
-% 
-% 
-% % which may not be the center of the data within the bin, causing bias and numerical loss.
-% % 
-% % Visual loss occurs in the graphical display of the data, as individual points are represented as tiles, and tiles are perceived according to their visual center, rather than according to the center of the data.
-% % 
-% %  
-% \end{itemize}
-% 
-% We explore the different types of loss in the remainder of this section in more detail.
-For clarity we will again be specifying loss using notation for a one dimensional binning process.
-This notation can then be extended to higher dimensions in applications.
-% 
-% \subsection{Numerical and Visual Loss}
-% Reducing data points $\mathbf{x}$ to the visual bin centers $\mathbf{x}^\ast$ results in a loss of information, as there is a reduction in the variability of the data when data points are transformed from individual values to bin centers on a grid.
-% If the visual bin centers are the same as the numerical centers for each bin, then any loss could be strictly attributed to difference between points and the bin centers. 
-% The problem is that the visual center and the numerical center for bins will rarely match, which introduces  visual bias  to the graphical representation of the data, as we perceive the bin center to be the physical middle of the symbol representing the data.
-% 
-% Therefore we distinguish two contributions to the total loss. 
-% Numerical loss results from the binning process as points are reduced to the numerical center point.
-The loss for an individual point, $L_i^\text{N}$, is the Euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$.
-The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
-% Visual loss is due to the distance between the numerical centers and the visual centers.
-The visual loss for each bin, $L_j^\text{V}$, is the Euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$.
-The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$).
-% This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.
-% This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions apply to either approach to binning.
-% 
-% The magnitude of the numerical and visual losses are clearly dictated by the scale on which the data was recorded, thus and the loss values have no direct interpretation.
-However, we can examine both the numerical and the visual loss from the perspective of proportion of the greatest possible loss.
-% To do this for numerical loss we consider the binning scenario that would always lead to greatest possible loss; allocating all data points into a single bin.
-Thus we can define the greatest possible numerical loss as
-% 
-% \begin{center}
-% $ L_0^\text{N} = \sum_{i=1}^{n} \sqrt{(x_i-\bar{x})^2 + (y_i-\bar{y})^2}  $, 
-% \end{center}
-% 
-% where $\bar{x}$ and $\bar{y}$ are the sample averages for x and y respectively.
-After we select a binning strategy and binwidth we can then assess the proportion of possible numerical loss incurred due to binning by computing $L^\text{N} / L_0^\text{N}$.
-
-%-------
-% Cut "total loss" stuff, because this was all a relic of "spatial loss" part of paper
-%Let us denote the difference between point $x_i$ from its bin center as $S_i = \left|x_i-b(x_i)\right|$ for each point in the data set $i=1, ..., n$.
-%We may then combine these losses from the entire data set to be defined as 
-%\[L_s(\mathbf{x}) = 1/S^2_{\emptyset} \cdot \sum_{i=1}^{n} S^2_i\] 
-%where $S_\emptyset^2$ is the loss which results when the entire data set is reduced to a single bin,
-%$S^2_{\emptyset} = \sum_i (x_i-\tilde{x})^2$ for the single bin center $\tilde{x}$.
-%The loss depends on the choice of bin center: candidates for bin centers are the mean of the points within the bin or the visual center of the bin. 
-%Fortunately we can show that the total loss perceived is partitioned by these two components of loss, %and it holds (see appendix \ref{proof:partition} for a proof):
-%\begin{eqnarray}\label{eqn:partition}
-%L_s(\mathbf{x}) =  1/S^2_{\emptyset} \cdot \left(\sum_i^n L^N_i + \sum_j^{N_x} L^V_j\right)
-%\end{eqnarray}
-%-------
-<<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
-set.seed(53022)
-data <- data.frame(x=rnorm(200, 0, 2), y=rnorm(200, 0, 2), Freq=1)
-data$xbin <- round(data$x)
-data$ybin <- round(data$y)
-data.means <- ddply(data, .(xbin,ybin), summarise, xmean=mean(x), ymean=mean(y))
-data <- merge(data, data.means)
-
-ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
-ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
-@
-
-%The decomposition also holds for the random binning algorithm as well as the standard binning algorithm. 
-%The expected total loss for either algorithm depends on the distribution the underlying data. 
-%The total loss due to binning is $\sum_{i=1}^n (x_i - b(x_i))^2$, where in the standard algorithm, $b(x)$ maps $x$ to the closest bin center, and in the random algorithm, $b(x)$ is assigned probabilistically.
-
-%---
-% I believe we showed that this is only true when "rebinning" previously binned data and not deterministic when randomly binning at doubled bins from original data, thus not practically useful information so commented out
-%It is worth noting that the loss function is a monotonically increasing function of binwidth.
-An interesting feature of the random binning algorithm is that the increase in loss that we experience by doubling the bin width is deterministic, even though the assignment of bins is not.
-This holds for repeated doubling of the bin width.
-A proof of this can be found in appendix \ref{proof:constantloss}.
-Doubling bin widths is, particularly in the framework of visualizations, a very natural step, making this an important property of the random binning.
-%---
-
-% \begin{figure}[hbtp]
-%     \centering
-% 		\subfloat[Numerical and Visual Loss]{\includegraphics[keepaspectratio=true, width=.46\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
-%     \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/FreqLoss.pdf}}
-% \caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.
-(b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}
-% 	\end{figure}
-% 
-% \subsection{Loss of Frequency Information (old writeup - to be reworked or removed)} 
-% Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently.
-By employing a color scheme for filling tiles according to the observed frequencies corresponds to a binning of the frequencies into a few different levels, with losses stemming from the same source as the previously discussed numerical loss.
-% We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_k\}$ with $1 \le k \le n_C$.
-An observed frequency $c_j$ with $1 \le j \le N_x \times N_y$, where $N_x$ and $N_y$ are the number of bin divisions in the X and Y directions respectively,  is assigned to the closest frequency level $c^\ast_k = b(c_j)$.
-% The \textit{unscaled} frequency loss is then defined as
-% \begin{eqnarray}\label{eqn:frequency}
-% L_\text{Freq} = \sum_{j=1}^{N_x N_y} \left(c_j - b(c_j)\right)^2
-% \end{eqnarray}
-% 
-% % When we consider loss of frequency information, we must consider both dimensions simultaneously, as frequency information is based on the combination of binning in $x$ and  $y$ direction.
-We define loss of frequency information as: \hh{what's that in words?}
-% % $$ L_\text{Freq} = \sum_j \left(c_j - \frac{1}{n_X n_Y}\sum_j c_j\right)^2,$$ 
-% % where $n_X, n_Y$ are the number of bins in $x$ and $y$, respectively. 
-% 
-% The loss of frequency information due to binning results from tiling over larger areas, thus results from {\it smoothing} the frequency variation (see Figure~\ref{fig:stdBinTutorial2}). 
-% 
-% Frequency data consists of counts, which most commonly exhibit skew densities, i.e.
-there are usually a lot of cells with small cell counts and a few cells with extremely large counts.
-A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution.
-This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}.
-Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph.
-This suggests that frequency loss be calculated as
-% %
-% \begin{eqnarray}\label{eqn:logfrequency}
-% L_{\log \text{Freq}} = \sum_{j=1}^{N_x N_y} \left(\log(c_j+1) - \log(b(c_j)+1)\right)^2
-% \end{eqnarray}
-% %
-% Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells. 
-% 
-% This numerical assessment of frequency loss does not account for limitations in human perceptual ability.
-Research suggests that under optimal conditions, we can effectively compare about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.
-% As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint. 
-% 
-% To examine this further, we simulated points from three different distributions: Uniform, Normal, and Exponential to determine under what circumstances log frequency scaling is performing better than linear frequency scaling.
-Figure~\ref{fig:logcolor} gives an overview of this simulation.
-In the top row binned scatterplots with a color scheme based on linear frequency scaling is used, the bottom row shows the same data based on log frequency scaled colors.
-% 
-% For uniformly distributed data (right column) the choice of color scheme does not have a huge impact, but under both the normal and exponential distribution, log frequency scaling results in categories which account for more uniform proportions of the data (as can be seen by the color strips in the second and third row of Figure~\ref{fig:logcolor}).
-This is desirable because it provides more discernible frequency information.
-% \begin{figure}[hbtp]
-% \centering
-% <<FrequencyCategories,echo=FALSE, results='hide', cache=TRUE>>=
-% options(stringsAsFactors=FALSE)
-% library(plyr)
-% library(ggplot2)
-% library(RColorBrewer)
-% set.seed(5032)
-% data <- rbind(data.frame(x=runif(1000, 0, 20), y=runif(1000, 0, 20), 
-%                          Freq=1, Distribution="Uniform"), 
-%               data.frame(x=rnorm(1000, 10, 3), y=rnorm(1000, 10, 3), 
-%                          Freq=1, Distribution="Normal"),
-%               data.frame(x=rexp(1000, .3), y=rexp(1000, .3), 
-%                          Freq=1, Distribution="Exponential"))
-% data <- data[which(abs(data$x-10)<10 & abs(data$y-10)<10),]
-% # qplot(data=data, x=x, y=y, geom="point")+facet_wrap(~Distribution)
-% 
-% data.bin <- data
-% data.bin[,1:3] <- round(data[,1:3])
-% data.bin <- ddply(data.bin, .(Distribution, x, y), summarise, Freq=sum(Freq))
-% data.bin <- data.bin[order(data.bin$Distribution, data.bin$Freq),]
-% # 
-% # 
-% # qplot(data=data.bin, x=Freq,geom="blank") + geom_density(aes(y=..scaled..),adjust=2) + facet_wrap(~Distribution, scales="free") + ylab("Density")
-% # 
-% # qplot(data=data.bin, x=Freq, geom="blank") + geom_density(aes(y=..scaled..), adjust=2) + facet_wrap(~Distribution,scales="free") + scale_x_continuous(trans="log", breaks=c(1, 5, 10, 20, 40, 60)) + ylab("Density")
-% # 
-% 
-% get.upper <- function(x){
-%   num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][2])
-%   num <- sapply(num, function(i) substr(i, 1, nchar(i)-1))
-%   return(as.numeric(num))
-% }
-% get.lower <- function(x){
-%   num <- sapply(x, function(i) strsplit(i, ",", fixed=TRUE)[[1]][1])
-%   num <- sapply(num, function(i) substr(i, 2, nchar(i)))
-%   return(as.numeric(num))
-% }
-% 
-% labelsToValues <- function(x){
-%   return(data.frame(lower=get.lower(x), upper=get.upper(x)))
-% }
-% 
-% 
-% linearbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(x$Freq, 7)), label=as.character(cut(x$Freq,7)), id=1:nrow(x))))[,2:4])
-% linearbins$fill <- factor(linearbins$fill)
-% 
-% x <- subset(linearbins, Distribution=="Uniform" & fill==1)
-% linearbins2 <- ddply(linearbins, .(Distribution, fill),
-%                      function(x){
-%                        temp <- labelsToValues(x$label[1])
-%                        return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
-%                      })
-% linearbins2 <- ddply(linearbins2, .(Distribution), transform, height=height/sum(height))
-% 
-% 
-% 
-% logbins <- cbind(data.bin, ddply(data.bin, .(Distribution), function(x) return(data.frame(fill=as.numeric(cut(log(x$Freq), 7)), label=as.character(cut(log(x$Freq),7)), id=1:nrow(x))))[,2:4])
-% logbins$fill <- factor(logbins$fill)
-% 
-% x <- subset(logbins, Distribution=="Uniform" & fill==1)
-% logbins2 <- ddply(logbins, .(Distribution, fill),
-%                   function(x){
-%                     temp <- exp(labelsToValues(x$label[1]))
-%                     newlabel <- paste(c("(", round(temp[1], 3), ",", round(temp[2], 3), "]"), sep="", collapse="")
-%                     return(cbind(temp, height=nrow(x)/(temp$upper-temp$lower), label=unique(x$label)))
-%                   })
-% 
-% logbins2 <- ddply(logbins2, .(Distribution), transform, height=height/sum(height))
-% 
-% colors <- brewer.pal(9, "Blues")[-c(1:2)]
-% 
-% multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, heights=NULL) {
-%   require(grid)
-%   
-%   # Make a list from the ... arguments and plotlist
-%   plots <- c(list(...), plotlist)
-%   
-%   numPlots = length(plots)
-%   
-%   # If layout is NULL, then use 'cols' to determine layout
-%   if (is.null(layout)) {
-%     # Make the panel
-%     # ncol: Number of columns of plots
-%     # nrow: Number of rows needed, calculated from # of cols
-%     nrow = ceiling(numPlots/cols)
-%     
-%     if (is.null(heights)) heights = unit(rep(1, numPlots), "null")
-%     layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-%                      ncol = cols, nrow = nrow)
-%   }
-%   
-%   if (numPlots==1) {
-%     print(plots[[1]])
-%     
-%   } else {
-%     # Set up the page
-%     grid.newpage()
-%     pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout), heights=heights)))
-%     
-%     # Make each plot, in the correct location
-%     for (i in 1:numPlots) {
-%       # Get the i,j matrix positions of the regions that contain this subplot
-%       matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-%       
-%       print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-%                                       layout.pos.col = matchidx$col))
-%     }
-%   }
-% }
-% 
-% p1 <- qplot(data=linearbins, x=x, y=y, geom="tile", fill=fill)+facet_wrap(~Distribution) +
-%   scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
-%   ggtitle("(a) Joint Density: Linear Binning") +  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-% 
-% X <- data.frame(fill=with(linearbins, rep(fill, Freq)), Distribution = with(linearbins, rep(Distribution, Freq)))
-% X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
-% p2 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(1)) + 
-%   scale_fill_manual(values=colors, name="Frequency", guide="none") + 
-%   facet_grid(.~Distribution, scales="free") +
-%   xlab("Count") + theme_bw() + ylab("") +
-%   ggtitle("(b) Color Frequency: Linear Binning") + 
-%   theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
-%   theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
-% # linear binning
-% pdf(file="p2.pdf", width=8, height=3)
-% p2
-% dev.off()
-% 
-% # p2 <- qplot(data=linearbins2, xmin=lower, xmax=upper, ymin=0, ymax=height*1000, fill=fill, geom="blank", asp=1.5) + 
-% #   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
-% #   scale_fill_manual("Tile Color", values=colors)+
-% #   facet_wrap(~Distribution, scales="free") + 
-% #   theme_bw() + scale_x_continuous("Bin Frequency") + 
-% #   scale_y_continuous("") + ggtitle("(b) Distribution of Frequency: Linear Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-% 
-% # p4 <- qplot(data=logbins2, xmin=lower, xmax=upper, ymin=0, ymax=1000*height, fill=fill,  geom="blank", asp=1.5) + 
-% #   geom_rect(colour=I("grey50"), show_guide=FALSE, size=0.5)+
-% #   scale_fill_manual("Tile Color", values=colors)+
-% #   facet_wrap(~Distribution, scales="free") + 
-% #   theme_bw() + scale_x_continuous("Bin Frequency") + 
-% #   scale_y_continuous("") + ggtitle("(c) Distribution of Frequency: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-% 
-% X <- data.frame(fill=with(logbins, rep(fill, Freq)), Distribution = with(logbins, rep(Distribution, Freq)))
-% X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
-% p4 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(NULL)) + 
-%   scale_fill_manual(values=colors, name="Frequency", guide="none") + 
-%   facet_grid(.~Distribution, scales="free") +
-%   xlab("Count") + theme_bw() + ylab("") +
-%   ggtitle("(c) Color Frequency: Log Binning") + 
-%   theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
-%   theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
-% 
-% 
-% p3 <- qplot(data=logbins, x=x, y=y, geom="tile", fill=fill, asp=1.0)+facet_wrap(~Distribution) +
-%   scale_fill_manual(values=colors, name="Frequency", guide="none") + theme_bw() +
-%   ggtitle("(d) Joint Density: Log Binning") + theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,0), "cm"))
-% 
-% multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
-% #ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
-% @
-% %\includegraphics[width=\textwidth]{figure/MultipleDensityFigure.pdf}
-% \caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
-% \end{figure}
-% 
-% Frequency loss is much more difficult to normalize, as we must balance the partial loss of frequency information due to over-plotting with the total loss of frequency information that results if we use only a single bin to categorize the data.
-% For the purposes of numerical assessment of frequency loss, we calculate the baseline frequency loss as the loss corresponding to a minimal bin width given by the smallest non-zero absolute difference of successive values in the original data set, as shown in Figure~\ref{binning}.
-
 %-----------------------------------------------------------------------------
@@
 
-Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with the  binning algorithm.
+Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm.
 This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
@@
 The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
-%\hh{XXX what other purpose does this scatterplot have besides showing the distributions?}
 
@@
 These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data.
-Real data is recorded to only the number of digits that the measurements allow, and in many cases rounded even further.
+Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
 
@@
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
-By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.
+By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.
 
@@
 For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension.
-To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning: For equal bin widths, the net spatial loss under standard binning is always less than or equal to the net spatial loss under random binning.
+To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning.
+For equal bin widths, the net spatial loss under standard binning is always less than or equal to the net spatial loss under random binning.
 This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.
 
-Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
+Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning the fine resolution simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
 It is apparent that the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
@@
 Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
-Another pertinent argument for using larger bins is that it will smooth over larger regions if visualizing the large scale density structure is of primary interest.
+Another pertinent argument for using larger bins is that it will smooth over larger regions if we are primary interested visualizing the large scale density structure.
 
@@
 However, as bin sizes become increasingly fine,  the bias decreases as the plot shows more precise spatial locations.
-When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate is spatially exact; thus making the binned scatterplot a consistent visual estimator.
+When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate is spatially exact.
+Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator.
 We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
@@
 These gaps of white space are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density.
-\hh{The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with an evaluation.}
+The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with visual evaluation.
 
@@
 
-To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension.
+To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension.
 Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
-Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.
+Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins.
 Note that random binning is effective at smoothing out the artificial striping patterns when non-integer multiples of the data resolution are selected.
@@
 For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density in outermost bins.
-Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data with two different origin choices at (0,0) and (-9,-9).
+Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data with two different origin choices: (0,0) and (-9,-9).
 The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions.
@@
 
-<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10 by 10 unit bins with origins at (0,0) and (-9,-9).
-The binning origin at (0,0) leads to 9\\% less net spatial loss than (-9,-9); with losses of 391,067 units and 420,458 units, respectively.">>=
+<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10X10 square bins with origins at (0,0) and (-9,-9).
+The bold lines at x=0 and y=0 denote the lower bounds of the data.">>=
 
@@
                      aspect.ratio=1, panel.grid= element_blank()) +
-  xlab("") + ylab("")
+  xlab("") + ylab("")+ geom_hline(yintercept=0) + geom_vline(xintercept=0)
+
 p2 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5offset[[1]]) +  
@@
                      aspect.ratio=1, panel.grid= element_blank()) +
-  xlab("") + ylab("")
+  xlab("") + ylab("") + geom_hline(yintercept=0) + geom_vline(xintercept=0)
 
@@
 where $x_{(1)}$ and $y_{(1)}$ are the minimal data values in $x$ and $y$. %, and $(o_x, o_y)$ is the origin offset. 
-Thus the offset indicates the number of units in each dimension to shift the origin below the highest possible origin.
+Thus the offset indicates the number of units in each dimension to shift the binning origin below the origin naturally encouraged by the data.
 
 It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units  minimizes the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension and $\alpha_y$ units in the $Y$ dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
-
-%\hh{`$\alpha/2$ origin offset' this an implicit naming.
-You have to make it either explicit or write this out.
-At the moment it is imprecise, because you mean this offset to be below the minimal value of the data.
-XXX }
-
 In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
-%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss.
-it's a bit confusing with the two minima floating around.
-It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
-Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm.
-Note that for simplicity, only origin offsets  with equal offsets in each dimension are included in this simulation.
-Since the data resolution  consists of one unit in each dimension, we  pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees.
+
+Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is shifted while using standard rectangular binning for the coarse simulated data sets.
+Note that for simplicity, changes to the origin offset are made equally in each dimension, thus shifting the origin at a 45 degree angle.
+Since the coarse data has a 2X2 unit resolution, we pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the theoretical assumptions to differing degrees.
 The round glyphs indicate the origin offset where the net loss reaches an absolute minimum in the simulation.
-For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.
-%\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
-For the bivariately skewed exponential data, the one unit origin offset minimizes the net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.
-
-<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=4, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a resolution of 2 units at various sized (2,4,6,8,10) square bins over the range of possible origin offsets.
-The vertical gray line in each facet indicates where the origin is offset 1 unit below the minimum data value.">>=
+For the two symmetrically distributed data sets, normal and uniform, the origin offset of (1,1) was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.
+For the bivariately skewed exponential data, the origin offset of (1,1) minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% above the minimum spatial losses for the 8X8 and 10X10 unit bins, respectively.
+
+<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=4, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a 2X2 unit resolution using various sized square bins over the range of possible origin offsets.
+The vertical gray line in each facet indicates the origin offset of (1,1).">>=
 ##  Check out the spatial loss for the data that is recorded to resolution = alpha at various origin offsets
@@
               minloss = min(totalSpatialLoss))
+lossOrigin$binsize <- factor(paste(lossOrigin$widthx,"X",lossOrigin$widthx,sep=""),
+                             levels=c("2X2","4X4","6X6","8X8","10X10"))
+mins$binsize <- paste(mins$widthx,"X",mins$widthx,sep="")
 
@@
       data=subset(lossOrigin,bintype=="standard"), size=I(1))+ 
-  facet_grid(distrib~widthx, scales="free_x", space="free_x") + theme_bw() + 
+  facet_grid(distrib~binsize, scales="free_x", space="free_x") + theme_bw() + 
   theme(panel.grid= element_blank(), legend.position="none")+
@@
 
-% 
-% \ktm{
-% \begin{itemize}
-% \item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytic proof)
-% \item Bin width: 
-%   \begin{itemize}
-%     \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
-%     \item for fine data any bin width is fine
-%     \item for coarse data need bin widths $\omega \ge \alpha$ (show plot with white space gaps)
-%     \item for coarse data bin widths should be scalar multiple of $\alpha$ (show plot with strips)
-%     \item Does the distribution type matter? 
-%   \end{itemize}
-% \item Origin: 
-%   \begin{itemize}
-%     \item Doesn't much matter for fine resolution data but it does for coarse data.
-%     \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
-%   \end{itemize} 
-% 
-% \end{itemize}
-% }
-
-%-----------------------------------------------------------------------------
-%\subsection{Choice of bin width}
-
-%-----------------------------------------------------------------------------
+% ----------------------------------------------------------------------------
 
 \subsection{Frequency Binning Specifications and Frequency Loss}
-
-%\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception.
-We use the binning to reflect what happens in practice.
-The binning then allows us to measure what kind of loss occurs because of the limitations of the human cognitive system.}
-
-Spatial binning reduces data to contain the center and frequency information for all bins.
-While bin frequencies could be mapped continuously to a precisely rendered shade, it is naive to believe that this information can be extracted at the same resolution due to limitations of our perception.
-Implicitly, we lose information  when (imperfectly) mapping the shade of a tile to its frequency.
-In order to make this loss explicit, we introduce a second stage of binning and discretize the color scale for frequencies in the spatially binned scatterplot.
-
-Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots.
-On the right frequency is mapped continuously to color; on the left  there is a varying numbers of  standard binned frequencies.
-It is trivial to discern differences between similarly shaded tiles
-when there are only four shades.
-This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colors better than in the continuous shading.
-This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. We will therefore use an upper limit of ten frequency bins in our exploration of frequency loss.
+\label{FreqBinningSpec}
+
+The reduced binned data from spatial binning contains the center and count information for all bins.
+The bin frequencies may be mapped continuously to a precisely rendered shade, however it is naive to believe that human perception will be able to perfectly extract that information.
+There is implicitly loss of frequency information occurring when the shade of a tile is visually mapped back to a frequency through the use of a shading scale index.
+Bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot, thus making the loss explicit.
+
+Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups.
+If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes much more difficult at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading.
+This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}.
+Our exploration of frequency loss will focus on frequency binning with at most ten bins because above this we experience implicit frequency from perceptual bounds that are not well reflected in the explicitly defined frequency loss.
 
@@
 The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
-
-There is a huge difference in the magnitude of frequency loss between the distributions: frequency losses are the highest in the scatterplot of exponentially distributed data, much lower for the normally distributed data and lowest in the case of the uniform data.
-For all distributions and both binning algorithms, frequency loss decreases as the number of frequency bins increases. This initially sharp decrease  largely flattens out at about four bins.
-Thus four and seven bins for color shades allow us to reduce frequency loss while allowing for easy distinction between frequency groups in the binned scatterplot.
-
-Lastly, for small number of bins, the frequency loss from standard frequency binning is higher than loss from quantile-based binning.
-However, in all three simulated data sets, the losses drop to similar levels for both algorithms when four or more frequency bins are used.
-
-The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data.
-Frequency binning using log counts exhibits a very similar behavior to binning of untransformed frequencies.
-The difference in scales makes it impossible to directly compare frequency loss and log frequency loss. 
-But the same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
+We first notice the large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data, and lowest for the uniform data.
+We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms.
+This relationship largely flattens out after the fourth frequency bin, each subsequent bin reducing the frequency loss less than the previous.
+Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot.
+Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used.
+But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.
+Therefore, it is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies.
+
+Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
+The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} counts for bins from the same sets of simulated data.
+Log frequency binning behaved very similarly to standard frequency binning, where log frequency loss decreased as the number of frequency bins increased.
+The same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
 
 <<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
+",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data.
 ">>=
@@
 
-%To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).
-Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.
-As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities \citep{colorperception}.
-
-%\hh{XXX I don't think that figure 12 is doing its job - and for that matter, we should talk about what its job is.
-There are too many different things going on in it, and the figure might feed better into your argument when things are separated out.
-}
-
-%\hh{XXX Let's try a first picture with a continuous scale and three discrete binned frequencies of four, seven, and ten bins.
-This should drive home the message that Healey is right, and we can't distinguish more than seven shades anyway, i.e.
-we don't see a difference between the continuously shaded plot and the one with discrete shades.
-This re-enforces our approach to treat color as a categorical variable, because we perceive it that way (otherwise why would we?)}
-
-%\hh{XXX figure 13 is a plot with a screaming message: never ever use a raw data mapping to show color.
-use quantiles or log frequencies.
-It doesn't scream that at the moment, so emphasize it in the text.}
-
-%\hh{XXX the rest of figure 12 should probably go into a figure 14 - to show that quantile binning and log frequency binning yield very similar results.
-We need to fix the legend.
-Right now they are atrocious (and I am aware that I am responsible for them, so I can say that :) )}
-
-Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem in binned scatterplots: the distribution of bin frequencies is heavily right skewed.
-High frequency bins visually overshadow any structure in low frequency bins.
-
+Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
 Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.
@@
 
-As losses from log-transformed and un-transformed frequencies are not directly comparable, the choice is guided by desired interpretation. 
-For standard log frequency binning, the shade is to be interpreted as an ordinal indicator of the groups based on equally spaced groupings of log bin frequencies, whereas for quantile frequency binning the shade denotes groups based on frequency quantiles.
-This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot in univariate visualization.
-
+Since the loss scales are not comparable, the choice is guided by desired interpretation. 
 Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
-\ktm{Both shading schemes are effective at reducing the visual impact of the highest density at the center, allowing for the differences in the surrounding bin frequencies to be emphasized.
-It should be noted that the frequency binning algorithm produced binned scatterplots that visually very similar, but this will not always be the case.
-It occured in this scenario because the distribution of log bin frequencies from the bivariate normal data is nearly uniform, thus the standard log binning aligns closely with the quantile binning.}
-
-%We also see that the shading of the binned scatterplot is invariant to the log transformation for quantile frequency binning but not for standard; because a monotone one-to-one transformation will not alter groupings based on order, but will change groupings based on value magnitude.
+For standard log frequency binning, the shade is to be interpreted as an ordinal indicator based on equally spaced groupings of log bin frequencies; whereas for quantile frequency binning the shade denotes groups based on frequency quantiles.
+This is analogous to the difference in interpreting a histogram of log transformed data and a boxplot of untransformed data in univariate visualization.
+Both shading schemes are effective at reducing the visual impact of the highest density bins near the center of each plot, allowing for the differences in the surrounding bin frequencies to be emphasized.
+It should be noted that both frequency binning algorithms produced very similar looking binned scatterplots, but this will not always be the case.
+It occurred in this scenario because the distribution of log bin frequencies from the bivariate normal data is nearly uniform, thus the standard log frequency binning aligns closely with groupings from the quantile frequency binning.
 
 \begin{figure}[hbtp]
-%  \begin{minipage}{.31\linewidth}
-% <<noFreqBin, echo=FALSE, out.width='.8\\linewidth',fig.width=4, fig.height=4>>=
-% binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
-% qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
-%             main="No Freq Binning") +
-%   xlab("") + ylab("")+
-%   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
-%                       guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
-%                                            title.position = "top")) + 
-%   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-%                      panel.grid= element_blank(),
-%                      plot.margin = unit(c(.1,.1,.1,.1),"cm") )
-% @
-%  \end{minipage}
-%\hfil\begin{minipage}{.685\linewidth}
 <<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=4.8>>=
@@
 @
-%\end{minipage}
-\caption{\label{freqbinning}Simulated bivariate normal data spatially binned using standard algorithm with 5X5 unit bins and origin at (0,0).
+\caption[Frequency binned scatterplots for simulated bivarate normal data.
+]{\label{freqbinning}Simulated bivariate normal data spatially binned using standard algorithm with 5X5 unit bins and origin at (0,0).
 Binned scatterplots with five frequency bins from quantile frequency binning (left) and standard log frequency binning (right).
@@
 
-The exploration of loss in spatial and frequency information in the process of creating a binned scatterplot yields a number of important properties and practical recommendations for their construction.
+The exploration of spatial and frequency loss in the process of creating a binned scatterplot yields a number of important properties and practical recommendations for their construction.
 These recommendations and loss properties will now be demonstrated through two examples using real bivariate data.
@@
 In a second example, we will investigate the relationship between scheduled and actual departure times for American commercial airline data.
-
-% \subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
-<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=
-# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
-computeLoss <- function(binning, data){
-  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
-  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
-  datplot <- rbind(cbind(unique(lRandom$NewData), Algorithm="Random"), cbind(unique(lMean$NewData), Algorithm="Standard"))
-  datplot$xbin <- round(datplot[,1]/binning[1],0)*binning[1]
-  datplot$ybin <- round(datplot[,2]/binning[2],0)*binning[2]
-  #datplot2 <- ddply(datplot, .(G, SO,Algorithm), summarize, count=length(G))
-    qplot(xbin, ybin, geom="tile", data = datplot, fill = Freq , 
-          ,main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
-        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
-    ggsave(paste("./RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
-  return(rbind(
-    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$Loss), 
-    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lMean$Loss)))
-}
-
-bins <- rbind(c(2,2), c(5,25))
-LossTable <- NULL      
-LossTable <- do.call("rbind", lapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
-LossTable <- as.data.frame(LossTable)
-names(LossTable) <- c("binX", "binY", "Function", "Numerical Loss G", "Numerical Loss SO", "Visual Loss G", "Visual Loss SO", "Total Loss G", "Total Loss SO", "Frequency Loss")
-tab <- xtable(LossTable, digits=c(0, 0, 0, 0, 8, 8, 8, 8, 8, 8, 8))
-@
-
-% In order to quantify the advantages of each algorithm, we first compare the three sources of loss at different bin widths.
-The numerical and visual losses inherent in creating a binned scatterplot are a result of our desire to reduce and summarize the data until it is more manageable in size, and hence is a factor in our decision to bin the data in the first place.
-% %Figure~\ref{fig:binning-comparison} shows clearly that the standard algorithm can produce misleading graphical summaries. 
-%  The random algorithm has higher numeric and visual loss than the standard algorithm, but in many cases the binned scatterplots are visually indistinguishable. 
-%  
-%  The numerical loss is a function of the distribution and bin width, and once the bin width is chosen, it is entirely fixed.
-The random binning algorithm will inherently have a larger numerical loss than the standard algorithm because it allows for random assignment of points to numeric bin centers further from the original data than does the standard algorithm.
-%  
-%  As the visual loss is the difference between the numerical center and visual center of the bin, it is a function of the distribution and bin width (and the specific random assignment, in the case of the random algorithm).
-We expect a larger visual loss under the random binning algorithms than in the standard binning because the numerical bin centers are allowed to be drawn further away from the visual center under the random binning algorithm.
-% 
-% Frequency loss is arguably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary.
-The standard and random algorithms are very similar when we compare frequency loss.
-The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticeable than the increase seen in numerical loss.
-%~\ref{fig:LossComparison}
-% 
-% Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues.
-Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a bin width that is a non-integer multiple of the incremental widths.
-In this scenario, we see that the non-synchronous data increments and bin widths will cause bins to contain inconstant numbers of possible data value.
-Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency.
-Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artificial frequency discrepancy that manifests in a striped pattern in the visualization.
-The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.
-
-%A similar situation can arise when discrete data values are recorded at consistant increments and binwidths are chosen that are a odd integer multiple of the data incremental widths.
-This causes the situation where every other bin will have a data value located on the upper bound.
-Since the standard binning algorithm always bins data values on the boundary between bins to the lower bin center we will have an additional possible data value contributing to the frequency of every other bin, causing a striped pattern in our visualization.
-Again, the random binning algorithm mitigates this problem because it will allocate points on bin boundaries to adjacent bin centers with probability 0.5, leading to a more accurate display of frequency information.
 
@@
 
-For this data example we revisit the baseball data used earlier.
-To construct the binned scatterplot to display the relationship between strikeouts and game counts, we begin by specifying the bin dimensions and origin for standard rectangular binning.
-The data are counts, each variable is therefore recorded as an integer, corresponding to a resolution of one.
+We now revisit the baseball data used earlier and construct binned scatterplots to display the relationship between strikeouts and game counts.
+We begin by specifying the bin dimensions and origin for standard rectangular binning.
+The data are counts, each variable is therefore recorded as an integer, corresponding to a one game by one strike-out data resolution.
 We use integer bin widths in each dimension of spatial bins to avoid artificial density stripes.
-The data has only 42,583 observations, so we can bin finely without taking an inordinate amount of computation time.
-We use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high.
+The data has only 42,583 observations, so we can bin on a fine grid without taking an inordinate amount of computation time.
+We use approximately 50 bins in each dimension; the data range from 1 to 106 games and from 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high.
 The origin will be set to a half unit below the minimum value in each dimension, at $(0.5,-0.5)$.
@@
 The most striking feature in the frequency distribution is the dark spot in the bottom right of the plot representing a large number of pitchers that played very few games and had very few strikeouts.
-It is nearly impossible to differentiate the density structure across the majority of bins representing better pitchers who played many games and earned many strikeouts.
+It is nearly impossible to distinguish the density structure across the remaining bins, representing better pitchers who played many games and earned many strikeouts.
 In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts.
 Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities.
-Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins then visualizes density on the log scale.
-Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure.
-
-<<BaseballBinnedScatters2X10,echo=F,include=T,eval=T,fig.width=12, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Binned Scatterplots for Games vs Strikeouts">>=
+Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins.
+Each frequency binning approach adds a layer of complexity to interpreting the shade, however they both effectively emphasize the forked ridges in the density structure.
+
+<<BaseballBinnedScatters2X10,echo=F,include=T,eval=T,fig.width=12, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Binned scatterplots for games versus strikeouts.">>=
  # raw frequencies plot
@@
 To remedy this, the two variables can be standardized then binned equivalently by rescaling the bin dimension, where the net spatial loss is interpreted as a distance in units of standard deviations.
-The net spatial loss for the standardized data is \Sexpr{round(standBaseballLoss2X10,2)} standard deviations, an average of \Sexpr{round(standBaseballLoss2X10PerObs,4)} standard deviation per data point.
-\hh{XXX for which of the two binnings is this loss?} We also find that we had approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification than had we naively set the binning origin at the minimum data values, $(1, 0)$.
-
-% Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.
-This is apparent in the order of colored lines based on the Strike Out binning size.
-This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as bin width increases.
-Whereas the visual loss function sums loss over bin, thus as the bin width increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.
-% 
-% 
-% % \begin{figure}[hbtp]
-% % \centering
-% % <<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
-% % 
-% % # connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
-% % #                      password="R R0cks", port=3306, dbname="baseball", 
-% % #                      host="headnode.stat.iastate.edu")
-% % # pitch <- new("dataDB", co=connect, table="Pitching")
-% % # d1 <- dbData(pitch, vars=c( "G", "SO"))
-% % # bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
-% % # 
-% % # library(multicore)
-% % # lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
-% % # lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
-% % # 
-% % # lossesRdm <- cbind(bins, lossesRdm)
-% % # lossesStd <- cbind(bins, lossesStd)
-% % # 
-% % # losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
-% % # write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
-% % 
-% % losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
-% % lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
-% % lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
-% % lossmelt$y <- as.factor(lossmelt$y)
-% % levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
-% % qplot(x=x, y=value,
-% %       group=interaction(y, Algorithm),
-% %       data=lossmelt, geom="line",
-% %       colour=y, size=I(1),
-% %       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
-% %         theme_bw() + theme(legend.position="bottom") + 
-% %         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-% %         facet_grid(variable ~ Algorithm, scales = "free") 
-% % 
-% % # qplot(x=x, y=value,
-% % #       group=interaction(y, Algorithm),
-% % #       data=lossmelt, geom="line",
-% % #       colour=y, size=I(1), linetype=Algorithm,
-% % #       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
-% % #         theme_bw() + theme(legend.position="bottom") + 
-% % #         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
-% % #         facet_grid(. ~ variable, scales = "free") 
-% % 
-% % @
-% % \caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
-% % \end{figure}
-% 
-% 
-% %\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
-% %\noindent
-% %Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.
-% 
-% \hh{the  code following this comment in the Rnw file is, I think, broken beyond repair - but we should include a table like the one it produced in the paper.
-- we will have different losses to report, though}
-% 
-% % 
-% % <<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
-% % loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
-% % loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
-% % loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
-% % loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)
-% % 
-% % loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% % loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% % loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% % loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
-% % 
-% % table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
-% %                                c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
-% %                                c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
-% %                                c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
-% %                                c(binwidth=1, loss1[5:7]*100, loss1[8])))
-% % names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
-% % print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
-% % @
-% 
-% 
-% We may also use this example to highlight the advantages of random binning.
-The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident.
-The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips.
-These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the bin widths are integer multiples of the data increments.
-This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
-% 
-% 
-% \textbf{Are we sure that this is because of an odd bin widths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
-% 
-% \hh{This is a valid concern, but we have to fix the notation at the beginning first, before we can address it.}
-% 
-% \begin{figure}[hbtp]
-% \centering
-% \subfloat[(2, 2) tiling of G, SO data]{
-% \includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd22.pdf}
-% }\\
-% \subfloat[(5, 25) tiling of G, SO data]{
-% \includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
-% }
-% \caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chosen bin width is not a multiple of the data resolution.}
-% \end{figure}
-% 
-%
+The net spatial loss for the standardized data is the same for each plot in Figure~\ref{fig:BaseballBinnedScatters2X10} because they do not differ in location, only in frequency shading.
+The net spatial loss is \Sexpr{round(standBaseballLoss2X10,2)} standard deviations, an average of \Sexpr{round(standBaseballLoss2X10PerObs,4)} standard deviation per data point.
+We also find that we have approximately 3\% lower net spatial loss, on the standardized scale, for our binning specification using the recommended origin offset than if we were to naively set the binning origin at the minimum data values of one game and zero strikeouts.
+
+%--------------------------------------------------------------------
 
@@
 
-%We obtained airline scheduled departure times and actual departure times from the Department of Transportation for the year 2011.
-These times do not include actual departure date, however, so a flight scheduled to leave at 23:00 which is delayed by two hours is recorded as leaving at 01:00.
-The uncompressed data file is about 450 MB, and consists of two columns of almost six million records.
-
 Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
@@
 Even while using alpha blending this results in a severely over-plotted graph.
-On the right is a binned scatterplot of the reduced data from standard binning at 1-minute intervals.
+On the right is a minimally binned scatterplot of the reduced data from standard binning at 1-minute intervals.
 If the origin is offset by a half minute for both departures and arrivals, this binning does not have any spatial loss, as all observations will be centered within bins.
-The 1-minute standard binning also reduces the data to 176,384 individual records, less than 3\% of the original data.
-
-\begin{figure}[hbtp]\centering
-\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfill
+The 1-minute standard binning also reduces the data to 176,384 reduced data triple, less than 3\% of the original data.
+
+\begin{figure}[hbtp]
+  \subfloat[Alpha-Blended Scatterplot of Airline Data]{\includegraphics[keepaspectratio=TRUE,width=.47\textwidth]{./figure/AlphaScatterAirline2014.png}} 
 %   Swap Back at later date (shrinks size of pdf to managable)
@@
 \ktm{This plot is not compiling and needs to be reconstructed (my apologies)}]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
-\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}
-\caption{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011.
+\subfloat[Minimally Binned of Airline Data]{\includegraphics[width=.53\linewidth, keepaspectratio=TRUE]{./figure/MinimalBinnedAirline2014.png}}
+
+\caption[Alpha-blended and minimally binned scatterplots for scheduled and actual departure times of airline flights.]{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011.
 The plot on the left is based on a sample of the data, the plot on the right shows all flights.
@@
 
-Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the line of identity.
-Scheduled departure times past 6 am in the morning are much more common than earlier flights.
+Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the identity line.
+Scheduled departure times past 6:00 am in the morning are much more common than earlier flights.
 It is much more likely for a  flight to be delayed than to leave early, leading to the wash-out effect above the line, that is getting thinner with increasing delays.
-The range of delays on usual days starts at about one hour at 6 am and increases during the day to about 2 hours.
-The few number of flights before 6 am are also visible in both plots.
-The triangle of observations on the bottom right visible in both plots is nothing but an artifact of the data collection consisting of flights that are scheduled before midnight, but are delayed to departures past midnight.
+The range of delays on usual days starts at about one hour at 6:00 am and increases during the day to about a two hour delay.
+The small number of flights before 6:00 am are also visible in both plots.
+The triangle of observations on the bottom right in both plots is nothing but an artifact of the data collection consisting of flights that are scheduled before midnight, but are delayed to departures past midnight.
 The cloud of outliers halfway between the two main structures is potentially interesting, since no immediate explanation comes to mind, and would be worthy of a follow-up investigation.
 
-What is not apparent in the alpha blended scatterplot, is some fine-level structure that the plot based on all of the data shows.
-\ktm{Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not cause any artificial density stripes.} A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
+What is not apparent in the alpha-blended scatterplot, is some fine-level structure that the minimally binned scatterplot shows.
+Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not cause any \textit{artificial} density stripes.
+A close inspection of the plot on the right hand side reveals darker colored vertical lines at 30 minute intervals.
 It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.
 
-\begin{figure}[hbtp]\centering
-\includegraphics[width=.49\linewidth,keepaspectratio=true]{images/AirlineStdBinning5mins.pdf}
-\includegraphics[width=.49\linewidth,keepaspectratio=true]{images/AirlineStdBinning15mins.pdf}
-\caption{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
+\begin{figure}[hbtp]
+\includegraphics[width=.49\linewidth,keepaspectratio=true]{./figure/AirlineStdBinning5mins.pdf}
+\includegraphics[width=.49\linewidth,keepaspectratio=true]{./figure/AirlineStdBinning15mins.pdf}
+\caption[Binned scatterplots for scheduled and actual departure times of airline flights using 5X5 and 15X15 minute bins.]{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
 15-minute bins produce an even more coarse summary of the data.}\label{airline5mins-binning}
@@
 
-Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points.
+Large bivariate data sets are very difficult to visualize in raw form, due to over-plotting of points.
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
-Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.
+Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in spatial and frequency information.
 
 We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
-The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution.
-However, standard binning algorithm is superior due to lower spatial loss.
-For data with resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.
-We were also able to show through simulation that  a reasonable default for the binning origin is an offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well even for skewed data.
+The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution.
+However, the standard binning algorithm is superior due to lower spatial loss.
+For data with a coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.
+We were also able to show through simulation that a reasonable default for the binning uses an origin offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well even for spatially skewed data.
 
@@
 
-If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.
+If we elect to use frequency binning to shade tiles ordinally, it is recommended to use between four and seven distinct shades.
 This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
-Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.
+Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts.
 
 Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
-The findings of this research provide suggestions for reasonable default settings for binning parameters that maintain of spatial and frequency information and have desirable visual properties.
+The findings of this research provide suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.
 
@@
 \begin{appendix}
-\section{Appendix}
-
+\section{Appendix for Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}  
 %----------------------------------------------------------------------------
-\subsection{Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}
+%\subsection{}
 \label{proof:offset}
-\hh{you need to use the same conditions for the proof as you are stating when you refer to it.
-A uniform distribution was not mentioned before.
-Also, make sure that this section is 'stand-alone', i.e.
-explain every single one of the parameters that you are using.}
-
-\ktm{The following proof assumes three conditions for standard rectangular binning of a set univariate data hold: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension, (ii) points are symmetric distributed within standard rectangular bins, (iii) the bin width, $\omega_x$, is an integer multiples of $\alpha_x$ (i.e.
-$\omega_x = k\alpha_x$ for some $k \in \{1,2,\dots\}$).
-Under these conditions it will be shown that spatial loss is minimized by setting the binning origin, $\beta_x$, to $\alpha_x/2$ units below the minimum data value (i.e.
-$\beta_x = x_{(1)} - \alpha_x/2$).\\ }
-
-\noindent Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha_x$ for some constant $\alpha_x \in \mathbb{R}$.
-Thus $x_j = x_1 + (j-1)\alpha_x$.\\
-
-\noindent Suppose then that we bin the data using standard rectangular binning with origin, $\beta_x = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data.
-Thus $b(x_j) = \beta_x + \omega/2 = (x_1 - \theta) + (k\alpha_x/2)$\\
-
-\noindent Spatial Loss, $L^S = \sum_{i=1}^{k} ||x_i-b_(x_i)|| $ is definitionally minimized when $b_(x_i)$ is the \textit{geometric median}.
+
+The following is proof that when univariate data is uniformly distributed at resolution $\alpha$ and standard rectangular binning is used with binwidths that are a scalar multiple of the data resolution (i.e.
+$\omega = k\alpha$ for some $k \in \{1,2,\dots\})$, then spatial loss is minimized by setting the binning origin to $\alpha/2$ units below the minimum data value; set $\beta = x_{(1)} - \alpha/2$.
+
+Let $x_1, x_2, \dots, x_k \in \mathbb{R}$ represent the values in a single bin such that $x_{i+1} = x_i + \alpha$ for some constant $\alpha \in \mathbb{R}$. Thus $x_j = x_1 + (j-1)\alpha$.
+
+Suppose then that we bin the data using standard rectangular binning with origin, $\beta = x_1 - \theta$, and binwidth $\omega$; where $\theta$ is the \textit{origin offset} from the data.
+Thus $b(x_j) = \beta + \omega/2 = (x_1 - \theta) + (k\alpha/2)$
+
+Spatial Loss, $L^S = \sum_{i=1}^{k} ||x_i-b_(x_i)|| $ is definitionally minimized when $b_(x_i)$ is the \textit{geometric median}.
 The geometric median for $x_1, \dots, x_k = Q_x(.5) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2$ , where $Q_x(\cdot)$ is the empirical quantile function.
-\\
-
-\noindent Thus the optimal offset is the $\theta$ such that\\
+
+Thus the optimal offset is the $\theta$ such that
 
 $ b(x_i) = Q_x(.5) $ \\
-$ \Rightarrow (x_1 - \theta) + (k\alpha_x/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
-$ \Rightarrow 2x_1 - 2\theta + k\alpha_x = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha_x ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha_x ) $ \\
-$ \Rightarrow -2\theta + k\alpha_x = (\lceil\frac{k+1}{2}\rceil - 1)\alpha_x  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha_x  $ \\
-$ \Rightarrow -2\theta + k\alpha_x = ((k+1) - 2)\alpha_x  $ \\
-$ \Rightarrow -2\theta = -\alpha_x $ \\
-$ \Rightarrow \theta = \alpha_x/2 $ \\
-
-Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha_x/2$.
+$ \Rightarrow (x_1 - \theta) + (k\alpha/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
+$ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
+$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha  $ \\
+$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha  $ \\
+$ \Rightarrow -2\theta = -\alpha $ \\
+$ \Rightarrow \theta = \alpha/2 $ \\
+
+Thus the optimal offset for reducing spatial loss in this scenario is $\theta = \alpha/2$.
 This result holds for data that is symmetrically distributed within the bin since the median will not change.
-It extends to multiple contiguous bins with resolution $\alpha_x$ data that has symmetrically distributed data withing each bin.
-\\
+It extends to multiple contiguous bins with resolution $\alpha$ data that has symmetrically distributed data withing each bin.
 
 If the same conditions are extended to the two dimensional case, then the origin for minimal spatial loss is at $(x_{(1)}-\alpha_x/2, y_{(1)}-\alpha_y/2)$ where $\alpha_x$ and $\alpha_y$  are the data resolution for each dimension, respectively.
-
-%----------------------------------------------------------------------------
-% 
-% \subsection{Partition of Spatial Loss}\label{proof:partition}
-% We want to show that spatial loss can be written as a partition of the form 
-% \[L_s(x) = 1/S^2_{\emptyset}\sum_i S^2_i = 1/S^2_{\emptyset}\left[\sum_i L^N_i + \sum_i L^V_i\right]\] 
-% 
-% For that let us consider the values $\{x_{i}: b(x_i) = x^\ast\}$ in a single bin with center $x^\ast$.
-% The numerical center of these points is given as  $1/n \sum x_i = \overline x$.
-% Then the total loss for each element $x_i$ in this bin is:
-% 
-% \begin{align*}
-% S_i^2 = \sum (x_i - x_i^\ast)^2 & = \sum (x_i - \overline x + \overline x - x_i^\ast)^2\\
-% & = \sum(x_i-\overline x)^2 + 2\sum (x_i-\overline x)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
-% & = \sum(x_i - \overline x)^2 + 2\sum(0)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
-% & = \sum(x_i - \overline x)^2 + \sum(\overline x-x_i^\ast)^2\\
-% & = L^N_i + L^V_i
-% \end{align*}
-% Adding over all elements in the bin and all bins gives the formula above.
-% 
-% \subsection{Constancy of Loss under Binary Splits}\label{proof:constantloss}
-% We want to show that under random binning the loss due to doubling of bin width is deterministic.
-% 
-% Let us consider the contribution $L_i$ of a single point $x_i$ to the loss when starting with the minimal bin width, i.e. each unique point is in a separate bin:
-% doubling the bin width leads to two  scenarios: 
-% a point is either at  the bin center or directly between two bins.
-% In the case that the point is at the bin center, its contribution to the total overall loss is zero,  $L_i = 0$, and the associated probability $p_i = 1$. 
-% 
-% Alternately, when the point is exactly half-way between two bin centers, the point is attributed with probability $p_i = \frac{1}{2}$ to either bin, leading to a loss of $L_i = \frac{1}{2} c_i$ regardless of which bin is assigned.
-Hence, in this case, loss is also entirely independent of which bin is assigned.
+%
+
 \end{appendix}
+
 \bibliographystyle{asa}
 \bibliography{references}
+
 \end{document}

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Start to update Flight example section to discuss new 2014 flights data

@@
 These are published online by the Bureau of Transportation Services at \url{http://www.transtats.bts.gov/DataIndex.asp}.
-Every day there are about 16,000 flights across the United States adding up to almost 6 Million flights a year.
-Scheduled and actual departure times for all flights in 2011 make up --in uncompressed form-- a file of about 450 MB.
+Every day there are about 16,000 flights across the United States which added up to over 5.8 million flights in 2014.
+The scheduled and actual departure times for flights in 2014 make up --in uncompressed comma separated form-- a file of about 53 MB.
 A comparison of scheduled and actual departure times allows us an investigation of on-time performance of air carriers.
@@
 If the origin is offset by a half minute for both departures and arrivals, this binning does not have any spatial loss, as all observations will be centered within bins.
-The 1-minute standard binning also reduces the data to 176,384 reduced data triple, less than 3\% of the original data.
+The 1-minute standard binning also reduces the data to around 190,000 reduced data triple, approximately 5\% of the original data size.
 
@@
 
-\caption[Alpha-blended and minimally binned scatterplots for scheduled and actual departure times of airline flights.]{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011.
-The plot on the left is based on a sample of the data, the plot on the right shows all flights.
+\caption[Alpha-blended and minimally binned scatterplots for scheduled and actual departure times of airline flights.]{\label{airline-scatter}Scheduled and actual departure times of flights across the United States in 2014.
 The large scale distributional patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Replace Database connect with more stable pitching.Rdata read-in

@@
 # dbDisconnect(connect)
-# save(d1, d1.ext, file="pitching.Rdata")
+# pitchdat <- d1.ext
+# save(d1, d1.ext,pitchdat, file="pitching.Rdata")
 load("pitching.Rdata")
@@
 <<BaseballBinningExamplePrep, echo=FALSE, eval=TRUE, include=FALSE, cache=TRUE>>=
-connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
-                     port=3306, user="dbaccess", dbname="baseball")
-pitchdat <- dbGetQuery(connect,  sprintf("select G, SO from Pitching"))
-dbDisconnect(connect)
+# connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
+#                      port=3306, user="dbaccess", dbname="baseball")
+# pitchdat <- dbGetQuery(connect,  sprintf("select G, SO from Pitching"))
+# dbDisconnect(connect)
 

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Edit Table
Continued work on Flights section rework

@@
 }
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 12 rows]{
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data]{
 \begin{tabular}{rr}\hline
@@
 \end{minipage}\hfil
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Centers, 12 rows]{
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Centers]{
 \begin{tabularx}{.7\textwidth}{rr} \hline
@@
 \end{minipage}\hfil
-\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Reduced Binned Data, 4 rows]{
+\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Reduced Binned Data]{
 \begin{tabularx}{.7\textwidth}{rrr}\hline
- \hspace{.2cm} $x^\ast$   &   \hspace{.2cm} $y^\ast$ & \hspace{.4cm}c \\ \hline
+ \hspace{.2cm} $x^\ast$   &  $y^\ast$ & \hspace{.4cm}c \\ \hline
 -5 & -5 & 5 \\ 
@@
 Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
-The plot on the left shows the scatterplot from a sample of one million of those records.
-Even while using alpha blending this results in a severely over-plotted graph.
+The plot on the left shows the traditional scatterplot; even while using alpha-blending this results in a severely over-plotted graph.
 On the right is a minimally binned scatterplot of the reduced data from standard binning at 1-minute intervals.
-If the origin is offset by a half minute for both departures and arrivals, this binning does not have any spatial loss, as all observations will be centered within bins.
+If the origin is offset by a half minute for both departures and arrivals, this binning does not have any spatial loss, because all observations are be centered within bins.
 The 1-minute standard binning also reduces the data to around 190,000 reduced data triple, approximately 5\% of the original data size.
+
+<<FlightTradScatterplot, echo=F,include=F,eval=F, cache=TRUE>>=
+source("RectangularBinningFunctions.R")
+load(file="DepartDelaysMinutes.RData")
+# Code takes a long long time to compile so ran once, saved to high res png, then set eval=FALSE
+ggplot(aes(x=Scheduled/60, y=Actual/60), data=DepartDelaysMinutes) +
+  geom_point(alpha=.05) + 
+  theme_bw() + theme(aspect.ratio=1,
+                     panel.grid= element_blank(),
+                     legend.text=element_text(size=12) )  +
+  scale_x_continuous("Scheduled Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+  scale_y_continuous("Actual Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+  ggtitle("Alpha-Blended Scatterplot of Airline Departure Times") 
+ggsave(filename = "./figure/AlphaScatterAirline2014.png", width= 6, height=6, dpi=900)
+@
+
+<<FlightMinBinScat, echo=F,include=F, cache=TRUE>>=
+source("RectangularBinningFunctions.R")
+load(file="DepartDelaysMinutes.RData")
+# Create minimal binning
+departure1minute <- RectBin2d(xs=DepartDelaysMinutes$Scheduled,ys=DepartDelaysMinutes$Actual, 
+                              originx = -0.5, originy = -0.5, 
+                              widthx = 1, widthy = 1, type="standard")
+# down to 3.261% of original number of rows
+# nrow(departure1minute)/nrow(DepartDelaysMinutes)
+ggplot(aes(x=binxs/60, y=binys/60), data=departure1minute) +
+  geom_tile( aes(fill=binfreq) ) + 
+  scale_x_continuous("Scheduled Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+  scale_y_continuous("Actual Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+  ggtitle("Minimally Binned Scatterplot of Airline Departure Times")+ 
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) +
+  theme_bw() + theme(aspect.ratio=1,
+                   panel.grid= element_blank(),
+                   legend.text=element_text(size=12) ) + 
+  geom_abline(intercept=0,slope=1, color="gray70", linetype=3)
+ggsave(filename = "./figure/MinimalBinnedAirline2014.png", width= 6.5, height=6, dpi=900)
+@
 

### 2015-12-15 versionForJCGS/jcgs-template.tex
Making adjustments to jcgs style. pretty easy switch... Much bigger concern is the 25 page max with 39 pages currently.

%&latex
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Title}
  \author{Author 1\thanks{
    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
    Department of YYY, University of XXX\\
    and \\
    Author 2 \\
    Department of ZZZ, University of WWW}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract.  200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  3 to 6 keywords, that do not appear in the title
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Body of paper.  Margins in this document are roughly 0.75 inches all
around, letter size paper.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig1.pdf}
\end{center}
\caption{Consistency comparison in fitting surrogate model in the tidal
power example. \label{fig:first}}
\end{figure}

\begin{table}
\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
\begin{center}
\begin{tabular}{rrrrr}
one & two & three & four & five\\\hline
1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
\end{tabular}
\end{center}
\end{table}

\begin{itemize}
\item Note that figures and tables (such as Figure~\ref{fig:first} and
Table~\ref{tab:tabone}) should appear in the paper, not at the end or
in separate files.
\item In the latex source, near the top of the file the command
\verb+\newcommand{\blind}{1}+ can be used to hide the authors and
acknowledgements, producing the required blinded version.
\item Remember that in the blind version, you should not identify authors
indirectly in the text.  That is, don't say ``In Smith et. al.  (2009) we
showed that ...''.  Instead, say ``Smith et. al. (2009) showed that ...''.
\item These points are only intended to remind you of some requirements.
Please refer to the instructions for authors
at \url{http://amstat.tandfonline.com/action/authorSubmission?journalCode=ubes20&page=instructions#}
\item For more about ASA\ style, please see \url{http://journals.taylorandfrancis.com/amstat/asa-style-guide/}
\item If you have supplementary material (e.g., software, data, technical
proofs), identify them in the section below.  In early stages of the
submission process, you may be unsure what to include as supplementary
material.  Don't worry---this is something that can be worked out at later stages.
\end{itemize}

\section{Methods}
\label{sec:meth}
Don't take any of these section titles seriously.  They're just for
illustration.

\section{Verifications}
\label{sec:verify}
This section will be just long enough to illustrate what a full page of
text looks like, for margins and spacing.

\cite{Campbell02}, \cite{Schubert13}, \cite{Chi81}

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
{\bf With this spacing we have 30 lines per page.}
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\section{Conclusion}
\label{sec:conc}

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Brief description. (file type)

\item[R-package for  MYNEW routine:] R-package ÒMYNEWÓ containing code to perform the diagnostic methods described in the article. The package also contains all datasets used as examples in the article. (GNU zipped tar file)

\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)

\end{description}

\section{BibTeX}

We hope you've chosen to use BibTeX!\ If you have, please feel free to use the package natbib with any bibliography style you're comfortable with. The .bst file Chicago was used here, and agsm.bst has been included here for your convenience.

\bibliographystyle{Chicago}

\bibliography{Bibliography-MM-MC}
\end{document}

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Making adjustments to jcgs style. pretty easy switch... Much bigger concern is the 25 page max with 39 pages currently.
small edits
Swap out slow loading pdf files for high resolution pngs

@@
-\documentclass[letterpaper,10pt]{article}
+%&latex
+\documentclass[12pt]{article}
+\usepackage{amsmath,amsthm,amssymb}
+\usepackage{graphicx,psfrag,epsf,float,wrapfig,subfig,tabularx,ulem}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
+
  \usepackage[utf8]{inputenc}
-\usepackage{amsmath,amsthm,amssymb}
  \usepackage{fullpage}
-\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
  %\graphicspath{{figure/}}
@@
  \usepackage{color}
-\usepackage{natbib}
  \usepackage{hyperref}
@@
 
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
 \newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
@@
 
+%-----------------------------------------------------------------------------------------------------
+%%%% Document
+%-----------------------------------------------------------------------------------------------------
+\begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%-----------------------------------------------------------------------------------------------------
+%%% Title
+%-----------------------------------------------------------------------------------------------------
+\if0\blind
+{
 %opening
@@
 \author{Karsten Maurer$^1$, Susan VanderPlas$^2$, Heike Hofmann$^{2,3}$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
-
-\begin{document}
   \maketitle
-
+} \fi
+
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Title}
+\end{center}
+  \medskip
+} \fi
+
+%-----------------------------------------------------------------------------------------------------
+%%% Abstract
+%-----------------------------------------------------------------------------------------------------
+\bigskip
 \begin{abstract}
@@
 
+\noindent%
+{\it Keywords:}  scatterplots, visual loss, aggregation, graphics
+\vfill
+
+%-----------------------------------------------------------------------------------------------------
+% Intro
+%-----------------------------------------------------------------------------------------------------
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
 \section{Introduction}
@@
 qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
-  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
+  ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=900)
 qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
-  ggsave("./figure/Overplotting.pdf", width=5, height=5)
+  ggsave("./figure/Overplotting.png", width=5, height=5, dpi=900)
 qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
-  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
+  ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=900)
 qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-  ggsave("./figure/HexBinning.pdf", width=5, height=5)
+  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=900)
 d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
@@
 
-  ggsave("./figure/BinScatter.pdf", width=5, height=5)
+  ggsave("./figure/BinScatter.png", width=5, height=5, dpi=900)
 
@@
 %
+\spacingset{1}
 \begin{table}[hbtp]
@@
 \end{table}
+\spacingset{1.45} % DON'T change the spacing!
 %
@@
 
-<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
+<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
 Density changes are impossible to make out in the solid black areas of the plots.
@@
   scale_y_continuous(breaks=c(0,25,50,75,100))
-@
+
+ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=900)
+@
+
+\begin{figure}[hbtp]
+\begin{center}
+ \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
+	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data.
+Density changes are impossible to make out in the solid black areas of the plots.
+97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.}
+	\end{center}
+\end{figure}
 

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Swap in pngs that I wrote (but forgot to include) earlier

@@
 \begin{figure}[hbtp]
-  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.pdf}}
-  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.pdf}} 
-	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.pdf}}
-
-  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.pdf}}
-  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.pdf}}
+  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.png}}
+  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.png}} 
+	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.png}}
+
+  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.png}}
+  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.png}}
 
@@
 \begin{center}
- \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
+ \includegraphics[keepaspectratio=TRUE,width=.99\textwidth]{./figure/SimScatters.png}
 	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data.

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Reorganize data structure in folder

@@
 # pitchdat <- d1.ext
-# save(d1, d1.ext,pitchdat, file="pitching.Rdata")
-load("pitching.Rdata")
+# save(d1, d1.ext,pitchdat, file="./data/pitching.Rdata")
+load("./data/pitching.Rdata")
 @
@@
 # write.csv(spatlossdatRound,"LossesByOriginBinwidth.csv", row.names=FALSE)
-lossOrigin <- read.csv("LossesByOriginBinwidth.csv",header=TRUE)
+lossOrigin <- read.csv("./data/LossesByOriginBinwidth.csv",header=TRUE)
 mins <- ddply(lossOrigin, .(widthx, bintype, distrib), summarise,
@@
 source("RectangularBinningFunctions.R")
-load(file="DepartDelaysMinutes.RData")
+load(file="./data/DepartDelaysMinutes.RData")
 # Code takes a long long time to compile so ran once, saved to high res png, then set eval=FALSE
@@
 source("RectangularBinningFunctions.R")
-load(file="DepartDelaysMinutes.RData")
+load(file="./data/DepartDelaysMinutes.RData")
 # Create minimal binning

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Update Baseball data to 2014

@@
 
-# connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
-#                      port=3306, user="dbaccess", dbname="baseball")
-# pitch <- new("dataDB", co=connect, table="Pitching")
-# 
-# d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs
-# 
-# xtd.data <- function(x) {
-# # replicate each line as often as given by the last entry
-#   k <- dim(x)[2]
-# 
-#   res <- data.frame(sapply(1:(k-1), function(var) {
-#   		return(rep(x[,var], x[,k]))
-# 		}
-# 	))
-# 	names(res) <- names(x)[1:(k-1)]
-# 	return(res)
-# }
-# d1.ext <- xtd.data(d1)
-# dbDisconnect(connect)
-# pitchdat <- d1.ext
-# save(d1, d1.ext,pitchdat, file="./data/pitching.Rdata")
+# # Downloaded from http://seanlahman.com/baseball-archive/statistics 2014 update
+# pitchdat <- read.csv("./data/Pitching2014.csv")[,c("G","SO")]
+# source("RectangularBinningFunctions.R")
+# d1 <- RectBin2d(xs=pitch2014$G,ys=pitch2014$SO, 
+#                      originx = .5, originy = -.5, 
+#                      widthx = 1, widthy = 1, type="standard")
+# names(d1) <- c( "G","SO","Freq")
+# save(d1, pitchdat, file="./data/pitching.Rdata")
+
 load("./data/pitching.Rdata")
@@
 <<scatterplots, echo=F,include=F>>=
-qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
+qplot(data=pitchdat, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=900)
@@
   ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=900)
-qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
+qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
   ggsave("./figure/HexBinning.png", width=5, height=5, dpi=900)
-d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
+d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
 d1.hex <- data.frame(hcell2xy(d1.hex.obj))
@@
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
-The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
+The scatterplot shows 144 seasons (from the years of 1871 -- 2014) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
 The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season.
-While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
+While the data set is only medium sized with \Sexpr{dim(pitchdat)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
 

### 2015-12-15 versionForJCGS/data/RectangularBinningFunctions.R

@@ -1,0 +1,90 @@
+### Preliminaries 
+library(dplyr)
+
+#----------------------------------------------------
+### Binning Functions
+#----------------------------------------------------
+
+## Standard 1d Rectangular Binning
+StandRectBin1d <- function(xs, origin, width){
+  binxs <- origin + width*( ceiling((xs-origin)/width) - 0.5 )
+  binxs[xs == origin] <- origin + width/2
+  return(binxs)
+}
+
+
+## Random 1d Rectangular Binning
+# Establish bounding bin centers for each observation
+# Then use Unif(0,1) draws compared to assignment probs to allocate
+# Reassignment for values below first center (bx1) or
+#     above highest center (bxJ)
+RandRectBin1d <- function(xs, origin, width){
+  bx1 <- origin + width/2
+  bxJ <- origin + width*(floor((max(xs)-bx1)/width) + .5)
+  lbs <- bx1 + width*floor((xs-bx1)/width)
+  ubs <- bx1 + width*ceiling((xs-bx1)/width)
+  # initially assign all values to upper bound
+  binxs <- ubs
+  # then use runif to reassign based on distance from 
+  plower <- (ubs - xs)/width
+  lowerindex <- (plower > runif(length(xs), 0, 1))
+  binxs[lowerindex] <- lbs[lowerindex]
+  binxs[xs < bx1] <- bx1
+  binxs[xs > bxJ] <- bxJ
+  return(binxs)
+}
+
+
+### Create binning function that does not calculate loss
+# to be used to track computation time for binning
+
+## 2d Rectangular Binning (for either Standard or Random)
+# standard is straight forward extension of 1d binning
+# random binning needs post processing to calculate minimum spatial loss
+RectBin2d <- function(xs,ys, originx, originy, widthx, widthy, type="standard"){
+  if(type=="standard"){
+    tempdat <- data.frame(xs = xs, ys=ys,
+                          binxs = StandRectBin1d(xs,originx,widthx),
+                          binys = StandRectBin1d(ys,originy,widthy))
+  }
+  if(type=="random"){
+    tempdat<- data.frame(xs = xs, ys=ys,
+                         binxs =  RandRectBin1d(xs,originx,widthx),
+                         binys =  RandRectBin1d(ys,originy,widthy))
+  }
+  outdat <- tempdat %>% 
+    group_by(binxs,binys) %>% 
+    summarise(binfreq = length(xs))
+  return(outdat[,1:3])
+}
+
+
+## Frequency Binning  
[…]

### 2015-12-15 versionForJCGS/data/UpdateSeasonsPitching.R

@@ -1,0 +1,17 @@
+### Add 2010 to 2013 to baseball data
+setwd("C:\\Users\\maurerkt\\Documents\\GitHub\\dbaccess\\versionForJCGS\\data")
+
+source("RectangularBinningFunctions.R")
+
+load(file="pitching.Rdata")
+
+pitch2013 <- read.csv("Pitching2013.csv")[,c("G","SO")]
+head(pitch2013)
+
+binout1 <- RectBin2d(xs=pitch2013$G,ys=pitch2013$SO, 
+                     originx = .5, originy = -.5, 
+                     widthx = 1, widthy = 1, type="standard")
+names(binout1) <- names(d1)
+head(binout1)
+
+# It turns out we already have the 2013 updated data version, we never updated the year in the text of the paper

### 2015-12-16 versionForJCGS/PaperPriorToShorteningForJCGS.Rnw
tightening up section 2
- cut repeated discussion of fig 1 parts a,b,c
- simplify discussion of aggregation based adaptations

--- 2015-12-15 versionForJCGS/PaperJCGS.Rnw
@@
 \begin{figure}[hbtp]
-  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.png}}
+  \subfloat[Traditional scatterplot ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.png}}
   \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.png}} 
@@
 	\caption[Traditional and adapted scatterplots for games vs.
-strikeouts data.]{\label{fig:scatter-alpha} Scatterplots of games versus strikeouts in major league baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05).
-Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively.}
+strikeouts data.]{\label{fig:scatter-alpha}Traditional and adapted scatterplots for games vs.
+strikeouts data.
+Note improvement in conveying visual density information in the aggregation based plots on the bottom row.}
 \end{figure}

### 2015-12-16 versionForJCGS/PaperJCGS.Rnw
shortening Section 3 intro

@@
 
-Figure~\ref{fig:scatter-alpha}(a) shows a traditional scatterplot with each observation is drawn with a filled circle.
-A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable.
+The top row of Figure~\ref{fig:scatter-alpha} demonstrates several varients of the traditional scatterplot where each observation is plotted using distinct points.
+The traditional, solid point, scatterplot in Figure~\ref{fig:scatter-alpha}(a) shows a triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable.
 \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}(b)) to mitigate the problem of over-plotting.
-Open circles make points that are close together more visually distinct; thus allowing for the perception of more density information than with filled points.
-A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}(c)).
-Alpha blending renders points as semi-transparent to provides more visibility of underlying points.
-
-\begin{figure}[hbtp]
-  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.png}}
-  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.png}} 
-	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.png}}
-
-  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.png}}
-  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.png}}
-
-	\caption[Traditional and adapted scatterplots for games vs.
-strikeouts data.]{\label{fig:scatter-alpha} Scatterplots of games versus strikeouts in major league baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05).
-Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively.}
-\end{figure}
-
-All of these methods fall short in the example.
+A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}(c)) which renders points as semi-transparent to provides more visibility of underlying points.
+%All of these methods fall short in the example.
 As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information.
@@
 The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.
+
+\begin{figure}[hbtp]
+  \subfloat[Traditional scatterplot ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.png}}
+  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.png}} 
+	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.png}}
+
+  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.png}}
+  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.png}}
+
+	\caption[Traditional and adapted scatterplots for games vs.
+strikeouts data.]{\label{fig:scatter-alpha} Traditional and adapted scatterplots for games vs.
+strikeouts data.
+Note improved density information in the aggregation-based plots.}
+\end{figure}
 
@@
 
-Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and binned scatterplots \citep{martin-gold}.
-Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin.
-Sunflower plots are particularly useful when the number of points in each bin remains reasonably small.
+%Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and binned scatterplots \citep{martin-gold}.
 Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion.
@@
 A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size or 3D renderings to provide joint density information within each grid cell, known as a tile.
+A {\it sunflower plot} uses symbols that increase in complexity proportional to the number of points in each bin.
+Due to difficulty in rendering descernable shapes in limited space, sunflower plots are only useful when the number of points in each bin remains reasonably small.
 A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.
@@
 
-Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (\ref{fig:scatter-alpha}(d)) and a bubble plot with frequency encoded as point size (\ref{fig:scatter-alpha}(e)).
+ Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) and a bubble plot with frequency encoded as point size (e).
 The hexagonally binned scatterplot and the bubble plot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha}(a-c).
@@
 The joint density consists two distinct ridges following two lines with very different slopes.
-The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game.
-The other line has a slope of about four times that rate.
-This high rate is also associated with fewer games played.
-Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and bats.
+A low ridge with high games and lower strikeout rates, and a high ridge with fewer games played but high strikeout rates.
+Closer investigation of additional variables reveals that this split in the density ridges corresponds mainly to pitchers in the pre-modern and modern era's of baseball; players in past had much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and pitchers are disadvantaged in the modern era due to substantial qualitative improvements in bats.
 
 For extremely large data sets, binned scatterplots are a more useful visualization of two-dimensional density information than the scatterplot, and are less computationally demanding, as not every single point in the data set has to be rendered separately.
-
-As with histograms, the width of bins (or the number of bins) is an important factor in the detail of the binned data and the resulting plot: if the bin width is too small in comparison to the amount of data available, there is little advantage to binning, but if the bin width is too large, interesting features of the joint distribution may be obscured by over-plotting.
+In order to explore the properties of binned scatterplots we must specify binning algorithms by which to aggregate.
+
+%As with histograms, the width of bins (or the number of bins) is an important factor in the detail of the binned data and the resulting plot: if the bin width is too small in comparison to the amount of data available, there is little advantage to binning, but if the bin width is too large, interesting features of the joint distribution may be obscured by over-plotting.
 
@@
 
-Note that this standard rectangular binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
+%Note that this standard rectangular binning procedure utilizes intervals that are open to the left and closes the outer bound of the leftmost bin.
 These specifications are consistent with the binning procedure used in the \texttt{hist()} function for creating histograms in base \texttt{R} \citep{R}.
@@
 
-\textit{Quantile binning} is another option that divides the range of the observations into bins each containing an equal number of points.
-The $j^{th}$ bin interval takes the form $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
-Note that this binning approach is \textit{not} desirable for spatially visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for binned scatterplots that employ a second stage of binning to create discrete shade scheme for displaying grouped bin frequencies, which will be discussed in Section~\ref{FreqLoss}.
-
-The bin boundaries and centers for each type of rectangular binning algorithm discussed above can be found in Table~\ref{tab:rectbinning}.
+An alternative to the rectangular binning processes discussed above, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
+In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
+The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
+More formally,
+%
+\begin{eqnarray}\label{randbin1}
+b^r_X(x_i) = \left\{\begin{array}{ll} 
+  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
+  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
+  \end{array}\right.
+\end{eqnarray}  
+%
+for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.
+In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.
+This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.
+The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.
+
+%The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the bin center.
+``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning in which adjacent bins are assigned a weight depending on the distance from the point to that bin, where all weights sum to one.
+With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
+Specifically, the expectation of the random binning algorithm is the same as for linear binning.
 
@@
 %-------------
-Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
+Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
 %-------------
-Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
+%Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
 \hline
@@
 %
-An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
-In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
-The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center.
-More formally,
-%
-\begin{eqnarray}\label{randbin1}
-b^r_X(x_i) = \left\{\begin{array}{ll} 
-  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
-  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
-  \end{array}\right.
-\end{eqnarray}  
-%
-for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.
-In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.
-This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.
-
-The deterministic standard binning algorithm is an example of a ``direct" binning algorithm, in which all points are assigned with weight one to the bin center.
-``Linear" binning  \citep{martin-gold} is a \textit{computationally intensive} alternative to direct binning in which adjacent bins are assigned a weight depending on the distance from the point to that bin, where all weights sum to one.
-With large data sets, the calculations required for linear binning become unwieldy, but the random binning algorithm can be considered an approximation to linear binning.
-Specifically, the expectation of the random binning algorithm is the same as for linear binning.
+
+\textit{Quantile binning} is another option that divides the range of the observations into bins each containing an equal number of points.
+The $j^{th}$ bin interval takes the form $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
+Note that this binning approach is \textit{not} desirable for spatially visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for binned scatterplots that employ a second stage of binning to create discrete shade scheme for displaying grouped bin frequencies, which will be discussed in Section~\ref{FreqLoss}.
 

### 2015-12-16 versionForJCGS/PaperJCGS.Rnw
Start to move rectangular and minimal binning to baseball example in section 2
next step: phase out hex binned scatter (figure 1d) in favor of rect. binning

@@
 <<scatterplots, echo=F,include=F>>=
-qplot(data=pitchdat, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
+# Traditional scatter
+qplot(data=pitchdat, x=G, y=SO, geom="point", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=900)
-qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
+
+# Open Circle Scatter
+qplot(data=d1, x=G, y=SO, geom="point", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/Overplotting.png", width=5, height=5, dpi=900)
-qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
+
+# Alpha Blended Scatter
+  qplot(data=d1, x=G, y=SO, geom="point", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
   ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=900)
+
+# Hex binned scatterplot
   qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
@@
 
+# hex binned bubbleplot
 qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
   scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-
-qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
-  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-
+ggsave("./figure/BinBubble.png", width=5, height=5, dpi=900)
+
+# minimally binned rectangular  
+qplot(G, SO, geom="tile", fill=Freq, data=d1) +
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
   ggsave("./figure/BinScatter.png", width=5, height=5, dpi=900)
@@
 
-\begin{figure}[hbtp]
-  \subfloat[Traditional scatterplot ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.png}}
-  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.png}} 
-	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.png}}
-
-  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.png}}
-  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.png}}
+\begin{figure}[h]
+  \subfloat[Traditional scatterplot ]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/Overplotting.png}}
+  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/OverplottingCircles.png}} 
+	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/OverplottingAlpha.png}}
+
+  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/HexBinning.png}}
+  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/BinBubble.png}}
+  \subfloat[Minimally binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/BinScatter.png}}
 
@@
 The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870}, where the white spaces between circles on the evenly spaced grid appear shaded due to an optical illusion; whereas this will not occur for a binned scatterplot on a tessellated grid.
+
+A traditional scatterplot is comparable to a \textit{minimally binned scatterplot} -- using small enough bins such that only a single unique coordinate pair exists within each bin --  however, the "bins" of a traditional scatterplot are shaded in a binary manner with no indication of overlapping observations.
+Alpha blending as used in Figure~\ref{fig:scatter-alpha}(c) extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
+The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is usually reached well before the maximum frequency, truncating the perceivable frequency information.
+By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which may represent more points.
 
@@
 
-A traditional scatterplot is comparable to a \textit{minimally binned scatterplot} -- using small enough bins such that only a single unique coordinate pair exists within each bin --  however, the "bins" of a traditional scatterplot are shaded in a binary manner with no indication of overlapping observations.
-Alpha blending as used in Figure~\ref{fig:scatter-alpha}(c) extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
-The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is usually reached well before the maximum frequency, truncating the perceivable frequency information.
-By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which may represent more points.
-This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}(d) and (e) under the same framework as plots (a)-(c).
-
 By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures.

### 2015-12-17 versionForJCGS/PaperJCGS.Rnw
Change hex binned tiles to minimally binned rectangular tiles to introduce minimal binning early in simple context to do double duty

@@
 
-\begin{figure}[h]
+\begin{figure}[hbtp]
   \subfloat[Traditional scatterplot ]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/Overplotting.png}}
@@
 
-  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/HexBinning.png}}
-  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/BinBubble.png}}
-  \subfloat[Minimally binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/BinScatter.png}}
+  \subfloat[Rectangular minimally binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.png}}  
+%  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/HexBinning.png}}
+  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinBubble.png}}
 
@@
 
- Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (d) and a bubble plot with frequency encoded as point size (e).
-The hexagonally binned scatterplot and the bubble plot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha}(a-c).
-The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870}, where the white spaces between circles on the evenly spaced grid appear shaded due to an optical illusion; whereas this will not occur for a binned scatterplot on a tessellated grid.
-
-A traditional scatterplot is comparable to a \textit{minimally binned scatterplot} -- using small enough bins such that only a single unique coordinate pair exists within each bin --  however, the "bins" of a traditional scatterplot are shaded in a binary manner with no indication of overlapping observations.
-Alpha blending as used in Figure~\ref{fig:scatter-alpha}(c) extends the binary shading of a standard scatterplot to an implicit shading according to frequency.
-The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is usually reached well before the maximum frequency, truncating the perceivable frequency information.
+ Figure~\ref{fig:scatter-alpha} contains examples of a rectangular binned scatterplot with frequency encoded as color (d) and a hexagonal binned bubble plot with frequency encoded as point size (e); both of which are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha}(a-c).
+Bubble plots is prone to suffer from the Hermann-grid illusion \citep{hermann:1870}, where the white spaces between circles on the evenly spaced grid appear shaded due to an optical illusion.
+The rectangular bins in Figure~\ref{fig:scatter-alpha}(d) are one game by one strikeout in size which matches the resolution of the recorded data; this is referred to as a \textit{minimally binned scatterplot}.
+With only a single unique coordinate pair exists within each bin the tiles are comparable to points in a traditional scatterplot; however, the unique coordinate pairs in a traditional scatterplot are shaded in a binary manner with no indication of overlapping observations.
+Alpha blending as used in Figure~\ref{fig:scatter-alpha}(c) is akin to bin shading; however the frequency mapping is imperfect because as soon as overlapping points surpass full opacity the perceivable frequency information is truncated.
 By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which may represent more points.

### 2015-12-17 versionForJCGS/PaperJCGS.Rnw
minor tightening of language and moving table captions under tables (no more awkward ISU thesis style placements)

@@
 For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.
-
 \textit{Rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
@@
 
-\textit{Standard rectangular binning} is a special cases of general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and a binwidth, $\omega_X$, need to be specified.
-Standard rectangular binning is necessarily used in the construction of all histograms;  the consistent binwidth makes the display of frequency proportional to density.
+\textit{Standard rectangular binning} is a special cases of general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and binwidth, $\omega_X$, need to be specified.
+Standard rectangular binning is necessarily used in the construction of histograms;  the consistent binwidth makes the display of frequency proportional to density.
 Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.
@@
 
-An alternative to the rectangular binning processes discussed above, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
+An alternative to the rectangular binning process, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.
@@
 \centering
-\caption{Rectangular and Random Binning Specifications}
 \begin{tabular}{lll} \hline
@@
 \end{tabular}
+\caption{Rectangular and Random Binning Specifications}
 \label{tab:rectbinning}
@@
 In cases of large data, binning greatly reduces the storage size for the information and the computation time needed to construct a binned scatterplot.
+Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
+Data reduction comes at the expense of spatial information of any of the individual points.
+The original spatial locations can only be recovered when minimally binned at the resolution of the data.
+The loss of information incurred from binning will be explored in following sections.
 
@@
 \centering
-\caption[Original, binned and reduced binned data table example.]{\label{DataReductionExample}Original, binned and reduced binned data tables, with data storage sizes.
-Binned using standard rectangular approach with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
-}
 \begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data]{
@@
 \end{minipage}
+\caption[Original, binned and reduced binned data table example.]{\label{DataReductionExample}Original, binned and reduced binned data tables, using standard rectangular binning with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
+}
 \end{table}
@@
 %
-Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
-Data reduction comes at the expense of spatial information of any of the individual points.
-After aggregation the original spatial locations cannot be recovered.
-The loss of information incurred from binning will be explored in following sections.
 

### 2015-12-17 versionForJCGS/PaperJCGS.Rnw
Simplify and condense in section for freq loss explanation

@@
 Making the data aggregation explicit allows us to calculate the loss we experience.
-
-By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures.
+%' 
+%' By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures.
 Using a small number of large bins may mask the real signal in the data, while an extremely large number of small bins may not sufficiently smooth over {\it noise} inherent in any real data set.
 Figure~\ref{binning} gives an overview of a data set and binned representations using different numbers of bins, demonstrating the loss of information with increasing bin size.
-
-\begin{figure}[hbtp]
-<<binning_setup, echo=FALSE,fig.width=2, fig.height=2, out.width='\\linewidth'>>=
-set.seed(46)
-x <- round(rnorm(200), digits=2)
-y <- round(rnorm(200), digits=2)
-X <- data.frame(x,y)
-@
-\begin{minipage}{.35\linewidth}
-<<first, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
-qplot(x,y, asp=1, main='Unbinned Data')+theme_bw()
-@
-\end{minipage}
-\hfil\begin{minipage}{.645\linewidth}
-\begin{minipage}{.495\linewidth}
-<<bin001, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
-binwidth=0.1
-binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
-qplot(x,y, asp=1, geom="blank") +
-  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
-  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
-  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
-                                by=binwidth)+binwidth/2), colour="grey90") + 
-  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
-                                by=binwidth)+binwidth/2), colour="grey90") +
-  theme_bw() + theme(panel.grid.major= element_blank(),
-                     panel.grid.minor= element_blank(), legend.position="bottom") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      trans="log", breaks=c(1, 2, 5, 10, 20))
-
-@
-\end{minipage}\hfil
-\begin{minipage}{.495\linewidth}
-<<bin025, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
-binwidth=0.25
-binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
-qplot(x,y, asp=1, geom="blank") +
-  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
-  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
-  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
-                                by=binwidth)+binwidth/2), colour="grey90") + 
-  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
-                                by=binwidth)+binwidth/2), colour="grey90") +
-  theme_bw() + theme(panel.grid.major= element_blank(),
-                     panel.grid.minor= element_blank(), legend.position="bottom") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      trans="log", breaks=c(1, 2, 5, 10, 20))
-@
-\end{minipage}\\
-\hfil\begin{minipage}{.495\linewidth}
-<<bin05, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
-binwidth=0.5
-binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
-qplot(x,y, asp=1, geom="blank") +
-  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
-  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
-  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
-                                by=binwidth)+binwidth/2), colour="grey90") + 
-  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
-                                by=binwidth)+binwidth/2), colour="grey90") +
-  theme_bw() + theme(panel.grid.major= element_blank(),
-                     panel.grid.minor= element_blank(), legend.position="bottom") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      trans="log", breaks=c(1, 2, 5, 10, 20))
-@
-\end{minipage}\hfil
-\begin{minipage}{.495\linewidth}
-<<bin1, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
-binwidth=1
-binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
-qplot(x,y, asp=1, geom="blank") +
-  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
-  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
-  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
-                                by=binwidth)+binwidth/2), colour="grey90") + 
-  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
-                                by=binwidth)+binwidth/2), colour="grey90") +
-  theme_bw() + theme(panel.grid.major= element_blank(),
-                     panel.grid.minor= element_blank(), legend.position="bottom") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
-                      trans="log", breaks=c(1, 2, 5, 10, 20))
-@
-\end{minipage}
-\end{minipage}
-\caption[Series of binned scatterplots of decreasing bin dimensions.]{\label{binning}Series of scatterplots showing the original data (scatterplot, left), and versions of the binned data for different bin widths.
+%' 
+%' \begin{figure}[hbtp]
+%' <<binning_setup, echo=FALSE,fig.width=2, fig.height=2, out.width='\\linewidth'>>=
+%' set.seed(46)
+%' x <- round(rnorm(200), digits=2)
+%' y <- round(rnorm(200), digits=2)
+%' X <- data.frame(x,y)
+%' @
+%' \begin{minipage}{.35\linewidth}
+%' <<first, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
+%' qplot(x,y, asp=1, main='Unbinned Data')+theme_bw()
+%' @
+%' \end{minipage}
+%' \hfil\begin{minipage}{.645\linewidth}
+%' \begin{minipage}{.495\linewidth}
+%' <<bin001, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
+%' binwidth=0.1
+%' binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+%' qplot(x,y, asp=1, geom="blank") +
+%'   geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+%'   ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+%'   geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") + 
+%'   geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") +
+%'   theme_bw() + theme(panel.grid.major= element_blank(),
+%'                      panel.grid.minor= element_blank(), legend.position="bottom") +
+%'   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+%'                       trans="log", breaks=c(1, 2, 5, 10, 20))
+%' 
+%' @
+%' \end{minipage}\hfil
+%' \begin{minipage}{.495\linewidth}
+%' <<bin025, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
+%' binwidth=0.25
+%' binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+%' qplot(x,y, asp=1, geom="blank") +
+%'   geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+%'   ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+%'   geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") + 
+%'   geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") +
+%'   theme_bw() + theme(panel.grid.major= element_blank(),
+%'                      panel.grid.minor= element_blank(), legend.position="bottom") +
+%'   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+%'                       trans="log", breaks=c(1, 2, 5, 10, 20))
+%' @
+%' \end{minipage}\\
+%' \hfil\begin{minipage}{.495\linewidth}
+%' <<bin05, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
+%' binwidth=0.5
+%' binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+%' qplot(x,y, asp=1, geom="blank") +
+%'   geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+%'   ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+%'   geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") + 
+%'   geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") +
+%'   theme_bw() + theme(panel.grid.major= element_blank(),
+%'                      panel.grid.minor= element_blank(), legend.position="bottom") +
+%'   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+%'                       trans="log", breaks=c(1, 2, 5, 10, 20))
+%' @
+%' \end{minipage}\hfil
+%' \begin{minipage}{.495\linewidth}
+%' <<bin1, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=5>>=
+%' binwidth=1
+%' binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
+%' qplot(x,y, asp=1, geom="blank") +
+%'   geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
+%'   ggtitle(sprintf("binwidth = %.2f", binwidth)) +
+%'   geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") + 
+%'   geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
+%'                                 by=binwidth)+binwidth/2), colour="grey90") +
+%'   theme_bw() + theme(panel.grid.major= element_blank(),
+%'                      panel.grid.minor= element_blank(), legend.position="bottom") +
+%'   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
+%'                       trans="log", breaks=c(1, 2, 5, 10, 20))
+%' @
+%' \end{minipage}
+%' \end{minipage}
+%' \caption[Series of binned scatterplots of decreasing bin dimensions.]{\label{binning}Series of scatterplots showing the original data (scatterplot, left), and versions of the binned data for different bin widths.
 The visual loss from binning at 0.1 is minimal, while a bin width of 1 gives a rough approximation.}
-\end{figure}
-
-In Figure~\ref{binning} the minimally binned scatterplot, with bin width equal to 0.1, is visually very similar to the traditional scatterplot; but importantly the binned scatterplot contains information about overlapping points.
+%' \end{figure}
+%' 
+%' In Figure~\ref{binning} the minimally binned scatterplot, with bin width equal to 0.1, is visually very similar to the traditional scatterplot; but importantly the binned scatterplot contains information about overlapping points.
 The second and third binned scatterplots, with bin width equal to 0.25 and 0.50 respectively, show higher-level summaries of the data but which may also provide more visually accessible information about the shape of the two-dimensional density distribution between $x$ and $y$.
@@
 
-Note that even though the losses from creating a binned scatterplot may turn out to be substantial, there is a huge gain with respect to an traditional scatterplot, where information can be masked in large data situations due to over-plotting of points.
+Note that while the losses from creating a binned scatterplot may turn out to be substantial, they present a huge gain with respect to an traditional scatterplot, where density information is implicitly masked in large data situations.
 The idea of loss from one-dimensional binning was explored by Scott using mean integrated squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}.
@@
 While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise.
-Color perception of shade is extremely context sensitive allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend.
-It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}).
-Figure~\ref{fig:context} shows an example of the context sensitivity of colors.
-Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.
-
-\begin{figure}[hbtp]
-\centering
-<<context, echo=F, fig.width=6, fig.height=4.5, out.width='.5\\textwidth'>>=
-#col <- c(rnorm(200, mean=0, sd=10), rnorm(10, mean=50), rnorm(200, mean=100, sd=10), rnorm(10, mean=50))
+Color perception of shade is extremely context sensitive -- see Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors -- allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend.
+It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.
+%Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.
+
+%' <<context, echo=F, include=F, fig.width=6, fig.height=4.5, out.width='.5\\textwidth'>>=
+%' #col <- c(rnorm(200, mean=0, sd=10), rnorm(10, mean=50), rnorm(200, mean=100, sd=10), rnorm(10, mean=50))
+%' Shade <- c(rbeta(200, 1,10),rep(.5,10), rbeta(200,10,1), rep(.5,10))
+%' x <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=.5, sd=0.075))
+%' y <- c(rnorm(210, mean=0, sd=0.075), rnorm(210, mean=.5, sd=0.075))
+%' Shape <- c(rep("circle",200),rep("triangle",10),rep("circle",200),rep("triangle",10))
+%' 
+%' require(ggplot2)
+%' qplot(x, y, colour=Shade, size=I(5),shape=Shape) + theme_bw() + 
+%'   xlab("") + ylab("")+
+%'   theme(aspect.ratio=1, axis.ticks=element_blank(), axis.text=element_blank()) +
+%'   scale_colour_gradient(limits=c(0,1), breaks=c(0,.5,1))
+%' 
+%' ggsave("./figure/ShadeContextDemo.png",width=5, height=5, dpi=600)
+%' @
+
+<<FreqLossDemoPlot,echo=F,include=T,eval=T,fig.width=c(6,6), fig.height=c(6,3), out.width=c('.49\\linewidth','.5\\linewidth'), fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="(Left) Q: Which triangles are darker?
+A: All triangles same shade = 500.
+\\newline (Right) Continuous and standard rectangular binned frequency color scales.", fig.scap="Continuous and discrete color scales for counts.">>=
+
+library(scales)
+#match color scales in each
+low = "#132B43"
+high = "#56B1F7"
+
 Shade <- c(rbeta(200, 1,10),rep(.5,10), rbeta(200,10,1), rep(.5,10))
@@
 require(ggplot2)
-qplot(x, y, colour=Shade, size=I(5),shape=Shape) + theme_bw() + 
+qplot(x, y, colour=1000*Shade, size=I(5),shape=Shape) + theme_bw() + 
   xlab("") + ylab("")+
-  theme(aspect.ratio=1, axis.ticks=element_blank(), axis.text=element_blank()) +
-  scale_colour_gradient(limits=c(0,1), breaks=c(0,.5,1))
-@
-\caption[Demonstration of the contextual sensitivity of shade perception.]{\label{fig:context}Assess the colors in this scatterplot.
-Are the triangles in the lighter cloud of points as dark as the triangles in the darker cloud of points?
-The answer is that all the triangles have the exact same shade!}
-\end{figure}
-
- Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
-We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, \mathscr{L} \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
-Figure~\ref{fig:FreqLossDemo} provides a visual example of a discrete color palette resulting from frequency binning.
-
-Research suggests that even under optimal conditions, we can effectively compare only about seven color hues simultaneously, and that we are even more limited in terms of distinguishing shade \citep{colorperception}.
-This provides a physical upper limit on the amount of frequency variation we can perceive through color.
-As a result, a frequency binning which produces seven of fewer frequency categories is preferable.
-
-<<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Continuous and discrete color scales for counts in the range 0 to 1000.
-Frequency binning done using  standard rectangular binning with origin $\\beta_0 = 0$ and binwidth $\\omega_c = 200$.", fig.scap="Continuous and discrete color scales for counts.">>=
-#J= 1
+  theme(plot.margin = unit(c(0.1,0.1,0.1,.1),"cm"), aspect.ratio=1,
+        axis.ticks=element_blank(), axis.text=element_blank()) +
+  scale_colour_gradient("Shade",limits=c(0,1000), breaks=c(0,500,1000), 
+                        low=high, high=low)
+
 J2 = 200
@@
 high = "#56B1F7"
-library(scales)
 blueGradient <- matrix(seq_gradient_pal(high, low, "Lab")(0:4400/4400), byrow=TRUE,
@@
   ylab("Continuous") 
-p2 <- qplot(id2, 2, geom="tile", fill=-fill2) + 
+p2 <- qplot(id2, 1.8, geom="tile", fill=-fill2) +  
   theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
         legend.position = "none",
-        axis.ticks=element_blank(),
-        axis.text.y=element_blank(),
+        axis.ticks=element_blank(),axis.text.y=element_blank(),
+        panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
         panel.background=element_blank())  + 
-  ylab("Discrete") + xlab("Count")
+  ylab("Discrete") + xlab("")
 
 grid.arrange(p1,p2,nrow=2)
-#ggsave("./figure/FreqBin.pdf",width=5, height=1.5)
+#ggsave("./figure/FreqBinDemo.png",width=5, height=1.5, dpi=600)
 @
+
+ Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information.
+We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, \mathscr{L} \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.
+Figure~\ref{fig:FreqLossDemoPlot} provides a visual example of a discrete color palette resulting from frequency binning.
+
+Research suggests that even under optimal conditions, we can effectively compare only about seven color hues simultaneously, and that we are even more limited in terms of distinguishing shade \citep{colorperception}.
+This provides a physical upper limit on the amount of frequency variation we can perceive through color.
+As a result, a frequency binning which produces seven of fewer frequency categories is preferable.
 
@@
 %
-where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.
-
+where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin.
+Note that this is effectively a sum of squared error from binning frequencies.
 While this numerical assessment of frequency loss does not exactly account for limitations in human perceptual ability, it does provide a more realistic model for the loss in perception that does occur.

### 2015-12-17 versionForJCGS/PaperJCGS.Rnw
Edits to Section 4.1

@@
 %
+
 \spacingset{1}
 \begin{table}[hbtp]
+\small
 \centering
@@
 \end{minipage}
+\normalsize
 \caption[Original, binned and reduced binned data table example.]{\label{DataReductionExample}Original, binned and reduced binned data tables, using standard rectangular binning with origin $(\beta_{0,x},\beta_{0,y})$ = (-10,-10) and binwidths $\omega_x = \omega_y = 10$.
@@
 The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all line segments represents the total spatial loss, $L^S$.
-
-<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms.
-Note that the total spatial loss for the standard algorithm is smaller than by random binning.
-The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>=
+For random assignment the total spatial loss can be calculated by simply replacing the standard binning function with the random binning function in Equation~\ref{totspatloss}.
+
+<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms." >>=
 n=12
@@
 
-For random assignment the total spatial loss can be calculated by simply replacing the standard binning function with the random binning function in Equation~\ref{totspatloss}.
 The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
-Special consideration should be paid to the two points with line segments highlighted in red in the panel for random binning.
-In standard binning these points are in neighboring bins but have been allocated to the opposite bin during random binning.
-We see that the binned scatterplot remains identical if these points are re-allocated back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
-
-The minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot, will be referred to as the \textit{net spatial loss}, $L^S_{net}$.
-The net spatial loss for standard binned data is always  equivalent to the total spatial loss due to the deterministic bin allocations.
-For random binning algorithms the net spatial loss is achieved by first exchanging bin assignments for all pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then calculating the total spatial loss from this processed binned data.
+We see that in random binning there are pairs of points -- like those with line segments highlighted in red -- that have been allocated in such a way that they are closer to their partner's bin center than their own.
+We see that the binned scatterplot remains identical if these points are swapped back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.
+To appropriately reflect the perceived loss of spatial information we use the \textit{net spatial loss}, $L^S_{net}$, which is the minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot.
+For standard binned data, the net spatial loss is always equivalent to the total spatial loss due to the deterministic bin allocations.
+For random binned data, the net spatial loss is achieved by first exchanging bin assignments for all pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then calculating the total spatial loss from this post-processed binned data.
 
@@
 
-It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates.
+% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates.
 The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets.
@@
 In binned scatterplots, we have traded the exact locations for exact frequencies within bins.
-
+% 
 Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly.
@@
 
-<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
+<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
 Density changes are impossible to make out in the solid black areas of the plots.
 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=
+library(scales) 
 ### Try 2d binning with MVN data
@@
   theme_bw() + 
+  theme(plot.margin =unit(c(0.1,0.1,0.1,.1),"cm")) +
   #ggtitle("Scatterplots of Simulated Data") +
@@
 \begin{center}
- \includegraphics[keepaspectratio=TRUE,width=.99\textwidth]{./figure/SimScatters.png}
+ \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
 	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data.

### 2015-12-17 versionForJCGS/PaperJCGS.Rnw
first edits to section 5

@@
 %
-The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
-
+The parameters were selected to have data values roughly span the region $[0,100]^2$.
+The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
+These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data.
+Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
+%
 <<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.
@@
 @
-
 \begin{figure}[hbtp]
@@
  \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
-	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data.
-Density changes are impossible to make out in the solid black areas of the plots.
-97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.}
+	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
 	\end{center}
 \end{figure}
-
-These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data.
-Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
-
+%
+%Density changes are impossible to make out in the solid black areas of the plots.
+97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
+%
 The \textit{data resolution} is defined as the smallest increment between successive data values.
@@
 
-<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=6, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
+<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
 ## create binned data for standard and random using variety of bin sizes (from 2 to 20, by 2)
@@
 spatlossdatSquareBins$bintype <- factor(spatlossdatSquareBins$bintype, levels=c("standard","random"))
-p1 <- qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
-      data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
+
+library(tidyr)
+spatlossdatSquareBins2 <- gather(spatlossdatSquareBins, measure, value, c(totalSpatialLoss,binTime))
+spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Net Spatial Loss","Binning Time (sec)"))
+# 
+# p1 <- qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
+#       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
+#   theme_bw() + ylab("Net Spatial Loss") +
+#   xlab("Width/Height of Square Bins") + ylim(c(0, 1100000))
+# 
+# p2 <- qplot(widthx,as.numeric(binTime),geom="path",colour=bintype, linetype=bintype,
+#       data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
+#   theme_bw() + ylab("Binning Time (sec)") +
+#   xlab("Width/Height of Square Bins")
+# 
+# grid.arrange(p1,p2, nrow=2)
+
+qplot(widthx,value,geom="path",colour=bintype, linetype=bintype,
+      data=spatlossdatSquareBins2,group=bintype, size=I(1)) +
+  facet_grid(measure~distrib, scales="free_y" )+
   theme_bw() + ylab("Net Spatial Loss") +
-  xlab("Width/Height of Square Bins") + ylim(c(0, 1100000))
-
-p2 <- qplot(widthx,as.numeric(binTime),geom="path",colour=bintype, linetype=bintype,
-      data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
-  theme_bw() + ylab("Binning Time (sec)") +
   xlab("Width/Height of Square Bins") 
-
-grid.arrange(p1,p2, nrow=2)
 @

### 2015-12-17 versionForJCGS/PaperJCGS.Rnw
Swap authorship order

@@
 \title{Binning Strategies and Related Loss for Large Data}
-\author{Karsten Maurer$^1$, Susan VanderPlas$^2$, Heike Hofmann$^{2,3}$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
+\author{Karsten Maurer$^1$, Heike Hofmann$^{2,3}$, Susan VanderPlas$^2$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
   \maketitle

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
title for anonymized version, paths to files

@@
   \begin{center}
-    {\LARGE\bf Title}
+    {\LARGE\bf Binning Strategies and Related Loss for Large Data}
 \end{center}
@@
 <<FlightMinBinScat, echo=F,include=F, cache=TRUE>>=
-source("RectangularBinningFunctions.R")
+source("./data/RectangularBinningFunctions.R")
 load(file="./data/DepartDelaysMinutes.RData")

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
-lower dpi on pngs to render faster while editing -Tons of minor edits to section 5.1

--- 2015-12-17 versionForJCGS/PaperJCGS.Rnw
@@
 qplot(data=pitchdat, x=G, y=SO, geom="point", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
-  ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=900)
+  ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=300)
 
@@
 qplot(data=d1, x=G, y=SO, geom="point", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
-  ggsave("./figure/Overplotting.png", width=5, height=5, dpi=900)
+  ggsave("./figure/Overplotting.png", width=5, height=5, dpi=300)
 
@@
   qplot(data=d1, x=G, y=SO, geom="point", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
-  ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=900)
+  ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=300)
 
@@
   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=900)
+  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
 d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
@@
   scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-ggsave("./figure/BinBubble.png", width=5, height=5, dpi=900)
+ggsave("./figure/BinBubble.png", width=5, height=5, dpi=300)
 
@@
   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-  ggsave("./figure/BinScatter.png", width=5, height=5, dpi=900)
+  ggsave("./figure/BinScatter.png", width=5, height=5, dpi=300)
 
@@
 Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics.
-The scatterplot shows 144 seasons (from the years of 1871 -- 2014) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).
-The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season.
+The scatterplot shows 144 seasons (from the years of 1871 -- 2014) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database.\footnote{Data Accessed 12/16/2015 at \url{http://www.seanlahman.com/baseball-archive/}}  The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season.
 While the data set is only medium sized with \Sexpr{dim(pitchdat)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
@@
 Note that numerical attributes other than frequency of the binned data may also be recorded during binning, however only frequency is required to construct a binned scatterplot.
-Data reduction comes at the expense of spatial information of any of the individual points.
-The original spatial locations can only be recovered when minimally binned at the resolution of the data.
+Data reduction comes at the expense of spatial information of any of the individual points, which can only be recovered when minimally binned at the resolution of the data.
 The loss of information incurred from binning will be explored in following sections.
@@
 %' 
-%' ggsave("./figure/ShadeContextDemo.png",width=5, height=5, dpi=600)
+%' ggsave("./figure/ShadeContextDemo.png",width=5, height=5, dpi=300)
 %' @
@@
 grid.arrange(p1,p2,nrow=2)
-#ggsave("./figure/FreqBinDemo.png",width=5, height=1.5, dpi=600)
+#ggsave("./figure/FreqBinDemo.png",width=5, height=1.5, dpi=300)
 @
@@
 
-ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=900)
+ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
 @
@@
 
-For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension.
+For rectangular binning specification options include the type of algorithm, location of the origin and binwidths for each dimension.
 To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning.
-For equal bin widths, the net spatial loss under standard binning is always less than or equal to the net spatial loss under random binning.
+Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning the fine resolution simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
+For equal bin sizes, the net spatial loss under standard binning is always less than or equal to the net spatial loss under random binning.
 This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.
-
-Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning the fine resolution simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
-It is apparent that the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases.
-The net spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.
-
-<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>=
+The net spatial loss from random binning becomes increasingly costly as bin sizes increase, as indicated by the widening gap between the lines and the steeper slopes.
+%
+<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning.">>=
 ## create binned data for standard and random using variety of bin sizes (from 2 to 20, by 2)
@@
   facet_grid(measure~distrib, scales="free_y" )+
-  theme_bw() + ylab("Net Spatial Loss") +
+  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,0.1, 0.1),"cm")) + 
+  ylab("Net Spatial Loss") +
   xlab("Width/Height of Square Bins") 
 @
-
-All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin.
+%
+%All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. 
 While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data.
-Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.
-Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.
-Another pertinent argument for using larger bins is that it will smooth over larger regions if we are primary interested visualizing the large scale density structure.
-
-If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
-Binned scatterplots are inherently biased displays of spatial information as they shift visual emphasis from the true location of individual points to the geometric centers of tiles.
-However, as bin sizes become increasingly fine,  the bias decreases as the plot shows more precise spatial locations.
-When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate is spatially exact.
-Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator.
-We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
-It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss.
-This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies.
-%Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
-
-While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artifacts in the binned scatterplot.
-For example, if the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins in the binned scatterplot.
-These gaps of white space are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density.
-The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with visual evaluation.
-
-There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial striping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.
-
-To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension.
+Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes.\footnote{Computation times using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz.} Larger bins may also be reasonable to use if we are primary interested visualizing the large scale density structure and wish to smooth over fine structural noise.
+
+While bin widths can be chosen as any positive real value, bin widths should be specified as an integer multiple of the resolution of the data because non-integer multiples of the data resolution lead to systematically different numbers of possible data values per bin.
+For bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of tiles in the binned scatterplot.
+While these gaps do exist in the data, they are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density and could also create the Hermann-grid optical illusion~\citep{hermann:1870, spillmann:1994}.
+More seriously, bin dimensions that are non-integer multiples larger than the data resolution lead to \textit{artificial striping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data.
+
+To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension.
 Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
@@
 
-<<UniformStripes, echo=F,include=T,eval=T,fig.width=8, fig.height=5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data with 1X1, 4X4 and 5X5 square bins.
-For standard binning, the 1X1 bins leaves white-space gaps between bins, and 5X5 bins cause artificial density stripes; whereas random binning smoothes density over these poor choices in bin dimensions.">>=
+<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>=
 # Commented out code for data generation: run once then recomment
@@
 allbindat$BinningType <- factor(allbindat$BinningType, levels=c("Standard","Random"))
-
-p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="1 X 1 Unit")) +
-  facet_grid(BinningType~BinSize)+
+allbindat$height <- 1 ; allbindat$width <- 1
+allbindat$height[which(allbindat$BinSize=="4 X 4 Unit")] <- 4
+allbindat$width[which(allbindat$BinSize=="4 X 4 Unit")] <- 4
+allbindat$height[which(allbindat$BinSize=="5 X 5 Unit")] <- 5
+allbindat$width[which(allbindat$BinSize=="5 X 5 Unit")] <- 5
+
+ggplot(aes(x=binxs, y=binys), data=allbindat) +
+  geom_tile(aes(fill=binfreq,height=height, width=width), data=allbindat)+
+  facet_grid(BinningType~BinSize, scales="free", space="free")+
   ggtitle(" ") +   xlab("") + ylab("") +
@@
                       limits=c(0, 450),name="Bin Count") + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
+                     aspect.ratio=1,
                      panel.grid= element_blank())
 
-p2 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="4 X 4 Unit")) +
-  facet_grid(BinningType~BinSize)+
-  ggtitle(" ") +   xlab("") + ylab("") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43",
-                      limits=c(0, 450),name="Bin Count") + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     panel.grid= element_blank())
-
-p3 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="5 X 5 Unit")) +
-  facet_grid(BinningType~BinSize)+
-  ggtitle(" ") +   xlab("") + ylab("") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43",
-                      limits=c(0, 450),name="Bin Count") + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     panel.grid= element_blank())
-
-grid.arrange(p1,p2,p3, nrow=1)
+# p2 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="4 X 4 Unit")) +
+#   facet_grid(BinningType~BinSize)+
+#   ggtitle(" ") +   xlab("") + ylab("") +
+#   scale_fill_gradient(low="#56B1F7", high="#132B43",
+#                       limits=c(0, 450),name="Bin Count") + 
+#   theme_bw() + theme(plot.margin =,
+#                      legend.position="none",aspect.ratio=1,
+#                      panel.grid= element_blank())
+# 
+# p3 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=subset(allbindat,BinSize=="5 X 5 Unit")) +
+#   facet_grid(BinningType~BinSize)+
+#   ggtitle(" ") +   xlab("") + ylab("") +
+#   scale_fill_gradient(low="#56B1F7", high="#132B43",
+#                       limits=c(0, 450),name="Bin Count") + 
+#   theme_bw() + theme(plot.margin = unit(c(0.1,0.1,0.1, 0.1),"cm"),
+#                      aspect.ratio=1,
+#                      panel.grid= element_blank())
+# 
+# grid.arrange(p1,p2,p3, nrow=1)
 
@@
 
+If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
+Binned scatterplots shift visual emphasis from true location of individual points to the geometric centers of tiles, making them visually biased displays of density.
+However, as bin sizes become increasingly fine, the bias decreases; with no visual bias when minimal binning perfectly matches bins with the data resolution the density estimate.
+Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator.
+We may also consider a minimal binned scatterplot as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
+It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss.
+This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies.
+%Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
+
 The exploration of spatial and frequency loss in the process of creating a binned scatterplot yields a number of important properties and practical recommendations for their construction.
@@
   ggtitle("Alpha-Blended Scatterplot of Airline Departure Times") 
-ggsave(filename = "./figure/AlphaScatterAirline2014.png", width= 6, height=6, dpi=900)
+ggsave(filename = "./figure/AlphaScatterAirline2014.png", width= 6, height=6, dpi=300)
 @
@@
   geom_abline(intercept=0,slope=1, color="gray70", linetype=3)
-ggsave(filename = "./figure/MinimalBinnedAirline2014.png", width= 6.5, height=6, dpi=900)
+ggsave(filename = "./figure/MinimalBinnedAirline2014.png", width= 6.5, height=6, dpi=300)
 @
@@
 # binnedRdm <- qplot(data=air.binned2.rdm, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
-# ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=600)
+# ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=300)
 @

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
finish tightening up section 5.1

@@
 
-<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>=
+<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>=
 # Commented out code for data generation: run once then recomment
@@
 The binning origin can also influence the spatial loss in the binned scatterplot.
-For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density in outermost bins.
-Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data with two different origin choices: (0,0) and (-9,-9).
-The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions.
-The binning origin at (0,0) is also superior in terms of the net spatial loss, which is about seven percent lower than for the (-9,-9) origin.
-For data with very fine resolution, it is a good idea to set the origin at the minimum value for each dimension.
-
-<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10X10 square bins with origins at (0,0) and (-9,-9).
-The bold lines at x=0 and y=0 denote the lower bounds of the data.">>=
+For data with fine resolution compared to the bin dimensions, the origin is only largely consequential for distributions with high density near a natural boundary -- e.g.
+weights of small items bounded below at 0 -- where the origin should align with the boundary so that the outermost bins do not cover empty density regions.
+As an example, Figure~\ref{fig:ExponentialOriginPlots} displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data where the binning origin at (0,0) incurs seven percent lower spatial loss than for the (-9,-9) origin due to the heavy overlap into negative regions.
+%The binning beginning at the (-9,-9) origin suffers visually from the illusion that the density drops off near the lower bounds due to the fact that these lowest bins in each dimension are largely empty due to the negative regions.
+
+<<ExponentialOriginPlots, echo=F,include=T,eval=T,fig.width=8, fig.height=3, out.width='.9\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots for the fine exponential data using standard binning with 10X10 square bins with origins at (0,0) and (-9,-9).
+The bold lines denote data lower bounds.">>=
 
@@
 p1 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5[[1]]) +  
-  ggtitle("Origin at (0,0)") +
+#  ggtitle("Origin at (0,0)") +
   scale_fill_gradient(low="#56B1F7", high="#132B43") + 
-  theme_bw() + theme(legend.position="none",
+  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
+                     legend.position="none",
                      aspect.ratio=1, panel.grid= element_blank()) +
@@
 p2 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5offset[[1]]) +  
-  ggtitle("Origin at (-9,-9)") + 
+#  ggtitle("Origin at (-9,-9)") + 
   scale_fill_gradient(low="#56B1F7", high="#132B43") + 
-  theme_bw() + theme(legend.position="none",
+  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
+                     legend.position="none",
                      aspect.ratio=1, panel.grid= element_blank()) +
@@
 
-For data with a coarse resolution the location of the origin is also important because the origin controls the proximity of possible data values to bin centers.
-We define the \textit{origin offset} for each dimension, as the tuple, ($o_x$, $o_y$), by which we shift the bivariate bin origin $(\beta_{0,x},\beta_{0,y})$ from the minimal values in $x$ and $y$:
-%
-\begin{eqnarray}\label{OriginOffset}
-(\beta_{0,x},\beta_{0,y}) = (x_{(1)}, y_{(1)}) - (o_x, o_y),
-\end{eqnarray}  
-%
-where $x_{(1)}$ and $y_{(1)}$ are the minimal data values in $x$ and $y$. %, and $(o_x, o_y)$ is the origin offset. 
-Thus the offset indicates the number of units in each dimension to shift the binning origin below the origin naturally encouraged by the data.
-
+For data with a coarse resolution the location of the origin is important because the origin controls the proximity of possible data values to bin centers.
+We define the \textit{origin offset} for each dimension, as the tuple, ($o_x$, $o_y$), by which we offset the bivariate bin origin from the data minima, ($x_{(1)}$, $y_{(1)}$).
+Thus the origin offset indicates the number of units in each dimension to shift the binning origin below the origin naturally encouraged by the data, resulting in $(\beta_{0,x},\beta_{0,y})= (x_{(1)}, y_{(1)}) - (o_x, o_y)$.
+% %
+% \begin{eqnarray}\label{OriginOffset}
+% (\beta_{0,x},\beta_{0,y}) = (x_{(1)}, y_{(1)}) - (o_x, o_y),
+% \end{eqnarray}  
+% %
  It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units  minimizes the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the $X$ dimension and $\alpha_y$ units in the $Y$ dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}).
-In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
-
+
+In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable binning choice for lowering spatial loss for coarse resolution bivariately symmetric data using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
 Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is shifted while using standard rectangular binning for the coarse simulated data sets.
-Note that for simplicity, changes to the origin offset are made equally in each dimension, thus shifting the origin at a 45 degree angle.
-Since the coarse data has a 2X2 unit resolution, we pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the theoretical assumptions to differing degrees.
+Note that for simplicity, changes to the origin offset are made equally in each dimension.
+Since the coarse data has a 2X2 unit resolution, we pay special attention to an origin offset of (1,1) in order to assess how well the proposed default origin offset at ($\alpha_x/2$,$\alpha_y/2$) works in each scenario.
 The round glyphs indicate the origin offset where the net loss reaches an absolute minimum in the simulation.
@@
 
-<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=4, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss for coarse simulated data at a 2X2 unit resolution using various sized square bins over the range of possible origin offsets.
-The vertical gray line in each facet indicates the origin offset of (1,1).">>=
+<<OriginOffsetSimulations, echo=F,include=T,eval=T,fig.width=8, fig.height=4, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Net spatial loss 2X2 unit resolution data using various sized square bins over the range of possible origin offsets.
+The vertical gray lines indicate origin offset of (1,1).">>=
 ##  Check out the spatial loss for the data that is recorded to resolution = alpha at various origin offsets
@@
 \bibliographystyle{asa}
+
+\spacingset{1} % reference spacing
+
 \bibliography{references}

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
remove another uni-code from ref (don't know why they are there)
Edits to section 5.2 on frequency loss

@@
 \begin{figure}[hbtp]
+\centering
   \subfloat[Traditional scatterplot ]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/Overplotting.png}}
@@
   \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinBubble.png}}
-
 	\caption[Traditional and adapted scatterplots for games vs.
@@
 The reduced binned data from spatial binning contains the center and count information for all bins.
-The bin frequencies may be mapped continuously to a precisely rendered shade, however it is naive to believe that human perception will be able to perfectly extract that information.
+The bin frequencies may be mapped continuously to a precisely rendered shade, however it is naive to believe that human perception will be able to perfectly extract that information as a numeric value.
 There is implicitly loss of frequency information occurring when the shade of a tile is visually mapped back to a frequency through the use of a shading scale index.
@@
 
-<<StandardFreqBinning4710none,echo=F,include=T,eval=T,fig.width=10, fig.height=3, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Binned scatterplots for the simulated bivarate normal data with varying numbers of standard binned frequency groups.
-At 10 frequency bins the shades are difficult to distinguish, with little difference from a binned scatterplot with continuous frequency shading.">>=
+<<StandardFreqBinning4710none,echo=F,include=T,eval=T,fig.width=10, fig.height=3, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Binned scatterplots for the simulated bivarate normal data with varying numbers of standard binned frequency groups.">>=
 binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
@@
 The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified.
-The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data.
-We first notice the large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data, and lowest for the uniform data.
-We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms.
-This relationship largely flattens out after the fourth frequency bin, each subsequent bin reducing the frequency loss less than the previous.
+Frequency loss is definitionally a decreasing function of the number of frequency bins for both standard and quantile frequency binning algorithms.
+The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using both binning algorithms, with between one and ten frequency bins, from each set of simulated data.
+We first note the large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data, and lowest for the uniform data.
+The decreasing frequency loss for both algorithms flattens out after the fourth frequency bin, each subsequent bin reducing the frequency loss less than the previous.
 Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot.
-Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used.
-But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.
-Therefore, it is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies.
-
-Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
-The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} counts for bins from the same sets of simulated data.
-Log frequency binning behaved very similarly to standard frequency binning, where log frequency loss decreased as the number of frequency bins increased.
-The same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
-
-<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution.
-",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data.
+
+Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
+It is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies due to the improved handling of skewed bin count distibutions while maintaining similar frequency loss to standard frequency binning when four to seven frequency bins are used, as seen in Figure~\ref{fig:FreqLossCompare}.
+Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.
+%In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
+
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively.",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data.
 ">>=
@@
 # find freq loss from standard, quantile, logCount
+detach("package:dplyr", unload=TRUE)
+source("BinningLossFunctions.R")
 freqLossSimDat <- NULL
@@
 freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels= c("Standard", "Quantile", "Standard Log", "Quantile Log"))
+# freqLossSimDat$FreqLossType <- "Total Freq Loss"
+# freqLossSimDat$FreqLossType[which(freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log" )] <- "Total Log Freq Loss"
+# 
+# qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
+#       data=freqLossSimDat , color = FreqBinType, size=I(1)) +
+#   facet_grid(FreqLossType~DistType, scales="free") +  theme_bw() +
+#   scale_x_continuous(breaks=1:10) +
+#   xlab("Freq Bins") + ylab("Total Freq Loss") + 
+#   theme(legend.position="bottom",plot.margin = unit(c(.1,.1,.1,.1),"cm"))
+#
 
@@
 
-Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
-Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.
-In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.
-
+Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared.
+The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} counts for bins from the same sets of simulated data.
+Log frequency binning behaved very similarly to standard frequency binning, where log frequency loss decreased as the number of frequency bins increased.
+The same advice to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.
 Since the loss scales are not comparable, the choice is guided by desired interpretation.
-Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
+%Figure~\ref{freqbinning} shows binned scatterplots for the simulated normal data with quantile frequency binning and standard log frequency binning.
 For standard log frequency binning, the shade is to be interpreted as an ordinal indicator based on equally spaced groupings of log bin frequencies; whereas for quantile frequency binning the shade denotes groups based on frequency quantiles.
@@
 Both shading schemes are effective at reducing the visual impact of the highest density bins near the center of each plot, allowing for the differences in the surrounding bin frequencies to be emphasized.
-It should be noted that both frequency binning algorithms produced very similar looking binned scatterplots, but this will not always be the case.
+%It should be noted that both frequency binning algorithms produced very similar looking binned scatterplots, but this will not always be the case.
 It occurred in this scenario because the distribution of log bin frequencies from the bivariate normal data is nearly uniform, thus the standard log frequency binning aligns closely with groupings from the quantile frequency binning.
 
-\begin{figure}[hbtp]
-<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=4.8>>=
-binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
-freqgrps = 5
-binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=FALSE)
-binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=FALSE)
-binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=TRUE)
-binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=TRUE)
-cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:freqgrps)/freqgrps)
-
-# #Plot with Binned Frequencies
-# labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
-# p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutFreqGroups[[1]],
-#             main="Standard Freq Binning") +
-#     xlab("") + ylab("")+
-#   scale_fill_manual("Frequencies", values=cols, 
-#                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-#                                            title.position = "top"),
-#                       breaks=unique(binoutFreqGroups[[1]]$freqgroup),
-#                       labels=labels) + 
-#   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-#                      legend.key.width = unit(1, "cm"),
-#                      panel.grid= element_blank(),
-#                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
-#                      legend.text=element_text(size=12) )
-
-# Plot with Quantile Binned Frequencies
-labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
-p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
-            main="Quantile Freq Binning") +
-    xlab("") + ylab("")+
-  scale_fill_manual("Frequencies", values=cols, 
-                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-                                           title.position = "top"),
-                      labels=labels) + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank(),
-                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
-                     legend.text=element_text(size=12) ) 
-# Plot with standard binned Log Frequencies
-labels = paste(unique(binoutLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutLogFreqGroups[[1]]$freqgroup), ")", sep="")
-p4 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutLogFreqGroups[[1]],
-            main="Standard Log Freq Binning") +
-    xlab("") + ylab("")+
-  scale_fill_manual("Log Frequencies", values=cols, 
-                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-                                           title.position = "top"),
-                      breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
-                      labels=labels) + 
-  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-                     legend.key.width = unit(1, "cm"),
-                     panel.grid= element_blank(),
-                     plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
-                     legend.text=element_text(size=12))
-
-# # Plot with quantile binned Log Frequencies
-# labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
-# p5 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup),
-#             data=binoutQuantLogFreqGroups[[1]],
-#             main="Quantile Log Freq Binning") +
-#     xlab("") + ylab("")+
-#   scale_fill_manual("Log Frequencies", values=cols, 
-#                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
-#                                            title.position = "top"),
-#                       breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
-#                       labels=labels) + 
-#   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
-#                      legend.key.width = unit(1, "cm"),
-#                      panel.grid= element_blank(),
-#                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
-#                      legend.text=element_text(size=12))
-
-grid.arrange(p3,p4, nrow=1)
-@
-\caption[Frequency binned scatterplots for simulated bivarate normal data.
+%' 
+%' \begin{figure}[hbtp]
+%' <<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=4.8>>=
+%' binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
+%' freqgrps = 5
+%' binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=FALSE)
+%' binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=FALSE)
+%' binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=TRUE)
+%' binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=TRUE)
+%' cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:freqgrps)/freqgrps)
+%'  
+%' # #Plot with Binned Frequencies
+%' # labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
+%' # p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutFreqGroups[[1]],
+%' #             main="Standard Freq Binning") +
+%' #     xlab("") + ylab("")+
+%' #   scale_fill_manual("Frequencies", values=cols, 
+%' #                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+%' #                                            title.position = "top"),
+%' #                       breaks=unique(binoutFreqGroups[[1]]$freqgroup),
+%' #                       labels=labels) + 
+%' #   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+%' #                      legend.key.width = unit(1, "cm"),
+%' #                      panel.grid= element_blank(),
+%' #                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+%' #                      legend.text=element_text(size=12) )
+%' 
+%' # Plot with Quantile Binned Frequencies
+%' labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
+%' p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
+%'             main="Quantile Freq Binning") +
+%'     xlab("") + ylab("")+
+%'   scale_fill_manual("Frequencies", values=cols, 
+%'                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+%'                                            title.position = "top"),
+%'                       labels=labels) + 
+%'   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+%'                      legend.key.width = unit(1, "cm"),
+%'                      panel.grid= element_blank(),
+%'                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+%'                      legend.text=element_text(size=12) ) 
+%' # Plot with standard binned Log Frequencies
+%' labels = paste(unique(binoutLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutLogFreqGroups[[1]]$freqgroup), ")", sep="")
+%' p4 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutLogFreqGroups[[1]],
+%'             main="Standard Log Freq Binning") +
+%'     xlab("") + ylab("")+
+%'   scale_fill_manual("Log Frequencies", values=cols, 
+%'                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+%'                                            title.position = "top"),
+%'                       breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
+%'                       labels=labels) + 
+%'   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+%'                      legend.key.width = unit(1, "cm"),
+%'                      panel.grid= element_blank(),
+%'                      plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
+%'                      legend.text=element_text(size=12)) 
+%' 
+%' # # Plot with quantile binned Log Frequencies
+%' # labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
+%' # p5 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup),
+%' #             data=binoutQuantLogFreqGroups[[1]],
+%' #             main="Quantile Log Freq Binning") +
+%' #     xlab("") + ylab("")+
+%' #   scale_fill_manual("Log Frequencies", values=cols, 
+%' #                       guide = guide_legend(label.position="bottom", label.hjust=0.5,  
+%' #                                            title.position = "top"),
+%' #                       breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
+%' #                       labels=labels) + 
+%' #   theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
+%' #                      legend.key.width = unit(1, "cm"),
+%' #                      panel.grid= element_blank(),
+%' #                      plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
+%' #                      legend.text=element_text(size=12)) 
+%'  
+%' grid.arrange(p3,p4, nrow=1)
+%' @
+%' \caption[Frequency binned scatterplots for simulated bivarate normal data.
 ]{\label{freqbinning}Simulated bivariate normal data spatially binned using standard algorithm with 5X5 unit bins and origin at (0,0).
@@
 }
-\end{figure}
+%' \end{figure}
 

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
Edits to baseball example part 2
commented out airline flights example. put in supplemental materials eventually

@@
 Technological advances have facilitated collection and dissemination of large data as records are digitized and our lives are increasingly lived online.
-According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes" (\url{http://www.emc.com/about/news/press/2014/20140409-01.htm}).
-This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: ``It's a revolution $\dots$ The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government.
+According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes".
+\footnote{Access 12/18/2015 at \url{http://www.emc.com/about/news/press/2014/20140409-01.htm}} This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: ``It's a revolution $\dots$ The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government.
 There is no area that is going to be untouched"-- Gary King, Harvard Institute.
@@
 The net spatial loss from random binning becomes increasingly costly as bin sizes increase, as indicated by the widening gap between the lines and the steeper slopes.
+
+If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
+Binned scatterplots shift visual emphasis from true location of individual points to the geometric centers of tiles, making them visually biased displays of density.
+However, as bin sizes become increasingly fine, the bias decreases; with no visual bias when minimal binning perfectly matches bins with the data resolution the density estimate.
+Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator.
+We may also consider a minimal binned scatterplot as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
+It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss.
+This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies.
+%Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
 %
@@
 
-<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively.",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data.
+<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=7, out.width='.99\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively.",fig.scap="Lineplots for total frequency losses from standard and quantile binning for simulated bivariate data.
 ">>=
@@
 %' \end{figure}
-
-%-----------------------------------------------------------------------------
-
-\section{Discussion and Examples}
-
-If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency.
-Binned scatterplots shift visual emphasis from true location of individual points to the geometric centers of tiles, making them visually biased displays of density.
-However, as bin sizes become increasingly fine, the bias decreases; with no visual bias when minimal binning perfectly matches bins with the data resolution the density estimate.
-Also, the density estimation more perfectly reflect the bivariate density as the sample size increases; thus making the binned scatterplot a consistent visual estimator.
-We may also consider a minimal binned scatterplot as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator.
-It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss.
-This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies.
-%Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.
-
-The exploration of spatial and frequency loss in the process of creating a binned scatterplot yields a number of important properties and practical recommendations for their construction.
+% 
+% %-----------------------------------------------------------------------------
+% 
+% \section{Discussion and Examples}
+% 
+% 
+% 
+% The exploration of spatial and frequency loss in the process of creating a binned scatterplot yields a number of important properties and practical recommendations for their construction.
 These recommendations and loss properties will now be demonstrated through two examples using real bivariate data.
@@
 
-\subsection{Binning Loss in Baseball Data: Strikeout and Game Counts}
+\subsection{Binning Loss in Baseball Data}
 
@@
 
-We now revisit the baseball data used earlier and construct binned scatterplots to display the relationship between strikeouts and game counts.
+We now revisit the baseball data used earlier and construct binned scatterplots using specification principles developed based on loss properties from the previous sections.
 We begin by specifying the bin dimensions and origin for standard rectangular binning.
-The data are counts, each variable is therefore recorded as an integer, corresponding to a one game by one strike-out data resolution.
-We use integer bin widths in each dimension of spatial bins to avoid artificial density stripes.
-The data has only 42,583 observations, so we can bin on a fine grid without taking an inordinate amount of computation time.
-We use approximately 50 bins in each dimension; the data range from 1 to 106 games and from 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high.
-The origin will be set to a half unit below the minimum value in each dimension, at $(0.5,-0.5)$.
-
-The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of raw bin frequencies.
-The most striking feature in the frequency distribution is the dark spot in the bottom right of the plot representing a large number of pitchers that played very few games and had very few strikeouts.
-It is nearly impossible to distinguish the density structure across the remaining bins, representing better pitchers who played many games and earned many strikeouts.
-In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts.
-Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities.
-Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins.
-Each frequency binning approach adds a layer of complexity to interpreting the shade, however they both effectively emphasize the forked ridges in the density structure.
+We use integer bin widths in each dimension of spatial bins to avoid artificial density stripes because each variable is recorded as an integer, corresponding to a one game by one strike-out data resolution.
+We desire relatively fine binning of approximately 50 bins in each dimension; the data range from 1 to 106 games and from 0 to 513 strike-outs, thus we set bin dimensions to two games by ten strikeouts.
+The origin offset of $(0.5,-0.5)$ is used to shift the binning back by a half game and down by one half strikeout to adjust for the data resolution.
+%The data has only 42,583 observations, so we can bin on a fine grid without taking an inordinate amount of computation time.
+We find that we have approximately 3\% lower net spatial loss, after standardizing measurement units, for our binning specification using the recommended origin offset than if we were to naively set the binning origin at the minimum data values of one game and zero strikeouts.
 
@@
 p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
-      main="Standard Freq Binning") +
+      main="Raw Freq Binning") +
   xlab("Games (count)") + ylab("Strike Outs (count)")+
@@
 
-The interpretability of the net spatial loss in this scenario suffers because the units and scales of the two variables differ.
+The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of raw bin frequencies.
+The most striking feature in the frequency distribution is the dark spot in the bottom right of the plot representing a large number of pitchers that played very few games and had very few strikeouts.
+It is nearly impossible to distinguish the density structure across the remaining bins, representing better pitchers who played many games and earned many strikeouts.
+%In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts. 
+Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities.
+Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins.
+Each frequency binning approach adds a layer of complexity to interpreting the shade, however they both effectively emphasize the forked ridges in the density structure.
+
+%The interpretability of the net spatial loss in this scenario suffers because the units and scales of the two variables differ.
 To remedy this, the two variables can be standardized then binned equivalently by rescaling the bin dimension, where the net spatial loss is interpreted as a distance in units of standard deviations.
@@
 
-%--------------------------------------------------------------------
-
-\subsection{Big Data: Airline Departure Times}
-The Federal Aviation Association (FAA) requires all airlines based in the United States to report   details for every single flight.
+\section{Conclusions and Future Work}
+
+Large bivariate data sets are very difficult to visualize in raw form, due to over-plotting of points.
+Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
+Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in spatial and frequency information.
+
+We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
+The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution.
+However, the standard binning algorithm is superior due to lower spatial loss.
+For data with a coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.
+We were also able to show through simulation that a reasonable default for the binning uses an origin offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well for heavily skewed data.
+
+Spatial binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
+It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure.
+%; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
+
+If we elect to use frequency binning to shade tiles ordinally, it is recommended to use between four and seven distinct shades.
+This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
+Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts.
+
+Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
+The findings of this research provide suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.
+
+%' 
+%' %--------------------------------------------------------------------
+%' 
+%' \subsection{Big Data: Airline Departure Times}
+%' The Federal Aviation Association (FAA) requires all airlines based in the United States to report   details for every single flight.
 These are published online by the Bureau of Transportation Services at \url{http://www.transtats.bts.gov/DataIndex.asp}.
-Every day there are about 16,000 flights across the United States which added up to over 5.8 million flights in 2014.
+%' Every day there are about 16,000 flights across the United States which added up to over 5.8 million flights in 2014.
 The scheduled and actual departure times for flights in 2014 make up --in uncompressed comma separated form-- a file of about 53 MB.
 A comparison of scheduled and actual departure times allows us an investigation of on-time performance of air carriers.
-
-Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
+%' 
+%' Figure~\ref{airline-scatter} shows two plots of the relationship between scheduled versus actual departure times.
 The plot on the left shows the traditional scatterplot; even while using alpha-blending this results in a severely over-plotted graph.
@@
 The 1-minute standard binning also reduces the data to around 190,000 reduced data triple, approximately 5\% of the original data size.
-
-<<FlightTradScatterplot, echo=F,include=F,eval=F, cache=TRUE>>=
-source("RectangularBinningFunctions.R")
-load(file="./data/DepartDelaysMinutes.RData")
-# Code takes a long long time to compile so ran once, saved to high res png, then set eval=FALSE
-ggplot(aes(x=Scheduled/60, y=Actual/60), data=DepartDelaysMinutes) +
-  geom_point(alpha=.05) + 
-  theme_bw() + theme(aspect.ratio=1,
-                     panel.grid= element_blank(),
-                     legend.text=element_text(size=12) )  +
-  scale_x_continuous("Scheduled Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
-  scale_y_continuous("Actual Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
-  ggtitle("Alpha-Blended Scatterplot of Airline Departure Times") 
-ggsave(filename = "./figure/AlphaScatterAirline2014.png", width= 6, height=6, dpi=300)
-@
-
-<<FlightMinBinScat, echo=F,include=F, cache=TRUE>>=
-source("./data/RectangularBinningFunctions.R")
-load(file="./data/DepartDelaysMinutes.RData")
-# Create minimal binning
-departure1minute <- RectBin2d(xs=DepartDelaysMinutes$Scheduled,ys=DepartDelaysMinutes$Actual, 
-                              originx = -0.5, originy = -0.5, 
-                              widthx = 1, widthy = 1, type="standard")
-# down to 3.261% of original number of rows
-# nrow(departure1minute)/nrow(DepartDelaysMinutes)
-ggplot(aes(x=binxs/60, y=binys/60), data=departure1minute) +
-  geom_tile( aes(fill=binfreq) ) + 
-  scale_x_continuous("Scheduled Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
-  scale_y_continuous("Actual Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
-  ggtitle("Minimally Binned Scatterplot of Airline Departure Times")+ 
-  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) +
-  theme_bw() + theme(aspect.ratio=1,
-                   panel.grid= element_blank(),
-                   legend.text=element_text(size=12) ) + 
-  geom_abline(intercept=0,slope=1, color="gray70", linetype=3)
-ggsave(filename = "./figure/MinimalBinnedAirline2014.png", width= 6.5, height=6, dpi=300)
-@
-
-\begin{figure}[hbtp]
-  \subfloat[Alpha-Blended Scatterplot of Airline Data]{\includegraphics[keepaspectratio=TRUE,width=.47\textwidth]{./figure/AlphaScatterAirline2014.png}} 
-%   Swap Back at later date (shrinks size of pdf to managable)
-%\subfloat[Million-point sample of Airline Data.
+%' 
+%' <<FlightTradScatterplot, echo=F,include=F,eval=F, cache=TRUE>>=
+%' source("RectangularBinningFunctions.R")
+%' load(file="./data/DepartDelaysMinutes.RData")
+%' # Code takes a long long time to compile so ran once, saved to high res png, then set eval=FALSE
+%' ggplot(aes(x=Scheduled/60, y=Actual/60), data=DepartDelaysMinutes) +
+%'   geom_point(alpha=.05) + 
+%'   theme_bw() + theme(aspect.ratio=1,
+%'                      panel.grid= element_blank(),
+%'                      legend.text=element_text(size=12) )  +
+%'   scale_x_continuous("Scheduled Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+%'   scale_y_continuous("Actual Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+%'   ggtitle("Alpha-Blended Scatterplot of Airline Departure Times") 
+%' ggsave(filename = "./figure/AlphaScatterAirline2014.png", width= 6, height=6, dpi=300)
+%' @
+%' 
+%' <<FlightMinBinScat, echo=F,include=F, cache=TRUE>>=
+%' source("./data/RectangularBinningFunctions.R")
+%' load(file="./data/DepartDelaysMinutes.RData")
+%' # Create minimal binning
+%' departure1minute <- RectBin2d(xs=DepartDelaysMinutes$Scheduled,ys=DepartDelaysMinutes$Actual, 
+%'                               originx = -0.5, originy = -0.5, 
+%'                               widthx = 1, widthy = 1, type="standard")
+%' # down to 3.261% of original number of rows
+%' # nrow(departure1minute)/nrow(DepartDelaysMinutes)
+%' ggplot(aes(x=binxs/60, y=binys/60), data=departure1minute) +
+%'   geom_tile( aes(fill=binfreq) ) + 
+%'   scale_x_continuous("Scheduled Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+%'   scale_y_continuous("Actual Departure Time", limits=c(0,24), breaks=seq(0,24,by=6)) +
+%'   ggtitle("Minimally Binned Scatterplot of Airline Departure Times")+ 
+%'   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) +
+%'   theme_bw() + theme(aspect.ratio=1,
+%'                    panel.grid= element_blank(),
+%'                    legend.text=element_text(size=12) ) + 
+%'   geom_abline(intercept=0,slope=1, color="gray70", linetype=3)
+%' ggsave(filename = "./figure/MinimalBinnedAirline2014.png", width= 6.5, height=6, dpi=300)
+%' @
+%' 
+%' \begin{figure}[hbtp]
+%'   \subfloat[Alpha-Blended Scatterplot of Airline Data]{\includegraphics[keepaspectratio=TRUE,width=.47\textwidth]{./figure/AlphaScatterAirline2014.png}} 
+%' %   Swap Back at later date (shrinks size of pdf to managable)
+%' %\subfloat[Million-point sample of Airline Data.
 \ktm{This plot is not compiling and needs to be reconstructed (my apologies)}]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
-\subfloat[Minimally Binned of Airline Data]{\includegraphics[width=.53\linewidth, keepaspectratio=TRUE]{./figure/MinimalBinnedAirline2014.png}}
-
-\caption[Alpha-blended and minimally binned scatterplots for scheduled and actual departure times of airline flights.]{\label{airline-scatter}Scheduled and actual departure times of flights across the United States in 2014.
+%' \subfloat[Minimally Binned of Airline Data]{\includegraphics[width=.53\linewidth, keepaspectratio=TRUE]{./figure/MinimalBinnedAirline2014.png}}
+%' 
+%' \caption[Alpha-blended and minimally binned scatterplots for scheduled and actual departure times of airline flights.]{\label{airline-scatter}Scheduled and actual departure times of flights across the United States in 2014.
 The large scale distributional patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}
-\end{figure}
-
-Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the identity line.
+%' \end{figure}
+%' 
+%' Both plots show the same large scale distributional patterns: scheduled and actual arrival times are highly correlated, recognizable from the conglomeration of points along the identity line.
 Scheduled departure times past 6:00 am in the morning are much more common than earlier flights.
@@
 The cloud of outliers halfway between the two main structures is potentially interesting, since no immediate explanation comes to mind, and would be worthy of a follow-up investigation.
-
-What is not apparent in the alpha-blended scatterplot, is some fine-level structure that the minimally binned scatterplot shows.
+%' 
+%' What is not apparent in the alpha-blended scatterplot, is some fine-level structure that the minimally binned scatterplot shows.
 Note that because we have bin widths equal to the resolution to which the data is recorded in each dimension, we know that the spatial binning algorithm will not cause any \textit{artificial} density stripes.
@@
 It is obvious that more flights are scheduled with departures on the hour and at 30 minutes past the hour.
-
-\begin{figure}[hbtp]
-\includegraphics[width=.49\linewidth,keepaspectratio=true]{./figure/AirlineStdBinning5mins.pdf}
-\includegraphics[width=.49\linewidth,keepaspectratio=true]{./figure/AirlineStdBinning15mins.pdf}
-\caption[Binned scatterplots for scheduled and actual departure times of airline flights using 5X5 and 15X15 minute bins.]{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
+%' 
+%' \begin{figure}[hbtp]
+%' \includegraphics[width=.49\linewidth,keepaspectratio=true]{./figure/AirlineStdBinning5mins.pdf}
+%' \includegraphics[width=.49\linewidth,keepaspectratio=true]{./figure/AirlineStdBinning15mins.pdf}
+%' \caption[Binned scatterplots for scheduled and actual departure times of airline flights using 5X5 and 15X15 minute bins.]{5-minute bins produce a higher-level summary of the data than shown in Figure~\ref{airline-scatter}b.
 15-minute bins produce an even more coarse summary of the data.}\label{airline5mins-binning}
-\end{figure}
-
-Figure~\ref{airline5mins-binning} displays the binned scatterplots for the flight departure data with larger bins.
+%' \end{figure}
+%' 
+%' Figure~\ref{airline5mins-binning} displays the binned scatterplots for the flight departure data with larger bins.
 Binning data by five-minute intervals produces a more high-level summary of the relationship between actual and scheduled departure time, though it necessarily obscures some of the finer details.
@@
 Binning by 15-minute intervals reduces the data set to a nearly-trivial 3,575 reduced binned data triples, but the graphical summary becomes granular and less appealing at that resolution.
-
-<<airlinesetup,echo=FALSE, eval=FALSE>>=
-setwd("./data/")
-if (!file.exists("airline.csv.gz")) {
-  # HH: I'm not sure, which files these should be ... 
-  # airline <- do.call("rbind", lapply(list.files(), function(i) read.csv(i)))
-
-  # write.csv(airline, "airline.csv", row.names=FALSE)
-}
-airline <- read.csv("airline.csv.gz")
-na.departure <- (is.na(airline[,8]) + is.na(airline[,9]))> 0
-airline <- airline[!na.departure,8:9]
-
-airline$CRS_DEP_TIME <- floor(airline$CRS_DEP_TIME/100)*60 + airline$CRS_DEP_TIME%%100
-airline$DEP_TIME <- floor(airline$DEP_TIME/100)*60 + airline$DEP_TIME%%100
-
-library(lubridate)
-convert.time <- function(x){
-  floor(x/60) + (x%%60)/60
-}
-
-library(multicore)
-
-if (!file.exists("ReducedAirlineData.csv")) {
-  airline.freq <- ddply(airline, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=nrow(i)))
-  write.csv(airline.freq, "ReducedAirlineData.csv", row.names=FALSE)
-}
-airline.freq <- read.csv("ReducedAirlineData.csv")
-airline.freq.plot <- airline.freq
-airline.freq.plot[,1] <- convert.time(airline.freq.plot[,1])
-airline.freq.plot[,2] <- convert.time(airline.freq.plot[,2])
-airline.freq.plot <- airline.freq.plot[order(airline.freq.plot$Freq),]
-airline.dummy <- airline.freq.plot[1,]
-airline.dummy[1,1] <- NA
-qplot(data=airline.freq.plot, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="blank", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (1 minute bins)") + geom_rect(aes(xmin=CRS_DEP_TIME-3/60, xmax=CRS_DEP_TIME+3/60, ymin=DEP_TIME-3/60, ymax=DEP_TIME+3/60, fill=Freq), alpha=I(.5)) + geom_rect(data=airline.dummy, aes(xmin=CRS_DEP_TIME-3/60, xmax=CRS_DEP_TIME+3/60, ymin=DEP_TIME-3/60, ymax=DEP_TIME+3/60, fill=Freq), aes.inherit=FALSE) + theme_bw()+ theme(legend.position="right") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
-ggsave("AirlineStdBinning1min.pdf", width=7.5, height=6, dpi=2304)
-
-library(ggplot2)
-library(dbData)
-air.sample <- airline.freq.plot[sample(1:nrow(airline.freq.plot), 1000000, replace=TRUE, prob=airline.freq.plot$Freq),1:2]
-
-qplot(data=air.sample, x=CRS_DEP_TIME, y=DEP_TIME, geom="point", alpha=I(.05), xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Sampled Airline Departure Times") + theme_bw()+ theme(legend.position="right") + scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
-ggsave("MillionSample.pdf", width=6, height=6, dpi=576)
-
-if (!file.exists("airlineBin5Std.csv")) {
-  air.binned5 <- binStd(airline.freq, c(5, 5))
-  air.binned5 <- ddply(air.binned5, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(convert.time(unique(i[,1:2])), Freq=sum(i$Freq)))
-  write.csv(air.binned5, "airlineBin5Std.csv", row.names=FALSE)
-}
-# air.binned5.rdm <- binRdm(cbind(airline, Freq=1), c(5, 5))
-# air.binned5.rdm2 <- ddply(air.binned5.rdm, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(convert.time(unique(i[,1:2]), Freq=sum(i$Freq)))
-# write.csv(air.binned5.rdm2, "airlineBin5Rdm.csv", row.names=FALSE)
-
-air.binned5 <- read.csv("airlineBin5Std.csv")
-qplot(data=air.binned5, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (5 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
-ggsave("AirlineStdBinning5mins.pdf", width=6, height=6, dpi=576)
-# 
-# qplot(data=air.binned5.rdm2, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (5 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B4FB", high="#183347", guide="legend", trans="log")
-# ggsave("AirlineRdmBinning5mins.png", binned, width=6, height=6)
-
-# air.binned <- data.frame(apply(airline, 2, function(i) 15*round(i/15, 0)))
-# air.binned.rdm <- binRdm(airline.freq, c(15, 15))
-# air.binned2 <- ddply(air.binned, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=sum(i$Freq)))
-# write.csv(air.binned2, "airlineBin15.csv", row.names=FALSE)
-air.binned <- read.csv("airlineBin15.csv")
-air.binned2 <- air.binned
-air.binned2[,1] <- convert.time(air.binned[,1])
-air.binned2[,2] <- convert.time(air.binned[,2])
-air.binned.rdm <- read.csv("airlineBin15Rdm.csv")
-air.binned2.rdm <- air.binned.rdm
-air.binned2.rdm[,1] <- convert.time(air.binned.rdm[,1])
-air.binned2.rdm[,2] <- convert.time(air.binned.rdm[,2])
-
-# air.binned2.rdm <- ddply(air.binned.rdm, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=sum(i$Freq)))
-# write.csv(air.binned2.rdm, "airlineBin15Rdm.csv", row.names=FALSE)
-
-binned <- qplot(data=air.binned2, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
-ggsave("AirlineStdBinning15mins.pdf", binned, width=6, height=6, dpi=576)
-
-# binnedRdm <- qplot(data=air.binned2.rdm, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
-# ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=300)
-@
-
-%--------------------------------------------------------------------------------------
-
-\section{Conclusions and Future Work}
-
-Large bivariate data sets are very difficult to visualize in raw form, due to over-plotting of points.
-Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
-Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in spatial and frequency information.
-
-We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
-The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution.
-However, the standard binning algorithm is superior due to lower spatial loss.
-For data with a coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.
-We were also able to show through simulation that a reasonable default for the binning uses an origin offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well even for spatially skewed data.
-
-Spatial binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure.
-It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure.
-%; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
-
-If we elect to use frequency binning to shade tiles ordinally, it is recommended to use between four and seven distinct shades.
-This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
-Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts.
-
-Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
-The findings of this research provide suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.
+%' 
+%' <<airlinesetup,echo=FALSE, eval=FALSE>>=
+%' setwd("./data/")
+%' if (!file.exists("airline.csv.gz")) {
+%'   # HH: I'm not sure, which files these should be ... 
+%'   # airline <- do.call("rbind", lapply(list.files(), function(i) read.csv(i)))
+%'   
+%'   # write.csv(airline, "airline.csv", row.names=FALSE)
+%' }
+%' airline <- read.csv("airline.csv.gz")
+%' na.departure <- (is.na(airline[,8]) + is.na(airline[,9]))> 0
+%' airline <- airline[!na.departure,8:9]
+%' 
+%' airline$CRS_DEP_TIME <- floor(airline$CRS_DEP_TIME/100)*60 + airline$CRS_DEP_TIME%%100
+%' airline$DEP_TIME <- floor(airline$DEP_TIME/100)*60 + airline$DEP_TIME%%100
+%' 
+%' library(lubridate)
+%' convert.time <- function(x){
+%'   floor(x/60) + (x%%60)/60
+%' }
+%' 
+%' library(multicore)
+%' 
+%' if (!file.exists("ReducedAirlineData.csv")) {
+%'   airline.freq <- ddply(airline, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=nrow(i)))
+%'   write.csv(airline.freq, "ReducedAirlineData.csv", row.names=FALSE)
+%' }
+%' airline.freq <- read.csv("ReducedAirlineData.csv")
+%' airline.freq.plot <- airline.freq
+%' airline.freq.plot[,1] <- convert.time(airline.freq.plot[,1])
+%' airline.freq.plot[,2] <- convert.time(airline.freq.plot[,2])
+%' airline.freq.plot <- airline.freq.plot[order(airline.freq.plot$Freq),]
+%' airline.dummy <- airline.freq.plot[1,]
+%' airline.dummy[1,1] <- NA
+%' qplot(data=airline.freq.plot, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="blank", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (1 minute bins)") + geom_rect(aes(xmin=CRS_DEP_TIME-3/60, xmax=CRS_DEP_TIME+3/60, ymin=DEP_TIME-3/60, ymax=DEP_TIME+3/60, fill=Freq), alpha=I(.5)) + geom_rect(data=airline.dummy, aes(xmin=CRS_DEP_TIME-3/60, xmax=CRS_DEP_TIME+3/60, ymin=DEP_TIME-3/60, ymax=DEP_TIME+3/60, fill=Freq), aes.inherit=FALSE) + theme_bw()+ theme(legend.position="right") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
+%' ggsave("AirlineStdBinning1min.pdf", width=7.5, height=6, dpi=2304)
+%'     
+%' 
+%' library(ggplot2)
+%' library(dbData)
+%' air.sample <- airline.freq.plot[sample(1:nrow(airline.freq.plot), 1000000, replace=TRUE, prob=airline.freq.plot$Freq),1:2]
+%' 
+%' qplot(data=air.sample, x=CRS_DEP_TIME, y=DEP_TIME, geom="point", alpha=I(.05), xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Sampled Airline Departure Times") + theme_bw()+ theme(legend.position="right") + scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
+%' ggsave("MillionSample.pdf", width=6, height=6, dpi=576)
+%' 
+%' if (!file.exists("airlineBin5Std.csv")) {
+%'   air.binned5 <- binStd(airline.freq, c(5, 5))
+%'   air.binned5 <- ddply(air.binned5, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(convert.time(unique(i[,1:2])), Freq=sum(i$Freq)))
+%'   write.csv(air.binned5, "airlineBin5Std.csv", row.names=FALSE)
+%' }
+%' # air.binned5.rdm <- binRdm(cbind(airline, Freq=1), c(5, 5))
+%' # air.binned5.rdm2 <- ddply(air.binned5.rdm, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(convert.time(unique(i[,1:2]), Freq=sum(i$Freq)))
+%' # write.csv(air.binned5.rdm2, "airlineBin5Rdm.csv", row.names=FALSE)
+%' 
+%' air.binned5 <- read.csv("airlineBin5Std.csv")
+%' qplot(data=air.binned5, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (5 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
+%' ggsave("AirlineStdBinning5mins.pdf", width=6, height=6, dpi=576)
+%' # 
+%' # qplot(data=air.binned5.rdm2, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (5 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B4FB", high="#183347", guide="legend", trans="log")
+%' # ggsave("AirlineRdmBinning5mins.png", binned, width=6, height=6)
+%' 
+%' # air.binned <- data.frame(apply(airline, 2, function(i) 15*round(i/15, 0)))
+%' # air.binned.rdm <- binRdm(airline.freq, c(15, 15))
+%' # air.binned2 <- ddply(air.binned, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=sum(i$Freq)))
+%' # write.csv(air.binned2, "airlineBin15.csv", row.names=FALSE)
+%' air.binned <- read.csv("airlineBin15.csv")
+%' air.binned2 <- air.binned
+%' air.binned2[,1] <- convert.time(air.binned[,1])
+%' air.binned2[,2] <- convert.time(air.binned[,2])
+%' air.binned.rdm <- read.csv("airlineBin15Rdm.csv")
+%' air.binned2.rdm <- air.binned.rdm
+%' air.binned2.rdm[,1] <- convert.time(air.binned.rdm[,1])
+%' air.binned2.rdm[,2] <- convert.time(air.binned.rdm[,2])
+%' 
+%' # air.binned2.rdm <- ddply(air.binned.rdm, .(CRS_DEP_TIME, DEP_TIME), function(i) cbind(unique(i), Freq=sum(i$Freq)))
+%' # write.csv(air.binned2.rdm, "airlineBin15Rdm.csv", row.names=FALSE)
+%' 
+%' binned <- qplot(data=air.binned2, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
+%' ggsave("AirlineStdBinning15mins.pdf", binned, width=6, height=6, dpi=576)
+%' 
+%' # binnedRdm <- qplot(data=air.binned2.rdm, x=CRS_DEP_TIME, y=DEP_TIME, fill=Freq, asp=1, geom="tile", xlab="Scheduled Departure Time", ylab="Actual Departure Time", main="Airline Departure Times (15 minute bins)") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradientn(colours=c("#56B1F7", "#132B43"), guide="legend", trans="log", breaks=c(1, 10, 100, 1000, 10000))+ scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) + scale_y_continuous(breaks=c(0, 6, 12, 18, 24))
+%' # ggsave("AirlineRdmBinning15mins.png", binnedRdm, width=8, height=8, dpi=300)
+%' @
+%' 
+%' %--------------------------------------------------------------------------------------
 

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
little edit to wording

@@
 
-We now revisit the baseball data used earlier and construct binned scatterplots using specification principles developed based on loss properties from the previous sections.
+We now revisit the baseball data used earlier and construct binned scatterplots using the loss-based specification principles developed in the previous sections.
 We begin by specifying the bin dimensions and origin for standard rectangular binning.

### 2015-12-18 versionForJCGS/PaperJCGS.Rnw
minor edits to allow compile on laptop I will be bringing to MN for break

@@
 %
-<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning.">>=
+<<SpatialLossCompare,echo=F,warning=FALSE,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.9\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss and computation times over a range of bin sizes for standard and random binning.">>=
 ## create binned data for standard and random using variety of bin sizes (from 2 to 20, by 2)
@@
 # find freq loss from standard, quantile, logCount
-detach("package:dplyr", unload=TRUE)
 source("BinningLossFunctions.R")

### 2016-01-07 versionForJCGS/PaperJCGS.Rnw
Blind For Submission
Spell Check

@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
+\newcommand{\blind}{1}
 \newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
@@
 {
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf Binning Strategies and Related Loss for Large Data}
-\end{center}
-  \medskip
+\title{Binning Strategies and Related Loss for Large Data}
+\author{Author A$^1$, Author B$^{2,3}$, Author C$^2$\\$^1$Affiliation X \\$^2$Affiliation Y \\$^3$Affiliation Z}
+  \maketitle
 } \fi
@@
 
-The top row of Figure~\ref{fig:scatter-alpha} demonstrates several varients of the traditional scatterplot where each observation is plotted using distinct points.
+The top row of Figure~\ref{fig:scatter-alpha} demonstrates several variants of the traditional scatterplot where each observation is plotted using distinct points.
 The traditional, solid point, scatterplot in Figure~\ref{fig:scatter-alpha}(a) shows a triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable.
@@
 An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}.
-Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points.
+Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and re-size the rows of cells according to density of points.
 This variable binning fragments the continuity of the axes into segments on different scales and also does not deal with points at identical coordinates.
@@
 A {\it sunflower plot} uses symbols that increase in complexity proportional to the number of points in each bin.
-Due to difficulty in rendering descernable shapes in limited space, sunflower plots are only useful when the number of points in each bin remains reasonably small.
+Due to difficulty in rendering discernible shapes in limited space, sunflower plots are only useful when the number of points in each bin remains reasonably small.
 A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.
@@
 Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins.
-It is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies due to the improved handling of skewed bin count distibutions while maintaining similar frequency loss to standard frequency binning when four to seven frequency bins are used, as seen in Figure~\ref{fig:FreqLossCompare}.
+It is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies due to the improved handling of skewed bin count distributions while maintaining similar frequency loss to standard frequency binning when four to seven frequency bins are used, as seen in Figure~\ref{fig:FreqLossCompare}.
 Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.
@@
 \begin{appendix}
-\section{Appendix for Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}  
+\section{Appendix for Origin Offset Proof}  
 %----------------------------------------------------------------------------
@@
 $ \Rightarrow -2\theta = -\alpha $ \\
-$ \Rightarrow \theta = \alpha/2 $ \\
+$ \Rightarrow \theta = \alpha/2 $
 

### 2016-01-07 versionForJCGS/PaperJCGS.Rnw
changes for submission

@@
 %opening
-\title{Binning Strategies and Related Loss for Large Data}
+\title{Binning Strategies and Related Loss for Binned Scatterplots of Large Data}
 \author{Karsten Maurer$^1$, Heike Hofmann$^{2,3}$, Susan VanderPlas$^2$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
@@
 {
-\title{Binning Strategies and Related Loss for Large Data}
+\title{Binning Strategies and Related Loss for Binned Scatterplots of Large Data}
 \author{Author A$^1$, Author B$^{2,3}$, Author C$^2$\\$^1$Affiliation X \\$^2$Affiliation Y \\$^3$Affiliation Z}
@@
 \noindent%
-{\it Keywords:}  scatterplots, visual loss, aggregation, graphics
+{\it Keywords:}  binned scatterplots, visual loss, aggregation, graphics
 \vfill
@@
   geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor)) + 
+  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor, linetype=linecolor)) + 
   scale_colour_manual(values = c("darkblue","red"))+
@@
   geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
-  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor)) + 
+  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor, linetype=linecolor)) + 
   scale_colour_manual(values = c("darkblue","red"))+
@@
 The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}.
-We see that in random binning there are pairs of points -- like those with line segments highlighted in red -- that have been allocated in such a way that they are closer to their partner's bin center than their own.
+We see that in random binning there are pairs of points -- like those with dashed line segments -- that have been allocated in such a way that they are closer to their partner's bin center than their own.
 We see that the binned scatterplot remains identical if these points are swapped back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed.

### 2016-06-08 versionForJCGS/PaperJCGS.Rnw
add space to check new setup working
Gather and alphabetize library calls

@@
 <<setup4,echo=F,include=F,eval=T>>=
+
+library(dbData)
+library(devtools)
 library(ggplot2)
 library(gridExtra)
-library(dbData)
+library(hexbin)
+library(lubridate)
+library(MASS)
+library(plyr)
+library(RColorBrewer)
 library(reshape2)
-#library(multicore)
-library(devtools)
+library(scales)
+library(tidyr)
 library(xtable)
-library(RColorBrewer)
-library(hexbin)
-library(MASS)
-library(scales)
-library(plyr)
+
 #source in binning functions
@@
 
-library(scales)
 #match color scales in each
@@
 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=
-library(scales) 
+
 ### Try 2d binning with MVN data
@@
 
-library(tidyr)
 spatlossdatSquareBins2 <- gather(spatlossdatSquareBins, measure, value, c(totalSpatialLoss,binTime))
@@
 
-%' 
+%' THE FOLLOWING SUBSECTION IS OMITTED FOR JCGS PAGE LIMIT CONSIDERATIONS
 %' %--------------------------------------------------------------------

### 2016-06-08 versionForJCGS/PaperJCGS.Rnw
cleanup support documentation for JCGS step 1.1

@@
 library(hexbin)
+library(Hmisc)
 library(lubridate)
@@
 <<DataReduction,echo=FALSE,eval=FALSE,results='asis'>>=
+  # Generate tables for raw, binned and reduced binned data
 n=12
@@
 
-require(ggplot2)
 qplot(x, y, colour=1000*Shade, size=I(5),shape=Shape) + theme_bw() + 
@@
        nrow=1000, ncol=4401)
-g <- rasterGrob(blueGradient, interpolate=TRUE)
+g <- grid::rasterGrob(blueGradient, interpolate=TRUE)
 

### 2016-10-01 versionForJCGS/PaperJCGS.Rnw
Push updates

@@
 
-library(dbData)
+# library(dbData)
 library(devtools)
@@
 
+# Load all data: pitching and binning loss demo
 load("./data/pitching.Rdata")
+load("./data/BinningLossDemos.Rdata")
 @
@@
 
-# Hex binned scatterplot
-  qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
-  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
+# # Hex binned scatterplot
+#   qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
+#   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+#   ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
 d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
@@
 # # write.csv(spatlossdat,"SpatialLossData.csv", row.names=FALSE)
-spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE)
+# spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE) 
+# NOTE: spatlossdatSquareBins saved to BinningLossDemos.Rdata which is read in chunk 1
 spatlossdatSquareBins$bintype <- factor(spatlossdatSquareBins$bintype, levels=c("standard","random"))
@@
 # write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
-
-allbindat <- read.csv("StripeBinData.csv",header=TRUE)
+# allbindat <- read.csv("StripeBinData.csv",header=TRUE)
+# NOTE: allbindat saved to BinningLossDemos.Rdata which is read in chunk 1
+
 allbindat$BinningType <- factor(allbindat$BinningType, levels=c("Standard","Random"))
@@
 # write.csv(spatlossdatRound,"LossesByOriginBinwidth.csv", row.names=FALSE)
-lossOrigin <- read.csv("./data/LossesByOriginBinwidth.csv",header=TRUE)
+# lossOrigin <- read.csv("./data/LossesByOriginBinwidth.csv",header=TRUE)
+# Save the loss binning data objects into central Rdata file for simplicity
+# save(lossOrigin, spatlossdatSquareBins, allbindat, file="BinningLossDemos.Rdata")
+
+# NOTE: lossOrigin saved to BinningLossDemos.Rdata which is read in chunk 1
+
 mins <- ddply(lossOrigin, .(widthx, bintype, distrib), summarise,
@@
 # find freq loss from standard, quantile, logCount
-source("BinningLossFunctions.R")
 freqLossSimDat <- NULL

### 2016-10-01 versionForJCGS/PaperJCGS.Rnw
Begin adding scagnostics discussion

@@
 library(xtable)
+library(scagnostics)
 
@@
 Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm.
-This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
+This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of distributional scenarios.
 Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning.
@@
 ### round all data to resolution alpha
-alpha <- 2
+alpha <- 5
 for(i in 1:3){
@@
 \end{figure}
+
+<<SimDataScagnostics,echo=F,include=F,eval=F>>= 
+# help(package="scagnostics")
+head(alldat)
+scags <- data.frame(0,0,0,0,0,0,0,0,0,0,0)
+counter <- 0
+for (i in c("Exponential","Normal","Uniform")){
+  for(j in c("Fine","Coarse")){
+    counter <- counter + 1
+    tmp <- filter(alldat, type==i, record=="Fine")
+    print(dim(tmp))
+    if(j=="Fine") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$x,tmp$y)),2))
+    if(j=="Coarse") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$roundx,tmp$roundy)),2))   
+    names(scags) <- c("type","record",names(scagnostics(tmp$x,tmp$y)))
+  }
+}
+head(scags)
+
+print(xtable(scags, digits=c(0, 0, 3,3,3,3,3,3,3,3,3,3)), include.rownames=FALSE)
+@
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Sat Oct 01 11:29:18 2016
+\begin{table}[ht]
+\ktm{
+\centering
+\begin{tabular}{lrrrrrrrr}
+  \hline
+Dist-Res & Outlie & Skewed & Clumpy & Sparse & Striate & Convex & Skinny & Stringy \\ 
+  \hline
+  Exp-Fine & 0.15 & 0.67 & 0.01 & 0.02 & 0.02 & 0.47 & 0.55 & 0.36 \\ 
+  Exp-Coarse & 0.09 & 0.65 & 0.04 & 0.04 & 0.18 & 0.53 & 0.35 & 0.30 \\ 
+  Norm-Fine & 0.07 & 0.55 & 0.00 & 0.02 & 0.01 & 0.60 & 0.31 & 0.39  \\ 
+  Norm-Coarse & 0.04 & 0.65 & 0.01 & 0.04 & 0.05 & 0.62 & 0.18 & 0.31 \\ 
+  Unif-Fine & 0.00 & 0.52 & 0.00 & 0.03 & 0.00  & 0.68 & 0.15 & 0.36 \\ 
+  Unif-Coarse & 0.00 & 0.62 & 0.00 & 0.04 & 0.01 & 0.69 & 0.11 & 0.29 \\ 
+   \hline
+\end{tabular}
+} % end ktm highlighting
+\normalsize
+\caption{\ktm{Scagnostics metrics for the six simulated datasets used in empirical exploration of spatial loss in binned scatterplots.}}
+\label{tab:scagnostics}
+\end{table}
+
 %

### 2016-10-01 AuthorResponseToReviewers.Rmd
Add document for review responses, edits to paper with new references and content for scagnostics.

---
title: "Authors Response to Reviewers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## *Binning Strategies and Related Loss for Binned Scatterplots of Large Data*

The following document strives to address comments, questions and concerns made by the reviewers on a point-by-point basis. The reviewer statement are included verbatim through *italic text blocks*; our responses are interspersed to address each point more directly. In the revised article that accompanies, the edits are made in red text to make changes more easily identifiable. The revisions will be reverted to black after the revision process.

### Responses to Reviewer 1 Comments/Questions

*The authors discuss different binning strategies and the resulting aggregation- induced errors. Both errors, loss of spatial resolution and loss of frequency information are introduced and formalized. Binning is an important preprocessing step and research in this area is highly relevant. However, there are some major points missing, making this work less convincing.*

*The most crucial point is in my opinion the missing discussion of the influence of different data distributions. In the visualization community exists research on different descriptors for two-dimensional data called Scagnostics. When I started reading this article, I expected a thorough discussion of different data distributions/properties and the influence on different binning strategies and error outcomes.*

- We are very grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots. Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution. You will find that a discussion of the relevance of scagnostics has been added immediately preceeding the start of subsection 5.1. It is interesting that the striation score is so small, given that the data resolution issue can lead to artificially induced stripping is seen in so prominently in Figure 6.

*In Section 4.2, a discussion on contrast effects hindering the perception of the encoded values is missing. Nearby colored points, will change the color impression. These effects can be modelled and estimated and maybe interesting to investigate a little further.*

*As a computer scientist, it is not clear to me, why the runtime is larger with lower numbers of bins (Section 5.1. lines 33ff.)  In my naïve implementation, I would iterate once over all data points and assign the respective bins by an adopted integer division. Consequently, the runtime would be in O(n d) with n being the number of data points and d the number of dimensions. Maybe, the authors could explain the runtime experiments more or discuss the used algorithm more. If the number of bins is 1, I would expect the fastest runtime.*

*In the application example, why are 50 bins chosen and why is standard rectangular binning chosen? How would other binning results differ from the presented binning. The different variants in Figure 11 are only different color mappings of density. I would include here a discussion of data properties (Scagnostics) together with some error measurements described by the authors beforehand.*

*Minor remarks:Add color legend to Figure 7*

### Responses to Reviewer 2 Comments/Questions

*This paper reviews visualization methods for scatterplots and their strengths and weaknesses in handling large datasets. Section 2 notes that binned aggregation is the most widely used and effective way to visualize big bivariate data. In Section 3, the paper introduces three types of binning algorithms: standard rectangular, random and quantile. The main idea in this paper is presented in Section 4, where loss of information is characterized using two metrics: spatial loss and frequency loss. Spatial loss accounts for the loss of information about each individual point’s location; and frequency loss addresses the limitation of human perception in distinguishing frequencies represented using hues or saturation. Section 5 then presents some empirical results on how various factors may affect spatial loss and frequency loss based on three synthetic datasets. These factors include bin width and bin origin for spatial loss, and number of bins for frequency loss.*

*In general, I think this paper addresses an important topic. Guidelines on choosing binning algorithms will be immensely helpful for practitioners dealing with big data. I am, however, reserved in recommending acceptance because of two main concerns: lack of novelty and unconvincing results.*

*Lack of Novelty:*
*Section 2 and 3 do a good job in reviewing the concepts of binned aggregation and varieties of binning algorithms. However these are fairly well known and the discussion does not seem to contribute novel ideas. The only thing that I haven’t seen before is the random binning approach. Is this a new idea proposed by you? If so, please state clearly. I also wonder how and why random binning can be useful. Please provide scenarios in which it would make sense to use random binning.*

*For the other ideas covered in these two sections, please properly cite relevant work.*

*Unconvincing Results:*
*In Section 4.1, you propose the idea of spatial loss as a metric to account for the loss of information. Please explain how this approach is related to Scott’s approach of using mean integrated square error. What makes your definition a better choice?*

*I do not understand how binned frequencies (as compared to true frequencies) are calculated in Section 4.2 on frequency loss. Please provide an example. Also please clarify is binned frequencies are measures of frequencies perceived by a human.*

*The first paragraph of Section 5.1 shows that spatial loss increases as bin size increases, this seems hardly surprising. In the second paragraph, you talk about three properties of visual estimator: unbiasedness, consistency and efficiency, I think these properties are interesting, but the rest of the paper does not address them sufficiently.*

*I also fail to understand how the discussion on artificial striping in Section 5.1 is relevant for spatial loss.*

*Please provide quantified results on spatial loss when you talk about how binning origin affects spatial loss in Figure 7.*

*In visualization practices, it is common to use both ordinal scales as well as continuous scales when mapping numbers to colors. How might you address the issue of frequency loss when continuous scales are used?*

*I don’t buy the advice to “use between four and seven frequency”. There simply isn’t empirical evidence in the paper to support this claim.*

*From a human-computer interaction perspective, it is also limiting when the paper talks about spatial loss and frequency loss without any mention of user tasks (e.g. comparison, outlier detection). Do these metric matter for all the tasks involved in reading the scatterplots?*

### 2016-10-01 AuthorResponseToReviewers.Rmd
Figure 3 cleanup and response

@@
 If the number of bins is 1, I would expect the fastest runtime.*
+
+- Your intuition is correct, but there was a misunderstanding about figure 3.
+The x-axis relates the the *size* of bins not the *number* of bins, meaning that the binning takes less time as the bin widths grows (ie the number of bins decreases).
+x=1 means using many 1X1 bins, not one large bin.
+The x-axis label has been edited to more clearly reflect that this is the size of the bin dimensions.
 

### 2016-10-01 versionForJCGS/PaperJCGS.Rnw
Figure 3 cleanup and response

--- 2016-10-01 versionForJCGS/PaperJCGS.Rnw
@@
 Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
-These distributions were selected for their variety in terms of shape, center and spread.
 100,000 observation pairs were simulated for each data set from the following distributions:
@@
 Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
+
 %
@@
 The \textit{data resolution} is defined as the smallest increment between successive data values.
-To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension.
+To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
 By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.
+
+\ktm{These bivariate distributions were selected for the variety in their defining characteristics.
+\textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness and monotonicity (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}).
+The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}.
+They show that these examples explore a variety of skewnesses, convexity and skinniness.
+This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples.
+The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution.
+These minor striations will still prove problematic when binning is inconsistent with the data resolution.
+}
+
+<<SimDataScagnostics,echo=F,include=F,eval=F>>= 
+# help(package="scagnostics")
+head(alldat)
+scags <- data.frame(0,0,0,0,0,0,0,0,0,0,0)
+counter <- 0
+for (i in c("Exponential","Normal","Uniform")){
+  for(j in c("Fine","Coarse")){
+    counter <- counter + 1
+    tmp <- filter(alldat, type==i, record=="Fine")
+    print(dim(tmp))
+    if(j=="Fine") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$x,tmp$y)),2))
+    if(j=="Coarse") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$roundx,tmp$roundy)),2))   
+    names(scags) <- c("type","record",names(scagnostics(tmp$x,tmp$y)))
+  }
+}
+head(scags)
+
+print(xtable(scags, digits=c(0, 0, 3,3,3,3,3,3,3,3,3,3,3)), include.rownames=FALSE)
+@
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Sat Oct 01 11:29:18 2016
+\begin{table}[ht]
+\ktm{
+\centering \small
+\begin{tabular}{lrrrrrrrrr}
+  \hline
+Dist-Res & Outlie & Skew & Clump & Sparse & Striate & Convex & Skinny & String & Mono. \\ 
+  \hline
+  Exp-Fine &   0.15 & 0.67 & 0.01 & 0.02 & 0.02 & 0.47 & 0.55 & 0.36 & 0.00 \\ 
+  Exp-Coarse &  0.15 & 0.71 & 0.02 & 0.02 & 0.03 & 0.51 & 0.52 & 0.27  & 0.00  \\ 
+  Norm-Fine &   0.07 & 0.55 & 0.00 & 0.02 & 0.01 & 0.60 & 0.31 & 0.39 & 0.00  \\ 
+  Norm-Coarse &   0.08 & 0.59 & 0.01 & 0.02 & 0.02 & 0.60 & 0.33 & 0.35  & 0.00  \\ 
+  Unif-Fine &   0.00 & 0.52 & 0.00 & 0.03 & 0.00 & 0.68 & 0.15 & 0.36 & 0.00  \\ 
+  Unif-Coarse & 0.00 & 0.56 & 0.00 & 0.03 & 0.01 & 0.68 & 0.14 & 0.38  & 0.00  \\ 
+   \hline
+\end{tabular}
+} % end ktm highlighting
+\normalsize
+\caption{\ktm{Scagnostics scores for the six simulated datasets used in empirical exploration of spatial loss in binned scatterplots.}}
+\label{tab:scagnostics}
+\end{table}
 
@@
 spatlossdatSquareBins2 <- gather(spatlossdatSquareBins, measure, value, c(totalSpatialLoss,binTime))
-spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Net Spatial Loss","Binning Time (sec)"))
+spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Binning Time (sec)","Net Spatial Loss"))
 # 
@@
   ylab("Net Spatial Loss") +
-  xlab("Width/Height of Square Bins") 
+  xlab("Size of Bin Dimensions") 
 @
@@
                              levels=c("2X2","4X4","6X6","8X8","10X10")) 
-mins$binsize <- paste(mins$widthx,"X",mins$widthx,sep="")
+mins$binsize <- factor(paste(mins$widthx,"X",mins$widthx,sep=""),
+                       levels=c("2X2","4X4","6X6","8X8","10X10"))
 

### 2016-10-01 versionForJCGS/PaperJCGS.Rnw
Added color legend to figure 7 as per review request

@@
 binoutExp5offset <- RectBin2d(datlist[[3]]$x,datlist[[3]]$y,-9,-9,10,10,type="standard")
-
-p1 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5[[1]]) +  
+options(scipen=7)
+p1 <- qplot(binxs, binys, geom="tile", fill=log10(binfreq), data=binoutExp5[[1]]) +  
 #  ggtitle("Origin at (0,0)") +
-  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+scale_fill_gradient("", low="#56B1F7", high="#132B43", 
+                    limits=c(0,5),breaks=0:5, labels=10^(0:5)) + 
   theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
-                     legend.position="none",
+                     # legend.position="none",
                      aspect.ratio=1, panel.grid= element_blank()) +
@@
 
-p2 <- qplot(binxs, binys, geom="tile", fill=log(binfreq), data=binoutExp5offset[[1]]) +  
+p2 <- qplot(binxs, binys, geom="tile", fill=log10(binfreq), data=binoutExp5offset[[1]]) +
 #  ggtitle("Origin at (-9,-9)") + 
-  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
+scale_fill_gradient("", low="#56B1F7", high="#132B43", 
+                    limits=c(0,5),breaks=0:5, labels=10^(0:5)) + 
   theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
-                     legend.position="none",
+                     # legend.position="none",
                      aspect.ratio=1, panel.grid= element_blank()) +

### 2016-10-01 versionForJCGS/PaperJCGS.Rnw
Revision to baseball example spatial binning justification. Still need to clean up freq binning justification.

@@
 
+### Remaining code for comparison will be moot for minimal binning since loss after offset = 0
+
 # how much better is loss under origin at recommended offset?
@@
 standSO <- (pitchdat$SO - mean(pitchdat$SO))/sd(pitchdat$SO)
-binoutSimpOrigin <- RectBin2d(standG,standSO,min(standG),min(standSO),2/mean(pitchdat$G),10/mean(pitchdat$SO),type="standard")
-binoutRecOrigin <- RectBin2d(standG,standSO,min(standG)-.5/mean(pitchdat$G),min(standSO)-.5/mean(pitchdat$SO),2/mean(pitchdat$G),10/mean(pitchdat$SO),type="standard")
-binoutRecOrigin[[2]]$totalSpatialLoss / binoutSimpOrigin[[2]]$totalSpatialLoss
-# spatial loss down 3% 
+binoutSimpOrigin <- RectBin2d(standG,standSO,min(standG),min(standSO),2/sd(pitchdat$G),10/sd(pitchdat$SO),type="standard")
+binoutRecOrigin <- RectBin2d(standG,standSO,min(standG)-.5/sd(pitchdat$G),min(standSO)-.5/sd(pitchdat$SO),2/sd(pitchdat$G),10/sd(pitchdat$SO),type="standard")
+binoutRecOrigin[[2]]$totalSpatialLoss / binoutSimpOrigin[[2]]$totalSpatialLoss # spatial loss down 9%
+
+# # how about loss over minimum binning?
+# binoutMinBin <- RectBin2d(standG,standSO,min(standG)-.5/sd(pitchdat$G),min(standSO)-.5/sd(pitchdat$SO),1/sd(pitchdat$G),1/sd(pitchdat$SO),type="standard")
+# binoutMinBin[[2]]$totalSpatialLoss / binoutRecOrigin[[2]]$totalSpatialLoss #
+
 standBaseballLoss2X10 <- binoutRecOrigin[[2]]$totalSpatialLoss
@@
 We now revisit the baseball data used earlier and construct binned scatterplots using the loss-based specification principles developed in the previous sections.
-We begin by specifying the bin dimensions and origin for standard rectangular binning.
-We use integer bin widths in each dimension of spatial bins to avoid artificial density stripes because each variable is recorded as an integer, corresponding to a one game by one strike-out data resolution.
-We desire relatively fine binning of approximately 50 bins in each dimension; the data range from 1 to 106 games and from 0 to 513 strike-outs, thus we set bin dimensions to two games by ten strikeouts.
+\ktm{In the spatial binning step we learned three rules of thumb: (1) use smaller bins to achieve lower loss of spatial information, (2) standard binning has less loss than random binning but we need to be careful to choose bin dimensions that align with data resolution, and (3) offset the origin by a factor of half the data resolution to better center point in bins.}
+
+\ktm{For the baseball data this leads us to specify small standard rectangular bins for data that is recorded with a one game by one strike-out data resolution.
+The data has only 42,583 observations, so we can use minimal binning to match the data resolution without taking an inordinate amount of computation time.
+While minimal binning eliminates spatial loss, the miniscule physical size of the resulting tiles makes it difficult to see the frequency shade; Therefore slightly larger two game by ten strikout bins are used, providing approximately 50 bins per dimension.
 The origin offset of $(0.5,-0.5)$ is used to shift the binning back by a half game and down by one half strikeout to adjust for the data resolution.
-%The data has only 42,583 observations, so we can bin on a fine grid without taking an inordinate amount of computation time.
-We find that we have approximately 3\% lower net spatial loss, after standardizing measurement units, for our binning specification using the recommended origin offset than if we were to naively set the binning origin at the minimum data values of one game and zero strikeouts.
+This offset for reduces the spatial loss by approximately 8\% for this bin size.}
 

### 2016-10-04 versionForJCGS/PaperJCGS.Rnw
minor changes

--- 2016-10-01 versionForJCGS/PaperJCGS.Rnw
@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{1}
+\newcommand{\blind}{0}
 \newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
@@
 <<setup4,echo=F,include=F,eval=T>>=
-
-# library(dbData)
-library(devtools)
 library(ggplot2)
 library(gridExtra)
+library(dbData)
+library(reshape2)
+#library(multicore)
+library(devtools)
+library(xtable)
+library(RColorBrewer)
 library(hexbin)
-library(Hmisc)
-library(lubridate)
 library(MASS)
+library(scales)
 library(plyr)
-library(RColorBrewer)
-library(reshape2)
-library(scales)
-library(tidyr)
-library(xtable)
-library(scagnostics)
-
 #source in binning functions
@@
 
-# Load all data: pitching and binning loss demo
 load("./data/pitching.Rdata")
-load("./data/BinningLossDemos.Rdata")
 @
@@
 
-# # Hex binned scatterplot
-#   qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
-#   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-#   ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
+# Hex binned scatterplot
+ggplot(data=pitchdat, aes(x=G, y=SO))+
+    geom_hex()+
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
 d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
@@
 <<DataReduction,echo=FALSE,eval=FALSE,results='asis'>>=
-  # Generate tables for raw, binned and reduced binned data
 n=12
@@
 
+library(scales)
 #match color scales in each
@@
 
+require(ggplot2)
 qplot(x, y, colour=1000*Shade, size=I(5),shape=Shape) + theme_bw() + 
@@
        nrow=1000, ncol=4401)
-g <- grid::rasterGrob(blueGradient, interpolate=TRUE)
+g <- rasterGrob(blueGradient, interpolate=TRUE)
 
@@
 Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm.
-This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of distributional scenarios.
+This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.
 Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning.
@@
 Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
+These distributions were selected for their variety in terms of shape, center and spread.
 100,000 observation pairs were simulated for each data set from the following distributions:
@@
 Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
-
 %
@@
 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=
-
+library(scales) 
 ### Try 2d binning with MVN data
@@
 The \textit{data resolution} is defined as the smallest increment between successive data values.
-To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension.
+To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
 By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.
-
-\ktm{These bivariate distributions were selected for the variety in their defining characteristics.
-\textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness and monotonicity (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}).
-The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}.
-They show that these examples explore a variety of skewnesses, convexity and skinniness.
-This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples.
-The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution.
-These minor striations will still prove problematic when binning is inconsistent with the data resolution.
-}
-
-<<SimDataScagnostics,echo=F,include=F,eval=F>>= 
-# help(package="scagnostics")
-head(alldat)
-scags <- data.frame(0,0,0,0,0,0,0,0,0,0,0)
-counter <- 0
-for (i in c("Exponential","Normal","Uniform")){
-  for(j in c("Fine","Coarse")){
-    counter <- counter + 1
-    tmp <- filter(alldat, type==i, record=="Fine")
-    print(dim(tmp))
-    if(j=="Fine") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$x,tmp$y)),2))
-    if(j=="Coarse") scags[counter,] <- c(i,j,round(as.vector(scagnostics(tmp$roundx,tmp$roundy)),2))   
-    names(scags) <- c("type","record",names(scagnostics(tmp$x,tmp$y)))
-  }
-}
-head(scags)
-
-print(xtable(scags, digits=c(0, 0, 3,3,3,3,3,3,3,3,3,3,3)), include.rownames=FALSE)
-@
-
-% latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Sat Oct 01 11:29:18 2016
-\begin{table}[ht]
-\ktm{
-\centering \small
-\begin{tabular}{lrrrrrrrrr}
-  \hline
-Dist-Res & Outlie & Skew & Clump & Sparse & Striate & Convex & Skinny & String & Mono. \\ 
-  \hline
-  Exp-Fine &   0.15 & 0.67 & 0.01 & 0.02 & 0.02 & 0.47 & 0.55 & 0.36 & 0.00 \\ 
-  Exp-Coarse &  0.15 & 0.71 & 0.02 & 0.02 & 0.03 & 0.51 & 0.52 & 0.27  & 0.00  \\ 
-  Norm-Fine &   0.07 & 0.55 & 0.00 & 0.02 & 0.01 & 0.60 & 0.31 & 0.39 & 0.00  \\ 
-  Norm-Coarse &   0.08 & 0.59 & 0.01 & 0.02 & 0.02 & 0.60 & 0.33 & 0.35  & 0.00  \\ 
-  Unif-Fine &   0.00 & 0.52 & 0.00 & 0.03 & 0.00 & 0.68 & 0.15 & 0.36 & 0.00  \\ 
-  Unif-Coarse & 0.00 & 0.56 & 0.00 & 0.03 & 0.01 & 0.68 & 0.14 & 0.38  & 0.00  \\ 
-   \hline
-\end{tabular}
-} % end ktm highlighting
-\normalsize
-\caption{\ktm{Scagnostics scores for the six simulated datasets used in empirical exploration of spatial loss in binned scatterplots.}}
-\label{tab:scagnostics}
-\end{table}
 
@@
 # # write.csv(spatlossdat,"SpatialLossData.csv", row.names=FALSE)
-# spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE) 
-# NOTE: spatlossdatSquareBins saved to BinningLossDemos.Rdata which is read in chunk 1
+spatlossdatSquareBins <- read.csv("SpatialLossData.csv",header=TRUE)
 spatlossdatSquareBins$bintype <- factor(spatlossdatSquareBins$bintype, levels=c("standard","random"))
 
+library(tidyr)
 spatlossdatSquareBins2 <- gather(spatlossdatSquareBins, measure, value, c(totalSpatialLoss,binTime))
-spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Binning Time (sec)","Net Spatial Loss"))
+spatlossdatSquareBins2$measure <- factor(spatlossdatSquareBins2$measure , labels=c("Net Spatial Loss","Binning Time (sec)"))
 # 
@@
   ylab("Net Spatial Loss") +
-  xlab("Size of Bin Dimensions") 
+  xlab("Width/Height of Square Bins") 
 @
@@
 # write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
-# allbindat <- read.csv("StripeBinData.csv",header=TRUE)
-# NOTE: allbindat saved to BinningLossDemos.Rdata which is read in chunk 1
-
+
+allbindat <- read.csv("StripeBinData.csv",header=TRUE)
 allbindat$BinningType <- factor(allbindat$BinningType, levels=c("Standard","Random"))
@@
 # write.csv(spatlossdatRound,"LossesByOriginBinwidth.csv", row.names=FALSE)
-# lossOrigin <- read.csv("./data/LossesByOriginBinwidth.csv",header=TRUE)
-# Save the loss binning data objects into central Rdata file for simplicity
-# save(lossOrigin, spatlossdatSquareBins, allbindat, file="BinningLossDemos.Rdata")
-
-# NOTE: lossOrigin saved to BinningLossDemos.Rdata which is read in chunk 1
-
+lossOrigin <- read.csv("./data/LossesByOriginBinwidth.csv",header=TRUE)
 mins <- ddply(lossOrigin, .(widthx, bintype, distrib), summarise,
@@
                              levels=c("2X2","4X4","6X6","8X8","10X10"))
-mins$binsize <- factor(paste(mins$widthx,"X",mins$widthx,sep=""),
-                       levels=c("2X2","4X4","6X6","8X8","10X10"))
+mins$binsize <- paste(mins$widthx,"X",mins$widthx,sep="")
 
@@
 # find freq loss from standard, quantile, logCount
+source("BinningLossFunctions.R")
 freqLossSimDat <- NULL
@@
 
-%' THE FOLLOWING SUBSECTION IS OMITTED FOR JCGS PAGE LIMIT CONSIDERATIONS
+%' 
 %' %--------------------------------------------------------------------

### 2016-10-04 versionForJCGS/PaperJCGS.Rnw
edit on file conflict

--- 2016-10-01 versionForJCGS/PaperJCGS.Rnw
@@
 <<setup4,echo=F,include=F,eval=T>>=
-
 # library(dbData)
@@
 load("./data/pitching.Rdata")
-load("./data/BinningLossDemos.Rdata")
+# load("./data/BinningLossDemos.Rdata")
 @
@@
 
-# # Hex binned scatterplot
-#   qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
-#   scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-#   ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
-d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
-d1.hex <- data.frame(hcell2xy(d1.hex.obj))
-names(d1.hex) <- c("G", "SO")
-d1.hex$Freq <-  d1.hex.obj@count
-
-# hex binned bubbleplot
-qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq+1) + 
-  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
-ggsave("./figure/BinBubble.png", width=5, height=5, dpi=300)
+# Hex binned scatterplot
+ggplot(data=pitchdat, aes(x=G, y=SO))+
+    geom_hex()+
+  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
+  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=300)
 

### 2016-10-04 versionForJCGS/PaperJCGS.Rnw
Minor rewording in section 5.3 and digit label work on figure 10

@@
 
+options(scipen=5, digits=5)
 #source in binning functions
@@
 load("./data/pitching.Rdata")
-# load("./data/BinningLossDemos.Rdata")
+load("./data/BinningLossDemos.Rdata")
 @
@@
 # find freq loss from standard, quantile, logCount
+options(digits=1, scipen=1) 
 freqLossSimDat <- NULL
@@
 It is nearly impossible to distinguish the density structure across the remaining bins, representing better pitchers who played many games and earned many strikeouts.
-%In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts. 
-Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities.
+\ktm{In this case there are the two frequency binning approaches from Section~\ref{FreqBinningSpec} that we can employ to deal with this skewness of the distribution of bin counts.} Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities.
 Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins.

### 2016-12-03 AuthorResponseToReviewers.Rmd
Added color legend to figure 7 as per review request
Revision to baseball example spatial binning justification. Still need to clean up freq binning justification.
Updates to citations in sections 2 and 3, plus added random-binning motivation. Addressing JCGS reviewer 2 comment.

@@
 
-- Your intuition is correct, but there was a misunderstanding about figure 3.
+- Your intuition is correct, but there was a misunderstanding about Figure 5.
 The x-axis relates the the *size* of bins not the *number* of bins, meaning that the binning takes less time as the bin widths grows (ie the number of bins decreases).
@@
 
+- The beginning of section 5.3 has been rewritten to more explicitly tie the binning choices for the baseball example to the lessons learned in 5.1 and 5.2.
+
 *Minor remarks:Add color legend to Figure 7*
+
+- Added as per request.
 
@@
 
+- You are correct that many of the ideas in Sections 2 and 3 are not original, with the exception of the random binning approach, where the methodology for random-binned aggregation is original to my best knowledge.
+The goal of these sections is to bring many binning related concepts under a unified notation for the benefit of clarity in the sections to follow.
+For the random binning, a statement has been added to address the motivation for its introduction: deterministic one-sided interval bins lead to lopsided allocation of points on bin boundaries, whereas a random assignment is expected to split the points evenly across bins.
+For the unoriginal ideas, additional citations have been made to place the unifying notation better within the existing literature.
+
 *Unconvincing Results:*

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Updates to citations in sections 2 and 3, plus added random-binning motivation. Addressing JCGS reviewer 2 comment.

@@
 Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}.
-More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}.
+\ktm{Sturges followed with in early work in formalizing histograms specification \citep{sturges1926choice}.}  More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}.
 \citet{scott1992} extends to the properties of multivariate binning strategies.
@@
 We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{Carr1987} are also common.
-While hexagonal binning has been shown to have slightly better graphical properties \citep{scott1992}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.
+While hexagonal binning has been shown to have slightly better graphical properties \citep{scott1992}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions \ktm{\citep{Gold2006}}.
 We will however only consider binning in up to two dimensions, $X$ and $Y$.
@@
 For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.
-\textit{Rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
+\ktm{What we will refer to as} \textit{General rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
 Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
@@
 \textit{Standard rectangular binning} is a special cases of general rectangular binning that uses intervals of equal size for all bins; thus only the origin of the first bin, $\beta_0$, and binwidth, $\omega_X$, need to be specified.
-Standard rectangular binning is necessarily used in the construction of histograms;  the consistent binwidth makes the display of frequency proportional to density.
+Standard rectangular binning is necessarily used in the construction of histograms \ktm{\citep{Pearson1895}};  the consistent binwidth makes the display of frequency proportional to density .
 Fixed width binning procedures are also highly computationally efficient \citep{Wickham2013Bin}.
@@
 This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.
-The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.
+\ktm{The impetus for developing the random binning assignment was to soften the hard breaks associated with one-sided interval assignment for scenarios where many points fall directly on bin boundaries by allowing points to be divided stochastically.} The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.
 
@@
 \textit{Quantile binning} is another option that divides the range of the observations into bins each containing an equal number of points.
-The $j^{th}$ bin interval takes the form $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function.
+The $j^{th}$ bin interval takes the form $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function \ktm{\citep{hyndman1996sample}}.
 Note that this binning approach is \textit{not} desirable for spatially visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for binned scatterplots that employ a second stage of binning to create discrete shade scheme for displaying grouped bin frequencies, which will be discussed in Section~\ref{FreqLoss}.
@@
 
-The standard and random binning algorithms are easily extendable to higher dimension.
+The standard and random binning algorithms are easily extended to higher dimensions when binned orthogonally \ktm{\citep{Gold2006}}.
 For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions.

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Added a comparison of Scott's MISE loss metric to our spatial loss using euclidean distances, as per reviewer 2 comment.

@@
 The idea of loss from one-dimensional binning was explored by Scott using mean integrated squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}.
-He later extended this discussion to two-dimensional binning, where he compared the mean integrated squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
-We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section~\ref{GenBinning} above.
+He later extended this discussion to two-dimensional binning, where he compared the mean integrated squared error (MISE) loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
+\ktm{Scott's MISE assesses the difference between the empirical density and the binned density approximation, as integrated the squared deviation over $\mathbb{R}^2$.
+Whereas, we quantify binning losses using the spatial displacement of the points to bin centers using euclidean distances, which better reflects the visual information lost in the shift from the traditional to binned scatterplots.}
 

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Small follow up wording edit to Scott MISE description.

@@
 He later extended this discussion to two-dimensional binning, where he compared the mean integrated squared error (MISE) loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}.
-\ktm{Scott's MISE assesses the difference between the empirical density and the binned density approximation, as integrated the squared deviation over $\mathbb{R}^2$.
+\ktm{Scott's MISE assesses the difference between the empirical density and the binned density approximation, via integration of the squared deviation over $\mathbb{R}^2$.
 Whereas, we quantify binning losses using the spatial displacement of the points to bin centers using euclidean distances, which better reflects the visual information lost in the shift from the traditional to binned scatterplots.}

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Clarifying example for freq loss

@@
         panel.background=element_blank()
-        ) + scale_x_continuous("", breaks=c(0,250,500,750,1000), labels=c(0,250,500,750,1000)) + 
+        ) + scale_x_continuous("", breaks=c(0,200,400,600,800,1000), labels=c(0,200,400,600,800,1000)) + 
   ylab("Continuous") 
@@
         panel.background=element_blank())  + 
-  ylab("Discrete") + xlab("")
+  ylab("Discrete") + scale_x_continuous("", breaks=c(0,200,400,600,800,1000), labels=c(0,200,400,600,800,1000))
 
@@
 where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin.
-Note that this is effectively a sum of squared error from binning frequencies.
-While this numerical assessment of frequency loss does not exactly account for limitations in human perceptual ability, it does provide a more realistic model for the loss in perception that does occur.
+Note that this is effectively a sum of squared \ktm{deviations between true frequencies to the centers of frequency bins.
+For example, in the discretized color scale in Figure~\ref{fig:FreqLossDemoPlot} that uses bins with bounds $\{0,200,400,600,800,1000\}$ the frequency deviation for each spatial bin would be compared to the corresponding bin centers at $\{100,300,500,700,900\}$.} While this numerical assessment of frequency loss does not exactly account for limitations in human perceptual ability, it does provide a more realistic model for the loss in perception that does occur.
 

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Update to Figure 6, adding spatial loss vs binsize to show how density striping is related to loss. Note that this still needs to be fixed since the plot uses spatial loss, not net spatial loss (random loss line wrong)

@@
 source("BinningLossFunctions.R")
+source("BinPackageFunctions.R")
 
@@
 Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
-
-Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
-100,000 observation pairs were simulated for each data set from the following distributions:
-
-\begin{itemize}
-\item Set I: $x_i \sim \text{iid Exp}(\lambda), y_i \sim \text{iid Exp}(\lambda)$ with $\lambda = 11$
-
-\item Set II: $x_i \sim \text{iid Normal}(\mu,\sigma^2), y_i \sim \text{iid Normal}(\mu,\sigma^2)$ with $\mu = 50$ and $\sigma=11$
-
-\item Set III: $x_i \sim \text{iid Uniform}(a,b), y_i \sim \text{iid Uniform}(a,b)$ with $a =0$ and $b=100$
-\end{itemize}
-%
-The parameters were selected to have data values roughly span the region $[0,100]^2$.
-The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
-These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data.
-Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
-
 %
@@
 
-ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
+ggsave("./figure/SimScatters.png", width=9, height=5.5, dpi=300)
 @
@@
 \begin{center}
- \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
+ \includegraphics[keepaspectratio=TRUE,width=.8\textwidth]{./figure/SimScatters.png}
 	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
@@
 %
+Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.
+100,000 observation pairs were simulated for each data set from the following distributions:
+
+\begin{itemize}
+\item Set I: $x_i \sim \text{iid Exp}(\lambda), y_i \sim \text{iid Exp}(\lambda)$ with $\lambda = 11$
+
+\item Set II: $x_i \sim \text{iid Normal}(\mu,\sigma^2), y_i \sim \text{iid Normal}(\mu,\sigma^2)$ with $\mu = 50$ and $\sigma=11$
+
+\item Set III: $x_i \sim \text{iid Uniform}(a,b), y_i \sim \text{iid Uniform}(a,b)$ with $a =0$ and $b=100$
+\end{itemize}
+%
+The parameters were selected to have data values roughly span the region $[0,100]^2$.
+The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.
+These simulated data sets are from continuous distributions, and thus the values are recorded to many decimal places; 6 decimal places in our simulate data.
+Real data is recorded to only the number of digits that measurement precision allows, and in many cases rounded even further.
+
 %Density changes are impossible to make out in the solid black areas of the plots.
@@
 By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.
-
-\ktm{These bivariate distributions were selected for the variety in their defining characteristics.
-\textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness and monotonicity (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}).
-The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}.
-They show that these examples explore a variety of skewnesses, convexity and skinniness.
-This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples.
-The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution.
-These minor striations will still prove problematic when binning is inconsistent with the data resolution.
-}
 
@@
 \end{table}
+
+\ktm{These bivariate distributions were selected for the variety in their defining characteristics.
+\textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: \textit{outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness} and \textit{monotonicity} (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}).
+The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}.
+They show that these examples explore a variety of \textit{skewnesses}, \textit{convexity} and \textit{skinniness}.
+This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples.
+The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution.
+These minor striations will still prove problematic when binning is inconsistent with the data resolution.
+}
 
@@
 
-<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>=
+<<UniformStripes, echo=F,include=F, cache=TRUE>>=
 # Commented out code for data generation: run once then recomment
@@
                      aspect.ratio=1,
-                     panel.grid= element_blank())
+                     panel.grid= element_blank()) +
+  theme(legend.position="bottom")
+
+  ggsave("./figure/CoarseUnif.png", width=6, height=4, dpi=300)
 
@@
 
+# NEED TO SAVE TO BinningLossDemos.Rdata which is read in chunk 1
+losses <- data.frame( bintype=NA, binsize=NA, type=NA, record=NA,
+                      avg_loss=NA, avg_standardized_loss=NA,
+                      avg_loss_x=NA,avg_loss_y=NA)
+counter=0
+
+for(type in "Uniform"){
+  for(record in "Coarse"){
+    for(binsize in seq(.5,8.5,by=.05)){
+      counter = counter + 1
+      alldatsub <- alldat[which(alldat$type==type & alldat$record==record),]
+      reduced_data <- rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
+                                  widthx=binsize,widthy=binsize, output="reduced")
+      losses[counter,] <- c("Standard",binsize,type,record, bin_loss(reduced_data))
+      counter = counter + 1
+      reduced_data <- rand_rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
+                                  widthx=binsize,widthy=binsize, output="reduced")
+      losses[counter,] <- c("Random",binsize,type,record, bin_loss(reduced_data))
+    }
+  }
+}
+
+ggplot()+ 
+  geom_line(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
+                color=bintype,group=bintype), data=losses,size=.5) +
+  geom_smooth(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
+                  color=bintype,group=bintype),size=.25, alpha=.2,
+              se=FALSE, data=losses,method="lm")+
+  geom_point(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
+                 color=bintype,group=bintype),
+             size=2,data=losses[losses$binsize %in% c(1,4,5), ])+
+#   geom_text(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
+#                 label=paste(binsize,"X",binsize,sep=""),color=bintype,group=bintype),
+#             size=10,data=losses[losses$binsize %in% c(1,4,5), ], 
+#             nudge_x=rep())
+  theme_bw() +
+  scale_color_manual("",values=c("red","black"))+
+  scale_x_continuous("Bin Sizes", breaks=1:8, labels=paste(1:8,"X",1:8,sep=""))+
+  annotate("text", x=c(3, 6.5), y=c(2.5,1.75), label=c("Random","Standard"),
+           color=c("red","black"),size=6)+
+  theme(legend.position="none", 
+         panel.grid.minor.x=element_blank()) + ylab("Average Spatial Loss")
+
+  ggsave("./figure/CoarseUnifLosses.png", width=4, height=5, dpi=300)
+
 @
+
+\begin{figure}[hbtp]
+\centering
+  \subfloat[Binned scatterplots]{\includegraphics[keepaspectratio=TRUE,width=.66\textwidth]{./figure/CoarseUnif.png}}
+  \subfloat[Spatial loss vs bin sizes]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/CoarseUnifLosses.png}} 
+  \caption[Traditional and adapted scatterplots for games vs.
+strikeouts data.]{\label{fig:UniformStripes} Binned scatterplots and spatial losses for various sized bins with coarse uniform data.
+Note the slight dip in loss at even integers, when bins are aligned with data resolution.}}
+\end{figure}
 

### 2016-12-03 versionForJCGS/BinPackageFunctions.R

@@ -1,0 +1,408 @@
+#------------------------------------------------------------------------------------------
+### Binning functions: 
+# 1d binning
+  # Binning by definition (bounds and centers)
+  # Creating bin definition lists from vectors of bounds and centers
+  # General rectangular binning
+  # Standard rectangular binning
+  # Quantile rectangular binning
+  # Random rectangular binning
+  # Frequency binning of reduced binned data
+
+# 2d binning
+  # Independant 1d binning grid
+  # Iterative conditional binning
+
+# Binned Scatterplot Functions
+  # Reduced binned data - one row per x/y bin
+  # Frequency binning - add freq bins to reduced data
+  # Plot construction from reduced binned
+  # Plot construction from raw data and bin specs
+
+# Binning Loss Functions and Diagnostic Plots
+  # Spatial Loss Summary
+  # Frequency Loss Summary
+  # Loss Screeplot for binning granularity
+
+#----------------------------------
+### univariate binning based on bins definition (potentially developed from different data source, i.e. train bins then bin test)
+# define bin boundaries, adapt outermost for the range of data to be binned, then use .bincode to drop to primitive to relabel
+bin_by_def <- function(new_data_vec, bin_definition, output="centers"){
+  # make bounds
+  bounds <- bin_definition$bin_bounds
+  bounds[1] <- min(min(new_data_vec),bounds[1])
+  bounds[length(bounds)] <- max(max(new_data_vec), bounds[length(bounds)])
+  # allocated to centers
+  data_bins <- bin_definition$bin_centers[.bincode(new_data_vec,bounds,right=FALSE, include.lowest=TRUE)]
+  names(data_bins) <- names(new_data_vec)
+  # Add bin labels
+  labs <-  paste(c("[",rep("(",(length(bin_definition$bin_bounds)-2))),
+                   bin_definition$bin_bounds[1:(length(bin_definition$bin_bounds)-1)],
+                   ",",bin_definition$bin_bounds[2:length(bin_definition$bin_bounds)],
+                   "]",sep="")
+  bin_labels <- factor(.bincode(new_data_vec,bounds,right=FALSE,include.lowest=TRUE),
+                       levels= 1:(length(bounds)-1),
[…]

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Fixed random binning net spatial loss problem from update to Figure 6.

@@
 
-# NEED TO SAVE TO BinningLossDemos.Rdata which is read in chunk 1
-losses <- data.frame( bintype=NA, binsize=NA, type=NA, record=NA,
-                      avg_loss=NA, avg_standardized_loss=NA,
-                      avg_loss_x=NA,avg_loss_y=NA)
-counter=0
-
-for(type in "Uniform"){
-  for(record in "Coarse"){
-    for(binsize in seq(.5,8.5,by=.05)){
-      counter = counter + 1
-      alldatsub <- alldat[which(alldat$type==type & alldat$record==record),]
-      reduced_data <- rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
-                                  widthx=binsize,widthy=binsize, output="reduced")
-      losses[counter,] <- c("Standard",binsize,type,record, bin_loss(reduced_data))
-      counter = counter + 1
-      reduced_data <- rand_rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
-                                  widthx=binsize,widthy=binsize, output="reduced")
-      losses[counter,] <- c("Random",binsize,type,record, bin_loss(reduced_data))
-    }
-  }
-}
+#   
+# lossesUnif <- data.frame( bintype=NA, binsize=NA, type=NA, record=NA,
+#                       avg_loss=NA, avg_standardized_loss=NA,
+#                       avg_loss_x=NA,avg_loss_y=NA)
+# counter=0
+# 
+# for(type in "Uniform"){
+#   for(record in "Coarse"){
+#     for(binsize in seq(.5,8.5,by=.05)){
+#       counter = counter + 1 
+#       alldatsub <- alldat[which(alldat$type==type & alldat$record==record),]
+#       reduced_data <- rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
+#                                   widthx=binsize,widthy=binsize, output="reduced")
+#       lossesUnif[counter,] <- c("Standard",binsize,type,record, bin_loss(reduced_data))
+#       counter = counter + 1
+#       reduced_data <- rand_rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
+#                                   widthx=binsize,widthy=binsize, output="reduced")
+#       lossesUnif[counter,] <- c("Random",binsize,type,record, bin_loss(reduced_data))
+#     }
+#   }
+# }
+# # lossesUnif saved to BinningLossDemos.Rdata which is loaded in startup chunck
 
@@
   geom_line(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
-                color=bintype,group=bintype), data=losses,size=.5) +
+                color=bintype,group=bintype), data=lossesUnif,size=.5) +
   geom_smooth(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
                   color=bintype,group=bintype),size=.25, alpha=.2,
-              se=FALSE, data=losses,method="lm")+
+              se=FALSE, data=lossesUnif,method="lm")+
   geom_point(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
                  color=bintype,group=bintype),
-             size=2,data=losses[losses$binsize %in% c(1,4,5), ])+
+             size=2,data=lossesUnif[lossesUnif$binsize %in% c(1,4,5), ])+
 #   geom_text(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
 #                 label=paste(binsize,"X",binsize,sep=""),color=bintype,group=bintype),
-#             size=10,data=losses[losses$binsize %in% c(1,4,5), ], 
+#             size=10,data=lossesUnif[lossesUnif$binsize %in% c(1,4,5), ], 
 #             nudge_x=rep())
@@
 # Save the loss binning data objects into central Rdata file for simplicity
-# save(lossOrigin, spatlossdatSquareBins, allbindat, file="BinningLossDemos.Rdata")
+# save(lossOrigin, spatlossdatSquareBins, allbindat, lossesUnif, file="./data/BinningLossDemos.Rdata")
 

### 2016-12-03 AuthorResponseToReviewers.Rmd
Added a comparison of Scott's MISE loss metric to our spatial loss using euclidean distances, as per reviewer 2 comment.
Clarifying example for freq loss
Update text to address addition of Figure 6(b). Also address reviewer in doc.

@@
 
+- The comparison and benefit over Scotts MISE is not included to the discussion leading into Section 4.
+
 *I do not understand how binned frequencies (as compared to true frequencies) are calculated in Section 4.2 on frequency loss.
@@
 Also please clarify is binned frequencies are measures of frequencies perceived by a human.*
+
+- A clarifying example using the visual from Figure 3 has been added to Section 4.2.
 
@@
 *I also fail to understand how the discussion on artificial striping in Section 5.1 is relevant for spatial loss.*
+
+- Artificial striping is a symptom of poorly aligned bin sizes and the data resolution, which will lead to imbalanced bin allocation and thus more spatial loss.
+A clarification of this idea and a visual comparison of bin sizes and spatial loss have been added to section 5.1 in the context of Figure 6.
 

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Update text to address addition of Figure 6(b). Also address reviewer in doc.

@@
 
-To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension.
-Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios.
+To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension \ktm{-- thus even integer sized bins are most desireable.
+Figure~\ref{fig:UniformStripes}(a) displays the binned scatterplots under several scenarios.
 Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins.
-Note that random binning is effective at smoothing out the artificial striping patterns when non-integer multiples of the data resolution are selected.
+Figure~\ref{fig:UniformStripes}(b) demonstates that with coarse data resolution there are departures from the linear relationship between spatial loss and bin sizes; with small spikes in spatial loss where the bin size is highly misaligned with the data resolution and slight drops in spatial loss when aligned perfectly.
+In this we can see that artificial striping is a symptom of poorly aligned bin specification.}
 
@@
   scale_fill_gradient(low="#56B1F7", high="#132B43",
-                      limits=c(0, 450),name="Bin Count") + 
+                      limits=c(0, 450),name="Bin \n Count") + 
   theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
                      aspect.ratio=1,
-                     panel.grid= element_blank()) +
-  theme(legend.position="bottom")
+                     panel.grid= element_blank())
 
@@
 \centering
-  \subfloat[Binned scatterplots]{\includegraphics[keepaspectratio=TRUE,width=.66\textwidth]{./figure/CoarseUnif.png}}
-  \subfloat[Spatial loss vs bin sizes]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/CoarseUnifLosses.png}} 
+  \subfloat[Binned scatterplots]{\includegraphics[keepaspectratio=TRUE,width=.63\textwidth]{./figure/CoarseUnif.png}} 
+  \hspace{.05\textwidth}
+  \subfloat[Spatial loss vs bin sizes]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/CoarseUnifLosses.png}} 
   \caption[Traditional and adapted scatterplots for games vs.
-strikeouts data.]{\label{fig:UniformStripes} Binned scatterplots and spatial losses for various sized bins with coarse uniform data.
+strikeouts data.]{\label{fig:UniformStripes} \ktm{Binned scatterplots and spatial losses for various sized bins with coarse uniform data.
 Note the slight dip in loss at even integers, when bins are aligned with data resolution.}}
 \end{figure}
+
+Note that random binning is effective at smoothing out the artificial striping patterns \ktm{when many point fell along bin boundaries -- accomplishing the motiviation for its development.
+However due to inflated spatial loss due to the stocastic bin assignment, standard binning with properly selected bin dimension is the better option}.
 
@@
 The data has only 42,583 observations, so we can use minimal binning to match the data resolution without taking an inordinate amount of computation time.
-While minimal binning eliminates spatial loss, the miniscule physical size of the resulting tiles makes it difficult to see the frequency shade; Therefore slightly larger two game by ten strikout bins are used, providing approximately 50 bins per dimension.
+While minimal binning eliminates spatial loss, the minuscule physical size of the resulting tiles makes it difficult to see the frequency shade; Therefore slightly larger two game by ten strikeout bins are used, providing approximately 50 bins per dimension.
 The origin offset of $(0.5,-0.5)$ is used to shift the binning back by a half game and down by one half strikeout to adjust for the data resolution.

### 2016-12-03 AuthorResponseToReviewers.Rmd
Added note on future work necessary to quantify implicit frequency loss. Also added rebuttal to reviewer 2's scoff at advice to restrict number of frequence bins.

@@
 
+- The statement addressing the Figure already contains a quantified comparison of spatial loss under the two scenarios.
+**"As an example, Figure 7 displays the binned scatterplots of ten-by-ten unit bins for the fine exponential data where the binning origin at (0,0) incurs seven percent lower spatial loss than for the (-9,-9) origin due to the heavy overlap into negative regions"** The difference is given as a percent decrease in spatial loss because the relative spatial loss change seemed most pertinent.
+
 *In visualization practices, it is common to use both ordinal scales as well as continuous scales when mapping numbers to colors.
@@
 
+- The need for future perception research to quantify the implicit frequency loss when using continuous shading scales is now noted in the discussion section of the paper.
+
 *I don’t buy the advice to “use between four and seven frequency”. There simply isn’t empirical evidence in the paper to support this claim.*
+
+- We do not want to allow more color bin than we can *easily* differentiate because the our expressed purpose for working with discretized color scales in this scenario is to have made loss of frequency information *entirely* explicit.
+While working with discretizing shades we found ourselves worrying about our ability to perfectly match bins to a legend after approximately 6 shades.
+We do not claim to have gathered empirical evidence, rather we defer to Healey and Enn's Theory on discernable colors (typically 4-7) to back out experience.
+Figure 9 was included for the reader to assess if they felt our suggestion was justifiable.
 

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Added note on future work necessary to quantify implicit frequency loss. Also added rebuttal to reviewer 2's scoff at advice to restrict number of frequence bins.
Clarify and soften rebuttal of 4-7 scoff slightly.

@@
 
-If we elect to use frequency binning to shade tiles ordinally, it is recommended to use between four and seven distinct shades.
+\ktm{Due to imperfect human perception may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit.
+If we aim to have highly accurate perceptitual mapping,} it is recommended to use at most between four and seven distinct shades.
 This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
 Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts.
+\ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots usign continuous shades.}
 

### 2016-12-04 AuthorResponseToReviewers.Rmd
Clarify and soften rebuttal of 4-7 scoff slightly.
Spell Check and minor edit

@@
 Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution.
-You will find that a discussion of the relevance of scagnostics has been added immediately preceeding the start of subsection 5.1.
+You will find that a discussion of the relevance of scagnostics has been added immediately preceding the start of subsection 5.1.
 It is interesting that the striation score is so small, given that the data resolution issue can lead to artificially induced stripping is seen in so prominently in Figure 6.
@@
 
-- You are correct that many of the ideas in Sections 2 and 3 are not original, with the exception of the random binning approach, where the methodology for random-binned aggregation is original to my best knowledge.
+- You are correct that many of the ideas in Sections 2 and 3 are not original, with the exception of the random binning approach, where the methodologies for random-binned aggregation and post processing for net spatial loss are original to our best knowledge.
 The goal of these sections is to bring many binning related concepts under a unified notation for the benefit of clarity in the sections to follow.
-For the random binning, a statement has been added to address the motivation for its introduction: deterministic one-sided interval bins lead to lopsided allocation of points on bin boundaries, whereas a random assignment is expected to split the points evenly across bins.
+For the random binning, clearer language of our idea and a statement has been added to address the motivation for its introduction: deterministic one-sided interval bins lead to lopsided allocation of points on bin boundaries, whereas a random assignment is expected to split the points evenly across bins.
 For the unoriginal ideas, additional citations have been made to place the unifying notation better within the existing literature.
@@
 While working with discretizing shades we found ourselves worrying about our ability to perfectly match bins to a legend after approximately 6 shades.
-We do not claim to have gathered empirical evidence, rather we defer to Healey and Enn's Theory on discernable colors (typically 4-7) to back out experience.
+We do not claim to have gathered empirical evidence, rather we defer to Healey and Enn's Theory on discernible colors (typically 4-7) to back out experience.
 Figure 9 was included for the reader to assess if they felt our suggestion was justifiable.
+A statement clarifying that the recommendations is to bolster the reliable perceptual mapping has been added to the discussion section.
 

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Spell Check and minor edit

@@
 
-An alternative to the rectangular binning process, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
+As an alternative to the rectangular binning process, \ktm{We propose} a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.

### 2016-12-04 AuthorResponseToReviewers.Rmd
Address reviewer 2's summary remarks

@@
 Do these metric matter for all the tasks involved in reading the scatterplots?*
+
+*Summary*
+*This paper feels like a survey paper but the references are not properly cited.
+As a research paper, it is not clear to me what are the main contributions, and how these contributions build on top of prior work (given that there is no “related work” section in the paper).
+While the two loss metrics seem potentially interesting, I think more empirical studies are needed with real human participants.
+It will also improve the paper if the authors can provide take-away guidelines or algorithms that can be applied to real-world problems.*
+
+- The main result from the paper is to provide simple adjustment recommendations for binning specification for reducing loss in binned scatterplots.
+To emphasize these recommendations, the baseball plot example in Section 5.3 has been reworked to highlight the take-home message.
+To your other point, we agree that future work to quantify the implicit loss of information (e.g.
+the imprecise ability to map continuously shaded tiles to a frequency legend) in reading binned scatterplots is needed; however, the current work strives to quantify explicit loss of information when the perception tasks are more simple (e.g.
+mapping distinguishably-discrete shaded tiles to an interval on the frequency legend).
+
+### Responses to Editors Comments/Questions
+*Editor's Comments to Author:*
+
+*Associate Editor - Comments to the Author:*
+*The paper in its current form does not provide a sufficiently thorough coverage of either review material or novel material to warrant publication.
+However, the reviewers have some useful suggestions for ways in which a different paper (or two) could possibly be produced by expanding on different parts of this submission.*

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Address reviewer 2's summary remarks

--- 2016-12-03 versionForJCGS/PaperJCGS.Rnw
@@
 For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers.
-\ktm{What we will refer to as} \textit{General rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
+\ktm{What we will refer to as} \textit{general rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data.
 Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval.
@@
 
-An alternative to the rectangular binning process, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
+As an alternative to the rectangular binning process, \ktm{we propose} a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins.
 In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$.

### 2016-12-04 AuthorResponseToReviewers.Rmd
Expanded on color contrast impact on perception a bit. Added Adelson and Fairchild citations.

@@
 These effects can be modelled and estimated and maybe interesting to investigate a little further.*
+
+- Citations to refer readers to Adelson and Fairchild as example of work on contrast effects has been added to Section 4.2 and related future work has been added to the discussion in Section 6.
 
@@
 I also wonder how and why random binning can be useful.
-Please provide scenarios in which it would make sense to use random binning.*
-
-*For the other ideas covered in these two sections, please properly cite relevant work.*
+Please provide scenarios in which it would make sense to use random binning.
+For the other ideas covered in these two sections, please properly cite relevant work.*
 
@@
 The goal of these sections is to bring many binning related concepts under a unified notation for the benefit of clarity in the sections to follow.
-For the random binning, clearer language of our idea and a statement has been added to address the motivation for its introduction: deterministic one-sided interval bins lead to lopsided allocation of points on bin boundaries, whereas a random assignment is expected to split the points evenly across bins.
-For the unoriginal ideas, additional citations have been made to place the unifying notation better within the existing literature.
+For the random binning, clearer language of proposal of the idea and a statement has been added to address the motivation for its introduction.
+For the unoriginal ideas, additional citations have been made to place our unifying notation better within the existing literature.
 
@@
 In the second paragraph, you talk about three properties of visual estimator: unbiasedness, consistency and efficiency, I think these properties are interesting, but the rest of the paper does not address them sufficiently.*
+
+- The interesting parallels with common properties of statistical estimators were used as a motivating concept for the construction of binning loss functions.
+We felt that an extended exploration of this analog would distract from the exploration on optimizing bin specification.
+The need for future work on exploring this topic has been added to the discussion in Section 6.
 
@@
 However, the reviewers have some useful suggestions for ways in which a different paper (or two) could possibly be produced by expanding on different parts of this submission.*
+
+- We hope that the revisions explanding review material and clarification of our contributions to the topic will meet the standards of your publication.
+If there are any additional comments, concerns or questions, our team is happy to address them.

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Expanded on color contrast impact on perception a bit. Added Adelson and Fairchild citations.

@@
 While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise.
-Color perception of shade is extremely context sensitive -- see Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors -- allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend.
+Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{adelson1993perceptual,fairchild2013color}}.
+See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors.
 It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.
@@
 
-Large bivariate data sets are very difficult to visualize in raw form, due to over-plotting of points.
+Large bivariate data sets are very difficult to visualize in raw form, \ktm{with the implicit loss of information due to} over-plotting of points.
 Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system.
 Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in spatial and frequency information.
+\ktm{We have introduced spatial and frequency loss functions to explicitly quantify the lost information in the construction of the binned scatterplot.
+This offers an optimizing criterion to guide binning specification and a foundation for future work on properties of a binned scatterplot as a visual estimator of the bivariate density -- such as unbiasedness, consistency and efficiency.}
 
 We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms.
-The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution.
+\ktm{Our proposed} random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution.
 However, the standard binning algorithm is superior due to lower spatial loss.
@@
 \ktm{Due to imperfect human perception may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit.
-If we aim to have highly accurate perceptitual mapping,} it is recommended to use at most between four and seven distinct shades.
+If we aim to have highly accurate perceptual mapping,} it is recommended to use at most between four and seven distinct shades.
 This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously.
 Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts.
-\ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots usign continuous shades.}
+\ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots using continuous shading and the impact of contrasting effects.}
 

### 2016-12-04 AuthorResponseToReviewers.Rmd
Add also Bartelson and Brenemon to shade contrast list.

@@
 
-- Citations to refer readers to Adelson and Fairchild as example of work on contrast effects has been added to Section 4.2 and related future work has been added to the discussion in Section 6.
+- Citations to refer readers to Bartelson & Breneman, Adelson, and Fairchild as example of work on contrast effects has been added to Section 4.2 and related future work has been added to the discussion in Section 6.
 

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Add also Bartelson and Brenemon to shade contrast list.

@@
 While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise.
-Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{adelson1993perceptual,fairchild2013color}}.
+Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{bartleson1967brightness,adelson1993perceptual,fairchild2013color}}.
 See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors.

### 2016-12-04 AuthorResponseToReviewers.Rmd
Cleaning up and putting header in R scripts to be submitted to JCGS

@@
 - We hope that the revisions explanding review material and clarification of our contributions to the topic will meet the standards of your publication.
+An R-package with binning, loss and plotting fuctions is available through an author's Github repository; however linking directly would not allow for blinded review.
+As such the functions used are provided in a supplementary R file.
+Additionally, a zipped folder containing an Rproj file for launching the Rpackage locally can be made available if desired.
 If there are any additional comments, concerns or questions, our team is happy to address them.

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Cleaning up and putting header in R scripts to be submitted to JCGS

@@
 set.seed(14)
-rxcent <- RandRectBin1d(x, -10, 10)
-rycent <- RandRectBin1d(y, -10, 10)
+
+rxcent <- rand_rect_bin_1d(x, -10, 10, output="centers")
+rycent <- rand_rect_bin_1d(y, -10, 10, output="centers")
 # point 1 and 11 are special 

### 2016-12-04 versionForJCGS/BinPackageFunctions.R

@@ -1,5 +1,5 @@
 #------------------------------------------------------------------------------------------
-### Binning functions: 
+### Binning functions:
 # 1d binning
   # Binning by definition (bounds and centers)
   # Creating bin definition lists from vectors of bounds and centers
@@ -12,10 +12,10 @@
 # 2d binning
   # Independant 1d binning grid
   # Iterative conditional binning
-
-# Binned Scatterplot Functions
   # Reduced binned data - one row per x/y bin
   # Frequency binning - add freq bins to reduced data
+
+# Binned Scatterplot Functions
   # Plot construction from reduced binned
   # Plot construction from raw data and bin specs
 
@@ -37,13 +37,13 @@ bin_by_def <- function(new_data_vec, bin_definition, output="centers"){
   names(data_bins) <- names(new_data_vec)
   # Add bin labels
   labs <-  paste(c("[",rep("(",(length(bin_definition$bin_bounds)-2))),
-                   bin_definition$bin_bounds[1:(length(bin_definition$bin_bounds)-1)],
-                   ",",bin_definition$bin_bounds[2:length(bin_definition$bin_bounds)],
-                   "]",sep="")
+                 bin_definition$bin_bounds[1:(length(bin_definition$bin_bounds)-1)],
+                 ",",bin_definition$bin_bounds[2:length(bin_definition$bin_bounds)],
+                 "]",sep="")
   bin_labels <- factor(.bincode(new_data_vec,bounds,right=FALSE,include.lowest=TRUE),
                        levels= 1:(length(bounds)-1),
                        labels= labs)
-  
+
   if(output=="centers") return(data_bins)
   if(output=="definition") return(bin_definition)
   if(output=="all") return(list(data_bins=data_bins,bin_labels=bin_labels,bin_definition=bin_definition))
@@ -106,7 +106,7 @@ rand_rect_bin_1d <- function(xs, origin, width, output="centers"){
   ubs <- bin_centers[1] + width*ceiling((xs-bin_centers[1])/width)
   # initially assign all values to upper bound
   data_bins <- ubs
-  # then use runif to reassign based on distance from 
+  # then use runif to reassign based on distance from
   plower <- (ubs - xs)/width
   lowerindex <- (plower > runif(length(xs), 0, 1))
[…]

### 2016-12-04 versionForJCGS/BinningLossFunctions.R

@@ -1,3 +1,10 @@
+### The following file contains functions were constructed in the process of writing the manuscipt and are required
+# to compile the PDF document.
+#
+# A cleaned subset of these functions have been packaged for dissemination as an R package designed to accompany
+# the paper. While the package is already available through Github it is not possible to anonymize the source for the
+# blinded review process, hence the are hosted in the BinPackageFunction.R file that also supplements the manuscript.
+# ---------------------------------------------------------------------------------------------------------------
 ### Preliminaries 
 
 library(ggplot2)
@@ -220,30 +227,4 @@ freqBin <- function(binout, binType="standard", ncolor, logCount=FALSE){
 # # Plot with LogCount Binned Frequencies
 # qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]]) +
 #   scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutLogFreqGroups[[1]]$freqgroup))
-
-
-
-
-
-### Create binning function that does not calculate loss
-# to be used to track computation time for binning
-
-## 2d Rectangular Binning (for either Standard or Random)
-# standard is straight forward extension of 1d binning
-# random binning needs post processing to calculate minimum spatial loss
-RectBin2dNoLoss <- function(xs,ys, originx, originy, widthx, widthy, type="standard"){
-  if(type=="standard"){
-  tempdat <- data.frame(xs = xs, ys=ys,
-                        binxs = StandRectBin1d(xs,originx,widthx),
-                        binys = StandRectBin1d(ys,originy,widthy))
-  }
-  if(type=="random"){
-    tempdat<- data.frame(xs = xs, ys=ys,
-                          binxs =  RandRectBin1d(xs,originx,widthx),
-                          binys =  RandRectBin1d(ys,originy,widthy))
-  }
-  outdat <- ddply(tempdat, .(binxs,binys), summarise,
-                  binfreq = length(xs))
-  return(outdat[,1:3])
-}
 

### 2016-12-05 AuthorResponseToReviewers.Rmd
final review updates

@@
 
-- We are very grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots.
+- We are grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots.
 Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution.
@@
 
+- A statement on the connection of spatially-related user takes has been added to the introduction of spatial loss in section 4.1.
+Also, the difficult comparison of bin frequency information through continuous shade has been added to the discussion on why frequency binning may be employed in Section 4.2.
+
 *Summary*

### 2016-12-05 versionForJCGS/PaperJCGS.Rnw
final review updates

@@
 Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}.
-\ktm{Sturges followed with in early work in formalizing histograms specification \citep{sturges1926choice}.}  More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}.
+\ktm{Sturges followed with an early work in formalizing histograms specification \citep{sturges1926choice}.}  More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}.
 \citet{scott1992} extends to the properties of multivariate binning strategies.
@@
 When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.
-This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers).
+\ktm{The tile size impacts the precision with which a user can read out location related attributes of the scatterplot; such as fine density patterns or outlier identification.} This loss can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers).
 Note that other distance metrics could be used, but the Euclidean distance has a desirable interpretability in $\mathbb{R}^2$.
@@
 See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors.
-It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.
+It is therefore not realistic, to expect readers to be able to decode frequency \ktm{or accurately compare bin frequencies} from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.
 %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.
@@
 The \textit{data resolution} is defined as the smallest increment between successive data values.
-To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension.
+To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension.
 This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}.
@@
 
-To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension \ktm{-- thus even integer sized bins are most desireable.
+To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension \ktm{-- thus even integer sized bins are most desirable.
 Figure~\ref{fig:UniformStripes}(a) displays the binned scatterplots under several scenarios.
 Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins.
-Figure~\ref{fig:UniformStripes}(b) demonstates that with coarse data resolution there are departures from the linear relationship between spatial loss and bin sizes; with small spikes in spatial loss where the bin size is highly misaligned with the data resolution and slight drops in spatial loss when aligned perfectly.
+Figure~\ref{fig:UniformStripes}(b) demonstrates that with coarse data resolution there are departures from the linear relationship between spatial loss and bin sizes; with small spikes in spatial loss where the bin size is highly misaligned with the data resolution and slight drops in spatial loss when aligned perfectly.
 In this we can see that artificial striping is a symptom of poorly aligned bin specification.}
@@
 
-Note that random binning is effective at smoothing out the artificial striping patterns \ktm{when many point fell along bin boundaries -- accomplishing the motiviation for its development.
-However due to inflated spatial loss due to the stocastic bin assignment, standard binning with properly selected bin dimension is the better option}.
+Note that random binning is effective at smoothing out the artificial striping patterns \ktm{when many point fell along bin boundaries -- accomplishing the motivation for its development.
+However due to inflated spatial loss due to the stochastic bin assignment, standard binning with properly selected bin dimension is the better option}.
 
@@
 We now revisit the baseball data used earlier and construct binned scatterplots using the loss-based specification principles developed in the previous sections.
-\ktm{In the spatial binning step we learned three rules of thumb: (1) use smaller bins to achieve lower loss of spatial information, (2) standard binning has less loss than random binning but we need to be careful to choose bin dimensions that align with data resolution, and (3) offset the origin by a factor of half the data resolution to better center point in bins.}
+\ktm{In the spatial binning step we learned three rules of thumb: (1) use smaller bins to achieve lower loss of spatial information, (2) standard binning has less loss than random binning but we need to be careful to choose bin dimensions that align with data resolution, and (3) offset the origin by a factor of half the data resolution to better center points within bins.}
 
@@
 
-\ktm{Due to imperfect human perception may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit.
+\ktm{Due to imperfect human perception, we may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit.
 If we aim to have highly accurate perceptual mapping,} it is recommended to use at most between four and seven distinct shades.
@@
 
-Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms.
-The findings of this research provide suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.
+Future implementations of software \ktm{should consider these loss-related findings in graphical tools for constructing binned scatterplots.
+This research provides suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.}
 

### 2016-12-05 versionForJCGS/BinPackageFunctions.R

@@ -1,3 +1,464 @@
+# This file contains a reworked set of binning, loss and plot building functions that are designed 
+# to accompany the paper in an associated R package. The package is currently in a development version hosted 
+# and available through Github
+#------------------------------------------------------------------------------------------
+### Binning functions: 
+# 1d binning
+# Binning by definition (bounds and centers)
+# Creating bin definition lists from vectors of bounds and centers
+# General rectangular binning
+# Standard rectangular binning
+# Quantile rectangular binning
+# Random rectangular binning
+# Frequency binning of reduced binned data
+
+# 2d binning
+# Independant 1d binning grid
+# Iterative conditional binning
+
+# Binned Scatterplot Functions
+# Reduced binned data - one row per x/y bin
+# Frequency binning - add freq bins to reduced data
+# Plot construction from reduced binned
+# Plot construction from raw data and bin specs
+
+# Binning Loss Functions and Diagnostic Plots
+# Spatial Loss Summary
+# Frequency Loss Summary
+# Loss Screeplot for binning granularity
+
+#----------------------------------
+### univariate binning based on bins definition (potentially developed from different data source, i.e. train bins then bin test)
+# define bin boundaries, adapt outermost for the range of data to be binned, then use .bincode to drop to primitive to relabel
+bin_by_def <- function(new_data_vec, bin_definition, output="centers"){
+  # make bounds
+  bounds <- bin_definition$bin_bounds
+  bounds[1] <- min(min(new_data_vec),bounds[1])
+  bounds[length(bounds)] <- max(max(new_data_vec), bounds[length(bounds)])
+  # allocated to centers
+  data_bins <- bin_definition$bin_centers[.bincode(new_data_vec,bounds,right=FALSE, include.lowest=TRUE)]
+  names(data_bins) <- names(new_data_vec)
+  # Add bin labels
+  labs <-  paste(c("[",rep("(",(length(bin_definition$bin_bounds)-2))),
+                 bin_definition$bin_bounds[1:(length(bin_definition$bin_bounds)-1)],
[…]

### 2016-12-05 AuthorResponseToReviewers.Rmd
Edit Letter to Reviewers

@@
 The revisions will be reverted to black after the revision process.
+Thank you very much for your time and helpful reviews.
 
@@
 
-- We hope that the revisions explanding review material and clarification of our contributions to the topic will meet the standards of your publication.
-An R-package with binning, loss and plotting fuctions is available through an author's Github repository; however linking directly would not allow for blinded review.
-As such the functions used are provided in a supplementary R file.
-Additionally, a zipped folder containing an Rproj file for launching the Rpackage locally can be made available if desired.
+- We hope that the revisions expanding review material and clarification of our contributions to the topic will meet the standards of your publication.
+An R-package with binning, loss and plotting functions is available through an author's Github repository; however linking directly would not allow for blinded review.
+As such the functions used are provided in a supplementary R file labeled "BinPackageFunctions.R".
+Additionally, a link to the R package on Github can be provided to the editor(s) upon request.
 If there are any additional comments, concerns or questions, our team is happy to address them.
